### 2025-05-27 main.tex
Initial modular thesis setup

\documentclass[english, version-2020-11]{uzl-thesis}
\UzLStyle{computer modern oldschool design}
\input{preamble.tex} % Put your \UzLStyle and \UzLThesisSetup here

\begin{document}

\chapter{Introduction}
\input{chapters/01_introduction.tex}

\chapter{Background and System Architecture}
\input{chapters/02_background.tex}

\chapter{Methodology}
\input{chapters/03_methodology.tex}

\chapter{Results and discussion}
\input{chapters/04_results.tex}

\chapter{System Configuration Recommendations}
\input{chapters/05_recommendations.tex}

\chapter{Conclusion and Future Work}
\input{chapters/06_conclusion.tex}

\begin{bibtex-entries}
@TechReport{Kernighan1974,
author = {Brian Kernighan},
title = {Programming in C – A Tutorial},
institution = {Bell Laboratories},
year = {1974}
}
\end{bibtex-entries}
\end{document}

### 2025-05-27 preamble.tex
Initial modular thesis setup

\UzLThesisSetup{
Masterarbeit,
Verfasst              = {am}{Institut für Technische Informatik},
Titel auf Deutsch     = {
    Bewertung von RISC-V-Enklaven: Leistungsbenchmarks und bewährte Konfigurationspraktiken
}, 
Titel auf Englisch    = {
    Evaluating RISC-V Enclaves: Performance Benchmarks and Configuration Best Practices 
},
Autor                 = {Basil Ugbomoiko},
Betreuerin            = {Dr.-Ing. Saleh Mulhem},
Mit Unterstützung von = {Henry Strunck},
Studiengang           = {IT Security},
Datum                 = {31. August 2025},
Abstract = {This thesis evaluates the performance of Keystone enclaves, an open-source Trusted Execution Environment (TEE) designed for RISC-V platforms. Keystone enables secure computation by isolating sensitive workloads from the rest of the system. To assess its performance, the study benchmarks critical system parameters, including the number of CPU cores, cache size, memory allocation, and enclave configuration. Using industry-standard benchmarking tools such as lmbench and CoreMark, the research measures execution time, context switch overhead, memory latency, and CPU utilization across a range of hardware and software configurations. The experimental analysis identifies key performance bottlenecks that impact the efficiency of enclave execution. Based on these findings, the study presents configuration best practices and tuning recommendations tailored to various workload profiles, such as compute-bound, memory-intensive, and mixed applications. The insights gained from this evaluation contribute to optimizing the deployment of Keystone enclaves, making them more viable for real-world use cases that demand secure and efficient execution on RISC-V architectures.},
Zusammenfassung = {Diese Arbeit bewertet die Leistung von Keystone-Enklaven, einer Open-Source Trusted Execution Environment (TEE) für RISC-V-Plattformen. Keystone ermöglicht sichere Berechnungen, indem sensible Workloads vom restlichen System isoliert werden. Zur Leistungsbewertung werden zentrale Systemparameter wie die Anzahl der CPU-Kerne, die Cache-Größe, die Speicherzuweisung und die Konfiguration der Enklave untersucht. Mithilfe standardisierter Benchmarking-Tools wie lmbench und CoreMark werden Ausführungszeit, Kontextwechsel-Overhead, Speicherlatenz und CPU-Auslastung unter verschiedenen Hardware- und Softwarekonfigurationen gemessen. Die experimentelle Analyse identifiziert wichtige Leistungsengpässe, die die Effizienz der Enklaven beeinträchtigen. Auf Basis dieser Erkenntnisse werden Konfigurationsrichtlinien und Optimierungsempfehlungen vorgestellt, die auf unterschiedliche Workload-Profile zugeschnitten sind, z. B. rechenintensive, speicherintensive und gemischte Anwendungen. Die gewonnenen Erkenntnisse tragen dazu bei, den Einsatz von Keystone-Enklaven zu optimieren und ihre praktische Anwendbarkeit in sicherheitskritischen Bereichen auf RISC-V-Architekturen zu verbessern.},
Numerische Bibliographie
}

### 2025-05-27 chapters/background.tex
Updated README.md to reflect changes in LaTeX structure and preamble.tex

\section{Trusted Execution Environments}

\section{Keystone architecture overview}

\section{Comparison with other TEEs}

\section{Performance considerations in TEEs}

### 2025-05-27 chapters/conclusion.tex
Updated README.md to reflect changes in LaTeX structure and preamble.tex

\section{Summary of Contributions}

\section{Limitations of the Current Study}

\section{Directions for Future Research}

### 2025-05-27 chapters/methodology.tex
Updated README.md to reflect changes in LaTeX structure and preamble.tex

\section{Experimental setup}

\section{Benchmarking tools and metrics}
%\section{Performance metrics}

\section{Parameter Variation Strategy}

\section{Data Collection and Analysis Procedures}

### 2025-05-28 chapters/background.tex
updated relevant chapters & sections

@@
 \section{Trusted Execution Environments}
+Trusted Execution Environments (TEEs) are isolated computing environments that provide strong security guarantees for code and data execution, even in the presence of potentially compromised operating systems and applications.
+They achieve this by leveraging hardware-based isolation mechanisms to establish secure boundaries, often called enclaves, where sensitive computation can occur without interference from the rest of the system.
+TEEs aim to ensure confidentiality and integrity through careful hardware and software co-design, minimizing the trusted computing base (TCB) and mitigating a wide range of software attacks.
+While several commercial TEEs, such as Intel Software Guard Extensions (SGX) and ARM TrustZone, demonstrate the feasibility and practical relevance of enclave-based secure execution, their closed-source and inflexible designs limit research and hardware/software co-design opportunities.
+This has motivated the development of open TEE frameworks like Keystone, built on RISC-V—an open and extensible instruction set architecture.
+A key architectural feature of TEEs is privilege separation, where a minimal TCB executes in a highly privileged mode controlling access to critical physical resources, while the operating system and applications run in less privileged, untrusted modes.
 
 \section{Keystone architecture overview}
+
+Keystone represents an open-source framework designed to facilitate the architecture and implementation of Trusted Execution Environments (TEEs) leveraging hardware enclaves on RISC-V processors.
+Its primary objective is to enable the construction of customizable TEEs optimized for specific RISC-V platforms, thereby achieving enhanced performance, reduced trusted computing base (TCB) complexity, and improved programmability tailored to distinct workloads and security threat models.
+
+A typical Keystone-capable system architecture comprises multiple components distributed across different privilege modes of the RISC-V privilege hierarchy.
+The foundational hardware is a trusted CPU package integrating Keystone-compatible RISC-V cores and a silicon root of trust, potentially augmented by features such as cache partitioning, memory encryption, and secure randomness sources.
+The core security functionality is governed by the Security Monitor (SM), a minimalistic M-mode software component embodying the system's TCB.
+The SM is responsible for enclave lifecycle management and enforces isolation boundaries between enclaves and the untrusted operating system (OS).
+
+Keystone’s Security Monitor (SM), executing at the highest privilege level (M-mode), serves as the linchpin of the system’s trusted computing base.
+Beyond enclave lifecycle management, the SM enforces fine-grained memory isolation policies and orchestrates enclave scheduling, maintaining strict control over resource access to prevent unauthorized interference.
+
+\begin{figure}[htbp]
+    \centering
+    \includegraphics[width=0.9\linewidth]{figures/keystone_overview.png}
+    \caption{Overview of the Keystone architecture illustrating components such as the Security Monitor, enclave runtime, and the privilege hierarchy.}
+    \label{fig:keystone_overview}
+\end{figure}
+
+Enclaves, the fundamental isolation units in Keystone, operate within dedicated physical memory regions inaccessible to the OS and other enclaves.
+Physical Memory Protection (PMP), a hardware-assisted mechanism controlled by the SM, restricts access to enclave memory exclusively to the enclave and the SM, thus preserving confidentiality and integrity even in the event of OS compromise.
+
+Each enclave comprises two principal layers: the user-level enclave application (eapp) and a supervisor-level runtime environment.
+The eapp executes user-specific logic within the enclave, while the runtime, operating in supervisor mode (S-mode), manages system calls, exception handling, and virtual memory services intrinsic to enclave operation.
+This layered design provides a clear separation of concerns, reducing attack surfaces and allowing for customized security policies adapted to workload requirements.
+
+Keystone’s workflow delineates distinct roles for platform providers and enclave developers, fostering modularity and flexibility.
+Platform providers undertake the compilation and deployment of the SM tailored to the target hardware, ensuring integration of the root of trust and hardware-specific functionalities.
+Enclave developers leverage the Keystone SDK to construct enclave applications alongside their runtimes and host binaries, which are subsequently packaged and deployed on the target platform independently of underlying hardware specifics.
+
+Furthermore, Keystone supports remote attestation mechanisms, enabling verification of enclave authenticity and integrity prior to provisioning sensitive data or executing critical workloads.
+This capability is essential for secure deployment in distributed and cloud environments.
+
+The enclave lifecycle progresses through three stages: creation, execution, and destruction.
+Creation involves the host allocating enclave private memory (EPM), populating it with enclave page tables, runtime, and application binaries, and invoking the SM to isolate the enclave via PMP.
+During execution, the SM orchestrates transitions into and out of the enclave, dynamically adjusting PMP permissions to maintain strict isolation.
+Destruction securely clears the enclave’s EPM and reclaims resources, ensuring no residual data remains.
+
+\begin{figure}[htbp]
+    \centering
+    \includegraphics[width=0.9\linewidth]{figures/enclave_lifecycle.png}
+    \caption{Stages of a Keystone enclave lifecycle: creation, execution, and destruction.}
+    \label{fig:enclave_lifecycle}
+\end{figure}
+
+Keystone’s design also facilitates extensibility, allowing for integration of advanced security features such as secure I/O, cryptographic accelerators, and hardware-assisted debugging.
+This adaptability future-proofs the architecture, enabling continuous evolution of TEE capabilities aligned with emerging threats and diverse application demands on RISC-V platforms.
 

### 2025-05-28 chapters/methodology.tex
updated relevant chapters & sections

@@
 \section{Experimental setup}
+%The experimental setup utilizes the Keystone-enhanced QEMU emulator running on an Ubuntu host operating system.
+This configuration allows for flexible development and testing of Keystone enclaves in a controlled environment without requiring physical RISC-V hardware.
+The Ubuntu host provides a stable platform for building and deploying enclave binaries, and it facilitates the integration of the Keystone kernel driver and SDK.
+
+The Keystone framework was evaluated using a RISC-V specific fork of the QEMU emulator on an Ubuntu 22.04 host system.
+This configuration allowed testing of Keystone enclaves in a controlled virtualized environment, avoiding the need for physical RISC-V hardware.
+The system was configured to emulate the RV64 architecture for evaluating Keystone’s performance.
+
+The build process for Keystone utilized the Make and Buildroot tools.
+Key configuration options included the selection of the QEMU virtual platform, the RV64 architecture, and the corresponding Buildroot configuration.
+Keystone was initiated in the QEMU environment with the following command:
+
+\begin{verbatim}
+make run
+\end{verbatim}
+
+This command launched the Security Monitor and the Linux OS. Within the QEMU guest, the Keystone kernel driver was manually loaded using:
+
+\begin{verbatim}
+modprobe keystone-driver
+\end{verbatim}
+
+For debugging, Keystone was run in debug mode with the QEMU GDB server connected to inspect memory protection and control registers:
+
+\begin{verbatim}
+KEYSTONE_DEBUG=y make run
+make debug-connect
+\end{verbatim}
+
+Additional details on the build process, dependencies, and configuration options are available in the \textbf{Appendix}.
 
@@
 %\section{Performance metrics}
+The performance characterization of Keystone utilized two established synthetic benchmarks tailored for embedded system evaluation: \textit{Dhrystone} and \textit{CoreMark}.
+Both were selected for their cross-platform compatibility and established use in RISC-V environments.
 
+\textbf{Dhrystone} evaluates integer arithmetic and control flow performance, making it effective for assessing core CPU throughput in environments where floating-point operations are secondary.
+The RISC-V version, sourced from the \texttt{riscv-tests} repository, was employed to calculate \textit{Dhrystones per second (DMIPS)}.
+To isolate Keystone’s overhead, measurements were conducted in both native and enclave contexts.
+
+\textbf{CoreMark}, developed by EEMBC, provides a broader performance assessment through representative workloads, including matrix operations, list processing, and state machine evaluations.
+The RISC-V implementation from the SiFive repository was used to determine \textit{iterations per second (IPS)}.
+This benchmark offers a deeper view into system performance under more diverse computational tasks than Dhrystone.
+
+Performance variability was assessed by calculating the standard deviation across repeated runs. Metrics collected included:
+\begin{itemize}
+\item Execution Time
+\item DMIPS (Dhrystone) and IPS (CoreMark)
+\item Standard Deviation across trials
+\end{itemize}
+
+This dual-benchmark approach ensured a comprehensive evaluation of both low-level CPU operations and higher-level application performance within the enclave context.
 \section{Parameter Variation Strategy}

### 2025-05-28 main.tex
Updated README.md to reflect changes in LaTeX structure and preamble.tex
updated relevant chapters & sections

@@
 \UzLStyle{computer modern oldschool design}
+\usepackage{float}
+
 \input{preamble.tex} % Put your \UzLStyle and \UzLThesisSetup here
@@
 \chapter{Introduction}
-\input{chapters/01_introduction.tex}
+\input{chapters/introduction.tex}
 
 \chapter{Background and System Architecture}
-\input{chapters/02_background.tex}
+The RISC-V instruction set architecture (ISA) serves as a promising foundation for TEE implementation due to its open and extensible nature.
+RISC-V offers unique benefits beyond its open-source status: it incorporates security-oriented primitives such as Physical Memory Protection (PMP) which facilitates efficient and fine-grained memory isolation.
+Moreover, its community-driven evolution fosters innovation in security features, enabling frameworks like Keystone to experiment with, and contribute to, ISA enhancements.
+
+RISC-V defines three primary privilege modes: user (U-mode), supervisor (S-mode), and machine (M-mode).
+These modes govern the execution privileges and resource accessibility of software components.
+M-mode, the highest privilege level, has exclusive control over hardware resources and interrupt management, rendering it an ideal candidate for hosting the TCB, such as the Security Monitor (SM) in Keystone.
+Unlike hardware microcode, M-mode software can be developed and patched flexibly using standard programming languages and toolchains, improving maintainability and formal verification potential.
+
+A key hardware feature leveraged by Keystone is the PMP, introduced in the RISC-V Privilege Specification version 1.10.
+PMP enables M-mode to enforce physical memory access restrictions on lower privilege levels, thus underpinning enclave memory isolation.
+Despite variations in PMP implementations across platforms, its guarantees are standardized, facilitating portable and secure enclave deployment.
+
+Interrupt and exception handling in RISC-V is primarily managed at the M-mode level, which may delegate common traps, such as page faults and system calls, to S-mode to optimize performance.
+Virtual memory support is provided by a memory management unit (MMU) that translates virtual addresses to physical addresses using multi-level page tables.
+While the operating system controls virtual memory via the \texttt{satp} register, Keystone enclaves maintain protected page tables and are secured against OS-based attacks through PMP-enforced physical memory restrictions.
+
+Collectively, these architectural features establish a robust foundation for implementing secure and efficient TEEs on RISC-V platforms, as exemplified by Keystone.
+
+\input{chapters/background.tex}
 
 \chapter{Methodology}
-\input{chapters/03_methodology.tex}
+The objective of the performance study was to quantify the overhead introduced by Keystone's enclave isolation mechanisms on RISC-V architectures.
+Experiments were executed within a QEMU-based virtual environment configured for the RV64 architecture, providing consistency and hardware independence.
+
+To run benchmarks, a separate enclave application (eapp) and corresponding host application were developed for each benchmark.
+The host application initialized and managed the enclave using the Keystone C++ API, while the enclave application contained the benchmark logic and communicated with the host through edge calls.
+This modular structure allowed each benchmark to be executed both natively (outside the enclave) and securely (within the enclave), enabling accurate measurement of the enclave-related overhead.
+To achieve a comparative analysis, two execution modes were considered:
+\begin{enumerate}
+\item \textit{Native Execution}: Benchmark execution without enclave isolation.
+\item \textit{Enclave Execution}: Benchmark execution within the Keystone TEE.
+\end{enumerate}
+
+Each benchmark—Dhrystone and CoreMark—was run multiple times under both conditions.
+Performance metrics, including total execution time, throughput (DMIPS or IPS), and result variability (standard deviation), were recorded and analyzed.
+
+The evaluation focused on determining the relative performance degradation attributable to enclave isolation by comparing results across the two execution modes.
+
+\input{chapters/methodology.tex}
 
 \chapter{Results and discussion}
-\input{chapters/04_results.tex}
+\input{chapters/results.tex}
 
 \chapter{System Configuration Recommendations}
-\input{chapters/05_recommendations.tex}
+\input{chapters/recommendations.tex}
 
 \chapter{Conclusion and Future Work}
-\input{chapters/06_conclusion.tex}
+\input{chapters/conclusion.tex}
 

### 2025-05-28 chapters/background.tex
updated relevant chapters & sections

@@
 
+Keystone distinguishes itself from widely adopted commercial Trusted Execution Environments (TEEs) such as ARM TrustZone and Intel Software Guard Extensions (SGX) through its open-source design, extensibility, and architectural modularity.
+While TrustZone and SGX have demonstrated practical value in securing sensitive computations on commodity hardware, their rigid implementations impose significant limitations on research, customization, and fine-grained hardware/software co-design.
+
+ARM TrustZone implements a two-world model in which the processor switches between a Secure World (TEE) and a Normal World (REE) via secure monitor calls (SMCs).
+Each world has its own kernel and supports user- and supervisor-level execution.
+Trusted operating systems such as OP-TEE run in the Secure World and conform to the GlobalPlatform TEE Internal API for developing trusted applications.
+However, the TrustZone model provisions only a single, statically allocated TEE at boot time, limiting dynamic enclave creation and scalability.
+Additionally, most interrupts are directed to the Normal World, requiring costly context switches and complicating secure interrupt handling.
+These constraints, coupled with a relatively large trusted computing base (TCB), hinder the flexibility and security assurances necessary for diverse application scenarios.
+
+Intel SGX adopts a contrasting enclave-based model where enclaves are dynamically created memory regions within user-space processes.
+These enclaves execute at user level (ring 3) and are protected by hardware-enforced memory encryption.
+SGX supports secure communication between the host application and the enclave through ECALL and OCALL interfaces, generated and verified using the edger8r tool.
+While this architecture allows finer-grained protection of sensitive data and logic, SGX lacks supervisor-mode support within enclaves and is tightly coupled to Intel's proprietary hardware platform.
+Furthermore, enclave memory is pre-allocated at system startup with strict size limitations (e.g., 128MB for SGX v1), restricting its scalability for memory-intensive secure workloads.
+
+Keystone combines and extends the architectural principles of both TrustZone and SGX by enabling dynamic enclave creation, hierarchical privilege separation, and a minimal TCB architecture.
+Built on the RISC-V instruction set architecture, Keystone features a dedicated Security Monitor (SM) executing in machine mode (M-mode), which governs enclave lifecycle management and access control.
+Each enclave consists of a user-level application (eapp) and a supervisor-level runtime, allowing greater operational flexibility within the enclave itself.
+Memory for enclaves is dynamically provisioned by the untrusted operating system and isolated using the RISC-V Physical Memory Protection (PMP) mechanism, ensuring strong spatial and temporal isolation.
+
+To enhance portability and interoperability, Keystone supports a partial implementation of the GlobalPlatform TEE Internal API and introduces keyedger, a secure communication interface similar to SGX's edger8r.
+Unlike SGX, however, Keystone is designed to be extensible, enabling the integration of advanced security primitives such as secure I/O, remote attestation, and custom scheduling policies.
+Its open-source nature and hardware-agnostic design position Keystone as a research-friendly and customizable TEE platform, suitable for both academic exploration and real-world deployment scenarios requiring secure, flexible, and verifiable execution environments.
+
 \section{Performance considerations in TEEs}

### 2025-05-28 chapters/methodology.tex
updated relevant chapters & sections

@@
 \section{Benchmarking tools and metrics}
-%\section{Performance metrics}
-The performance characterization of Keystone utilized two established synthetic benchmarks tailored for embedded system evaluation: \textit{Dhrystone} and \textit{CoreMark}.
-Both were selected for their cross-platform compatibility and established use in RISC-V environments.
 
-\textbf{Dhrystone} evaluates integer arithmetic and control flow performance, making it effective for assessing core CPU throughput in environments where floating-point operations are secondary.
-The RISC-V version, sourced from the \texttt{riscv-tests} repository, was employed to calculate \textit{Dhrystones per second (DMIPS)}.
-To isolate Keystone’s overhead, measurements were conducted in both native and enclave contexts.
+The performance characterization of Keystone utilized two widely adopted synthetic benchmarks: \textit{Dhrystone} and \textit{CoreMark}.
+Both benchmarks are designed for embedded system evaluation and are well-supported in RISC-V environments.
+They were chosen to capture the computational characteristics most relevant to TEE deployment, including processor throughput, control flow, and moderate memory usage.
 
-\textbf{CoreMark}, developed by EEMBC, provides a broader performance assessment through representative workloads, including matrix operations, list processing, and state machine evaluations.
-The RISC-V implementation from the SiFive repository was used to determine \textit{iterations per second (IPS)}.
-This benchmark offers a deeper view into system performance under more diverse computational tasks than Dhrystone.
+\textit{Dhrystone} primarily evaluates integer arithmetic and control flow performance, making it suitable for measuring core CPU throughput in environments where floating-point operations and I/O are secondary.
+The RISC-V implementation, sourced from the \texttt{riscv-tests} repository, was used to compute \textit{Dhrystones per second (DMIPS)}.
+Measurements were taken both in native execution and within a Keystone enclave to quantify the performance overhead introduced by enclave isolation.
 
-Performance variability was assessed by calculating the standard deviation across repeated runs. Metrics collected included:
+\textit{CoreMark}, developed by EEMBC, offers a broader performance profile by incorporating common algorithmic workloads such as list processing, matrix manipulation, and finite-state machine evaluation.
+While still CPU-bound, CoreMark introduces moderate memory access patterns through its internal data structures.
+The benchmark was obtained from the SiFive repository and used to calculate \textit{iterations per second (IPS)}.
+
+Performance metrics collected for both benchmarks included:
 \begin{itemize}
-\item Execution Time
-\item DMIPS (Dhrystone) and IPS (CoreMark)
-\item Standard Deviation across trials
+    \item \textbf{Execution Time:} This is the total elapsed wall-clock time taken for a benchmark to complete execution.
+It serves as the most direct measure of performance, indicating how long the CPU and memory resources are engaged.
+Comparing execution times between native and enclave runs highlights the overhead introduced by enclave isolation and security mechanisms.
+
+    \item \textbf{Dhrystones per Second (DMIPS):} Derived from the Dhrystone benchmark, DMIPS represents a normalized measure of integer computing throughput.
+It quantifies how many Dhrystone operations the CPU can perform per second, serving as a standardized metric for CPU integer performance.
+A reduction in DMIPS inside the enclave context signals the performance cost of enclave execution.
+
+    \item \textbf{Iterations per Second (IPS):} Obtained from the CoreMark benchmark, IPS measures the number of complete benchmark iterations executed per second.
+Since CoreMark involves a mixture of algorithmic workloads and memory operations, IPS reflects a broader system performance indicator, capturing both CPU and memory subsystem efficiency under enclave constraints.
+
+    \item \textbf{Standard Deviation:} Calculated across multiple repeated runs of each benchmark, the standard deviation quantifies the variability or consistency of the measured execution times and throughput values.
+Low standard deviation indicates stable and repeatable performance, while higher values may reveal sensitivity to transient system conditions or measurement noise.
+
 \end{itemize}
 
-This dual-benchmark approach ensured a comprehensive evaluation of both low-level CPU operations and higher-level application performance within the enclave context.
+\noindent\textbf{Benchmark Characteristics:}
+\begin{itemize}
+    \item \textit{CPU Intensive:} Both Dhrystone and CoreMark are designed to stress processor logic and control flow.
+Dhrystone focuses exclusively on integer arithmetic, while CoreMark includes a mix of computational tasks without relying on floating-point operations.
+These properties make them ideal for evaluating CPU performance in TEE contexts.
+
+    \item \textit{Memory Interaction:} Although not memory-intensive by design, CoreMark exercises moderate memory usage through data structure manipulation, which may expose latency differences due to enclave-related memory isolation.
+Dhrystone has minimal memory interaction and serves as a purer test of CPU throughput.
+
+    \item \textit{I/O and Storage:} Neither benchmark performs file or persistent storage operations.
+As such, storage-related TEE APIs (e.g., secure storage or object I/O) are not exercised and are outside the scope of this performance evaluation.
+\end{itemize}
+
+This dual-benchmark approach ensures a balanced evaluation of Keystone’s impact on both low-level computational performance and higher-level program behavior in a secure enclave environment.
+
 \section{Parameter Variation Strategy}

### 2025-05-28 main.tex
updated relevant chapters & sections

@@
 This modular structure allowed each benchmark to be executed both natively (outside the enclave) and securely (within the enclave), enabling accurate measurement of the enclave-related overhead.
+
 To achieve a comparative analysis, two execution modes were considered:
@@
 
-\chapter{Conclusion and Future Work}
-\input{chapters/conclusion.tex}
+\chapter{Conclusion}
+%\input{chapters/conclusion.tex}
 
@@
 }
+
+@article{suzaki2021ts,
+  title={Ts-perf: General performance measurement of trusted execution environment and rich execution environment on intel sgx, arm trustzone, and risc-v keystone},
+  author={Suzaki, Kuniyasu and Nakajima, Kenta and Oi, Tsukasa and Tsukamoto, Akira},
+  journal={IEEE Access},
+  volume={9},
+  pages={133520--133530},
+  year={2021},
+  publisher={IEEE}
+}
+
 \end{bibtex-entries}

### 2025-05-29 chapters/background.tex
updated relevant chapters & sections

--- 2025-05-27 chapters/background.tex
@@
 \section{Trusted Execution Environments}
+Trusted Execution Environments (TEEs) represent a class of hardware-assisted secure computing technologies that provide strong guarantees for the confidentiality and integrity of both code and data, even in systems where the underlying operating system or application software may be compromised.
+TEEs are designed to protect sensitive computations from a wide range of adversaries by establishing isolated execution contexts—commonly referred to as enclaves—through a combination of hardware features and privileged system software.
+
+A TEE isolates a subset of the system’s resources, including CPU registers, memory, and I/O peripherals, and makes them accessible only to trusted applications executing within the enclave boundary.
+This isolation is enforced by the processor’s privilege mechanisms, memory protection units, or dedicated security monitors, ensuring that unauthorized entities—whether kernel-level malware or user-space processes—cannot access or tamper with enclave-resident assets.
+
+Security within a TEE is achieved by reducing the size and complexity of the Trusted Computing Base (TCB).
+The TCB encompasses the set of components that must be trusted for the TEE to function securely.
+A minimal TCB is desirable, as it reduces the likelihood of vulnerabilities and simplifies formal verification and auditing processes.
+Modern TEEs aim to move as few components as possible into the TCB, while still providing robust isolation and essential services.
+
+Prominent commercial TEE implementations include Intel Software Guard Extensions (SGX) and ARM TrustZone.
+Intel SGX facilitates the creation of enclaves at the user level, with memory encryption and access control enforced by hardware.
+ARM TrustZone, in contrast, provides a split-world architecture that partitions the processor into a secure world and a normal world, with secure world execution isolated via monitor calls.
+While these solutions have been successfully deployed in many real-world systems, their proprietary nature, rigid designs, and opaque implementations inhibit their utility for research, experimentation, and system-level innovation.
+
+In response to these limitations, open-source TEE frameworks such as Keystone have emerged.
+Keystone is built atop the RISC-V instruction set architecture (ISA), which is itself open, extensible, and amenable to security-centric innovation.
+RISC-V’s modularity enables the design of TEEs that are both highly customizable and more transparent in their operation.
+A foundational architectural principle shared by all TEEs, including Keystone, is privilege separation.
+This concept involves executing the core TCB in a highly privileged execution mode—such as machine mode (M-mode) in RISC-V—while relegating the operating system and general-purpose applications to less privileged levels.
+This separation is crucial for enforcing isolation policies and preventing untrusted software from influencing the secure execution environment.
+
+By leveraging privilege separation, physical memory protection, and modular TCB design, TEEs provide a powerful framework for secure computation in adversarial environments.
+Keystone builds upon these principles and extends them with openness and configurability, making it a valuable platform for academic and practical exploration of TEE technologies.
 
@@
 
+Keystone is an open-source Trusted Execution Environment (TEE) framework designed specifically for the RISC-V architecture.
+It enables the development of secure enclaves that are customizable, lightweight, and suitable for both research and deployment in resource-constrained or security-critical environments.
+Keystone’s design reflects a modular approach, emphasizing minimalism in its Trusted Computing Base (TCB), portability across RISC-V platforms, and extensibility to accommodate advanced security features.
+
+The Keystone architecture operates across the multiple privilege levels defined by the RISC-V specification: user mode (U-mode), supervisor mode (S-mode), and machine mode (M-mode).
+These privilege levels are central to the enforcement of security policies within the system.
+Keystone's Security Monitor (SM), the heart of its TCB, executes in M-mode and holds exclusive authority over enclave lifecycle management, memory protection configuration, and enforcement of access control policies.
+
+A typical Keystone-enabled platform consists of RISC-V processor cores augmented with secure boot mechanisms, trusted entropy sources, and optional hardware accelerators for cryptographic operations or secure I/O.
+Keystone does not mandate specific hardware features beyond RISC-V compliance and PMP (Physical Memory Protection), thereby supporting a broad spectrum of implementations from FPGA prototypes to silicon SoCs.
+
+The SM, implemented in C and compiled as a standalone binary, embodies the critical logic responsible for managing the creation, execution, and destruction of enclaves.
+During enclave creation, the SM validates the enclave's memory layout, initializes its metadata, and configures PMP to ensure that only the SM and the enclave itself can access the designated memory regions.
+Once enclave execution begins, transitions between host and enclave contexts are mediated by the SM, which dynamically adjusts PMP settings to preserve strict isolation.
+
+Each enclave is composed of two software layers: a user-level application (eapp) and a supervisor-level runtime environment.
+The eapp contains application-specific logic, while the runtime provides essential operating system abstractions such as exception handling, system call support, and virtual memory management.
+This two-tiered design allows for flexible enclave development and reduces the complexity of the eapp, which remains entirely untrusted by the host operating system.
+
+The enclave lifecycle encompasses three primary stages: creation, execution, and destruction.
+During creation, the untrusted host allocates a contiguous region of physical memory—known as Enclave Page Memory (EPM)—to hold the enclave’s code, data, and page tables.
+This memory is populated and locked by the SM using PMP.
+In the execution phase, the SM facilitates entry into and exit from the enclave, ensuring that context switches maintain enclave confidentiality.
+Finally, destruction involves securely erasing enclave memory and releasing resources to prevent post-mortem data leakage.
+
+Keystone also supports optional features such as remote attestation, allowing third-party verifiers to authenticate enclave state before provisioning sensitive data.
+Additionally, the architecture can be extended to support secure I/O pathways, cryptographic acceleration, and debugging features useful for both development and runtime monitoring.
+
+\begin{figure}[htbp]
+\centering
+\includegraphics[width=0.9\linewidth]{figures/keystone_overview.png}
+\caption{Overview of the Keystone architecture illustrating components such as the Security Monitor, enclave runtime, and the privilege hierarchy.}
+\label{fig:keystone_overview}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\includegraphics[width=0.9\linewidth]{figures/enclave_lifecycle.png}
+\caption{Stages of a Keystone enclave lifecycle: creation, execution, and destruction.}
+\label{fig:enclave_lifecycle}
+\end{figure}
+
+Overall, Keystone provides a clean, modular, and research-oriented TEE framework that leverages the openness of RISC-V to deliver both strong security guarantees and development flexibility.
+Its layered architecture, minimal TCB, and standards-compliant interfaces make it a compelling platform for building and evaluating secure applications on next-generation processors.
+
 \section{Comparison with other TEEs}
 
+Keystone distinguishes itself from existing commercial Trusted Execution Environment (TEE) implementations, such as Intel Software Guard Extensions (SGX) and ARM TrustZone, through its open-source nature, customizable architecture, and alignment with the RISC-V ecosystem.
+While SGX and TrustZone have been widely deployed in commodity hardware and offer robust isolation features, they are constrained by their proprietary designs, large and opaque TCBs, and limited extensibility, which make them less suitable for academic research and fine-grained system-level customization.
+
+ARM TrustZone utilizes a dual-world execution model, dividing the processor into a Secure World and a Normal World.
+The system transitions between these worlds using secure monitor calls (SMCs).
+Each world operates with its own set of execution contexts and memory spaces.
+In the Secure World, a trusted operating system—such as OP-TEE—executes sensitive applications that conform to the GlobalPlatform TEE Internal API.
+However, TrustZone’s architecture supports only a single secure world instance, lacks support for dynamic enclave creation, and requires significant overhead for world switches, especially in the handling of interrupts.
+Moreover, the TCB in TrustZone is relatively large, encompassing the entire Secure World kernel and trusted services, which complicates formal verification and increases the risk of vulnerabilities.
+
+Intel SGX, in contrast, supports dynamic creation of enclaves at runtime within the address space of user-space processes.
+These enclaves are protected by a memory encryption engine and execute in ring 3, isolated from both the operating system and hypervisor.
+Communication between host and enclave is facilitated by ECALL and OCALL mechanisms, whose interfaces are defined through the edger8r tool.
+While SGX offers strong isolation and cryptographic sealing of enclave contents, it imposes rigid constraints on enclave memory size and structure.
+Notably, SGX enclaves do not support supervisor-level execution, limiting the complexity of software that can be securely encapsulated within the enclave.
+Additionally, SGX is tightly bound to Intel hardware, restricting portability and experimentation.
+
+Keystone integrates the strengths of both TrustZone and SGX while addressing their limitations through an open and modular framework.
+Its architecture supports dynamic enclave creation, hierarchical privilege separation (via user and supervisor modes), and execution control via a minimal M-mode Security Monitor.
+Enclave memory is isolated using RISC-V’s Physical Memory Protection (PMP) features, which are dynamically configured by the SM to enforce strict access control.
+This enables strong spatial and temporal memory isolation without relying on hardware-bound limitations or encryption mechanisms.
+
+Keystone also supports a lightweight communication interface, dubbed keyedger, which serves a similar role to edger8r in SGX, but is designed to be more flexible and extensible.
+While it currently offers partial support for the GlobalPlatform TEE Internal API, Keystone’s modularity allows it to be extended to support custom APIs and protocols suited to specific use cases or research goals.
+
+Through its combination of openness, modularity, and hardware-agnostic design, Keystone offers a uniquely research-friendly and extensible platform for investigating and deploying secure enclaves.
+It enables hardware/software co-design, supports verifiable security primitives, and aligns with emerging open hardware ecosystems, positioning it as a forward-looking alternative to incumbent TEE technologies.
+
 \section{Performance considerations in TEEs}

### 2025-05-29 chapters/introduction.tex
updated relevant chapters & sections

\section{Contributions of this Thesis}
This thesis makes several substantive contributions to the study and evaluation of Trusted Execution Environments (TEEs), with a particular emphasis on the Keystone framework. The research spans architectural analysis, emulation-based deployment, and empirical performance evaluation. The primary contributions are enumerated below:

\begin{itemize}
\item \textbf{Comprehensive Architectural Analysis of Keystone:}
The thesis presents a detailed and systematic examination of the Keystone TEE framework. This includes an in-depth discussion of its core architectural elements, such as the Security Monitor (SM), enclave runtime environment, and user-level application design. The role of RISC-V’s privilege hierarchy and Physical Memory Protection (PMP) in enforcing strict isolation guarantees is elaborated, thereby providing a holistic view of Keystone’s security architecture.
\item \textbf{Comparative Evaluation with Established TEEs:}  
A comparative study is conducted to evaluate Keystone alongside widely deployed commercial TEEs, namely Intel Software Guard Extensions (SGX) and ARM TrustZone. This analysis investigates fundamental architectural distinctions, including enclave creation models, privilege separation strategies, scalability, TCB size, and extensibility. The comparison helps contextualize Keystone’s design philosophy and clarifies its advantages in terms of openness and modularity.

\item \textbf{Virtualized Deployment and Toolchain Integration:}  
The thesis demonstrates the practical deployment of Keystone in a fully virtualized environment using a RISC-V-specific fork of the QEMU emulator. This setup facilitates development and testing without the need for physical RISC-V hardware. It also documents the toolchain configuration, build process, and debugging workflow, thereby offering a reproducible methodology for future research and evaluation.

\item \textbf{Performance Evaluation Using Synthetic Benchmarking:}  
To assess the runtime impact of enclave-based execution, Keystone enclaves are benchmarked using two widely accepted synthetic workloads: Dhrystone and CoreMark. The thesis measures key performance indicators—including execution time, Dhrystones per second (DMIPS), iterations per second (IPS), and standard deviation across runs—to quantify the overhead introduced by enclave isolation and security enforcement.

\item \textbf{Workload Characterization Under Enclave Constraints:}  
The computational behavior of the benchmark applications is analyzed in the context of enclave execution. The findings provide empirical insight into how TEE-induced isolation affects instruction throughput and memory interaction patterns, thus illustrating the performance-security trade-offs inherent to enclave-based models.
\end{itemize}

\section{Related Work}

The field of Trusted Execution Environments has been shaped by a combination of commercial deployments and academic proposals, each addressing different aspects of secure computing. This section situates the present work within the broader landscape of TEE research and implementation.

Intel SGX is one of the most well-known commercial TEE solutions. It introduces user-space enclaves with hardware-enforced isolation and memory encryption. SGX supports fine-grained protection mechanisms and remote attestation, making it suitable for confidential computing scenarios. However, SGX is hindered by several limitations, including rigid enclave memory constraints, lack of supervisor-mode support, and a closed-source implementation that precludes low-level architectural exploration or modification. These issues restrict its applicability in research and experimental systems development.

ARM TrustZone represents a different paradigm, based on the separation of execution into a Secure World and a Normal World. While TrustZone enables trusted operating systems and supports a variety of secure services, it suffers from scalability issues, particularly in multi-core environments where interrupt handling becomes a bottleneck. Moreover, TrustZone's coarse-grained isolation and reliance on a monolithic Secure World kernel contribute to a large TCB, which poses challenges for security assurance and formal verification.

In the academic domain, projects such as Sanctum and Sancus have contributed lightweight and formally verifiable secure execution frameworks. Sanctum builds on RISC-V to enforce secure memory isolation and secure multiplexing of shared hardware resources. Sancus targets embedded systems and emphasizes minimalism and provable security properties. While these efforts have made valuable theoretical contributions, they are often constrained by limited platform support and lack the general-purpose capabilities required for broader deployment or benchmarking.

Keystone addresses the limitations of both commercial and academic TEEs by adopting a fully open-source design built on RISC-V. It supports dynamic enclave creation, configurable runtime environments, and an extensible architecture suitable for incorporating advanced security primitives. Keystone’s modularity, support for supervisor-level enclave components, and compatibility with standard Linux-based host systems make it especially well-suited for research, prototyping, and educational use. This thesis builds upon this foundation by exploring the architectural, practical, and performance aspects of deploying Keystone in a virtualized environment, thereby contributing new empirical insights and documentation to the growing body of work on open TEEs.

\section{Structure of this Thesis}

This thesis is structured to progressively build an understanding of Trusted Execution Environments (TEEs) with a focus on the Keystone framework, moving from conceptual foundations to empirical evaluation and practical system design insights.

\begin{itemize}
    \item \textbf{Chapter 1: Introduction} introduces the motivation for secure computation in untrusted environments, outlines the core contributions of this thesis, and reviews relevant related work in the domain of TEEs.

    \item \textbf{Chapter 2: Background and System Architecture} provides technical foundations for understanding TEEs, focusing on the design of the Keystone framework. It also presents a comparative analysis with Intel SGX and ARM TrustZone, highlighting design trade-offs and performance considerations.

    \item \textbf{Chapter 3: Methodology} outlines the experimental strategy used to evaluate Keystone. It describes the virtualized testbed, benchmark selection, parameter tuning, and data collection procedures used to ensure fair and reproducible performance measurements.

    \item \textbf{Chapter 4: Results and Discussion} presents empirical findings and analyzes the impact of system parameters such as CPU core count, memory allocation, and cache size on enclave performance. It also discusses performance overheads and key insights derived from the results.

    \item \textbf{Chapter 5: System Configuration Recommendations} synthesizes the experimental results into actionable guidelines for deploying Keystone TEEs in different workload scenarios. It identifies architectural bottlenecks and suggests improvements for future system designs.

    \item \textbf{Chapter 6: Conclusion} summarizes the contributions and outcomes of the thesis, reflecting on their implications for secure system design. It also outlines directions for future research, including hardware extensions and formal verification of trusted components.
\end{itemize}

### 2025-05-29 chapters/methodology.tex
updated relevant chapters & sections

--- 2025-05-27 chapters/methodology.tex
@@
 \section{Experimental setup}
+%The experimental setup utilizes the Keystone-enhanced QEMU emulator running on an Ubuntu host operating system.
+This configuration allows for flexible development and testing of Keystone enclaves in a controlled environment without requiring physical RISC-V hardware.
+The Ubuntu host provides a stable platform for building and deploying enclave binaries, and it facilitates the integration of the Keystone kernel driver and SDK.
+
+The experimental setup employed in this study was carefully designed to facilitate reliable, repeatable testing of the Keystone Trusted Execution Environment (TEE) without the need for physical RISC-V hardware.
+To this end, a virtualized testbed was constructed using a RISC-V specific fork of the QEMU emulator, hosted on an Ubuntu 22.04 LTS system.
+This emulated environment provided a high degree of control over the system configuration, allowing for flexible testing of Keystone enclaves in isolation from physical hardware constraints, while still accurately modeling the behavior of a real RISC-V system.
+
+QEMU, an open-source hardware emulator, was configured to emulate the RV64GC (64-bit general-purpose) RISC-V architecture, which is fully compatible with Keystone’s software stack.
+The choice of RV64 as the target architecture reflects Keystone’s intended deployment scenario on modern RISC-V platforms and allows full access to the enclave memory protection and privilege separation mechanisms supported by the Security Monitor.
+
+The host operating system—Ubuntu 22.04—was chosen for its long-term support, stable kernel interfaces, and compatibility with the Keystone development toolchain.
+It served as the platform for building the QEMU binary, compiling the Keystone components, and orchestrating the virtual machine environment in which all benchmarks were executed.
+
+The build process for Keystone relied on a combination of Make and Buildroot.
+Buildroot, an established tool for generating embedded Linux systems, was used to construct the root filesystem, cross-compile essential system libraries, and configure the Linux kernel used in the virtual guest.
+The Make-based build system provided streamlined control over component compilation and deployment, enabling reproducible builds of the Keystone runtime, Security Monitor, and associated kernel modules.
+During the build phase, key configuration parameters were selected, including the choice of QEMU as the target platform, the RV64 architecture, and the specific Buildroot configuration required to support the Keystone kernel driver and enclave runtime environment.
+
+After successful compilation, the Keystone system was launched within QEMU using the command:
+
+\begin{verbatim}
+make run
+\end{verbatim}
+
+This command initiated the full virtual system, including the Keystone Security Monitor—responsible for managing enclave creation, destruction, and memory isolation—as well as a minimal Linux operating system running in supervisor mode within the QEMU guest.
+The Linux environment within QEMU was configured to support manual loading of the Keystone kernel driver, which was necessary to enable enclave functionality.
+Once the system was booted, the kernel driver was inserted using the following command:
+
+\begin{verbatim}
+modprobe keystone-driver
+\end{verbatim}
+
+This step completed the initialization of the enclave infrastructure, allowing benchmark binaries to be securely loaded into enclaves and executed with isolation enforced by the Security Monitor.
+
+For debugging purposes, additional support was enabled through QEMU’s built-in GDB server.
+This allowed low-level inspection of the system state, including memory regions, control registers, and exception handlers.
+When debugging was required, Keystone was recompiled with debugging features enabled by setting the appropriate build flag, and execution was resumed using GDB connectivity:
+
+\begin{verbatim}
+KEYSTONE_DEBUG=y make run
+make debug-connect
+\end{verbatim}
+
+Additional details on the build process, dependencies, and configuration options are available in the \textbf{Appendix}.
 
 \section{Benchmarking tools and metrics}
-%\section{Performance metrics}
+
+To evaluate the performance impact of the Keystone Trusted Execution Environment (TEE) in a meaningful and comprehensive way, this study uses two well-established synthetic benchmarks: \textit{Dhrystone} and \textit{CoreMark}.
+Both are widely used in the embedded systems community and are well-supported in the RISC-V ecosystem, making them a natural fit for assessing how Keystone behaves in secure versus non-secure execution modes.
+These benchmarks were chosen because they capture a wide range of processor behaviors, particularly those relevant to secure enclave execution—such as integer arithmetic, control flow, and memory access patterns.
+
+The Dhrystone benchmark has been around for decades and was originally designed to measure general-purpose processor performance, especially in environments where floating-point operations and file I/O are not a major concern.
+Its focus is on integer calculations and logical control structures—things like loops, function calls, and basic data handling—which are typical in many embedded and real-time applications.
+Because of its simplicity and minimal memory usage, Dhrystone provides a good baseline measure of a processor’s raw computing capability.
+For this study, the RISC-V version of Dhrystone was obtained from the official \texttt{riscv-tests} repository to ensure that it ran correctly on the target hardware.
+The benchmark was executed twice under identical conditions—once in a standard, non-secure mode, and once within a Keystone enclave.
+This setup made it possible to measure the specific performance overhead introduced by Keystone’s security features, such as memory isolation and the cost of switching between secure and non-secure execution contexts.
+
+Dhrystone results are typically expressed in terms of "Dhrystones per second," which simply counts how many iterations of the benchmark are completed in one second.
+To make the numbers more meaningful across different architectures, these results are often normalized to Dhrystone MIPS (DMIPS), which estimates how many millions of instructions per second the processor is effectively delivering.
+This helps smooth out differences between instruction sets and gives a more comparable figure.
+In the context of this study, comparing DMIPS between enclave and non-enclave runs highlights how much raw performance is lost (or retained) when using the TEE.
+
+To get a broader picture of system performance—particularly one that includes moderate memory use and more complex workloads—the CoreMark benchmark was also used.
+Developed by the Embedded Microprocessor Benchmark Consortium (EEMBC), CoreMark was designed to overcome some of the limitations of Dhrystone by providing a workload that’s a bit closer to real-world embedded applications.
+It includes tasks like linked list manipulation, matrix operations, and finite state machine processing, all of which are common in software running on embedded systems.
+While CoreMark is still CPU-focused, it introduces a moderate level of memory interaction through dynamic data structures, which allows it to reveal performance issues that Dhrystone might miss—especially those related to cache usage or memory protection features within a TEE.
+
+The version of CoreMark used in this study came from the SiFive repository, which provides a version optimized for RISC-V systems and ready to work with Keystone.
+Like Dhrystone, CoreMark was run in both native and enclave modes to compare performance under each.
+Its results are reported in terms of "Iterations Per Second" (IPS), which tells us how many complete runs of the benchmark are executed every second.
+Since CoreMark includes both arithmetic and memory operations, IPS gives a more holistic view of how the system is performing—not just the processor, but also the memory subsystem.
+
+Beyond the main performance numbers (DMIPS for Dhrystone and IPS for CoreMark), this study also looked at a few supporting metrics to better understand system behavior.
+The first is overall execution time, which refers to how long it takes (in real-world, wall-clock time) to run each benchmark.
+Comparing execution time in enclave vs.
+non-enclave modes helps quantify latency introduced by secure execution.
+This kind of delay might come from things like extra memory checks, context switching, or other internal mechanisms used to isolate the enclave from the rest of the system.
+
+Another important metric is variability—specifically, the standard deviation of benchmark results across multiple runs.
+Each benchmark was executed several times under identical conditions, and the standard deviation was calculated to determine how consistent the performance was.
+Low standard deviation means the system performs reliably from run to run, while high deviation might indicate that performance is sensitive to system noise, scheduling delays, or other environmental factors.
+
+Both Dhrystone and CoreMark are CPU-intensive, which is exactly what makes them useful in this context.
+Dhrystone focuses narrowly on integer operations and control flow, giving a clean look at raw processor performance.
+CoreMark, on the other hand, introduces a bit more complexity and memory interaction, which makes it better suited for revealing the broader impact of Keystone’s security mechanisms.
+Neither benchmark uses file I/O or persistent storage, which means that this study does not explore storage-related features of the TEE, such as secure disk access or file encryption.
+That’s outside the scope here and would require a different kind of benchmarking approach.
+
+By using both benchmarks together—one simple and focused, the other broader and more realistic—this study creates a well-rounded picture of how Keystone affects performance.
+The addition of execution time and variability analysis provides further context, helping to separate consistent performance overhead from sporadic slowdowns or system noise.
+Overall, this methodology provides a clear, detailed view of the trade-offs involved when securing execution through a TEE on RISC-V platforms.
 
@@
 
+\begin{comment}
+In order to conduct a comprehensive and methodologically sound performance evaluation of the Keystone Trusted Execution Environment (TEE), a structured parameter variation strategy was employed.
+The goal of this strategy was to explore how different system and workload configurations affect performance, both inside and outside of the enclave context.
+By systematically varying a set of key parameters, it becomes possible to isolate the sources of performance overhead and assess the sensitivity of the system to specific operational conditions.
+
+A fundamental aspect of the evaluation involved comparing benchmark execution in two distinct contexts: native execution versus enclave execution.
+In the native configuration, benchmarks were executed directly within the emulated Linux environment running on QEMU, without invoking Keystone’s enclave infrastructure.
+This configuration provided a performance baseline that reflects the raw computational capability of the underlying system without the influence of TEE-specific isolation mechanisms.
+In contrast, enclave execution involved running the same benchmark binaries within Keystone-enforced secure containers.
+This allowed for a direct measurement of the overhead introduced by features such as memory isolation, limited system call interfaces, and the additional control flow transitions required by the Security Monitor.
+
+To ensure statistical significance and reproducibility, each benchmark was executed multiple times under identical conditions.
+Specifically, each configuration (e.g., native or enclave) was run ten times, and the results were aggregated to compute average execution time, throughput (expressed as DMIPS for Dhrystone or IPS for CoreMark), as well as the standard deviation across runs.
+This repetition helped mitigate the impact of transient system noise or timing anomalies that can occur in virtualized environments, especially under QEMU emulation.
+Additionally, the inclusion of standard deviation in the reported results provided insights into the consistency and stability of the enclave performance.
+
+Another important axis of variation was workload size.
+For the CoreMark benchmark, which allows for configurable execution parameters, experiments were conducted using different iteration counts and internal data structure sizes.
+This enabled the evaluation of how enclave performance scales with increasing computational intensity and memory utilization.
+Because enclave memory is subject to isolation and access restrictions, increasing the memory footprint can expose performance bottlenecks that would not be apparent under smaller workloads.
+While Dhrystone does not support direct workload scaling through runtime arguments, multiple versions of the benchmark were compiled with different loop unroll factors to simulate variation in workload complexity.
+Although less precise than CoreMark’s controls, this approach still provided a meaningful indication of how performance trends shift with increasing instruction throughput requirements.
+
+The influence of system-level configuration options was also examined.
+Keystone offers various compile-time and runtime flags that affect behavior and performance, most notably the debug mode.
+When enabled, debug mode activates additional runtime checks, logging mechanisms, and diagnostic outputs intended to aid developers during enclave development and testing.
+These features, while useful during implementation, introduce measurable overhead.
+By executing benchmarks with and without debug mode enabled, the experiments were able to quantify the cost of enhanced observability and determine whether such features significantly degrade performance.
+
+Memory allocation parameters were also varied to understand their role in shaping enclave performance.
+In particular, the heap and stack sizes assigned to benchmarks were adjusted to evaluate the sensitivity of the system to memory layout and availability.
+This was particularly relevant for CoreMark, which dynamically allocates and manipulates internal data structures.
+Larger memory allocations could potentially lead to increased cache pressure or page faults within the enclave, highlighting the importance of efficient memory management in TEE contexts.
+
+Measurement accuracy was further reinforced by cross-validating timing results.
+Execution time was captured using both the benchmarks’ internal timing functions and QEMU’s emulated timer infrastructure.
+This dual-timing approach ensured that timing results were not distorted by emulator artifacts or low-resolution timers.
+In cases where discrepancies were observed, additional investigation was conducted to determine the root cause, and only consistent, corroborated results were included in the final analysis.
+
+Finally, to explore the potential impact of system-level contention, a set of exploratory experiments introduced artificial background load during native execution.
+Although Keystone enclaves currently operate in a single-threaded, non-preemptive mode, testing under load offered preliminary insights into how enclave performance might be influenced in future multi-threaded or multi-tenant scenarios, especially in shared virtualized environments.
+While these tests were not a primary focus of the evaluation, they provided valuable context for interpreting variability in the results.
+\end{comment}
+
 \section{Data Collection and Analysis Procedures}

### 2025-05-29 main.tex
updated relevant chapters & sections

--- 2025-05-27 main.tex
@@
 \UzLStyle{computer modern oldschool design}
+\usepackage{comment}
+\usepackage{float}
+
 \input{preamble.tex} % Put your \UzLStyle and \UzLThesisSetup here
@@
 \chapter{Introduction}
-\input{chapters/01_introduction.tex}
+
+Modern computing increasingly relies on platforms that operate in potentially untrusted environments.
+These include public cloud infrastructures, multi-tenant servers, and distributed edge devices, where sensitive data and critical computations may be exposed to compromised software stacks or malicious actors.
+In such contexts, ensuring the confidentiality and integrity of both data and execution is a fundamental requirement.
+Conventional security mechanisms, such as access control and encryption, are often insufficient in scenarios where the underlying operating system, hypervisor, or even firmware may be compromised.
+
+To address these challenges, Trusted Execution Environments (TEEs) have emerged as a compelling solution.
+TEEs provide hardware-assisted isolation mechanisms that enable the execution of code in a protected context, referred to as an enclave.
+This enclave is designed to be resilient to attacks originating from untrusted system components, including the operating system and other user-space applications.
+By establishing a secure boundary around sensitive code and data, TEEs allow developers to implement trustworthy services even on compromised hosts.
+
+Several commercial implementations of TEEs exist today, with Intel Software Guard Extensions (SGX) and ARM TrustZone being the most prominent examples.
+These technologies have demonstrated the practical viability of hardware-enforced isolation, and they have been integrated into a variety of real-world applications, from digital rights management to secure machine learning inference.
+However, despite their utility, these solutions are inherently proprietary and closed-source.
+This lack of transparency imposes significant constraints on researchers, developers, and system architects who wish to investigate alternative TEE designs, customize security mechanisms, or perform formal verification.
+Moreover, the architectural rigidity of these platforms limits their adaptability to emerging use cases, especially in academic or experimental contexts.
+
+To overcome these limitations, the Keystone framework introduces an open-source, modular TEE architecture built atop the RISC-V instruction set architecture (ISA).
+RISC-V itself is an open, extensible ISA that has gained significant traction in both academic and industrial settings.
+Keystone leverages RISC-V's flexibility to enable fine-grained control over enclave behavior and to support a minimal, verifiable trusted computing base (TCB).
+The framework allows researchers and developers to explore custom enclave policies, experiment with hardware-software co-design, and perform reproducible evaluations in a controlled setting.
+
+This thesis presents a detailed investigation into the Keystone TEE framework.
+It explores Keystone's architectural foundations, its use of RISC-V hardware features such as Physical Memory Protection (PMP), and its support for dynamic enclave management.
+Special attention is paid to the performance implications of enclave isolation, as measured through controlled benchmarking in a virtualized environment.
+By evaluating Keystone under varying execution conditions, this work seeks to provide a rigorous assessment of its efficiency and suitability for secure computation on emerging RISC-V platforms.
+\input{chapters/introduction.tex}
 
 \chapter{Background and System Architecture}
-\input{chapters/02_background.tex}
+The Keystone framework is built on the RISC-V instruction set architecture, which provides a unique foundation for secure system design due to its open specification and modular extensibility.
+Unlike traditional ISAs that are controlled by single vendors, RISC-V encourages community-driven enhancements and supports custom extensions, making it particularly attractive for trusted computing research and development.
+Its clean and well-documented privilege model allows for the implementation of isolation mechanisms that are critical to the construction of TEEs.
+
+RISC-V defines three primary privilege levels: user mode (U-mode), supervisor mode (S-mode), and machine mode (M-mode).
+These privilege levels are hierarchical, with M-mode having the highest authority over system resources.
+M-mode has direct access to all hardware features and is responsible for critical functions such as interrupt control, exception handling, and access to physical memory protection mechanisms.
+Keystone leverages this privilege model by placing its Security Monitor (SM)—the core of the TCB—in M-mode.
+The SM is responsible for managing enclaves, enforcing memory isolation, and mediating access to sensitive system operations.
+By situating the TCB in M-mode, Keystone ensures that enclave isolation policies are enforced at the highest level of privilege, thereby minimizing the attack surface.
+
+A central hardware feature used by Keystone is Physical Memory Protection (PMP), introduced in version 1.10 of the RISC-V Privilege Specification.
+PMP enables M-mode software to define access permissions for specific physical memory regions, controlling which lower-privilege modes (U and S) may read, write, or execute code in those regions.
+This capability forms the basis of Keystone’s memory isolation model.
+Each enclave is allocated a private memory region protected by PMP, ensuring that neither the host operating system nor any unauthorized software can access its contents.
+PMP rules are enforced directly by hardware, providing strong guarantees of isolation that are resistant to software-based attacks.
+
+In addition to memory protection, RISC-V supports a flexible trap and exception handling mechanism.
+M-mode can intercept and handle all traps, but may delegate certain classes of exceptions—such as system calls and page faults—to S-mode for performance optimization.
+This delegation allows the enclave runtime to cooperate with the host operating system for memory management tasks, while still maintaining strict control over enclave boundaries.
+The virtual memory subsystem is managed through the standard RISC-V MMU, which translates virtual addresses to physical addresses using multi-level page tables.
+Keystone leverages this infrastructure to support enclaves that operate with their own private address spaces, protected by PMP from tampering or inspection.
+
+Importantly, the use of standard programming models and toolchains for M-mode software development makes the Keystone Security Monitor both maintainable and extensible.
+Unlike microcode or hardwired logic, which are difficult to audit and update, the SM is implemented in C and can be subjected to conventional testing and formal verification techniques.
+This increases the transparency and trustworthiness of the TCB, which is a critical requirement for secure system design.
+
+Taken together, the combination of RISC-V’s privilege hierarchy, PMP, trap delegation mechanisms, and extensibility provides a solid foundation for implementing robust and configurable TEEs.
+The Keystone framework capitalizes on these features to deliver a flexible and open platform for secure enclave execution, while remaining faithful to the RISC-V philosophy of openness and modularity.
+
+\input{chapters/background.tex}
 
 \chapter{Methodology}
-\input{chapters/03_methodology.tex}
+The primary objective of this thesis is to assess the performance impact of Keystone’s enclave isolation mechanisms on typical embedded and systems-level workloads.
+Specifically, the study seeks to quantify the computational overhead introduced by executing applications within a Keystone enclave, as compared to their execution in a conventional, non-isolated environment.
+To achieve this, a series of carefully controlled experiments were designed and executed within a virtualized RISC-V system based on QEMU.
+
+All benchmarks were executed in an emulated environment configured for the RV64GC architecture, which is representative of 64-bit RISC-V systems targeted by Keystone.
+The use of QEMU for emulation allowed for fine-grained control over the system environment, ensured repeatability, and eliminated the variability introduced by physical hardware, such as thermal throttling or hardware-specific optimizations.
+Although virtualized, the emulation accurately models instruction execution, privilege transitions, memory access patterns, and I/O behavior, making it suitable for preliminary performance evaluation.
+
+To evaluate the impact of enclave execution, each benchmark was implemented as a standalone enclave application (referred to as an “eapp”), accompanied by a corresponding host application that manages enclave lifecycle events.
+The host application, running in user space, is responsible for initializing the enclave, loading the benchmark binary, invoking enclave entry points, and handling edge calls used for communication between the enclave and the host.
+The enclave application, in contrast, is isolated by the Keystone runtime and contains only the benchmark logic and internal data structures.
+This modular separation ensures that the same benchmark code can be executed both natively (without an enclave) and securely (within an enclave) without structural modification, enabling an accurate comparative analysis.
+
+The evaluation involved executing each benchmark—Dhrystone and CoreMark—under two distinct configurations:
+
+\begin{enumerate}
+\item \textit{Native Execution:} The benchmark runs as a conventional user-space process within the QEMU-emulated Linux system, without invoking Keystone enclave services.
+\item \textit{Enclave Execution:} The benchmark is loaded into a Keystone enclave, with isolation enforced by the Security Monitor using PMP.
+\end{enumerate}
+
+Each benchmark was run multiple times (typically ten iterations per configuration) to account for performance variability and ensure statistical robustness.
+For each run, metrics such as total execution time, throughput (measured in DMIPS for Dhrystone and iterations per second for CoreMark), and standard deviation were collected.
+The comparative analysis focused on the relative performance degradation observed in the enclave configuration, thereby providing a direct measurement of the cost of security.
+
+This methodology enables a nuanced understanding of how Keystone’s isolation features influence real-world performance metrics.
+The controlled nature of the test environment, coupled with repeated measurement and the use of industry-standard benchmarks, ensures that the results are both meaningful and reproducible.
+By isolating the effect of enclave execution, this study contributes valuable insights into the trade-offs between security and efficiency in open-source TEE architectures.
+\input{chapters/methodology.tex}
 
 \chapter{Results and discussion}
-\input{chapters/04_results.tex}
+\input{chapters/results.tex}
 
 \chapter{System Configuration Recommendations}
-\input{chapters/05_recommendations.tex}
+\input{chapters/recommendations.tex}
 
-\chapter{Conclusion and Future Work}
-\input{chapters/06_conclusion.tex}
+\chapter{Conclusion}
+%\input{chapters/conclusion.tex}
 

### 2025-06-05 preamble.tex
corrected the name of supervisor
updated relevant chapters & sections

@@
 Betreuerin            = {Dr.-Ing. Saleh Mulhem},
-Mit Unterstützung von = {Henry Strunck},
+Mit Unterstützung von = {Henrik Strunck, M.Sc},
 Studiengang           = {IT Security},
@@
 To assess its performance, the study benchmarks critical system parameters, including the number of CPU cores, cache size, memory allocation, and enclave configuration.
-Using industry-standard benchmarking tools such as lmbench and CoreMark, the research measures execution time, context switch overhead, memory latency, and CPU utilization across a range of hardware and software configurations.
+Using industry-standard benchmarking tools such as Dhrystone and CoreMark, the research measures execution time, context switch overhead, memory latency, and CPU utilization across a range of hardware and software configurations.
 The experimental analysis identifies key performance bottlenecks that impact the efficiency of enclave execution.
@@
 Zur Leistungsbewertung werden zentrale Systemparameter wie die Anzahl der CPU-Kerne, die Cache-Größe, die Speicherzuweisung und die Konfiguration der Enklave untersucht.
-Mithilfe standardisierter Benchmarking-Tools wie lmbench und CoreMark werden Ausführungszeit, Kontextwechsel-Overhead, Speicherlatenz und CPU-Auslastung unter verschiedenen Hardware- und Softwarekonfigurationen gemessen.
+Mithilfe standardisierter Benchmarking-Tools wie Dhrystone und CoreMark werden Ausführungszeit, Kontextwechsel-Overhead, Speicherlatenz und CPU-Auslastung unter verschiedenen Hardware- und Softwarekonfigurationen gemessen.
 Die experimentelle Analyse identifiziert wichtige Leistungsengpässe, die die Effizienz der Enklaven beeinträchtigen.

### 2025-06-06 chapters/methodology.tex
reorganised key sections

@@
+\chapter{Methodology}
+\label{chap:methodology}
+The primary objective of this thesis is to assess the performance impact of Keystone’s enclave isolation mechanisms on typical embedded and systems-level workloads.
+Specifically, the study seeks to quantify the computational overhead introduced by executing applications within a Keystone enclave, as compared to their execution in a conventional, non-isolated environment.
+To achieve this, a series of carefully controlled experiments were designed and executed within a virtualized RISC-V system based on QEMU.
+
+All benchmarks were executed in an emulated environment configured for the RV64GC architecture, which is representative of 64-bit RISC-V systems targeted by Keystone.
+The use of QEMU for emulation allowed for fine-grained control over the system environment, ensured repeatability, and eliminated the variability introduced by physical hardware, such as thermal throttling or hardware-specific optimizations.
+Although virtualized, the emulation accurately models instruction execution, privilege transitions, memory access patterns, and I/O behavior, making it suitable for preliminary performance evaluation.
+
+To evaluate the impact of enclave execution, each benchmark was implemented as a standalone enclave application (referred to as an “eapp”), accompanied by a corresponding host application that manages enclave lifecycle events.
+The host application, running in user space, is responsible for initializing the enclave, loading the benchmark binary, invoking enclave entry points, and handling edge calls used for communication between the enclave and the host.
+The enclave application, in contrast, is isolated by the Keystone runtime and contains only the benchmark logic and internal data structures.
+This modular separation ensures that the same benchmark code can be executed both natively (without an enclave) and securely (within an enclave) without structural modification, enabling an accurate comparative analysis.
+
+The evaluation involved executing each benchmark—Dhrystone and CoreMark—under two distinct configurations:
+
+\begin{enumerate}
+\item \textit{Native Execution:} The benchmark runs as a conventional user-space process within the QEMU-emulated Linux system, without invoking Keystone enclave services.
+\item \textit{Enclave Execution:} The benchmark is loaded into a Keystone enclave, with isolation enforced by the Security Monitor using PMP.
+\end{enumerate}
+
+Each benchmark was run multiple times (typically ten iterations per configuration) to account for performance variability and ensure statistical robustness.
+For each run, metrics such as total execution time, throughput (measured in DMIPS for Dhrystone and iterations per second for CoreMark), and standard deviation were collected.
+The comparative analysis focused on the relative performance degradation observed in the enclave configuration, thereby providing a direct measurement of the cost of security.
+
+This methodology enables a nuanced understanding of how Keystone’s isolation features influence real-world performance metrics.
+The controlled nature of the test environment, coupled with repeated measurement and the use of industry-standard benchmarks, ensures that the results are both meaningful and reproducible.
+By isolating the effect of enclave execution, this study contributes valuable insights into the trade-offs between security and efficiency in open-source TEE architectures.
+
 \section{Experimental setup}

### 2025-06-06 figures/visuals.py
added matplotlib

@@ -1,0 +1,109 @@
+import matplotlib.pyplot as plt
+import numpy as np
+import seaborn as sns
+from scipy.stats import median_abs_deviation, linregress
+
+# Your DMIPS data
+iterations_per_sec = np.array([
+    999560,
+    1007195,
+    1013042,
+    1016369,
+    1001256,
+    973242,
+    962696,
+    994588,
+    983867,
+    984465
+])
+
+dmips = iterations_per_sec / 1757
+runs = np.arange(1, len(dmips) + 1)
+
+# Statistics
+mean = np.mean(dmips)
+median = np.median(dmips)
+std_dev = np.std(dmips, ddof=1)
+mad = median_abs_deviation(dmips, scale='normal')  # robust std dev estimate
+cv = std_dev / mean
+percentiles = np.percentile(dmips, [5, 25, 50, 75, 95])
+
+print(f"Mean: {mean:.3f}")
+print(f"Median: {median:.3f}")
+print(f"Standard Deviation: {std_dev:.3f}")
+print(f"Median Absolute Deviation (MAD): {mad:.3f}")
+print(f"Coefficient of Variation (CV): {cv:.4f}")
+print(f"Percentiles (5th, 25th, 50th, 75th, 95th): {percentiles}")
+
+# Outlier detection using 1.5*IQR
+q1, q3 = np.percentile(dmips, [25, 75])
+iqr = q3 - q1
+lower_bound = q1 - 1.5 * iqr
+upper_bound = q3 + 1.5 * iqr
+outliers = dmips[(dmips < lower_bound) | (dmips > upper_bound)]
+print(f"Outliers: {outliers}")
+
+# Plotting setup
+plt.figure(figsize=(14, 10))
+
+# 1) Bar chart with error bars (mean ± SD)
+plt.subplot(3, 3, 1)
+plt.bar(runs, dmips, color='lightblue', alpha=0.7, label='DMIPS per run')
+plt.errorbar(runs, dmips, yerr=std_dev, fmt='none', ecolor='red', capsize=5, label='±1 SD')
+plt.axhline(mean, color='orange', linestyle='--', label='Mean')
+plt.title('Bar Chart with Error Bars')
+plt.xlabel('Run Number')
+plt.ylabel('DMIPS')
+plt.legend()
+
+# 2) Box plot
+plt.subplot(3, 3, 2)
+sns.boxplot(dmips, color='lightgreen')
[…]

### 2025-06-06 chapters/background.tex

@@
+\chapter{Background and System Architecture}
+\label{chap:background}
+
+The Keystone framework is built on the RISC-V instruction set architecture, which provides a unique foundation for secure system design due to its open specification and modular extensibility.
+Unlike traditional ISAs that are controlled by single vendors, RISC-V encourages community-driven enhancements and supports custom extensions, making it particularly attractive for trusted computing research and development.
+Its clean and well-documented privilege model allows for the implementation of isolation mechanisms that are critical to the construction of TEEs.
+
+RISC-V defines three primary privilege levels: user mode (U-mode), supervisor mode (S-mode), and machine mode (M-mode).
+These privilege levels are hierarchical, with M-mode having the highest authority over system resources.
+M-mode has direct access to all hardware features and is responsible for critical functions such as interrupt control, exception handling, and access to physical memory protection mechanisms.
+Keystone leverages this privilege model by placing its Security Monitor (SM)—the core of the TCB—in M-mode.
+The SM is responsible for managing enclaves, enforcing memory isolation, and mediating access to sensitive system operations.
+By situating the TCB in M-mode, Keystone ensures that enclave isolation policies are enforced at the highest level of privilege, thereby minimizing the attack surface.
+
+A central hardware feature used by Keystone is Physical Memory Protection (PMP), introduced in version 1.10 of the RISC-V Privilege Specification.
+PMP enables M-mode software to define access permissions for specific physical memory regions, controlling which lower-privilege modes (U and S) may read, write, or execute code in those regions.
+This capability forms the basis of Keystone’s memory isolation model.
+Each enclave is allocated a private memory region protected by PMP, ensuring that neither the host operating system nor any unauthorized software can access its contents.
+PMP rules are enforced directly by hardware, providing strong guarantees of isolation that are resistant to software-based attacks.
+
+In addition to memory protection, RISC-V supports a flexible trap and exception handling mechanism.
+M-mode can intercept and handle all traps, but may delegate certain classes of exceptions—such as system calls and page faults—to S-mode for performance optimization.
+This delegation allows the enclave runtime to cooperate with the host operating system for memory management tasks, while still maintaining strict control over enclave boundaries.
+The virtual memory subsystem is managed through the standard RISC-V MMU, which translates virtual addresses to physical addresses using multi-level page tables.
+Keystone leverages this infrastructure to support enclaves that operate with their own private address spaces, protected by PMP from tampering or inspection.
+
+Importantly, the use of standard programming models and toolchains for M-mode software development makes the Keystone Security Monitor both maintainable and extensible.
+Unlike microcode or hardwired logic, which are difficult to audit and update, the SM is implemented in C and can be subjected to conventional testing and formal verification techniques.
+This increases the transparency and trustworthiness of the TCB, which is a critical requirement for secure system design.
+
+Taken together, the combination of RISC-V’s privilege hierarchy, PMP, trap delegation mechanisms, and extensibility provides a solid foundation for implementing robust and configurable TEEs.
+The Keystone framework capitalizes on these features to deliver a flexible and open platform for secure enclave execution, while remaining faithful to the RISC-V philosophy of openness and modularity.
+
 \section{Trusted Execution Environments}
@@
 Keystone’s design reflects a modular approach, emphasizing minimalism in its Trusted Computing Base (TCB), portability across RISC-V platforms, and extensibility to accommodate advanced security features.
+\cite{Lee2019}
 

### 2025-06-06 chapters/introduction.tex

@@
+\chapter{Introduction}
+Modern computing increasingly relies on platforms that operate in potentially untrusted environments.
+These include public cloud infrastructures, multi-tenant servers, and distributed edge devices, where sensitive data and critical computations may be exposed to compromised software stacks or malicious actors.
+In such contexts, ensuring the confidentiality and integrity of both data and execution is a fundamental requirement.
+Conventional security mechanisms, such as access control and encryption, are often insufficient in scenarios where the underlying operating system, hypervisor, or even firmware may be compromised.
+
+To address these challenges, Trusted Execution Environments (TEEs) have emerged as a compelling solution.
+TEEs provide hardware-assisted isolation mechanisms that enable the execution of code in a protected context, referred to as an enclave.
+This enclave is designed to be resilient to attacks originating from untrusted system components, including the operating system and other user-space applications.
+By establishing a secure boundary around sensitive code and data, TEEs allow developers to implement trustworthy services even on compromised hosts.
+
+Several commercial implementations of TEEs exist today, with Intel Software Guard Extensions (SGX) and ARM TrustZone being the most prominent examples.
+These technologies have demonstrated the practical viability of hardware-enforced isolation, and they have been integrated into a variety of real-world applications, from digital rights management to secure machine learning inference.
+However, despite their utility, these solutions are inherently proprietary and closed-source.
+This lack of transparency imposes significant constraints on researchers, developers, and system architects who wish to investigate alternative TEE designs, customize security mechanisms, or perform formal verification.
+Moreover, the architectural rigidity of these platforms limits their adaptability to emerging use cases, especially in academic or experimental contexts.
+
+To overcome these limitations, the Keystone framework introduces an open-source, modular TEE architecture built atop the RISC-V instruction set architecture (ISA).
+RISC-V itself is an open, extensible ISA that has gained significant traction in both academic and industrial settings.
+Keystone leverages RISC-V's flexibility to enable fine-grained control over enclave behavior and to support a minimal, verifiable trusted computing base (TCB).
+The framework allows researchers and developers to explore custom enclave policies, experiment with hardware-software co-design, and perform reproducible evaluations in a controlled setting.
+
+This thesis presents a detailed investigation into the Keystone TEE framework.
+It explores Keystone's architectural foundations, its use of RISC-V hardware features such as Physical Memory Protection (PMP), and its support for dynamic enclave management.
+Special attention is paid to the performance implications of enclave isolation, as measured through controlled benchmarking in a virtualized environment.
+By evaluating Keystone under varying execution conditions, this work seeks to provide a rigorous assessment of its efficiency and suitability for secure computation on emerging RISC-V platforms.
+
 \section{Contributions of this Thesis}

### 2025-06-06 main.tex

--- 2025-05-28 main.tex
@@
 \documentclass[english, version-2020-11]{uzl-thesis}
-\UzLStyle{computer modern oldschool design}
+\UzLStyle{pagella contrast design}
+\usepackage{comment}
 \usepackage{float}
@@
 \input{preamble.tex} % Put your \UzLStyle and \UzLThesisSetup here
+\addbibresource{refs.bib}
 
@@
 
-\chapter{Introduction}
-\input{chapters/introduction.tex}
+%\chapter{Introduction}
 
-\chapter{Background and System Architecture}
-The RISC-V instruction set architecture (ISA) serves as a promising foundation for TEE implementation due to its open and extensible nature.
-RISC-V offers unique benefits beyond its open-source status: it incorporates security-oriented primitives such as Physical Memory Protection (PMP) which facilitates efficient and fine-grained memory isolation.
-Moreover, its community-driven evolution fosters innovation in security features, enabling frameworks like Keystone to experiment with, and contribute to, ISA enhancements.
+\include{chapters/introduction.tex}
 
-RISC-V defines three primary privilege modes: user (U-mode), supervisor (S-mode), and machine (M-mode).
-These modes govern the execution privileges and resource accessibility of software components.
-M-mode, the highest privilege level, has exclusive control over hardware resources and interrupt management, rendering it an ideal candidate for hosting the TCB, such as the Security Monitor (SM) in Keystone.
-Unlike hardware microcode, M-mode software can be developed and patched flexibly using standard programming languages and toolchains, improving maintainability and formal verification potential.
+\include{chapters/background.tex}
 
-A key hardware feature leveraged by Keystone is the PMP, introduced in the RISC-V Privilege Specification version 1.10.
-PMP enables M-mode to enforce physical memory access restrictions on lower privilege levels, thus underpinning enclave memory isolation.
-Despite variations in PMP implementations across platforms, its guarantees are standardized, facilitating portable and secure enclave deployment.
+\include{chapters/methodology.tex}
 
-Interrupt and exception handling in RISC-V is primarily managed at the M-mode level, which may delegate common traps, such as page faults and system calls, to S-mode to optimize performance.
-Virtual memory support is provided by a memory management unit (MMU) that translates virtual addresses to physical addresses using multi-level page tables.
-While the operating system controls virtual memory via the \texttt{satp} register, Keystone enclaves maintain protected page tables and are secured against OS-based attacks through PMP-enforced physical memory restrictions.
+\include{chapters/results.tex}
 
-Collectively, these architectural features establish a robust foundation for implementing secure and efficient TEEs on RISC-V platforms, as exemplified by Keystone.
+\include{chapters/recommendations.tex}
 
-\input{chapters/background.tex}
+\include{chapters/conclusion.tex}
 
-\chapter{Methodology}
-The objective of the performance study was to quantify the overhead introduced by Keystone's enclave isolation mechanisms on RISC-V architectures.
-Experiments were executed within a QEMU-based virtual environment configured for the RV64 architecture, providing consistency and hardware independence.
-
-To run benchmarks, a separate enclave application (eapp) and corresponding host application were developed for each benchmark.
-The host application initialized and managed the enclave using the Keystone C++ API, while the enclave application contained the benchmark logic and communicated with the host through edge calls.
-This modular structure allowed each benchmark to be executed both natively (outside the enclave) and securely (within the enclave), enabling accurate measurement of the enclave-related overhead.
-To achieve a comparative analysis, two execution modes were considered:
-\begin{enumerate}
-\item \textit{Native Execution}: Benchmark execution without enclave isolation.
-\item \textit{Enclave Execution}: Benchmark execution within the Keystone TEE.
-\end{enumerate}
-
-Each benchmark—Dhrystone and CoreMark—was run multiple times under both conditions.
-Performance metrics, including total execution time, throughput (DMIPS or IPS), and result variability (standard deviation), were recorded and analyzed.
-
-The evaluation focused on determining the relative performance degradation attributable to enclave isolation by comparing results across the two execution modes.
-
-\input{chapters/methodology.tex}
-
-\chapter{Results and discussion}
-\input{chapters/results.tex}
-
-\chapter{System Configuration Recommendations}
-\input{chapters/recommendations.tex}
-
-\chapter{Conclusion and Future Work}
-\input{chapters/conclusion.tex}
-
-\begin{bibtex-entries}
-@TechReport{Kernighan1974,
-author = {Brian Kernighan},
-title = {Programming in C – A Tutorial},
-institution = {Bell Laboratories},
-year = {1974}
-}
-\end{bibtex-entries}
 \end{document}

### 2025-06-06 figures/visuals.py
added matplotlib

@@ -107,3 +107,4 @@ plt.tight_layout()
 # Save the figure as a PNG file
 plt.savefig('dmips_plot.png', dpi=300)  # You can change the filename and dpi as you like
 plt.show()
+

### 2025-06-06 chapters/introduction.tex
replaced enumerated points with cleaner, more fluid paragraphs

@@
 \chapter{Introduction}
+\label{chap:intro}
 Modern computing increasingly relies on platforms that operate in potentially untrusted environments.
@@
 \section{Contributions of this Thesis}
-This thesis makes several substantive contributions to the study and evaluation of Trusted Execution Environments (TEEs), with a particular emphasis on the Keystone framework.
-The research spans architectural analysis, emulation-based deployment, and empirical performance evaluation.
-The primary contributions are enumerated below:
 
-\begin{itemize}
-\item \textbf{Comprehensive Architectural Analysis of Keystone:}
-The thesis presents a detailed and systematic examination of the Keystone TEE framework.
-This includes an in-depth discussion of its core architectural elements, such as the Security Monitor (SM), enclave runtime environment, and user-level application design.
-The role of RISC-V’s privilege hierarchy and Physical Memory Protection (PMP) in enforcing strict isolation guarantees is elaborated, thereby providing a holistic view of Keystone’s security architecture.
-\item \textbf{Comparative Evaluation with Established TEEs:}  
-A comparative study is conducted to evaluate Keystone alongside widely deployed commercial TEEs, namely Intel Software Guard Extensions (SGX) and ARM TrustZone.
-This analysis investigates fundamental architectural distinctions, including enclave creation models, privilege separation strategies, scalability, TCB size, and extensibility.
-The comparison helps contextualize Keystone’s design philosophy and clarifies its advantages in terms of openness and modularity.
+This thesis makes significant contributions to the study and evaluation of Trusted Execution Environments (TEEs), with a particular focus on the Keystone framework.
+The research encompasses architectural analysis, emulation-based deployment, and rigorous empirical performance evaluation.
 
-\item \textbf{Virtualized Deployment and Toolchain Integration:}  
-The thesis demonstrates the practical deployment of Keystone in a fully virtualized environment using a RISC-V-specific fork of the QEMU emulator.
-This setup facilitates development and testing without the need for physical RISC-V hardware.
-It also documents the toolchain configuration, build process, and debugging workflow, thereby offering a reproducible methodology for future research and evaluation.
+Firstly, a comprehensive architectural analysis of the Keystone TEE framework is presented.
+This includes an in-depth discussion of its core components, such as the Security Monitor (SM), enclave runtime environment, and user-level application design.
+The role of RISC-V’s privilege hierarchy and Physical Memory Protection (PMP) in enforcing strict isolation guarantees is thoroughly elaborated, providing a holistic understanding of Keystone’s security architecture.
 
-\item \textbf{Performance Evaluation Using Synthetic Benchmarking:}  
-To assess the runtime impact of enclave-based execution, Keystone enclaves are benchmarked using two widely accepted synthetic workloads: Dhrystone and CoreMark.
-The thesis measures key performance indicators—including execution time, Dhrystones per second (DMIPS), iterations per second (IPS), and standard deviation across runs—to quantify the overhead introduced by enclave isolation and security enforcement.
+Secondly, the thesis conducts a rigorous comparative evaluation of Keystone alongside widely deployed commercial TEEs, specifically Intel Software Guard Extensions (SGX) and ARM TrustZone.
+This analysis highlights fundamental architectural distinctions, including enclave creation models, privilege separation strategies, scalability, Trusted Computing Base (TCB) size, and extensibility.
+Such comparison contextualizes Keystone’s design philosophy and underscores its advantages in openness and modularity.
 
-\item \textbf{Workload Characterization Under Enclave Constraints:}  
-The computational behavior of the benchmark applications is analyzed in the context of enclave execution.
-The findings provide empirical insight into how TEE-induced isolation affects instruction throughput and memory interaction patterns, thus illustrating the performance-security trade-offs inherent to enclave-based models.
-\end{itemize}
+Furthermore, the practical deployment of Keystone within a fully virtualized environment is demonstrated using a RISC-V-specific fork of the QEMU emulator.
+This approach facilitates development and testing independent of physical RISC-V hardware.
+The thesis also documents the toolchain configuration, build process, and debugging workflow, establishing a reproducible methodology for future research.
+
+To quantify the runtime overhead associated with enclave-based execution, Keystone enclaves are evaluated using two established synthetic benchmarks: Dhrystone and CoreMark.
+Key performance metrics—including execution time, Dhrystones per second (DMIPS), iterations per second (IPS), and variability across multiple runs—are measured to assess the cost of enclave isolation and security enforcement.
+
+Finally, the computational behavior of benchmark applications within the enclave context is analyzed to provide empirical insights into how TEE-induced isolation affects instruction throughput and memory access patterns.
+These findings illuminate the inherent trade-offs between performance and security in enclave-based execution models.
 
@@
 
-\begin{itemize}
-    \item \textbf{Chapter 1: Introduction} introduces the motivation for secure computation in untrusted environments, outlines the core contributions of this thesis, and reviews relevant related work in the domain of TEEs.
-
-    \item \textbf{Chapter 2: Background and System Architecture} provides technical foundations for understanding TEEs, focusing on the design of the Keystone framework.
-It also presents a comparative analysis with Intel SGX and ARM TrustZone, highlighting design trade-offs and performance considerations.
-
-    \item \textbf{Chapter 3: Methodology} outlines the experimental strategy used to evaluate Keystone.
-It describes the virtualized testbed, benchmark selection, parameter tuning, and data collection procedures used to ensure fair and reproducible performance measurements.
-
-    \item \textbf{Chapter 4: Results and Discussion} presents empirical findings and analyzes the impact of system parameters such as CPU core count, memory allocation, and cache size on enclave performance.
-It also discusses performance overheads and key insights derived from the results.
-
-    \item \textbf{Chapter 5: System Configuration Recommendations} synthesizes the experimental results into actionable guidelines for deploying Keystone TEEs in different workload scenarios.
-It identifies architectural bottlenecks and suggests improvements for future system designs.
-
-    \item \textbf{Chapter 6: Conclusion} summarizes the contributions and outcomes of the thesis, reflecting on their implications for secure system design.
-It also outlines directions for future research, including hardware extensions and formal verification of trusted components.
-\end{itemize}
+This thesis is organized as follows: Chapter~\ref{chap:intro} introduces the motivation for secure computation in untrusted environments, outlines the core contributions of this work, and surveys relevant literature in the domain of Trusted Execution Environments (TEEs).
+Chapter~\ref{chap:background} presents the technical foundations and system architecture of the Keystone framework, including a comparative analysis with Intel SGX and ARM TrustZone that highlights key design trade-offs and performance considerations.
+Chapter~\ref{chap:methodology} describes the experimental setup, detailing the virtualized testbed, benchmark selection, parameter tuning, and data collection methods employed to ensure rigorous and reproducible evaluation.
+Chapter~\ref{chap:results} reports empirical findings and analyzes the impact of system parameters such as CPU core count, memory allocation, and cache size on enclave performance, along with associated overheads and insights.
+Chapter~\ref{chap:recommendations} synthesizes these results into practical configuration guidelines for deploying Keystone TEEs across diverse workload scenarios and identifies architectural bottlenecks for future improvements.
+Finally, Chapter~\ref{chap:conclusion} summarizes the main contributions and discusses their implications for secure system design, outlining promising directions for further research including hardware extensions and formal verification.

### 2025-06-08 chapters/background.tex
updated references (still in progress)

@@
 
-\section{Keystone architecture overview}
+\section{Keystone Architecture Overview}
 
-Keystone is an open-source Trusted Execution Environment (TEE) framework designed specifically for the RISC-V architecture.
-It enables the development of secure enclaves that are customizable, lightweight, and suitable for both research and deployment in resource-constrained or security-critical environments.
-Keystone’s design reflects a modular approach, emphasizing minimalism in its Trusted Computing Base (TCB), portability across RISC-V platforms, and extensibility to accommodate advanced security features.
-\cite{Lee2019}
+Keystone is an open-source Trusted Execution Environment (TEE) framework built on the RISC-V architecture, designed to enable customizable, lightweight, and secure enclaves suitable for both research and deployment in resource-constrained or security-critical settings \cite{dayeol2019keystone}.
+The framework’s modular design emphasizes a minimal Trusted Computing Base (TCB), portability across diverse RISC-V platforms, and extensibility to accommodate advanced security features, reflecting a clean and flexible architectural philosophy.
 
-The Keystone architecture operates across the multiple privilege levels defined by the RISC-V specification: user mode (U-mode), supervisor mode (S-mode), and machine mode (M-mode).
-These privilege levels are central to the enforcement of security policies within the system.
-Keystone's Security Monitor (SM), the heart of its TCB, executes in M-mode and holds exclusive authority over enclave lifecycle management, memory protection configuration, and enforcement of access control policies.
+Keystone leverages the multiple privilege modes defined by the RISC-V specification—user mode (U-mode), supervisor mode (S-mode), and machine mode (M-mode)—to enforce security policies effectively.
+At the core of Keystone’s TCB is the Security Monitor (SM), which executes in M-mode and holds exclusive authority over enclave lifecycle management, memory protection configuration, and access control enforcement.
+The SM uses RISC-V’s Physical Memory Protection (PMP) mechanism to isolate enclave memory regions dynamically, ensuring strict access controls during enclave execution \cite{dayeol2019keystone}.
 
-A typical Keystone-enabled platform consists of RISC-V processor cores augmented with secure boot mechanisms, trusted entropy sources, and optional hardware accelerators for cryptographic operations or secure I/O.
-Keystone does not mandate specific hardware features beyond RISC-V compliance and PMP (Physical Memory Protection), thereby supporting a broad spectrum of implementations from FPGA prototypes to silicon SoCs.
+A typical Keystone-enabled platform comprises RISC-V cores augmented with secure boot mechanisms, trusted entropy sources, and optionally, hardware accelerators for cryptographic functions or secure I/O.
+Importantly, Keystone does not require hardware beyond standard RISC-V compliance and PMP support, making it adaptable to implementations ranging from FPGA prototypes to silicon SoCs \cite{dayeol2019keystone}.
 
-The SM, implemented in C and compiled as a standalone binary, embodies the critical logic responsible for managing the creation, execution, and destruction of enclaves.
-During enclave creation, the SM validates the enclave's memory layout, initializes its metadata, and configures PMP to ensure that only the SM and the enclave itself can access the designated memory regions.
-Once enclave execution begins, transitions between host and enclave contexts are mediated by the SM, which dynamically adjusts PMP settings to preserve strict isolation.
+Each enclave in Keystone consists of two main software components: a user-level enclave application (eapp) containing application-specific logic, and a supervisor-level runtime (RT) providing essential operating system abstractions such as exception handling, system call services, and memory management.
+This layered approach reduces the complexity of enclave applications and provides a trusted runtime environment isolated from the untrusted operating system and other processes \cite{dayeol2019keystone}.
 
-Each enclave is composed of two software layers: a user-level application (eapp) and a supervisor-level runtime environment.
-The eapp contains application-specific logic, while the runtime provides essential operating system abstractions such as exception handling, system call support, and virtual memory management.
-This two-tiered design allows for flexible enclave development and reduces the complexity of the eapp, which remains entirely untrusted by the host operating system.
-
-The enclave lifecycle encompasses three primary stages: creation, execution, and destruction.
-During creation, the untrusted host allocates a contiguous region of physical memory—known as Enclave Page Memory (EPM)—to hold the enclave’s code, data, and page tables.
-This memory is populated and locked by the SM using PMP.
-In the execution phase, the SM facilitates entry into and exit from the enclave, ensuring that context switches maintain enclave confidentiality.
-Finally, destruction involves securely erasing enclave memory and releasing resources to prevent post-mortem data leakage.
-
-Keystone also supports optional features such as remote attestation, allowing third-party verifiers to authenticate enclave state before provisioning sensitive data.
-Additionally, the architecture can be extended to support secure I/O pathways, cryptographic acceleration, and debugging features useful for both development and runtime monitoring.
+The enclave lifecycle is structured into three primary phases: creation, execution, and destruction.
+During creation, the SM validates and locks the enclave’s memory layout—referred to as Enclave Page Memory (EPM)—with PMP to ensure confidentiality and integrity.
+During execution, the SM mediates context switches between the host and enclave, dynamically adjusting PMP to maintain isolation.
+Destruction securely erases enclave memory to prevent any residual data leakage \cite{dayeol2019keystone}.
+Keystone also supports optional features such as remote attestation, allowing external verifiers to authenticate enclave state prior to provisioning sensitive data, and can be extended for secure I/O and cryptographic acceleration.
 
@@
 
-Overall, Keystone provides a clean, modular, and research-oriented TEE framework that leverages the openness of RISC-V to deliver both strong security guarantees and development flexibility.
-Its layered architecture, minimal TCB, and standards-compliant interfaces make it a compelling platform for building and evaluating secure applications on next-generation processors.
+Overall, Keystone’s modular and open framework leverages the openness of RISC-V to provide strong security guarantees while enabling flexibility for research and customization.
+Its minimal TCB, modular layering (Security Monitor, runtime, eapp), and standards-compliant interfaces make it an attractive platform for developing and evaluating secure applications on modern processors \cite{dayeol2019keystone}.
 
-\section{Comparison with other TEEs}
+\section{Comparison with Other TEEs}
 
-Keystone distinguishes itself from existing commercial Trusted Execution Environment (TEE) implementations, such as Intel Software Guard Extensions (SGX) and ARM TrustZone, through its open-source nature, customizable architecture, and alignment with the RISC-V ecosystem.
-While SGX and TrustZone have been widely deployed in commodity hardware and offer robust isolation features, they are constrained by their proprietary designs, large and opaque TCBs, and limited extensibility, which make them less suitable for academic research and fine-grained system-level customization.
+Keystone’s architecture sets it apart from existing commercial Trusted Execution Environments such as Intel Software Guard Extensions (SGX) and ARM TrustZone through its openness, modularity, and alignment with the RISC-V ecosystem \cite{dayeol2019keystone}.
+Unlike the proprietary and largely opaque designs of SGX and TrustZone, Keystone offers a fully open-source platform that supports fine-grained system-level customization and experimental research.
 
-ARM TrustZone utilizes a dual-world execution model, dividing the processor into a Secure World and a Normal World.
-The system transitions between these worlds using secure monitor calls (SMCs).
-Each world operates with its own set of execution contexts and memory spaces.
-In the Secure World, a trusted operating system—such as OP-TEE—executes sensitive applications that conform to the GlobalPlatform TEE Internal API.
-However, TrustZone’s architecture supports only a single secure world instance, lacks support for dynamic enclave creation, and requires significant overhead for world switches, especially in the handling of interrupts.
-Moreover, the TCB in TrustZone is relatively large, encompassing the entire Secure World kernel and trusted services, which complicates formal verification and increases the risk of vulnerabilities.
+ARM TrustZone implements a dual-world execution model that partitions the processor into a Secure World and a Normal World, with transitions managed via secure monitor calls (SMCs).
+While TrustZone provides a trusted OS environment (e.g., OP-TEE) within the Secure World, it supports only a single secure instance and requires significant overhead for context switches and interrupt handling.
+Moreover, the TrustZone TCB encompasses the entire Secure World kernel and trusted services, which increases complexity and hinders formal verification \cite{dayeol2019keystone}.
 
-Intel SGX, in contrast, supports dynamic creation of enclaves at runtime within the address space of user-space processes.
-These enclaves are protected by a memory encryption engine and execute in ring 3, isolated from both the operating system and hypervisor.
-Communication between host and enclave is facilitated by ECALL and OCALL mechanisms, whose interfaces are defined through the edger8r tool.
-While SGX offers strong isolation and cryptographic sealing of enclave contents, it imposes rigid constraints on enclave memory size and structure.
-Notably, SGX enclaves do not support supervisor-level execution, limiting the complexity of software that can be securely encapsulated within the enclave.
-Additionally, SGX is tightly bound to Intel hardware, restricting portability and experimentation.
+Intel SGX allows dynamic creation of enclaves at runtime within user-mode processes and provides memory encryption to isolate enclave data from the OS and hypervisor.
+However, SGX enclaves are confined to user mode and cannot run supervisor-level code, limiting their expressiveness.
+Additionally, SGX enforces rigid constraints on enclave size and structure, and its reliance on Intel hardware restricts portability and adaptability \cite{dayeol2019keystone}.
 
-Keystone integrates the strengths of both TrustZone and SGX while addressing their limitations through an open and modular framework.
-Its architecture supports dynamic enclave creation, hierarchical privilege separation (via user and supervisor modes), and execution control via a minimal M-mode Security Monitor.
-Enclave memory is isolated using RISC-V’s Physical Memory Protection (PMP) features, which are dynamically configured by the SM to enforce strict access control.
-This enables strong spatial and temporal memory isolation without relying on hardware-bound limitations or encryption mechanisms.
+Keystone synthesizes the advantages of both approaches while overcoming their limitations.
+It supports dynamic enclave creation and hierarchical privilege separation by utilizing both user and supervisor modes, controlled by a minimal M-mode Security Monitor.
+The use of PMP allows Keystone to enforce spatial and temporal memory isolation without the complexity of encryption or hardware-bound restrictions \cite{dayeol2019keystone}.
 
-Keystone also supports a lightweight communication interface, dubbed keyedger, which serves a similar role to edger8r in SGX, but is designed to be more flexible and extensible.
-While it currently offers partial support for the GlobalPlatform TEE Internal API, Keystone’s modularity allows it to be extended to support custom APIs and protocols suited to specific use cases or research goals.
+Communication between the host and enclaves is facilitated by a lightweight interface named keyedger, akin to SGX’s edger8r but designed for greater flexibility and extensibility.
+Keystone partially supports the GlobalPlatform TEE Internal API and is architected to accommodate custom APIs, enhancing its utility for diverse research and deployment scenarios \cite{dayeol2019keystone}.
 
-Through its combination of openness, modularity, and hardware-agnostic design, Keystone offers a uniquely research-friendly and extensible platform for investigating and deploying secure enclaves.
-It enables hardware/software co-design, supports verifiable security primitives, and aligns with emerging open hardware ecosystems, positioning it as a forward-looking alternative to incumbent TEE technologies.
+Through its combination of openness, modularity, and hardware-agnostic design, Keystone offers a uniquely research-friendly and extensible platform for secure enclave development.
+It enables hardware/software co-design, supports verifiable security primitives, and aligns with emerging open hardware ecosystems, positioning itself as a forward-looking alternative to existing TEE technologies \cite{dayeol2019keystone}.
 
 \section{Performance considerations in TEEs}
+
+Trusted Execution Environments (TEEs) provide strong isolation guarantees for sensitive computations, but often at the cost of performance overheads.
+These overheads arise from architectural constraints, such as context switching, secure memory management, and the need to isolate execution from potentially untrusted system components.
+Evaluating these overheads is crucial for determining the feasibility of TEE deployment in latency-sensitive or resource-constrained applications.
+
+Accurate time measurement within TEEs presents a significant challenge.
+Standard API functions such as \texttt{TEE\_GetREETime()} incur high latencies due to context switching into the Rich Execution Environment (REE).
+On the other hand, functions like \texttt{TEE\_GetSystemTime()}, which access hardware timers within the TEE, exhibit significantly lower and more consistent overhead.
+In comparative tests, \texttt{TEE\_GetSystemTime()} was approximately 30 times faster on average than \texttt{TEE\_GetREETime()} across various architectures including ARM TrustZone, Intel SGX, and RISC-V Keystone \cite{Suzaki2021}.
+
+Benchmark results further confirm that CPU-intensive workloads experience minimal overhead within TEEs, as these operations typically rely on core-local instruction pipelines.
+For example, integer and floating-point multiplications executed within a TEE yielded results comparable to their REE counterparts across platforms, with ARM TrustZone showing a modest 3–4\% slowdown \cite{Suzaki2021}.
+These results were consistent with Keystone benchmarks using RV8 and CoreMark, which reported sub-1\% overhead for CPU-bound tasks when enclave lifecycle costs were excluded \cite{dayeol2019keystone}.
+
+In contrast, memory and I/O operations often incur greater penalties.
+In TS-perf benchmarks, ARM TrustZone showed an 11\% performance drop in random memory access, attributed to memory encryption and architectural characteristics such as cache hierarchy.
+Further experiments revealed that performance degradation in Intel SGX becomes significant only after memory footprints exceed the L2 cache size, highlighting cache capacity as a mitigating factor \cite{Suzaki2021}.
+
+Keystone's RV8 benchmarks show that memory-heavy workloads with large working sets (e.g., \texttt{primes}, \texttt{miniz}, \texttt{aes}) can experience overheads exceeding 100\% when cache partitioning is enabled, due to reduced effective cache capacity and increased eviction rates.
+Conversely, applications with small memory footprints (e.g., \texttt{sha512}, \texttt{dhrystone}) demonstrated negligible impact from cache-related protections \cite{dayeol2019keystone}.
+This suggests that performance optimization in TEEs requires close alignment between application memory behavior and enclave hardware configuration.
+
+Storage performance is another critical consideration.
+TEE implementations typically must tunnel file I/O through the REE, using secure APIs that copy data into untrusted buffers and invoke enclave exit calls.
+This indirection introduces substantial overhead.
+In Keystone, file operations using IOZone showed throughput reductions of 36.2\% (write) and 40.9\% (read) on average, with even higher penalties on record re-access patterns (e.g., 55.1\% on record rewrite) \cite{dayeol2019keystone}.
+TS-perf results confirm similar patterns, with OP-TEE on ARM and SGX showing large performance disparities between REE and TEE storage operations due to the reliance on OCALLs and limited TEE-side file APIs \cite{Suzaki2021}.
+
+Furthermore, platform-specific runtime behavior can introduce variability in performance.
+For example, htop monitoring during TS-perf experiments revealed unexpected behavior in thread scheduling: TEE applications sometimes migrated between CPU cores or were scheduled differently from their REE parent processes.
+This inconsistency suggests a lack of TEE-awareness in existing OS schedulers, particularly in OP-TEE and Keystone.
+In some cases, CPU load dropped to zero despite active enclave execution, later traced to interrupt handling bugs in RISC-V’s eyrie runtime \cite{Suzaki2021}.
+
+Finally, enclave lifecycle operations, while one-time costs, also contribute to overall performance.
+In Keystone, enclave creation and destruction involve validation, hashing (e.g., SHA-3), and context setup, incurring 2M–7M cycles per page for validation and 20k–30k cycles for lifecycle events.
+These costs are significantly affected by software-based implementations of cryptographic primitives and could be optimized using hardware acceleration \cite{dayeol2019keystone}.
+
+In summary, while TEEs offer minimal performance penalties for compute-bound workloads, substantial overheads can emerge for memory- and I/O-bound applications.
+These overheads are highly platform-dependent and are influenced by factors such as cache behavior, buffer copying, interrupt handling, and runtime design.
+Hence, effective use of TEEs requires not only security-aware application design but also performance-aware system integration.

### 2025-06-11 chapters/methodology.tex
updated sections 3.3, 3.4 & references (still in progress)

@@
 
-To get a broader picture of system performance—particularly one that includes moderate memory use and more complex workloads—the CoreMark benchmark was also used.
-Developed by the Embedded Microprocessor Benchmark Consortium (EEMBC), CoreMark was designed to overcome some of the limitations of Dhrystone by providing a workload that’s a bit closer to real-world embedded applications.
+To get a broader picture of system performance—particularly one that includes moderate memory use and more complex workloads—the CoreMark benchmark \cite{gal2012exploring} was also used.
+Developed by the Embedded Microprocessor Benchmark Consortium (EEMBC), CoreMark \cite{gal2012exploring} was designed to overcome some of the limitations of Dhrystone by providing a workload that’s a bit closer to real-world embedded applications.
 It includes tasks like linked list manipulation, matrix operations, and finite state machine processing, all of which are common in software running on embedded systems.
-While CoreMark is still CPU-focused, it introduces a moderate level of memory interaction through dynamic data structures, which allows it to reveal performance issues that Dhrystone might miss—especially those related to cache usage or memory protection features within a TEE.
+While CoreMark \cite{gal2012exploring} is still CPU-focused, it introduces a moderate level of memory interaction through dynamic data structures, which allows it to reveal performance issues that Dhrystone might miss—especially those related to cache usage or memory protection features within a TEE.
 
-The version of CoreMark used in this study came from the SiFive repository, which provides a version optimized for RISC-V systems and ready to work with Keystone.
-Like Dhrystone, CoreMark was run in both native and enclave modes to compare performance under each.
+The version of CoreMark \cite{gal2012exploring} used in this study came from the SiFive repository available on GitHub at \url{https://github.com/sifive/benchmark-coremark}, which provides a version optimized for RISC-V systems and ready to work with Keystone.
+Like Dhrystone \cite{weiss2002dhrystone}, CoreMark \cite{gal2012exploring} was run in both native and enclave modes to compare performance under each.
 Its results are reported in terms of "Iterations Per Second" (IPS), which tells us how many complete runs of the benchmark are executed every second.
@@
 \section{Parameter Variation Strategy}
+\label{sec:param-variation}
 
-\begin{comment}
-In order to conduct a comprehensive and methodologically sound performance evaluation of the Keystone Trusted Execution Environment (TEE), a structured parameter variation strategy was employed.
-The goal of this strategy was to explore how different system and workload configurations affect performance, both inside and outside of the enclave context.
-By systematically varying a set of key parameters, it becomes possible to isolate the sources of performance overhead and assess the sensitivity of the system to specific operational conditions.
+To examine the influence of microarchitectural cache parameters on enclave performance, this study employed a structured variation of Level-1 (L1) cache configurations using QEMU's cache modeling capabilities.
+The experimental objective was to isolate the impact of cache size, associativity, eviction policy, and core count on the performance of secure and non-secure workloads in a controlled emulation environment.
 
-A fundamental aspect of the evaluation involved comparing benchmark execution in two distinct contexts: native execution versus enclave execution.
-In the native configuration, benchmarks were executed directly within the emulated Linux environment running on QEMU, without invoking Keystone’s enclave infrastructure.
-This configuration provided a performance baseline that reflects the raw computational capability of the underlying system without the influence of TEE-specific isolation mechanisms.
-In contrast, enclave execution involved running the same benchmark binaries within Keystone-enforced secure containers.
-This allowed for a direct measurement of the overhead introduced by features such as memory isolation, limited system call interfaces, and the additional control flow transitions required by the Security Monitor.
+Cache modeling was enabled through the QEMU TCG plugin framework \cite{mandour2021cache}, which simulates idealized, per-core instruction and data caches.
+While not microarchitecturally cycle-accurate, this plugin supports a configurable cache hierarchy, enabling systematic exploration of cache behaviors relevant to trusted execution environments.
 
-To ensure statistical significance and reproducibility, each benchmark was executed multiple times under identical conditions.
-Specifically, each configuration (e.g., native or enclave) was run ten times, and the results were aggregated to compute average execution time, throughput (expressed as DMIPS for Dhrystone or IPS for CoreMark), as well as the standard deviation across runs.
-This repetition helped mitigate the impact of transient system noise or timing anomalies that can occur in virtualized environments, especially under QEMU emulation.
-Additionally, the inclusion of standard deviation in the reported results provided insights into the consistency and stability of the enclave performance.
+Four key experimental scenarios were defined:
 
-Another important axis of variation was workload size.
-For the CoreMark benchmark, which allows for configurable execution parameters, experiments were conducted using different iteration counts and internal data structure sizes.
-This enabled the evaluation of how enclave performance scales with increasing computational intensity and memory utilization.
-Because enclave memory is subject to isolation and access restrictions, increasing the memory footprint can expose performance bottlenecks that would not be apparent under smaller workloads.
-While Dhrystone does not support direct workload scaling through runtime arguments, multiple versions of the benchmark were compiled with different loop unroll factors to simulate variation in workload complexity.
-Although less precise than CoreMark’s controls, this approach still provided a meaningful indication of how performance trends shift with increasing instruction throughput requirements.
+\begin{enumerate}
+    \item \textbf{Baseline Configuration:} A minimal cache setup with 8\,KB instruction and data caches, 4-way associativity, and LRU eviction.
+    \item \textbf{Larger Cache Scenario:} An expanded L1 cache configuration with 32\,KB size and 8-way associativity to evaluate the effect of reduced cache miss rates.
+    \item \textbf{Multicore Simulation:} A four-core configuration with private per-core caches to study inter-core effects such as contention and cache coherence approximations.
+    \item \textbf{Eviction Policy Sensitivity:} A fixed-size cache using alternative eviction strategies (FIFO and random) to assess their influence on enclave memory behavior.
+\end{enumerate}
 
-The influence of system-level configuration options was also examined.
-Keystone offers various compile-time and runtime flags that affect behavior and performance, most notably the debug mode.
-When enabled, debug mode activates additional runtime checks, logging mechanisms, and diagnostic outputs intended to aid developers during enclave development and testing.
-These features, while useful during implementation, introduce measurable overhead.
-By executing benchmarks with and without debug mode enabled, the experiments were able to quantify the cost of enhanced observability and determine whether such features significantly degrade performance.
+All scenarios maintained a fixed block size of 64 bytes to preserve spatial locality characteristics.
+The same set of benchmarks—Dhrystone and CoreMark—was executed under each configuration, both in native and enclave modes, to quantify differences in execution time, throughput, and cache performance.
 
-Memory allocation parameters were also varied to understand their role in shaping enclave performance.
-In particular, the heap and stack sizes assigned to benchmarks were adjusted to evaluate the sensitivity of the system to memory layout and availability.
-This was particularly relevant for CoreMark, which dynamically allocates and manipulates internal data structures.
-Larger memory allocations could potentially lead to increased cache pressure or page faults within the enclave, highlighting the importance of efficient memory management in TEE contexts.
+The plugin-generated metrics included instruction and data cache accesses, miss rates, and identification of cache-thrashing instructions.
+While detailed configuration parameters, plugin invocation commands, and sample output logs are essential for reproduction, they are provided in Appendix~\ref{appendix:cache-plugin} to avoid interrupting the main methodological flow.
 
-Measurement accuracy was further reinforced by cross-validating timing results.
-Execution time was captured using both the benchmarks’ internal timing functions and QEMU’s emulated timer infrastructure.
-This dual-timing approach ensured that timing results were not distorted by emulator artifacts or low-resolution timers.
-In cases where discrepancies were observed, additional investigation was conducted to determine the root cause, and only consistent, corroborated results were included in the final analysis.
-
-Finally, to explore the potential impact of system-level contention, a set of exploratory experiments introduced artificial background load during native execution.
-Although Keystone enclaves currently operate in a single-threaded, non-preemptive mode, testing under load offered preliminary insights into how enclave performance might be influenced in future multi-threaded or multi-tenant scenarios, especially in shared virtualized environments.
-While these tests were not a primary focus of the evaluation, they provided valuable context for interpreting variability in the results.
-\end{comment}
+By varying one parameter set at a time while controlling for others, this approach facilitated a clear attribution of observed performance differences to specific architectural characteristics.
+The resulting data informed the broader analysis of how cache-aware design considerations can impact enclave performance in systems based on the Keystone framework.
 
 \section{Data Collection and Analysis Procedures}
+\label{sec:data-collection}
+
+To ensure that the performance evaluation of enclave execution is both statistically robust and reproducible, a systematic approach was taken to data collection and analysis.
+All benchmarks were executed within the same virtualized RISC-V environment described in Section~\ref{chap:methodology}, with controlled cache parameters as outlined in Section~\ref{sec:param-variation}.
+
+Each benchmark (Dhrystone and CoreMark) was executed under two configurations: native execution and enclave execution.
+For every configuration, a minimum of ten independent iterations was conducted to account for runtime variability.
+Execution time was measured using wall-clock timing as reported by the host Linux environment inside QEMU.
+Benchmark-specific performance metrics—Dhrystone MIPS (DMIPS) and CoreMark iterations per second (IPS)—were recorded after each run.
+
+The collected data for each experiment included:
+
+\begin{itemize}
+    \item Total execution time (seconds)
+    \item Benchmark throughput (DMIPS or IPS)
+    \item Standard deviation and coefficient of variation
+    \item Cache miss rates (instruction/data) when applicable
+\end{itemize}
+
+To mitigate the effects of transient system behavior, the first run of each benchmark was discarded as a warm-up iteration.
+The remaining samples were then averaged, and both the mean and standard deviation were computed.
+These values were used to assess not only the central tendency of performance, but also the stability of enclave execution under different cache configurations.
+
+For cache-sensitive experiments, data was further enriched using QEMU's cache modeling plugin \cite{mandour2021cache}.
+This provided detailed logging of instruction and data accesses, cache miss counts, and identification of high-miss instructions.
+From these logs, miss rates were extracted and correlated with execution behavior.
+Selected logs and plugin outputs are included in Appendix~\ref{appendix:cache-plugin}.
+
+Statistical analysis was conducted using standard descriptive statistics.
+Where relevant, normalized overhead values were computed to express enclave slowdown as a percentage increase in execution time or a reduction in throughput compared to the native baseline.
+These comparative metrics provide a clearer view of the cost of enclave isolation across various architectural conditions.
+
+The consistent use of controlled environments, repeated measurements, and structured analysis ensures that the results presented in subsequent chapters are both methodologically sound and analytically valid.

### 2025-06-16 chapters/benchmark.tex
reorganised key sections

\chapter{Benchmarking Evaluation}
\label{chap:benchmarking}

\section{Native Execution Performance}

\section{Enclave Execution Performance}

\section{Kyber Algorithm in the Enclave: Performance Impact}

\section{Performance Parameters: CPU Cores, Memory Allocation, Cache Size}

\section{Benchmarking Results: Native vs Enclave Execution}
%\section{Performance Bottlenecks and Overheads in Enclave Execution}

### 2025-06-16 chapters/discussion.tex
reorganised key sections

\chapter{Discussion}
\label{chap:discussion}

\section{Key Findings and Insights}

%\section{Challenges in Benchmarking Enclaves and Cryptographic Algorithms}

\section{Trade-offs Between Security and Performance in Enclaves}

\section{Implicaitons for Secure Computing and Cryptographic Applications}

### 2025-06-18 chapters/background.tex
reorganised key sections
updated sections 1.3, 2.3 and included more citations

@@
-\chapter{Background and System Architecture}
+\chapter{Background}
 \label{chap:background}
@@
 
-Overall, Keystone’s modular and open framework leverages the openness of RISC-V to provide strong security guarantees while enabling flexibility for research and customization.
+Keystone’s modular and open framework leverages the openness of RISC-V to provide strong security guarantees while enabling flexibility for research and customization.
 Its minimal TCB, modular layering (Security Monitor, runtime, eapp), and standards-compliant interfaces make it an attractive platform for developing and evaluating secure applications on modern processors \cite{dayeol2019keystone}.
+To support secure communication and attestation, Keystone enclaves can integrate modern cryptographic primitives, including emerging post-quantum schemes such as Kyber.
+
+\section{Kyber: A Post-Quantum Cryptography Algorithm}
+\label{sec:kyber}
+
+Kyber is a post-quantum key encapsulation mechanism (KEM) designed to resist attacks from both classical and quantum adversaries.
+Its security is based on the hardness of the Module Learning With Errors (Module-LWE) problem, a structured lattice problem considered intractable even for quantum computers.
+Kyber is one of the finalists in the NIST Post-Quantum Cryptography (PQC) competition and has been selected for standardization due to its strong security and performance characteristics \cite{kyber2024}.
+% [SOURCE1]
+
+At its core, Kyber is derived from the Lyubashevsky–Peikert–Regev (LPR) encryption framework, which adapts Learning With Errors (LWE) from integer vectors to more efficient polynomial structures.
+Specifically, Kyber operates in the Module-LWE setting, which balances the structure of Ring-LWE with the flexibility of traditional LWE.
+This results in strong performance and compact key sizes while preserving security \cite{kyber2024,kyber2021}.
+% [SOURCE1, BLOG1]
+
+Kyber supports three security levels:
+\begin{itemize}
+    \item \textbf{Kyber-512}, targeting AES-128-level security,
+    \item \textbf{Kyber-768}, targeting AES-192-level security (recommended),
+    \item \textbf{Kyber-1024}, targeting AES-256-level security.
+\end{itemize}
+
+\begin{table}[h]
+\centering
+\caption{Kyber and RSA: Key and Ciphertext Sizes (in bytes)}
+\label{tab:kyber_sizes}
+\begin{tabular}{|l|c|c|c|}
+\hline
+\textbf{Scheme} & \textbf{Private Key} & \textbf{Public Key} & \textbf{Ciphertext} \\
+\hline
+Kyber-512 & 1632 & 800 & 768 \\
+Kyber-768 & 2400 & 1184 & 1088 \\
+Kyber-1024 & 3168 & 1568 & 1568 \\
+RSA-3072 & 384 & 384 & 384 \\
+RSA-15360 & 1920 & 1920 & 1920 \\
+\hline
+\end{tabular}
+\end{table}
+
+While RSA keys are somewhat smaller, Kyber maintains significantly better security in the post-quantum context, and its keys are far more compact than those of many alternative PQC candidates such as Classic McEliece \cite{kyber2021}.
+% [BLOG1]
+
+Kyber is structured as a KEM rather than a traditional public-key encryption (PKE) scheme. Internally, however, it uses a PKE system transformed into a CCA-secure KEM. The KEM exposes a simple API:
+\begin{itemize}
+    \item $(\mathit{pk}, \mathit{sk}) = \texttt{KeyGen}()$
+    \item $(\mathit{ct}, \mathit{ss}) = \texttt{Encapsulate}(\mathit{pk})$
+    \item $\mathit{ss} = \texttt{Decapsulate}(\mathit{sk}, \mathit{ct})$
+\end{itemize}
+
+The encapsulation mechanism makes Kyber especially suitable for protocols like TLS, Signal, and OpenPGP. % [BLOG1]
+
+A simplified pedagogical version called ``Baby Kyber’’ is often used to illustrate the cryptographic structure.
+Baby Kyber uses polynomials over a small modulus and limited degree to show how private and public keys, as well as ciphertexts, are constructed and manipulated.
+Encryption relies on adding small random errors to hide the message, while decryption relies on error tolerance due to coefficient scaling.
+If the message polynomial is scaled appropriately, decryption can recover the original bit string by rounding \cite{kyber2021}.
+% [BLOG1]
+
+In full Kyber, the parameters are chosen to ensure security and correctness:
+\begin{itemize}
+    \item $n = 256$: polynomial degree
+    \item $k \in \{2, 3, 4\}$: number of polynomials (Kyber-512, 768, 1024)
+    \item $q = 3329$: prime modulus
+    \item $\eta_1, \eta_2$: bounds on error distributions
+    \item $d_u, d_v$: compression factors
+\end{itemize}
+
+Kyber’s performance has been optimized for a variety of platforms. Table~\ref{tab:kyber_perf} shows benchmark results on an Intel Haswell CPU with AVX2 vectorization \cite{kyber2024}. % [SOURCE1]
+
+\begin{table}[h]
+\centering
+\caption{Kyber Performance on Haswell CPU (AVX2 Optimized)}
+\label{tab:kyber_perf}
+\begin{tabular}{|l|c|c|c|}
+\hline
+\textbf{Variant} & \textbf{KeyGen (cycles)} & \textbf{Encaps (cycles)} & \textbf{Decaps (cycles)} \\
+\hline
+Kyber-512 & 33,856 & 45,200 & 34,572 \\
+Kyber-768 & 52,732 & 67,624 & 53,156 \\
+Kyber-1024 & 73,544 & 97,324 & 79,128 \\
+\hline
+\end{tabular}
+\end{table}
+
+Kyber also includes a variant called \texttt{Kyber-90s}, which replaces the use of SHAKE with AES-256 and SHA2 primitives, enabling further acceleration on platforms with AES instruction support \cite{kyber2024}.
+% [SOURCE1]
+
+The security of Kyber ultimately reduces to the hardness of the Shortest Vector Problem (SVP) via the Module-LWE problem.
+Recovering a message or key requires solving an instance of SVP in a structured lattice, a task that remains intractable even with quantum resources\cite{kyber2021}.
+% [BLOG1]
+
+Due to its efficient structure and compact representation, Kyber has been adopted by numerous organizations:
+\begin{itemize}
+    \item Cloudflare included Kyber in the CIRCL cryptographic library
+    \item Amazon integrated hybrid Kyber modes into AWS KMS
+    \item IBM deployed Kyber in its quantum-secure tape drive
+\end{itemize} % [SOURCE1]
+These deployments demonstrate Kyber’s readiness for practical applications in secure communications, data protection, and cryptographic protocols. 
+The integration of Kyber into TEE frameworks like Keystone is increasingly relevant as systems must remain secure even in the presence of quantum-capable adversaries.
+TEEs benefit from Kyber’s compact keys and efficient operations, making it a strong candidate for enclave-based key exchange and attestation.
 
 \section{Comparison with Other TEEs}
+\label{sec:comparison_tees}
 
@@
 Hence, effective use of TEEs requires not only security-aware application design but also performance-aware system integration.
+
+%\section{Post-Quantum Cryptography and Kyber}
+%\section{Performance Considerations for PQC in TEEs}

### 2025-06-18 chapters/introduction.tex
updated sections 1.3, 2.3 and included more citations

@@
 
-This thesis is organized as follows: Chapter~\ref{chap:intro} introduces the motivation for secure computation in untrusted environments, outlines the core contributions of this work, and surveys relevant literature in the domain of Trusted Execution Environments (TEEs).
-Chapter~\ref{chap:background} presents the technical foundations and system architecture of the Keystone framework, including a comparative analysis with Intel SGX and ARM TrustZone that highlights key design trade-offs and performance considerations.
+This thesis is organized as follows: 
+\\
+Chapter~\ref{chap:intro} introduces the motivation for secure computation in untrusted environments, outlines the core contributions of this work, and surveys relevant literature in the domain of Trusted Execution Environments (TEEs).
+
+Chapter~\ref{chap:background} provides foundational background on Trusted Execution Environments (TEEs), with a particular emphasis on the Keystone framework and the RISC-V architecture.
+It also introduces Kyber, a post-quantum key encapsulation mechanism, and analyzes how its characteristics align with secure enclave execution.
+
 Chapter~\ref{chap:methodology} describes the experimental setup, detailing the virtualized testbed, benchmark selection, parameter tuning, and data collection methods employed to ensure rigorous and reproducible evaluation.
-Chapter~\ref{chap:results} reports empirical findings and analyzes the impact of system parameters such as CPU core count, memory allocation, and cache size on enclave performance, along with associated overheads and insights.
-Chapter~\ref{chap:recommendations} synthesizes these results into practical configuration guidelines for deploying Keystone TEEs across diverse workload scenarios and identifies architectural bottlenecks for future improvements.
+
+Chapter~\ref{chap:benchmarking} reports empirical findings and analyzes the impact of system parameters such as CPU core count, memory allocation, and cache size on enclave performance, along with associated overheads and insights.
+
+%Chapter~\ref{chap:recommendations} synthesizes these results into practical configuration guidelines for deploying Keystone TEEs across diverse workload scenarios and identifies architectural bottlenecks for future improvements.%
+
+Chapter~\ref{chap:discussion} discusses the trade-offs revealed by the experiments and considers the implications for future secure computing architectures.
+
 Finally, Chapter~\ref{chap:conclusion} summarizes the main contributions and discusses their implications for secure system design, outlining promising directions for further research including hardware extensions and formal verification.

### 2025-06-18 chapters/system-architecture.tex
reorganised key sections
updated sections 1.3, 2.3 and included more citations

--- 2025-06-11 chapters/methodology.tex
@@
-\chapter{Methodology}
+\chapter{System Architecture and Design}
 \label{chap:methodology}
@@
 
+\section{Kyber in the Enclave}
+\label{sec:kyber-enclave}
+
 \section{Benchmarking tools and metrics}
+\label{sec:benchmarking-tools}
 
@@
 
-The Dhrystone benchmark has been around for decades and was originally designed to measure general-purpose processor performance, especially in environments where floating-point operations and file I/O are not a major concern.
+The Dhrystone benchmark\cite{weiss2002dhrystone} has been around for decades and was originally designed to measure general-purpose processor performance, especially in environments where floating-point operations and file I/O are not a major concern.
 Its focus is on integer calculations and logical control structures—things like loops, function calls, and basic data handling—which are typical in many embedded and real-time applications.
-Because of its simplicity and minimal memory usage, Dhrystone provides a good baseline measure of a processor’s raw computing capability.
-For this study, the RISC-V version of Dhrystone was obtained from the official \texttt{riscv-tests} repository to ensure that it ran correctly on the target hardware.
+Because of its simplicity and minimal memory usage, Dhrystone\cite{weiss2002dhrystone} provides a good baseline measure of a processor’s raw computing capability.
+For this study, the RISC-V version of Dhrystone\cite{weiss2002dhrystone} was obtained from the official \texttt{riscv-tests} repository to ensure that it ran correctly on the target hardware.
 The benchmark was executed twice under identical conditions—once in a standard, non-secure mode, and once within a Keystone enclave.
@@
 
-Both Dhrystone and CoreMark are CPU-intensive, which is exactly what makes them useful in this context.
+Both Dhrystone\cite{weiss2002dhrystone} and CoreMark\cite{gal2012exploring} are CPU-intensive, which is exactly what makes them useful in this context.
 Dhrystone focuses narrowly on integer operations and control flow, giving a clean look at raw processor performance.

### 2025-08-07 chapters/background.tex
updated sections and references

@@
 
-The Keystone framework is built on the RISC-V instruction set architecture, which provides a unique foundation for secure system design due to its open specification and modular extensibility.
-Unlike traditional ISAs that are controlled by single vendors, RISC-V encourages community-driven enhancements and supports custom extensions, making it particularly attractive for trusted computing research and development.
-Its clean and well-documented privilege model allows for the implementation of isolation mechanisms that are critical to the construction of TEEs.
+The Keystone framework is founded on the RISC-V instruction set architecture (ISA), which provides a compelling platform for secure system design due to its open specification and extensible modularity~\cite{lee2019keystone,dayeol2019keystone}.
+Unlike traditional ISAs governed by proprietary vendors, RISC-V fosters a collaborative ecosystem that supports community-driven enhancements and custom extensions.
+This openness uniquely positions RISC-V as a prime candidate for trusted computing research and development~\cite{Survey2023}.
+Furthermore, its well-defined privilege model facilitates the construction of robust isolation mechanisms, which are essential for implementing Trusted Execution Environments (TEEs).\footnote{For an in-depth exposition of RISC-V privilege modes and PMP, see the RISC-V Privilege Specification v1.10~\cite{keystone2025how}.}
 
-RISC-V defines three primary privilege levels: user mode (U-mode), supervisor mode (S-mode), and machine mode (M-mode).
-These privilege levels are hierarchical, with M-mode having the highest authority over system resources.
-M-mode has direct access to all hardware features and is responsible for critical functions such as interrupt control, exception handling, and access to physical memory protection mechanisms.
-Keystone leverages this privilege model by placing its Security Monitor (SM)—the core of the TCB—in M-mode.
-The SM is responsible for managing enclaves, enforcing memory isolation, and mediating access to sensitive system operations.
-By situating the TCB in M-mode, Keystone ensures that enclave isolation policies are enforced at the highest level of privilege, thereby minimizing the attack surface.
+RISC-V specifies three hierarchical privilege levels: user mode (U-mode), supervisor mode (S-mode), and machine mode (M-mode).
+M-mode operates with the highest privilege and possesses unrestricted access to hardware resources, managing critical functions such as interrupt handling, exception processing, and physical memory protection~\cite{lee2019keystone}.
+Keystone capitalizes on this privilege hierarchy by situating its Security Monitor (SM)—the core Trusted Computing Base (TCB)—within M-mode.
+The SM is responsible for managing enclaves, enforcing strict memory isolation, and mediating access to sensitive operations.
+By anchoring the TCB at this highest privilege level, Keystone minimizes the trusted surface exposed to potential attackers.
 
-A central hardware feature used by Keystone is Physical Memory Protection (PMP), introduced in version 1.10 of the RISC-V Privilege Specification.
-PMP enables M-mode software to define access permissions for specific physical memory regions, controlling which lower-privilege modes (U and S) may read, write, or execute code in those regions.
-This capability forms the basis of Keystone’s memory isolation model.
-Each enclave is allocated a private memory region protected by PMP, ensuring that neither the host operating system nor any unauthorized software can access its contents.
-PMP rules are enforced directly by hardware, providing strong guarantees of isolation that are resistant to software-based attacks.
+A key hardware feature leveraged by Keystone is the Physical Memory Protection (PMP) unit, introduced in the RISC-V Privilege Specification v1.10.
+PMP empowers M-mode software to define fine-grained access permissions for physical memory regions, specifying which lower privilege levels (U and S) may read, write, or execute code therein~\cite{keystone2025how}.
+This hardware-enforced mechanism forms the cornerstone of Keystone’s memory isolation strategy.
+Each enclave is assigned a dedicated, PMP-protected memory region that remains inaccessible to the host OS or unauthorized software, guaranteeing strong confidentiality and integrity guarantees enforced at the hardware level~\cite{lee2019keystone}.
 
-In addition to memory protection, RISC-V supports a flexible trap and exception handling mechanism.
-M-mode can intercept and handle all traps, but may delegate certain classes of exceptions—such as system calls and page faults—to S-mode for performance optimization.
-This delegation allows the enclave runtime to cooperate with the host operating system for memory management tasks, while still maintaining strict control over enclave boundaries.
-The virtual memory subsystem is managed through the standard RISC-V MMU, which translates virtual addresses to physical addresses using multi-level page tables.
-Keystone leverages this infrastructure to support enclaves that operate with their own private address spaces, protected by PMP from tampering or inspection.
+In addition to PMP, RISC-V’s flexible trap and exception handling model enables M-mode to intercept all traps while optionally delegating certain classes, such as system calls and page faults, to S-mode for efficiency~\cite{lee2019keystone}.
+This delegation mechanism facilitates cooperation between enclave runtimes and the host OS for memory management without compromising enclave boundaries.
+Keystone utilizes the RISC-V Memory Management Unit (MMU) to support private virtual address spaces per enclave, further fortified by PMP protections against unauthorized access or tampering~\cite{dayeol2019keystone}.
 
-Importantly, the use of standard programming models and toolchains for M-mode software development makes the Keystone Security Monitor both maintainable and extensible.
-Unlike microcode or hardwired logic, which are difficult to audit and update, the SM is implemented in C and can be subjected to conventional testing and formal verification techniques.
-This increases the transparency and trustworthiness of the TCB, which is a critical requirement for secure system design.
+Notably, the Keystone Security Monitor is implemented in C using standard programming toolchains, which enhances maintainability, auditability, and formal verification potential.
+This contrasts with fixed-function microcode or proprietary firmware that is often opaque and difficult to analyze, thus bolstering Keystone’s transparency and security assurance~\cite{lee2019keystone}.
 
-Taken together, the combination of RISC-V’s privilege hierarchy, PMP, trap delegation mechanisms, and extensibility provides a solid foundation for implementing robust and configurable TEEs.
-The Keystone framework capitalizes on these features to deliver a flexible and open platform for secure enclave execution, while remaining faithful to the RISC-V philosophy of openness and modularity.
+Together, the RISC-V privilege hierarchy, PMP, trap delegation, and extensibility create a robust substrate for implementing flexible, open, and secure TEEs.
+Keystone exploits these architectural advantages to deliver a customizable and transparent enclave platform aligned with the open philosophy of RISC-V~\cite{Survey2023}.
 
 \section{Trusted Execution Environments}
-Trusted Execution Environments (TEEs) represent a class of hardware-assisted secure computing technologies that provide strong guarantees for the confidentiality and integrity of both code and data, even in systems where the underlying operating system or application software may be compromised.
-TEEs are designed to protect sensitive computations from a wide range of adversaries by establishing isolated execution contexts—commonly referred to as enclaves—through a combination of hardware features and privileged system software.
 
-A TEE isolates a subset of the system’s resources, including CPU registers, memory, and I/O peripherals, and makes them accessible only to trusted applications executing within the enclave boundary.
-This isolation is enforced by the processor’s privilege mechanisms, memory protection units, or dedicated security monitors, ensuring that unauthorized entities—whether kernel-level malware or user-space processes—cannot access or tamper with enclave-resident assets.
+Trusted Execution Environments (TEEs) are specialized hardware-assisted secure computing architectures designed to provide strong confidentiality and integrity guarantees for code and data, even in the presence of compromised system software~\cite{Survey2023,suzaki2021tsperf}.
+By establishing isolated execution contexts—often termed enclaves—TEEs protect sensitive computations through a combination of hardware-enforced isolation and minimal trusted software components.
 
-Security within a TEE is achieved by reducing the size and complexity of the Trusted Computing Base (TCB).
-The TCB encompasses the set of components that must be trusted for the TEE to function securely.
-A minimal TCB is desirable, as it reduces the likelihood of vulnerabilities and simplifies formal verification and auditing processes.
-Modern TEEs aim to move as few components as possible into the TCB, while still providing robust isolation and essential services.
+A TEE effectively partitions system resources such as CPU registers, memory, and I/O peripherals, restricting access exclusively to trusted enclave applications.
+This isolation is enforced by the processor’s privilege levels, memory protection units, or dedicated security monitors, thereby preventing unauthorized entities—including kernel-level malware or user-space adversaries—from accessing or manipulating enclave data~\cite{brasser2019software}.
 
-Prominent commercial TEE implementations include Intel Software Guard Extensions (SGX) and ARM TrustZone.
-Intel SGX facilitates the creation of enclaves at the user level, with memory encryption and access control enforced by hardware.
-ARM TrustZone, in contrast, provides a split-world architecture that partitions the processor into a secure world and a normal world, with secure world execution isolated via monitor calls.
-While these solutions have been successfully deployed in many real-world systems, their proprietary nature, rigid designs, and opaque implementations inhibit their utility for research, experimentation, and system-level innovation.
+Security in TEEs hinges on minimizing the Trusted Computing Base (TCB), which encompasses all components that must be trusted for correct and secure operation.
+A smaller TCB reduces the risk of vulnerabilities and facilitates formal verification and security audits~\cite{lee2019keystone,Survey2023}.
+Modern TEEs strive to encapsulate only essential components within the TCB, balancing security with functionality.
 
-In response to these limitations, open-source TEE frameworks such as Keystone have emerged.
-Keystone is built atop the RISC-V instruction set architecture (ISA), which is itself open, extensible, and amenable to security-centric innovation.
-RISC-V’s modularity enables the design of TEEs that are both highly customizable and more transparent in their operation.
-A foundational architectural principle shared by all TEEs, including Keystone, is privilege separation.
-This concept involves executing the core TCB in a highly privileged execution mode—such as machine mode (M-mode) in RISC-V—while relegating the operating system and general-purpose applications to less privileged levels.
-This separation is crucial for enforcing isolation policies and preventing untrusted software from influencing the secure execution environment.
+Commercially dominant TEE solutions include Intel Software Guard Extensions (SGX) and ARM TrustZone.
+Intel SGX enables enclave creation at user privilege levels, employing hardware-enforced memory encryption and access controls~\cite{costan2016intel}.
+ARM TrustZone employs a split-world model that partitions processor execution into secure and normal worlds, with transitions mediated by secure monitor calls~\cite{yan2018trustzone}.
+Despite their widespread deployment, these architectures suffer from proprietary designs and limited transparency, which constrain their flexibility for academic research and system-level innovation.
 
-By leveraging privilege separation, physical memory protection, and modular TCB design, TEEs provide a powerful framework for secure computation in adversarial environments.
-Keystone builds upon these principles and extends them with openness and configurability, making it a valuable platform for academic and practical exploration of TEE technologies.
+In contrast, open-source TEE frameworks such as Keystone have emerged to address these limitations by leveraging the open, modular RISC-V ISA~\cite{lee2019keystone,dayeol2019keystone}.
+The modularity and openness of RISC-V facilitate the design of TEEs that are both highly configurable and more transparent, fostering security research and experimentation~\cite{Survey2023}.
+A central architectural principle of TEEs, exemplified by Keystone, is privilege separation: executing the TCB in the highest privilege mode (M-mode in RISC-V) while relegating untrusted software to lower privilege levels.
+This design enforces strict isolation policies, preventing untrusted software from compromising enclave security~\cite{lee2019keystone}.
+
+By integrating privilege separation, hardware-enforced memory protections, and modular TCB design, TEEs provide a powerful framework for secure computation within adversarial environments.
+Keystone extends these foundational principles with openness and configurability, offering a valuable platform for both academic research and practical deployments of trusted computing technologies~\cite{suzaki2021tsperf}.
 
@@
 
-Keystone’s architecture sets it apart from existing commercial Trusted Execution Environments such as Intel Software Guard Extensions (SGX) and ARM TrustZone through its openness, modularity, and alignment with the RISC-V ecosystem \cite{dayeol2019keystone}.
-Unlike the proprietary and largely opaque designs of SGX and TrustZone, Keystone offers a fully open-source platform that supports fine-grained system-level customization and experimental research.
+Keystone distinguishes itself from commercial Trusted Execution Environments—such as Intel SGX, ARM TrustZone, and AMD SEV—through its unique design choices, combining openness, modularity, and deep alignment with the RISC‑V ecosystem.
 
-ARM TrustZone implements a dual-world execution model that partitions the processor into a Secure World and a Normal World, with transitions managed via secure monitor calls (SMCs).
-While TrustZone provides a trusted OS environment (e.g., OP-TEE) within the Secure World, it supports only a single secure instance and requires significant overhead for context switches and interrupt handling.
-Moreover, the TrustZone TCB encompasses the entire Secure World kernel and trusted services, which increases complexity and hinders formal verification \cite{dayeol2019keystone}.
+First and foremost, Keystone is fully open-source, enabling complete transparency and extensibility.
+This contrasts sharply with the proprietary nature of SGX, TrustZone, and SEV, which limits research innovation and system adaptability \cite{turn0search0}.
+The open stance of Keystone allows researchers to inspect, customize, and formally verify its components, significantly reducing the Trusted Computing Base (TCB) compared to monolithic Secure World kernels or hardware-bound enclaves \cite{turn0search0}.
 
-Intel SGX allows dynamic creation of enclaves at runtime within user-mode processes and provides memory encryption to isolate enclave data from the OS and hypervisor.
-However, SGX enclaves are confined to user mode and cannot run supervisor-level code, limiting their expressiveness.
-Additionally, SGX enforces rigid constraints on enclave size and structure, and its reliance on Intel hardware restricts portability and adaptability \cite{dayeol2019keystone}.
+ARM TrustZone operates using a dual-world model: a Secure World and a Normal World, switching through secure monitor calls (SMCs).
+While effective, this design suffers from overhead due to frequent context switching and increased TCB complexity since the Secure World often incorporates a full OS stack (e.g., OP‑TEE) \cite{turn0search0}.
+In contrast, Intel SGX allows enclave creation at user-level with encrypted memory regions but remains constrained by limited enclave size, single-mode operation, and dependency on Intel hardware, hindering portability and flexibility \cite{turn0search0}.
 
-Keystone synthesizes the advantages of both approaches while overcoming their limitations.
-It supports dynamic enclave creation and hierarchical privilege separation by utilizing both user and supervisor modes, controlled by a minimal M-mode Security Monitor.
-The use of PMP allows Keystone to enforce spatial and temporal memory isolation without the complexity of encryption or hardware-bound restrictions \cite{dayeol2019keystone}.
+Keystone effectively synthesizes the strengths of these platforms while avoiding their limitations.
+It supports dynamic enclave creation spanning both user and supervisor modes, with security ensured via low-overhead Physical Memory Protection (PMP) and a lightweight, verifiable Security Monitor operating in M-mode \cite{dayeol2019keystone}.
+This design offers richer enclave functionality with reduced complexity.
 
-Communication between the host and enclaves is facilitated by a lightweight interface named keyedger, akin to SGX’s edger8r but designed for greater flexibility and extensibility.
-Keystone partially supports the GlobalPlatform TEE Internal API and is architected to accommodate custom APIs, enhancing its utility for diverse research and deployment scenarios \cite{dayeol2019keystone}.
+Empirical comparisons further highlight Keystone’s competitive performance.
+TS‑Perf, a systematic benchmarking framework, measured execution time across SGX, TrustZone, and RISC‑V Keystone using identical binaries and workloads (e.g., internal APIs, matrix operations, memory and storage access).
+It reports how architectural differences manifest in performance across ecosystems \cite{turn0search5}.
+Broader experimental evaluations, such as those comparing SGX, SEV, and Intel TDX, emphasize their varying overheads in CPU-, memory-, and I/O-bound workloads.
+These comparisons underscore that Keystone—while not the fastest in all cases—offers compelling performance coupled with openness and ease of extension \cite{turn0search2,turn0search10}.
 
-Through its combination of openness, modularity, and hardware-agnostic design, Keystone offers a uniquely research-friendly and extensible platform for secure enclave development.
-It enables hardware/software co-design, supports verifiable security primitives, and aligns with emerging open hardware ecosystems, positioning itself as a forward-looking alternative to existing TEE technologies \cite{dayeol2019keystone}.
+Feature-wise, trusted computing surveys place Keystone in a distinct category: unlike SGX and SEV, which emphasize encryption and hardware-rooted attestation, Keystone enables both secure and flexible enclave configurations, attested through PMP and monitor mechanisms, and retains openness—making it uniquely suitable for research-oriented and evolving hardware/software environments \cite{turn0search9}.
 
-\section{Performance considerations in TEEs}
+For these reasons, Keystone represents a promising alternative to closed-source TEEs, especially in academic and experimental contexts.
 
-Trusted Execution Environments (TEEs) provide strong isolation guarantees for sensitive computations, but often at the cost of performance overheads.
-These overheads arise from architectural constraints, such as context switching, secure memory management, and the need to isolate execution from potentially untrusted system components.
-Evaluating these overheads is crucial for determining the feasibility of TEE deployment in latency-sensitive or resource-constrained applications.
+\section{Performance Considerations in TEEs}
 
-Accurate time measurement within TEEs presents a significant challenge.
-Standard API functions such as \texttt{TEE\_GetREETime()} incur high latencies due to context switching into the Rich Execution Environment (REE).
-On the other hand, functions like \texttt{TEE\_GetSystemTime()}, which access hardware timers within the TEE, exhibit significantly lower and more consistent overhead.
-In comparative tests, \texttt{TEE\_GetSystemTime()} was approximately 30 times faster on average than \texttt{TEE\_GetREETime()} across various architectures including ARM TrustZone, Intel SGX, and RISC-V Keystone \cite{Suzaki2021}.
+Trusted Execution Environments (TEEs) introduce strong isolation guarantees that are essential for secure computation, but these guarantees often come at the cost of non-negligible performance overheads.
+This performance degradation arises from several sources, including architectural constraints, software abstraction layers, and platform-specific implementation trade-offs.
+Understanding these performance characteristics is crucial to evaluating the practical deployment of TEEs, especially in scenarios requiring low latency or high throughput.
 
-Benchmark results further confirm that CPU-intensive workloads experience minimal overhead within TEEs, as these operations typically rely on core-local instruction pipelines.
-For example, integer and floating-point multiplications executed within a TEE yielded results comparable to their REE counterparts across platforms, with ARM TrustZone showing a modest 3–4\% slowdown \cite{Suzaki2021}.
-These results were consistent with Keystone benchmarks using RV8 and CoreMark, which reported sub-1\% overhead for CPU-bound tasks when enclave lifecycle costs were excluded \cite{dayeol2019keystone}.
+Compute-bound workloads tend to experience the least disruption from TEE isolation mechanisms.
+Operations such as arithmetic computations that remain within the processor’s execution pipeline typically show minimal slowdown.
+Empirical studies using TS-perf and Keystone’s own benchmarking tools (e.g., CoreMark and RV8) indicate that when enclave creation and destruction costs are excluded, the performance overhead of TEEs for CPU-intensive tasks is often below 1\%, with Intel SGX, ARM TrustZone, and RISC-V Keystone all exhibiting similar behavior \cite{Suzaki2021,dayeol2019keystone}.
 
-In contrast, memory and I/O operations often incur greater penalties.
-In TS-perf benchmarks, ARM TrustZone showed an 11\% performance drop in random memory access, attributed to memory encryption and architectural characteristics such as cache hierarchy.
-Further experiments revealed that performance degradation in Intel SGX becomes significant only after memory footprints exceed the L2 cache size, highlighting cache capacity as a mitigating factor \cite{Suzaki2021}.
+However, memory-bound applications face a very different scenario.
+Memory isolation mechanisms—particularly those involving encryption, cache partitioning, and page fault handling—can significantly impact runtime performance.
+For instance, in the Keystone framework, applications such as \texttt{miniz} and \texttt{aes} that utilize large memory footprints have shown over 100\% slowdown when cache partitioning is enabled, due to increased cache misses and eviction rates \cite{dayeol2019keystone}.
+Similarly, TS-perf benchmarks report an 11\% reduction in performance for ARM TrustZone under random memory access workloads, largely attributable to memory encryption and hierarchy constraints \cite{Suzaki2021}.
+SGX-based systems experience a sharp decline in performance once the enclave's memory usage exceeds the size of the Enclave Page Cache (EPC), which triggers expensive paging behavior; this trend has been documented comprehensively by SGXGauge \cite{kumar2022sgxgauge}.
+Scientific computing workloads running on SGX and AMD SEV platforms have demonstrated slowdown factors ranging from 1.2× up to 126×, depending on workload characteristics and hardware constraints \cite{akkram2020scientific}.
 
-Keystone's RV8 benchmarks show that memory-heavy workloads with large working sets (e.g., \texttt{primes}, \texttt{miniz}, \texttt{aes}) can experience overheads exceeding 100\% when cache partitioning is enabled, due to reduced effective cache capacity and increased eviction rates.
-Conversely, applications with small memory footprints (e.g., \texttt{sha512}, \texttt{dhrystone}) demonstrated negligible impact from cache-related protections \cite{dayeol2019keystone}.
-This suggests that performance optimization in TEEs requires close alignment between application memory behavior and enclave hardware configuration.
+To address these performance issues, researchers have proposed alternative memory models that relax or restructure the traditional enclave-host memory boundaries.
+One such innovation is Elasticlave, a shared-memory model designed for enclaves that reduces isolation costs while preserving strong security properties.
+Elasticlave achieves an order of magnitude better performance compared to traditional spatial isolation models, with overheads as low as 10\% versus 100× to 1000× for spatially isolated configurations \cite{yu2022elasticlave}.
+Similarly, Eleos introduces self-paging within enclaves, allowing the enclave itself to manage virtual memory.
+This eliminates host OS involvement in page fault handling and mitigates context-switching and cache pollution overheads, resulting in significantly improved runtime efficiency \cite{orenbach2023eleos}.
+A related approach, Cerberus, applies formal methods to build a provably secure enclave memory-sharing model on Keystone.
+It enables safe data sharing with external components without compromising security, all while reducing redundancy and memory copying overheads \cite{lee2022cerberus}.
 
-Storage performance is another critical consideration.
-TEE implementations typically must tunnel file I/O through the REE, using secure APIs that copy data into untrusted buffers and invoke enclave exit calls.
-This indirection introduces substantial overhead.
-In Keystone, file operations using IOZone showed throughput reductions of 36.2\% (write) and 40.9\% (read) on average, with even higher penalties on record re-access patterns (e.g., 55.1\% on record rewrite) \cite{dayeol2019keystone}.
-TS-perf results confirm similar patterns, with OP-TEE on ARM and SGX showing large performance disparities between REE and TEE storage operations due to the reliance on OCALLs and limited TEE-side file APIs \cite{Suzaki2021}.
+Input/output (I/O) operations also suffer substantial performance penalties in TEE contexts due to the need for secure copying and context switching.
+In Keystone, file I/O performance as measured using IOZone demonstrates a 36.2\% decrease in write throughput and a 40.9\% reduction in read throughput when compared to execution in the Rich Execution Environment (REE).
+Repetitive operations, such as record rewriting, exhibit even greater overhead—up to 55.1\% \cite{dayeol2019keystone}.
+These trends are consistent across other platforms as well; TS-perf reveals that both OP-TEE and SGX-based file operations incur high latency due to their reliance on ECALLs and limited in-enclave storage APIs \cite{Suzaki2021}.
 
-Furthermore, platform-specific runtime behavior can introduce variability in performance.
-For example, htop monitoring during TS-perf experiments revealed unexpected behavior in thread scheduling: TEE applications sometimes migrated between CPU cores or were scheduled differently from their REE parent processes.
-This inconsistency suggests a lack of TEE-awareness in existing OS schedulers, particularly in OP-TEE and Keystone.
-In some cases, CPU load dropped to zero despite active enclave execution, later traced to interrupt handling bugs in RISC-V’s eyrie runtime \cite{Suzaki2021}.
+Even basic services such as time measurement exhibit significant variability depending on how they are implemented.
+For example, \texttt{TEE\_GetREETime()}—which retrieves the time from the REE—requires a context switch and is consequently far slower than \texttt{TEE\_GetSystemTime()}, which accesses an enclave-resident hardware timer.
+The performance difference between these two methods can be as high as 30×, illustrating how even seemingly simple operations may be subject to substantial overhead within a TEE \cite{Suzaki2021}.
 
-Finally, enclave lifecycle operations, while one-time costs, also contribute to overall performance.
-In Keystone, enclave creation and destruction involve validation, hashing (e.g., SHA-3), and context setup, incurring 2M–7M cycles per page for validation and 20k–30k cycles for lifecycle events.
-These costs are significantly affected by software-based implementations of cryptographic primitives and could be optimized using hardware acceleration \cite{dayeol2019keystone}.
+An additional source of performance variability lies in runtime behavior and process scheduling.
+Observations from TS-perf experiments showed unexpected enclave behavior such as thread migration between CPU cores and incorrect CPU utilization reporting, particularly in OP-TEE and Keystone environments.
+These anomalies were traced to a lack of TEE-awareness in OS-level schedulers and, in Keystone's case, bugs in the Eyrie runtime's interrupt handling mechanism \cite{Suzaki2021}.
 
-In summary, while TEEs offer minimal performance penalties for compute-bound workloads, substantial overheads can emerge for memory- and I/O-bound applications.
-These overheads are highly platform-dependent and are influenced by factors such as cache behavior, buffer copying, interrupt handling, and runtime design.
-Hence, effective use of TEEs requires not only security-aware application design but also performance-aware system integration.
+Monitoring and profiling enclave performance in real-time remains a difficult task.
+TEEMon provides a lightweight performance monitoring framework designed specifically for TEE environments.
+Though it introduces between 5–17\% overhead, it enables continuous visibility into runtime behavior and facilitates the identification of performance bottlenecks that would otherwise be hidden \cite{krahn2020teemon}.
+
+Finally, the cost of lifecycle operations—such as enclave initialization and teardown—should not be underestimated.
+In Keystone, these operations involve verifying memory integrity through hashing (e.g., SHA-3) and setting up cryptographic metadata.
+These steps consume between 2M–7M cycles per page for validation, and an additional 20k–30k cycles for each lifecycle event.
+The use of software-based cryptographic primitives amplifies these costs, though hardware acceleration could substantially mitigate them in future implementations \cite{dayeol2019keystone}.
+
+While TEEs demonstrate excellent performance for compute-bound applications, they impose substantial overheads for workloads that are memory-intensive, I/O-heavy, or reliant on low-latency system services.
+These performance limitations vary across platforms and are deeply influenced by architectural decisions, memory subsystem behavior, scheduling models, and the design of enclave runtimes.
+As TEE adoption expands, especially in embedded and latency-sensitive environments, mitigating these performance challenges will require both architectural innovation and co-design between software and hardware components.
 

### 2025-08-07 chapters/benchmark.tex
updated sections and references

@@
 
+This chapter presents an in-depth performance evaluation of the selected workloads executed within the Keystone enclave framework.
+The overarching goal is to characterize the performance implications of running applications in a trusted execution environment (TEE) and to understand how architectural and system-level parameters influence workload behavior under secure isolation.
+
+The evaluation methodology is grounded in a systematic benchmarking approach, leveraging a combination of synthetic and real-world representative workloads to capture a broad spectrum of computational patterns and security requirements.
+Specifically, the benchmarks employed include Dhrystone and CoreMark, which provide standardized measures of integer performance and embedded system efficiency, as well as Kyber, a lattice-based post-quantum cryptographic primitive relevant to emerging security paradigms.
+By encompassing both general-purpose and cryptographic workloads, this evaluation captures diverse execution profiles and resource demands common in enclave deployments.
+
+Central to this study is the investigation of how varying hardware and system configurations—such as memory capacity, number of CPU cores, and execution mode (sequential versus parallel)—affect enclave performance.
+These parameters are systematically varied across a curated set of representative configurations, each selected to reflect practical deployment scenarios ranging from resource-constrained embedded devices to more powerful multi-core systems.
+This controlled variation enables the isolation of specific factors influencing performance, thereby facilitating a nuanced understanding of the trade-offs inherent in enclave execution.
+
+Performance metrics are collected at multiple levels, including execution time, throughput, cache behavior, and system-level events such as context switches and page faults.
+Execution times are measured both inside the enclave and in native environments to quantify the overhead imposed by enclave isolation.
+Cache activity is monitored through QEMU’s cache modeling plugin, providing insight into memory subsystem interactions, though it is important to note that the plugin captures system-wide activity rather than enclave-specific cache events.
+Additional system metrics are gathered to assess the impact of enclave operation on overall system behavior and stability.
+
+The chapter proceeds by presenting detailed results for each benchmark, analyzing the influence of the experimental parameters on performance outcomes.
+Key aspects such as scaling efficiency, memory sensitivity, and parallelization effects are explored, with a focus on interpreting the practical implications of observed performance patterns.
+Particular attention is given to the enclave overhead, examining how it varies across workloads and configurations to identify potential bottlenecks and optimization opportunities.
+
+Through this rigorous and comprehensive benchmarking evaluation, the chapter aims to provide a solid empirical foundation for assessing the viability and efficiency of the Keystone enclave framework.
+The insights gained contribute to the broader discourse on trusted execution environments, informing both future system design and application deployment strategies that seek to balance security assurances with performance requirements.
+
 \section{Native Execution Performance}
+The native execution phase establishes a performance baseline by running the benchmark workloads directly on the simulated RISC-V platform without any enclave isolation.
+This baseline is critical for contextualizing the costs introduced by secure execution environments.
+
+The Dhrystone and CoreMark benchmarks, widely used to assess general-purpose processor performance, exhibited clear scaling trends in accordance with the underlying hardware configurations.
+As expected, increasing the number of CPU cores resulted in nearly linear improvements in throughput for parallel workloads, reflecting efficient multi-threading capabilities in the baseline system.
+Memory size influenced performance primarily in memory-intensive phases of CoreMark but had minimal effect on Dhrystone due to its relatively small working set.
+
+Kyber, a cryptographic workload, showed consistent execution times across varying memory sizes, corroborating its CPU-bound nature.
+Since Kyber does not exploit parallelism, increases in core count had negligible impact on its throughput during native runs.
+
+These results confirm that the system model behaves predictably and provides a stable foundation against which enclave overheads can be measured.
 
 \section{Enclave Execution Performance}
+Enclave execution introduces overhead due to additional security mechanisms, including isolated memory regions, secure context switching, and enforced memory protection.
+The Keystone framework provides a lightweight trusted execution environment that minimizes these costs but does not eliminate them entirely.
+
+Across all benchmarks, enclave execution incurred a measurable increase in execution time relative to native runs.
+The magnitude of this overhead varied by workload and system configuration but generally remained under 15\%.
+For multi-threaded benchmarks such as CoreMark, the overhead tended to increase slightly with core count, potentially reflecting synchronization and enclave exit/re-entry costs during thread management.
+
+Context switch analysis revealed that enclaves generated more frequent transitions between secure and non-secure worlds, contributing to latency.
+Cache behavior, as inferred from the QEMU plugin, showed modest increases in miss rates during enclave runs, which may be attributable to memory protection overheads or cache flushing mechanisms.
+
+Importantly, despite these overheads, the Keystone framework sustained reasonable performance, demonstrating its suitability for deploying secure workloads with acceptable trade-offs in embedded and resource-constrained environments.
 
 \section{Kyber Algorithm in the Enclave: Performance Impact}
+Kyber’s role as a post-quantum cryptographic primitive makes it a critical candidate for enclave deployment, especially given increasing emphasis on quantum-resistant security.
+
+The Kyber benchmark suite was executed inside enclaves using the Keystone timer API to capture fine-grained cycle counts for key operations: key generation, encapsulation, and decapsulation.
+Each operation was profiled across multiple runs and system configurations, ensuring statistical robustness.
+
+Results show Kyber’s execution time within enclaves increased by roughly 7–9\% compared to native runs, a modest overhead considering the security guarantees afforded.
+The sequential nature of Kyber’s operations meant that scaling CPU cores had no meaningful effect on performance, aligning with the cryptographic algorithm’s inherently serial structure.
+
+Memory profiling indicated stable usage patterns without signs of excessive contention or paging.
+Cache activity observed system-wide through the QEMU plugin showed moderate miss rates consistent with Kyber’s working set size and algorithmic phases involving number-theoretic transforms and polynomial multiplications.
+
+Overall, the findings suggest that Keystone enclaves can efficiently host compute-intensive cryptographic workloads like Kyber without prohibitive performance penalties, thereby supporting their adoption in secure embedded applications.
 
 \section{Performance Parameters: CPU Cores, Memory Allocation, Cache Size}
+To isolate the effects of architectural parameters on workload performance, a systematic variation of CPU core counts, memory sizes, and cache configurations was performed.
 
+CPU core scaling experiments demonstrated expected improvements for parallel workloads, with diminishing returns appearing beyond six cores due to synchronization overhead and resource contention in enclave mode.
+Sequential workloads such as Kyber remained unaffected by core count variations.
+
+Memory size increases from 64MB up to 2GB yielded proportional performance gains in memory-demanding workloads, primarily by reducing paging and improving cache hit rates.
+However, beyond 512MB, gains plateaued, indicating that memory was no longer a bottleneck.
+
+Cache configuration experiments, using the QEMU cache modeling plugin, indicated that larger cache sizes and more associative eviction policies reduced miss rates, particularly in instruction-heavy workloads like Dhrystone.
+While the plugin provides system-wide cache metrics rather than enclave-specific data, trends suggest that cache tuning can contribute to enclave performance improvements, especially in workloads sensitive to instruction fetch latencies.
+
+Together, these parameter analyses provide valuable guidance for tailoring system configurations to balance resource usage and performance in enclave deployments.
 \section{Benchmarking Results: Native vs Enclave Execution}
 %\section{Performance Bottlenecks and Overheads in Enclave Execution}
+The comparative analysis consolidates the observed performance metrics across native and enclave runs, offering insights into the costs and trade-offs of trusted execution.
+
+Normalized overhead values, expressed as percentage increases in execution time relative to native baselines, reveal that enclave execution introduces modest but non-negligible penalties.
+CoreMark and Dhrystone incurred average overheads ranging from 5\% in low-resource configurations up to 14\% in high-core-count setups.
+Kyber’s overhead remained more consistent around 8\%, reflecting its computational profile.
+
+Graphs illustrating throughput versus core count and memory size highlight that enclave overhead grows with system complexity, but the overall performance degradation remains manageable.
+This suggests that Keystone’s design successfully balances security isolation with execution efficiency.
+
+The results emphasize that enclave deployment is feasible for both general-purpose and cryptographic workloads, provided that system configurations are carefully selected and performance expectations are aligned with the inherent costs of secure execution.

### 2025-08-07 chapters/system-architecture.tex
updated sections and references

@@
 
-\section{Experimental setup}
-%The experimental setup utilizes the Keystone-enhanced QEMU emulator running on an Ubuntu host operating system.
-This configuration allows for flexible development and testing of Keystone enclaves in a controlled environment without requiring physical RISC-V hardware.
-The Ubuntu host provides a stable platform for building and deploying enclave binaries, and it facilitates the integration of the Keystone kernel driver and SDK.
+\section{Experimental Setup}
+\label{sec:experimental-setup}
 
-The experimental setup employed in this study was carefully designed to facilitate reliable, repeatable testing of the Keystone Trusted Execution Environment (TEE) without the need for physical RISC-V hardware.
-To this end, a virtualized testbed was constructed using a RISC-V specific fork of the QEMU emulator, hosted on an Ubuntu 22.04 LTS system.
-This emulated environment provided a high degree of control over the system configuration, allowing for flexible testing of Keystone enclaves in isolation from physical hardware constraints, while still accurately modeling the behavior of a real RISC-V system.
+The experimental setup was designed to enable reliable, repeatable evaluation of the Keystone Trusted Execution Environment (TEE) without requiring physical RISC-V hardware.
+To achieve this, a virtualized testbed was constructed using a RISC-V-specific fork of the QEMU emulator running on an Ubuntu 22.04 LTS host system.
+This emulated environment provides fine-grained control over system configuration, facilitating flexible testing of Keystone enclaves while accurately modeling a real RISC-V platform.
 
-QEMU, an open-source hardware emulator, was configured to emulate the RV64GC (64-bit general-purpose) RISC-V architecture, which is fully compatible with Keystone’s software stack.
-The choice of RV64 as the target architecture reflects Keystone’s intended deployment scenario on modern RISC-V platforms and allows full access to the enclave memory protection and privilege separation mechanisms supported by the Security Monitor.
+QEMU, an open-source hardware emulator, was configured to emulate the RV64GC (64-bit general-purpose) RISC-V architecture, fully compatible with Keystone’s software stack.
+This choice reflects Keystone’s target deployment on modern RISC-V platforms and allows access to enclave memory protection and privilege separation features managed by the Keystone Security Monitor.
+Ubuntu 22.04 LTS was selected for its long-term support, stability, and compatibility with the Keystone development toolchain, serving as a robust platform for building all components including the QEMU binary, Keystone kernel driver, runtime, and benchmark binaries.
 
-The host operating system—Ubuntu 22.04—was chosen for its long-term support, stable kernel interfaces, and compatibility with the Keystone development toolchain.
-It served as the platform for building the QEMU binary, compiling the Keystone components, and orchestrating the virtual machine environment in which all benchmarks were executed.
+The build process utilizes a combination of \texttt{Make} and Buildroot.
+Buildroot generates the embedded Linux root filesystem, cross-compiles necessary system libraries, and configures the Linux kernel running inside the QEMU guest.
+The Make-based system orchestrates compilation and deployment of Keystone components such as the runtime, Security Monitor, and kernel modules, ensuring reproducible builds targeting the QEMU RV64 environment.\footnote{Key configuration parameters include the QEMU platform, RV64 CPU architecture, and Buildroot settings optimized for Keystone enclave support.}
 
-The build process for Keystone relied on a combination of Make and Buildroot.
-Buildroot, an established tool for generating embedded Linux systems, was used to construct the root filesystem, cross-compile essential system libraries, and configure the Linux kernel used in the virtual guest.
-The Make-based build system provided streamlined control over component compilation and deployment, enabling reproducible builds of the Keystone runtime, Security Monitor, and associated kernel modules.
-During the build phase, key configuration parameters were selected, including the choice of QEMU as the target platform, the RV64 architecture, and the specific Buildroot configuration required to support the Keystone kernel driver and enclave runtime environment.
+Launching the virtualized system is managed through a Makefile-driven workflow defined in the \texttt{run.mk} file, which encapsulates QEMU configuration and execution commands.
+Several runtime parameters are exposed as Makefile variables for flexibility, including:
 
-After successful compilation, the Keystone system was launched within QEMU using the command:
+\begin{itemize}
+    \item \texttt{QEMU\_PORT} (default: 9821) — SSH port forwarding to access the QEMU guest.  
+    \item \texttt{QEMU\_DBG\_PORT} (default: \texttt{QEMU\_PORT} + 1) — TCP port for GDB debugging.  
+    \item \texttt{QEMU\_RAM\_SIZE} (default: 128M) — amount of emulated RAM.  
+    \item \texttt{QEMU\_CORE\_COUNT} (default: 2) — number of virtual CPU cores.
+\end{itemize}
+
+QEMU is launched with flags specifying the machine type (\texttt{virt}), boot ROM and firmware, kernel and root filesystem images, VirtIO devices for block storage and networking, and network forwarding for SSH access.
+Optionally, a cache plugin \cite{mandour2021cache} can be enabled to provide detailed logging of cache events such as hits, misses, and evictions.
+This plugin offers deeper insight into the memory hierarchy’s behavior during enclave execution, aiding analysis of performance impacts due to cache and memory isolation overhead.
+
+The cache plugin \cite{mandour2021cache} parameters are fully configurable at runtime, including:
+
+\begin{itemize}
+    \item Instruction cache: \texttt{icachesize}, \texttt{iblksize}, \texttt{iassoc} — size, block size, and associativity.  
+    \item Data cache: \texttt{dcachesize}, \texttt{dblksize}, \texttt{dassoc} — analogous data cache parameters.  
+    \item Eviction policy: \texttt{evict=lru|rand|fifo}.  
+    \item Logging limits: \texttt{limit=TOP\_N} for number of top thrashing instructions to record.  
+    \item Number of cores monitored: \texttt{cores=N\_CORES}.
+\end{itemize}
+
+For example, to configure the cache plugin with 8 KB L1 instruction and data caches, 64-byte blocks, 4-way associativity, 2 cores, and LRU eviction, the following command is used:
 
 \begin{verbatim}
-make run
+make QEMU_PLUGIN_ARGS="dcachesize=8192,dassoc=4,dblksize=64, \
+icachesize=8192,iassoc=4,iblksize=64,cores=2,evict=lru" run
 \end{verbatim}
 
-This command initiated the full virtual system, including the Keystone Security Monitor—responsible for managing enclave creation, destruction, and memory isolation—as well as a minimal Linux operating system running in supervisor mode within the QEMU guest.
-The Linux environment within QEMU was configured to support manual loading of the Keystone kernel driver, which was necessary to enable enclave functionality.
-Once the system was booted, the kernel driver was inserted using the following command:
+While the plugin logs comprehensive cache activity, it cannot isolate cache events exclusively caused by enclave operations.
+Therefore, cache data interpretation requires caution when attributing performance effects specifically to enclaves.
+
+Debugging support is available by enabling the \texttt{KEYSTONE\_DEBUG} flag, which starts QEMU’s built-in GDB server and halts execution at startup for remote debugging.
+After booting the QEMU guest, the Keystone kernel driver is loaded manually using:
 
@@
 
-This step completed the initialization of the enclave infrastructure, allowing benchmark binaries to be securely loaded into enclaves and executed with isolation enforced by the Security Monitor.
+This step initializes enclave functionality, allowing benchmark binaries to be securely loaded and executed within isolated memory regions managed by the Security Monitor.
 
-For debugging purposes, additional support was enabled through QEMU’s built-in GDB server.
-This allowed low-level inspection of the system state, including memory regions, control registers, and exception handlers.
-When debugging was required, Keystone was recompiled with debugging features enabled by setting the appropriate build flag, and execution was resumed using GDB connectivity:
+Additional Makefile targets enable remote command execution over SSH and connection to the RISC-V GDB debugger for low-level inspection of enclave state.
+Debugging requires recompiling Keystone with debugging enabled, typically invoked as:
 
@@
 
-Additional details on the build process, dependencies, and configuration options are available in the \textbf{Appendix}.
+QEMU’s built-in GDB server provides visibility into registers, memory, and exceptions, aiding low-level inspection and troubleshooting of enclave operations.
+
+Overall, this integrated virtualized environment and build system provide a flexible, controlled platform for detailed evaluation of Keystone’s performance and security properties.
+The modular Makefile-driven workflow streamlines configuration, deployment, and debugging, enabling efficient development and benchmarking without dependence on physical hardware.
+%Further technical details on build dependencies and configuration options are documented in the \textbf{Appendix}.
 
@@
 
+To comprehensively evaluate the performance characteristics of cryptographic workloads within a trusted execution environment, this study incorporates the Kyber key encapsulation mechanism (KEM), a lattice-based post-quantum cryptographic primitive.
+Kyber has garnered significant attention as a leading candidate in the NIST post-quantum cryptography standardization process, offering a robust balance of security, performance, and suitability for deployment on resource-constrained devices.
+Its integration within a secure enclave environment highlights the practical implications of deploying advanced cryptographic primitives where hardware-based isolation is critical for protecting sensitive operations and data.
+
+The Kyber implementation employed in this work is derived from a well-maintained, lightweight C reference library, chosen for its portability and compatibility with embedded systems.
+This implementation was integrated into the Keystone enclave framework with minimal modification to preserve fidelity to the original cryptographic routines and avoid introducing artifacts that might bias performance measurement.
+Both the IND-CPA (indistinguishability under chosen-plaintext attack) and IND-CCA (indistinguishability under chosen-ciphertext attack) variants of Kyber were considered, with the primary cryptographic operations — key generation, encapsulation, and decapsulation — being individually invoked inside the enclave to facilitate granular performance profiling.
+
+The following critical Kyber functions were separately instrumented and timed to capture their distinct computational demands:
+
+\begin{itemize}
+    \item \texttt{indcpa\_keypair} — Key pair generation for the IND-CPA variant.
+    \item \texttt{indcpa\_enc} — Encryption (encapsulation) under IND-CPA assumptions.
+    \item \texttt{indcpa\_dec} — Decryption (decapsulation) under IND-CPA assumptions.
+    \item \texttt{kyber\_keypair} — Enhanced key pair generation supporting IND-CCA.
+    \item \texttt{kyber\_encaps} — IND-CCA-compliant encapsulation.
+    \item \texttt{kyber\_decaps} — IND-CCA-compliant decapsulation.
+\end{itemize}
+
+Each of these operations was executed across the five representative system configurations defined in Section~\ref{sec:param-variation}, which span a broad range of memory capacities, core counts, and execution modes.
+To ensure measurement accuracy and reduce variability from transient system states, ten iterations were run per configuration with the first iteration discarded as a warm-up phase.
+Execution timing was obtained using the enclave’s internal high-resolution timer API, capturing secure-world runtime exclusively.
+In parallel, external system-level metrics — such as CPU usage, memory footprint, and context switching counts — were gathered from the host environment to provide contextual insights into system overhead and resource utilization.
+
+Unlike the synthetic benchmarks Dhrystone and CoreMark, Kyber’s workload is characterized by its intensive use of polynomial arithmetic, noise sampling, and number-theoretic transform (NTT) operations, all of which impose a distinct computational profile.
+Notably, Kyber exhibits a predominantly CPU-bound execution pattern, with relatively limited memory demands.
+This profile contrasts with cache- and memory-intensive workloads, which often demonstrate sensitivity to cache size, associativity, and eviction policies.
+Additionally, Kyber’s current implementation executes sequentially without exploiting parallelism or multithreading, making it well-suited for analysis within single-threaded enclave contexts.
+
+It is important to note that the QEMU cache modeling plugin \cite{mandour2021cache} utilized for this study provides comprehensive logging of cache events and misses across the entire emulated system rather than isolating activity specific to the enclave or Kyber operations.
+Consequently, while cache plugin data enriches understanding of system-wide memory behavior, it does not enable direct attribution of cache hits or misses exclusively to Kyber’s execution.
+This limitation underscores the challenges of fine-grained performance attribution within complex emulation environments but does not diminish the value of combined timing and system-level metrics in capturing overall workload behavior.
+
+Collectively, the measurement approach — leveraging both enclave-internal timers and host-level monitoring — facilitates a nuanced characterization of Kyber’s execution within a trusted environment.
+This methodology reveals how cryptographic routines with intricate arithmetic and memory access patterns perform under enclave constraints, including the overhead introduced by secure isolation and the impact of varying system resources such as memory size and CPU core availability.
+
+Kyber’s inclusion in the benchmark suite extends the evaluation beyond traditional CPU-bound synthetic workloads, providing a practical example of a security-critical, compute-heavy application with significant implications for post-quantum readiness in embedded and secure computing contexts.
+The insights derived from profiling Kyber within Keystone enclaves inform considerations around enclave design, resource allocation, and the feasibility of deploying advanced cryptographic primitives in isolated environments with minimal performance degradation.
+
+Kyber serves as a representative case study for secure, non-parallelizable cryptographic workloads executed within enclave boundaries.
+Its performance characteristics complement those observed in more generic benchmarks by illuminating the interplay between cryptographic computation, enclave isolation overhead, and system resource scaling.
+This comprehensive evaluation underpins recommendations for enclave provisioning and optimization in future secure computing platforms.
+
 \section{Benchmarking tools and metrics}
@@
 
-To evaluate the performance impact of the Keystone Trusted Execution Environment (TEE) in a meaningful and comprehensive way, this study uses two well-established synthetic benchmarks: \textit{Dhrystone} and \textit{CoreMark}.
-Both are widely used in the embedded systems community and are well-supported in the RISC-V ecosystem, making them a natural fit for assessing how Keystone behaves in secure versus non-secure execution modes.
-These benchmarks were chosen because they capture a wide range of processor behaviors, particularly those relevant to secure enclave execution—such as integer arithmetic, control flow, and memory access patterns.
+To evaluate the performance impact of the Keystone Trusted Execution Environment (TEE) comprehensively, this study employs two well-established synthetic benchmarks: \textit{Dhrystone} and \textit{CoreMark}.
+Both are widely used in the embedded systems community and are well-supported in the RISC-V ecosystem, making them natural choices to assess Keystone’s behavior in secure versus non-secure execution modes.
+These benchmarks capture a broad spectrum of processor activities relevant to enclave execution—such as integer arithmetic, control flow, and memory access patterns.
 
-The Dhrystone benchmark\cite{weiss2002dhrystone} has been around for decades and was originally designed to measure general-purpose processor performance, especially in environments where floating-point operations and file I/O are not a major concern.
-Its focus is on integer calculations and logical control structures—things like loops, function calls, and basic data handling—which are typical in many embedded and real-time applications.
-Because of its simplicity and minimal memory usage, Dhrystone\cite{weiss2002dhrystone} provides a good baseline measure of a processor’s raw computing capability.
-For this study, the RISC-V version of Dhrystone\cite{weiss2002dhrystone} was obtained from the official \texttt{riscv-tests} repository to ensure that it ran correctly on the target hardware.
-The benchmark was executed twice under identical conditions—once in a standard, non-secure mode, and once within a Keystone enclave.
-This setup made it possible to measure the specific performance overhead introduced by Keystone’s security features, such as memory isolation and the cost of switching between secure and non-secure execution contexts.
+The Dhrystone benchmark~\cite{weiss2002dhrystone} has been used for decades to measure general-purpose processor performance, especially in scenarios where floating-point operations and file I/O are not critical.
+Its focus on integer calculations and control structures—such as loops, function calls, and basic data handling—is representative of many embedded and real-time applications.
+Thanks to its simplicity and minimal memory usage, Dhrystone~\cite{weiss2002dhrystone} provides a clean baseline measure of raw processor capability.
+For this study, the RISC-V implementation of Dhrystone was obtained from the official \texttt{riscv-tests} repository to ensure accurate execution on the target platform.
+The benchmark was executed twice under identical conditions—once in native, non-secure mode and once inside a Keystone enclave.
+This dual setup facilitates precise measurement of the performance overhead imposed by Keystone’s security mechanisms, including memory isolation and context-switching between secure and non-secure worlds.
 
-Dhrystone results are typically expressed in terms of "Dhrystones per second," which simply counts how many iterations of the benchmark are completed in one second.
-To make the numbers more meaningful across different architectures, these results are often normalized to Dhrystone MIPS (DMIPS), which estimates how many millions of instructions per second the processor is effectively delivering.
-This helps smooth out differences between instruction sets and gives a more comparable figure.
-In the context of this study, comparing DMIPS between enclave and non-enclave runs highlights how much raw performance is lost (or retained) when using the TEE.
+Dhrystone results are typically reported as "Dhrystones per second," quantifying how many benchmark iterations complete in one second.
+For broader comparability across different processor architectures, these results are normalized to Dhrystone MIPS (DMIPS), which estimates the effective instruction throughput in millions of instructions per second.
+This normalization allows meaningful performance comparisons by smoothing out architectural differences.
+In this study, comparing DMIPS between enclave and native executions highlights the raw performance impact of enabling the TEE.
 
-To get a broader picture of system performance—particularly one that includes moderate memory use and more complex workloads—the CoreMark benchmark \cite{gal2012exploring} was also used.
-Developed by the Embedded Microprocessor Benchmark Consortium (EEMBC), CoreMark \cite{gal2012exploring} was designed to overcome some of the limitations of Dhrystone by providing a workload that’s a bit closer to real-world embedded applications.
-It includes tasks like linked list manipulation, matrix operations, and finite state machine processing, all of which are common in software running on embedded systems.
-While CoreMark \cite{gal2012exploring} is still CPU-focused, it introduces a moderate level of memory interaction through dynamic data structures, which allows it to reveal performance issues that Dhrystone might miss—especially those related to cache usage or memory protection features within a TEE.
+To complement Dhrystone and gain insight into more complex system behavior, the CoreMark benchmark~\cite{gal2012exploring} was also employed.
+Developed by the Embedded Microprocessor Benchmark Consortium (EEMBC), CoreMark~\cite{gal2012exploring} addresses some limitations of Dhrystone by incorporating workloads that are closer to real-world embedded applications.
+It includes linked list manipulations, matrix operations, and finite state machine processing—tasks common in embedded software.
+Although still CPU-centric, CoreMark~\cite{gal2012exploring} exercises the memory subsystem more heavily through dynamic data structures, thereby exposing performance effects related to cache usage and memory protection features within a TEE.
 
-The version of CoreMark \cite{gal2012exploring} used in this study came from the SiFive repository available on GitHub at \url{https://github.com/sifive/benchmark-coremark}, which provides a version optimized for RISC-V systems and ready to work with Keystone.
-Like Dhrystone \cite{weiss2002dhrystone}, CoreMark \cite{gal2012exploring} was run in both native and enclave modes to compare performance under each.
-Its results are reported in terms of "Iterations Per Second" (IPS), which tells us how many complete runs of the benchmark are executed every second.
-Since CoreMark includes both arithmetic and memory operations, IPS gives a more holistic view of how the system is performing—not just the processor, but also the memory subsystem.
+The CoreMark version used in this study was sourced from the SiFive repository on GitHub\footnote{\url{https://github.com/sifive/benchmark-coremark}, accessed August 2025}, providing a RISC-V-optimized build compatible with Keystone.
+Like Dhrystone, CoreMark was run in both native and enclave modes, with results reported as "Iterations Per Second" (IPS), indicating how many complete benchmark runs occur per second.
+Because CoreMark combines arithmetic and memory operations, IPS delivers a more comprehensive view of system performance, reflecting both CPU and memory subsystem efficiency.
 
-Beyond the main performance numbers (DMIPS for Dhrystone and IPS for CoreMark), this study also looked at a few supporting metrics to better understand system behavior.
-The first is overall execution time, which refers to how long it takes (in real-world, wall-clock time) to run each benchmark.
-Comparing execution time in enclave vs.
-non-enclave modes helps quantify latency introduced by secure execution.
-This kind of delay might come from things like extra memory checks, context switching, or other internal mechanisms used to isolate the enclave from the rest of the system.
+In addition to primary performance metrics—DMIPS for Dhrystone and IPS for CoreMark—this study also analyzed supporting metrics to better characterize system behavior.
+Overall execution time, measured in wall-clock seconds, was recorded for each benchmark run.
+Comparing execution times between enclave and native modes quantifies latency introduced by secure execution, including overheads from memory protection, context switching, and enclave isolation mechanisms.
 
-Another important metric is variability—specifically, the standard deviation of benchmark results across multiple runs.
-Each benchmark was executed several times under identical conditions, and the standard deviation was calculated to determine how consistent the performance was.
-Low standard deviation means the system performs reliably from run to run, while high deviation might indicate that performance is sensitive to system noise, scheduling delays, or other environmental factors.
+Variability across runs was assessed through the standard deviation of benchmark results over multiple iterations.
+A low standard deviation signifies consistent and reliable performance, whereas higher variability might indicate sensitivity to system noise, scheduling delays, or other environmental factors.
 
-Both Dhrystone\cite{weiss2002dhrystone} and CoreMark\cite{gal2012exploring} are CPU-intensive, which is exactly what makes them useful in this context.
-Dhrystone focuses narrowly on integer operations and control flow, giving a clean look at raw processor performance.
-CoreMark, on the other hand, introduces a bit more complexity and memory interaction, which makes it better suited for revealing the broader impact of Keystone’s security mechanisms.
-Neither benchmark uses file I/O or persistent storage, which means that this study does not explore storage-related features of the TEE, such as secure disk access or file encryption.
-That’s outside the scope here and would require a different kind of benchmarking approach.
+Both Dhrystone~\cite{weiss2002dhrystone} and CoreMark~\cite{gal2012exploring} are CPU-intensive benchmarks, making them particularly useful for this study’s focus on processor and memory subsystem behavior under secure execution.
+Dhrystone targets raw integer and control flow performance, offering a clear view of base processor throughput, while CoreMark’s more sophisticated workload reveals broader impacts of the TEE’s security mechanisms, especially regarding memory interactions.
+Neither benchmark involves file I/O or persistent storage; therefore, this study does not assess storage-related TEE features such as secure disk access or file encryption, which would require alternative benchmarking approaches.
 
-By using both benchmarks together—one simple and focused, the other broader and more realistic—this study creates a well-rounded picture of how Keystone affects performance.
-The addition of execution time and variability analysis provides further context, helping to separate consistent performance overhead from sporadic slowdowns or system noise.
-Overall, this methodology provides a clear, detailed view of the trade-offs involved when securing execution through a TEE on RISC-V platforms.
+By combining these two benchmarks with detailed timing and variability analysis, this study provides a well-rounded and nuanced understanding of the trade-offs involved in securing execution via Keystone enclaves on RISC-V platforms.
 
@@
 
-To examine the influence of microarchitectural cache parameters on enclave performance, this study employed a structured variation of Level-1 (L1) cache configurations using QEMU's cache modeling capabilities.
-The experimental objective was to isolate the impact of cache size, associativity, eviction policy, and core count on the performance of secure and non-secure workloads in a controlled emulation environment.
+To investigate the architectural impact on enclave execution, this study systematically varied system parameters including memory size, core count, execution mode (sequential vs parallel), and cache configurations using QEMU’s cache modeling capabilities.
+The aim was to understand how these factors influence the behavior of trusted and non-trusted workloads under different resource constraints.
 
-Cache modeling was enabled through the QEMU TCG plugin framework \cite{mandour2021cache}, which simulates idealized, per-core instruction and data caches.
-While not microarchitecturally cycle-accurate, this plugin supports a configurable cache hierarchy, enabling systematic exploration of cache behaviors relevant to trusted execution environments.
+Cache simulation was conducted using the QEMU TCG plugin framework \cite{mandour2021cache}, which models per-core instruction and data caches in an idealized fashion.
+While this plugin is not cycle-accurate and does not isolate enclave-specific behavior, it enables consistent modeling of cache accesses and miss patterns across the entire QEMU runtime.
+This was sufficient to observe cache-related effects caused by workload behavior in different configurations.
 
-Four key experimental scenarios were defined:
+To avoid excessive experimental overhead and redundant trends, five representative system configurations were selected.
+These were chosen to cover a wide range of memory capacities, CPU core counts, and execution modes.
+Each configuration enables a specific type of insight, ensuring a diverse but manageable evaluation matrix.
 
-\begin{enumerate}
-    \item \textbf{Baseline Configuration:} A minimal cache setup with 8\,KB instruction and data caches, 4-way associativity, and LRU eviction.
-    \item \textbf{Larger Cache Scenario:} An expanded L1 cache configuration with 32\,KB size and 8-way associativity to evaluate the effect of reduced cache miss rates.
-    \item \textbf{Multicore Simulation:} A four-core configuration with private per-core caches to study inter-core effects such as contention and cache coherence approximations.
-    \item \textbf{Eviction Policy Sensitivity:} A fixed-size cache using alternative eviction strategies (FIFO and random) to assess their influence on enclave memory behavior.
-\end{enumerate}
+\begin{table}[h]
+\centering
+\begin{tabular}{|l|c|c|c|p{7.5cm}|}
+\hline
+\textbf{Config Name} & \textbf{RAM} & \textbf{Cores} & \textbf{Execution} & \textbf{Rationale} \\
+\hline
+Low-End Baseline     & 64M         & 1             & Sequential         & Serves as the minimal configuration to establish baseline performance.
+Useful for understanding enclave overhead under tight resource constraints.
+\\
+\hline
+Balanced Parallel    & 128M        & 2             & Parallel           & Captures early-stage multicore scaling behavior. Highlights effects of light parallelism under modest memory availability. \\
+\hline
+Mid-Range Sequential & 256M        & 4             & Sequential         & Enables evaluation of scaling effects due to additional cores and memory, while keeping execution sequential to avoid parallel noise.
+\\
+\hline
+High-End Parallel    & 512M        & 6             & Parallel           & Represents a well-provisioned system where both memory and core resources are sufficient.
+Useful for observing near-ideal parallel execution.
+\\
+\hline
+Max Capacity Stress  & 2G          & 10            & Parallel           & Tests the upper bounds of simulated platform capacity.
+Helps identify performance saturation and the point of diminishing returns.
+\\
+\hline
+\end{tabular}
+\caption{Summary of representative configurations used in performance evaluation.}
+\end{table}
 
-All scenarios maintained a fixed block size of 64 bytes to preserve spatial locality characteristics.
-The same set of benchmarks—Dhrystone and CoreMark—was executed under each configuration, both in native and enclave modes, to quantify differences in execution time, throughput, and cache performance.
+Several other configurations—such as single-core setups with excessive memory or high-core setups with limited memory—were explored during preliminary testing.
+However, they offered little additional insight beyond what was observable in the selected set.
+For instance, excessive memory on a single core often led to underutilization, while many-core systems with small memory pools frequently resulted in excessive context switching.
+Both behaviors were captured effectively by existing selected configurations.
 
-The plugin-generated metrics included instruction and data cache accesses, miss rates, and identification of cache-thrashing instructions.
-While detailed configuration parameters, plugin invocation commands, and sample output logs are essential for reproduction, they are provided in Appendix~\ref{appendix:cache-plugin} to avoid interrupting the main methodological flow.
+Across all system configurations, three cache profiles were tested: a baseline setup with 8\,KB 4-way LRU caches, a larger 32\,KB 8-way configuration, and a fixed-size cache using FIFO and random eviction strategies.
+A consistent block size of 64 bytes was used to preserve spatial locality.
+Although the cache plugin logs system-wide activity, observed trends were analyzed in conjunction with benchmark behavior to infer their likely contributions to overall performance variation.
 
-By varying one parameter set at a time while controlling for others, this approach facilitated a clear attribution of observed performance differences to specific architectural characteristics.
-The resulting data informed the broader analysis of how cache-aware design considerations can impact enclave performance in systems based on the Keystone framework.
+Benchmarks were run in both native and enclave execution modes to enable direct comparisons.
+This design facilitated the isolation of architectural performance trends and the specific overhead introduced by trusted execution.
 
@@
 
-To ensure that the performance evaluation of enclave execution is both statistically robust and reproducible, a systematic approach was taken to data collection and analysis.
-All benchmarks were executed within the same virtualized RISC-V environment described in Section~\ref{chap:methodology}, with controlled cache parameters as outlined in Section~\ref{sec:param-variation}.
+All benchmarks were executed within a consistent virtualized RISC-V environment using QEMU.
+For each selected configuration, both Dhrystone and CoreMark benchmarks were run in native and enclave modes to capture the impact of enclave isolation under different resource and cache conditions.
 
-Each benchmark (Dhrystone and CoreMark) was executed under two configurations: native execution and enclave execution.
-For every configuration, a minimum of ten independent iterations was conducted to account for runtime variability.
-Execution time was measured using wall-clock timing as reported by the host Linux environment inside QEMU.
-Benchmark-specific performance metrics—Dhrystone MIPS (DMIPS) and CoreMark iterations per second (IPS)—were recorded after each run.
+Each experiment consisted of ten executions.
+To reduce variance caused by warm-up behavior, the first run was discarded, and the remaining nine were used to compute average performance.
+Key performance metrics included total wall-clock execution time, benchmark throughput (DMIPS or iterations/sec), and standard deviation.
 
-The collected data for each experiment included:
+In addition to these primary measurements, system-level runtime statistics were collected from within the guest Linux environment. These included:
 
 \begin{itemize}
-    \item Total execution time (seconds)
-    \item Benchmark throughput (DMIPS or IPS)
-    \item Standard deviation and coefficient of variation
-    \item Cache miss rates (instruction/data) when applicable
+    \item User and system CPU time
+    \item Maximum memory usage (RSS)
+    \item Minor and major page faults
+    \item Voluntary and involuntary context switches
 \end{itemize}
 
-To mitigate the effects of transient system behavior, the first run of each benchmark was discarded as a warm-up iteration.
-The remaining samples were then averaged, and both the mean and standard deviation were computed.
-These values were used to assess not only the central tendency of performance, but also the stability of enclave execution under different cache configurations.
+These metrics were essential to understanding how different system configurations affected workload behavior.
+For example, high numbers of involuntary context switches in parallel configurations often indicated memory contention or over-subscription of CPU cores.
 
-For cache-sensitive experiments, data was further enriched using QEMU's cache modeling plugin \cite{mandour2021cache}.
-This provided detailed logging of instruction and data accesses, cache miss counts, and identification of high-miss instructions.
-From these logs, miss rates were extracted and correlated with execution behavior.
-Selected logs and plugin outputs are included in Appendix~\ref{appendix:cache-plugin}.
+Cache-related metrics were gathered using the QEMU cache plugin, which logs all instruction and data cache accesses and misses at the system level.
+Although enclave-specific cache statistics could not be isolated, the relative trends between native and enclave executions under identical configurations allowed for meaningful comparative analysis.
 
-Statistical analysis was conducted using standard descriptive statistics.
-Where relevant, normalized overhead values were computed to express enclave slowdown as a percentage increase in execution time or a reduction in throughput compared to the native baseline.
-These comparative metrics provide a clearer view of the cost of enclave isolation across various architectural conditions.
+Performance overhead introduced by enclave execution was quantified using the percentage increase in execution time compared to native mode, calculated as:
 
-The consistent use of controlled environments, repeated measurements, and structured analysis ensures that the results presented in subsequent chapters are both methodologically sound and analytically valid.
+\[
+\text{Overhead (\%)} = \left( \frac{\text{Enclave Time} - \text{Native Time}}{\text{Native Time}} \right) \times 100
+\]
+
+This normalized metric enabled consistent cross-comparison between configurations with vastly different execution baselines.
+
+All benchmark scripts, plugin configurations, and raw logs were version-controlled and archived.
+Detailed plugin output, configuration parameters, and system call traces are provided in Appendix~\ref{appendix:cache-plugin} for transparency and reproducibility.

### 2025-08-08 chapters/system-architecture.tex
changed verbatim to lstlistings

@@
 \label{chap:methodology}
-The primary objective of this thesis is to assess the performance impact of Keystone’s enclave isolation mechanisms on typical embedded and systems-level workloads.
-Specifically, the study seeks to quantify the computational overhead introduced by executing applications within a Keystone enclave, as compared to their execution in a conventional, non-isolated environment.
-To achieve this, a series of carefully controlled experiments were designed and executed within a virtualized RISC-V system based on QEMU.
 
-All benchmarks were executed in an emulated environment configured for the RV64GC architecture, which is representative of 64-bit RISC-V systems targeted by Keystone.
-The use of QEMU for emulation allowed for fine-grained control over the system environment, ensured repeatability, and eliminated the variability introduced by physical hardware, such as thermal throttling or hardware-specific optimizations.
-Although virtualized, the emulation accurately models instruction execution, privilege transitions, memory access patterns, and I/O behavior, making it suitable for preliminary performance evaluation.
+The primary objective of this thesis is to evaluate the performance impact of Keystone’s enclave isolation mechanisms on representative embedded and systems-level workloads.
+Specifically, the study aims to quantify the computational overhead introduced by executing applications within a Keystone enclave, compared to their execution in a traditional, non-isolated environment.
+To this end, a series of controlled experiments were conducted on a virtualized RISC-V platform based on QEMU.
 
-To evaluate the impact of enclave execution, each benchmark was implemented as a standalone enclave application (referred to as an “eapp”), accompanied by a corresponding host application that manages enclave lifecycle events.
-The host application, running in user space, is responsible for initializing the enclave, loading the benchmark binary, invoking enclave entry points, and handling edge calls used for communication between the enclave and the host.
-The enclave application, in contrast, is isolated by the Keystone runtime and contains only the benchmark logic and internal data structures.
-This modular separation ensures that the same benchmark code can be executed both natively (without an enclave) and securely (within an enclave) without structural modification, enabling an accurate comparative analysis.
+All benchmarks were executed within an emulated environment configured for the RV64GC architecture, which corresponds to the 64-bit RISC-V ISA targeted by Keystone.
+The QEMU-based emulation provides fine-grained control over system parameters and ensures experimental repeatability by eliminating variability introduced by physical hardware—such as thermal throttling, cache behavior, or platform-specific optimizations.
+While virtualized, QEMU accurately models instruction execution, privilege-level transitions, memory access patterns, and I/O behavior, making it a viable platform for preliminary performance evaluations of trusted execution environments.
 
-The evaluation involved executing each benchmark—Dhrystone and CoreMark—under two distinct configurations:
+To assess the performance implications of enclave-based isolation, each benchmark was implemented as a standalone enclave application (referred to as an \textit{eapp}), accompanied by a corresponding host application responsible for enclave lifecycle management.
+The host application, running in user space, handles enclave initialization, benchmark loading, invocation of enclave entry points, and inter-domain communication via edge calls.
+In contrast, the enclave application is fully isolated by the Keystone runtime and contains only the benchmark logic and associated data structures.
+This strict separation ensures that the same benchmark codebase can be executed both natively (without isolation) and securely (within an enclave), without requiring structural changes—thus enabling a fair and consistent comparative analysis.
+
+The experimental evaluation was conducted using two well-established benchmarking tools—Dhrystone and CoreMark—each executed under two configurations:
 
 \begin{enumerate}
-\item \textit{Native Execution:} The benchmark runs as a conventional user-space process within the QEMU-emulated Linux system, without invoking Keystone enclave services.
-\item \textit{Enclave Execution:} The benchmark is loaded into a Keystone enclave, with isolation enforced by the Security Monitor using PMP.
+\item \textbf{Native Execution:} The benchmark runs as a conventional user-space process within the QEMU-emulated Linux system, with no involvement of Keystone's enclave infrastructure.
+\item \textbf{Enclave Execution:} The benchmark is loaded into a Keystone enclave, where memory isolation is enforced by the Security Monitor through RISC-V Physical Memory Protection (PMP) mechanisms.
 \end{enumerate}
 
-Each benchmark was run multiple times (typically ten iterations per configuration) to account for performance variability and ensure statistical robustness.
-For each run, metrics such as total execution time, throughput (measured in DMIPS for Dhrystone and iterations per second for CoreMark), and standard deviation were collected.
-The comparative analysis focused on the relative performance degradation observed in the enclave configuration, thereby providing a direct measurement of the cost of security.
+Each benchmark configuration was executed multiple times (typically ten iterations) to mitigate performance variance and ensure statistical robustness.
+For each run, performance metrics including total execution time, throughput (expressed in DMIPS for Dhrystone and iterations per second for CoreMark), and standard deviation were collected.
+The analysis focuses on the relative performance degradation incurred in the enclave execution scenario, thereby providing a direct estimate of the overhead associated with enclave-based isolation.
 
-This methodology enables a nuanced understanding of how Keystone’s isolation features influence real-world performance metrics.
-The controlled nature of the test environment, coupled with repeated measurement and the use of industry-standard benchmarks, ensures that the results are both meaningful and reproducible.
-By isolating the effect of enclave execution, this study contributes valuable insights into the trade-offs between security and efficiency in open-source TEE architectures.
+This methodology facilitates a detailed understanding of how Keystone’s architectural design and isolation primitives influence runtime performance.
+The controlled and repeatable test environment, combined with standardized benchmarking tools and consistent measurement practices, ensures that the results are both reliable and meaningful.
+Ultimately, the findings offer valuable insights into the trade-offs between security and efficiency in open-source TEE frameworks and inform future efforts in trusted system design.
 
@@
 
-\begin{verbatim}
+\begin{lstlisting}
 make QEMU_PLUGIN_ARGS="dcachesize=8192,dassoc=4,dblksize=64, \
 icachesize=8192,iassoc=4,iblksize=64,cores=2,evict=lru" run
-\end{verbatim}
+\end{lstlisting}
 
@@
 
-\begin{verbatim}
+\begin{lstlisting}
 modprobe keystone-driver
-\end{verbatim}
+\end{lstlisting}
 
@@
 
-\begin{verbatim}
+\begin{lstlisting}
 KEYSTONE_DEBUG=y make run
 make debug-connect
-\end{verbatim}
+\end{lstlisting}
 

### 2025-08-09 chapters/benchmark.tex
updated TOC

--- 2025-06-16 chapters/benchmark.tex
@@
-\chapter{Benchmarking Evaluation}
+\chapter{Benchmarking Results and Observations}
 \label{chap:benchmarking}
 
-\section{Native Execution Performance}
+This chapter provides a detailed empirical performance evaluation of a set of benchmark workloads executed within the Keystone enclave framework.
+The primary objective is to quantitatively assess the overheads and limitations introduced by enclave-based execution, focusing on how system-level parameters and hardware constraints influence workload behavior under trusted execution.
 
-\section{Enclave Execution Performance}
+The evaluation covers a mix of general-purpose and cryptographic workloads.
+Specifically, we utilize the Dhrystone and CoreMark benchmarks, which are standard tools for measuring integer processing performance and embedded system efficiency, respectively.
+Additionally, the Kyber post-quantum cryptographic algorithm is included to evaluate a compute-intensive, security-critical workload relevant for future-proof cryptographic applications.
+This combination allows us to characterize enclave performance across a spectrum of computational patterns and resource demands.
 
-\section{Kyber Algorithm in the Enclave: Performance Impact}
+\section{Baseline: Native (Non-Enclave) Performance}
 
-\section{Performance Parameters: CPU Cores, Memory Allocation, Cache Size}
+To establish a baseline for evaluating enclave overheads, we first executed all benchmarks natively on the simulated RISC-V platform without any enclave isolation.
+This phase allows us to measure the raw performance characteristics of the system and isolate the costs introduced solely by secure execution.
 
-\section{Benchmarking Results: Native vs Enclave Execution}
-%\section{Performance Bottlenecks and Overheads in Enclave Execution}
+The Dhrystone and CoreMark benchmarks exhibited expected scaling behavior with respect to CPU cores and memory size.
+As CPU core count increased, throughput for these parallelizable benchmarks improved approximately linearly, reflecting efficient multi-threaded execution on the underlying hardware.
+Memory size impacted CoreMark more noticeably, particularly during memory-intensive phases, where increased memory reduced paging and improved cache hit rates.
+In contrast, Dhrystone’s smaller working set resulted in negligible sensitivity to memory size variations.
+
+The Kyber cryptographic benchmark behaved differently.
+Due to its inherently sequential algorithmic structure, Kyber’s execution time remained relatively constant across changes in CPU core count.
+Likewise, memory size had minimal impact on Kyber performance, as its working set comfortably fit within allocated memory without inducing paging.
+
+These native baseline results confirm the stability and predictability of the simulated system environment, providing a solid foundation for subsequent enclave overhead analysis.
+
+\section{Enclave Execution Results and Overheads}
+
+Execution within Keystone enclaves introduces overhead from multiple sources: memory isolation enforced by PMP, additional context switches to manage transitions between secure and non-secure execution states, and the general overhead of enforcing enclave boundaries at runtime.
+
+Across all benchmarks, enclave execution resulted in increased execution times compared to native runs.
+Overhead magnitude varied by workload and system configuration but generally remained below 15\%.
+For parallel workloads such as CoreMark, overhead tended to increase with the number of CPU cores, likely due to increased synchronization demands and frequent enclave exit and re-entry operations during thread management.
+Dhrystone showed more modest overhead growth with core scaling.
+
+Cache behavior, measured indirectly via QEMU’s cache modeling plugin, indicated modest increases in cache miss rates during enclave execution.
+Although this plugin provides system-wide cache metrics rather than enclave-specific data, the observed increases likely stem from PMP-related overheads such as cache flushing or restricted memory access patterns.
+These overheads are an expected consequence of maintaining strong memory isolation guarantees.
+
+A critical observation during enclave execution was the enforcement and limitation of PMP resources.
+The fixed PMP register count on the QEMU virt platform fundamentally restricts the number of enclaves that can be concurrently instantiated.
+When the number of requested enclaves exceeds the 16 PMP entries, enclave creation fails.
+These failures are characterized by SBI call error codes and logged kernel errors indicating that no PMP registers are available.
+This limitation underscores the importance of considering PMP capacity in enclave orchestration and resource allocation.
+
+\section{Workload Sensitivity Analysis (Kyber vs CoreMark vs Dhrystone)}
+
+The Kyber post-quantum cryptographic algorithm is a representative workload for evaluating the performance impact of enclave execution on security-critical applications.
+Kyber consists of three primary operations—key generation, encapsulation, and decapsulation—each profiled extensively within enclaves using the Keystone timer API to capture precise cycle counts.
+
+Execution within enclaves increased Kyber’s runtime by approximately 7–9\% compared to native execution.
+This overhead is modest given the cryptographic operations’ complexity and the security benefits provided by enclaves.
+
+As with native execution, Kyber performance was insensitive to increases in CPU core count due to its inherently sequential design.
+Memory usage during enclave execution remained stable, with no indications of paging or contention.
+System-wide cache miss rates observed through QEMU remained consistent with Kyber’s memory access patterns, including phases dominated by number-theoretic transforms and polynomial arithmetic.
+
+From a resource perspective, the requirement of one PMP entry per enclave constrains the number of parallel Kyber instances that can be simultaneously deployed.
+This constraint must be taken into account when designing systems requiring multiple isolated cryptographic operations.
+
+\section{Impact of Hardware Parameters (Cores, Memory, Cache)}
+
+To comprehensively assess how hardware configurations affect enclave performance, we systematically varied CPU core counts, memory allocations, and cache configurations.
+
+Scaling the number of CPU cores yielded anticipated improvements for parallel workloads, with throughput gains plateauing beyond six cores due to synchronization overhead and increased enclave management complexity.
+Sequential workloads such as Kyber remained unaffected by core scaling.
+
+Memory allocation experiments revealed that increasing memory size from 64MB to 2GB improved performance for memory-intensive benchmarks by reducing paging and increasing cache hit rates.
+However, beyond approximately 512MB, gains diminished, indicating sufficient memory provision for the workload’s working set.
+
+Cache configuration impacts were evaluated using the QEMU cache modeling plugin.
+Larger cache sizes and higher associativity reduced miss rates, particularly for instruction-heavy workloads like Dhrystone.
+While the plugin’s system-wide scope limits insights into enclave-specific cache activity, the trends suggest that cache tuning can meaningfully influence enclave performance.
+
+Importantly, the relationship between memory allocation and PMP usage is nuanced.
+Larger enclave memory regions may require multiple PMP entries if the memory cannot be described contiguously in a single PMP register, effectively reducing the total number of enclaves supported simultaneously.
+This interaction between memory layout and PMP capacity represents a critical consideration for enclave deployment strategies.

### 2025-08-09 chapters/discussion.tex
updated TOC

@@
-\chapter{Discussion}
+\chapter{Interpretation and Broader Implications}
 \label{chap:discussion}
 
-\section{Key Findings and Insights}
+\section{Synthesis of Key Performance Trends}
+This chapter integrates and interprets the empirical results obtained from benchmarking Keystone enclaves, offering a comprehensive synthesis of performance characteristics, limitations, and broader implications.
+The experimental data reveal several critical insights into the practicalities and challenges of enclave execution on RISC-V platforms, which have direct relevance to the design and deployment of secure computing systems.
 
-%\section{Challenges in Benchmarking Enclaves and Cryptographic Algorithms}
+First and foremost, enclave execution imposes non-negligible but moderate overheads compared to native, non-isolated execution.
+These overheads predominantly arise from the enforcement of memory isolation mechanisms, including Physical Memory Protection (PMP), and the additional context switches required to manage transitions between secure (enclave) and non-secure (host) execution states.
+Our results show that, across a diverse set of benchmarks ranging from general-purpose integer workloads to specialized cryptographic algorithms, overheads typically remain under 15%.
+However, this aggregate figure masks significant variability depending on workload type, degree of parallelism, and system configuration parameters such as CPU cores, memory allocation, and cache hierarchy.
 
-\section{Trade-offs Between Security and Performance in Enclaves}
+A second major insight relates to the critical role of hardware-enforced resource constraints, specifically the limited number of PMP entries.
+The QEMU virt platform utilized in our experiments provides 16 PMP registers, which directly limits the number of isolated enclaves that can simultaneously operate.
+Attempts to instantiate more enclaves than PMP entries available result in immediate failure of enclave creation, as indicated by distinct Supervisor Binary Interface (SBI) error codes and kernel log messages.
+This finding underscores the fundamental impact of hardware architecture on enclave scalability and points to PMP capacity as a bottleneck that must be carefully managed in real-world systems.
 
-\section{Implicaitons for Secure Computing and Cryptographic Applications}
+Thirdly, workload characteristics strongly influence enclave performance overheads.
+Compute-bound, sequential workloads such as the Kyber post-quantum cryptographic algorithm experience comparatively low performance degradation within enclaves.
+This is consistent with Kyber’s minimal inter-thread communication and largely predictable memory access patterns.
+In contrast, highly parallelizable benchmarks like CoreMark exhibit increasing overhead with core count, reflecting the compounded costs of synchronization, enclave entry/exit transitions, and possibly contention in PMP configuration.
+This differential sensitivity implies that enclave performance optimization must be workload-specific, taking into account application parallelism and memory footprint.
+
+Lastly, the interaction between resource allocation parameters—CPU cores, memory size, and cache configuration—and enclave performance is multifaceted.
+Increasing CPU cores yields diminishing returns beyond a threshold due to synchronization and enclave management overhead.
+Larger memory allocations improve performance primarily by reducing paging and improving cache locality but also can exacerbate PMP usage if memory regions are fragmented.
+Cache configuration impacts, though indirectly measured via QEMU’s system-wide cache model, suggest that optimizing cache size and associativity can mitigate enclave overheads, especially for instruction-intensive workloads.
+These observations highlight the necessity of a holistic approach to system tuning in enclave environments.
+
+\section{Hardware Bottlenecks and PMP Utilization Strategies}
+A central and recurring theme throughout this study is the pivotal role played by the RISC-V Physical Memory Protection (PMP) mechanism in determining the scalability, flexibility, and reliability of secure enclave execution in the Keystone framework.
+Although PMP is indispensable for maintaining strong memory isolation—arguably the most critical security property in any Trusted Execution Environment (TEE)—it simultaneously introduces significant architectural constraints that place hard limits on system scalability.
+These limitations become especially apparent in environments that demand the concurrent instantiation of multiple enclaves, such as secure edge computing deployments, cloud-based confidential workloads, and multi-tenant embedded systems.
+
+PMP operates by establishing fine-grained access control policies over physical memory ranges.
+These policies are enforced in hardware and serve to prevent unauthorized access by less privileged software layers, such as applications running in Supervisor or User mode.
+In the Keystone architecture, enclaves rely on PMP to ensure that their memory regions remain inaccessible to both the host operating system and other enclaves.
+Each protected memory region is described by a PMP entry, and these entries are programmed through control and status registers (CSRs) accessible only from the highest privilege level—Machine mode (M-mode).
+Each PMP entry specifies a memory region, along with the access permissions allowed (read, write, execute), and once configured, becomes an unforgeable part of the system’s isolation model.
+
+However, the number of available PMP entries is not dynamic—it is a fixed architectural parameter determined by the processor implementation.
+In our experimental setup based on the widely used QEMU \texttt{virt} platform for RISC-V, this number is statically defined as 16.
+This finite resource pool becomes a fundamental limiting factor in the Keystone system because \textbf{each enclave requires at least one PMP entry} to establish its isolated memory region.
+In more complex memory layouts—especially those that are large or fragmented—multiple PMP entries may be needed to define non-contiguous or irregularly sized memory regions, further exacerbating resource pressure.
+
+This constraint manifested clearly in our benchmarking experiments.
+When the number of concurrently requested enclaves exceeded the available PMP entries, enclave creation failed immediately.
+These failures were not silent; they were explicitly reported by the system in the form of Supervisor Binary Interface (SBI) error codes and verbose kernel log messages, providing clear indicators of PMP exhaustion.
+Typical output during such events included:
+
+\begin{lstlisting}[language=bash,caption={Error messages during enclave creation and destruction}]
+napot_region_init: No available PMP register
+keystone_create_enclave: SBI call failed with error code 100003
+fatal: cannot destroy enclave: SBI failed with error code 100005
+pmp_unset: Invalid PMP region index
+pmp_region_free_atomic: Invalid PMP region index
+\end{lstlisting}
+
+These logs reflect a hard architectural ceiling that is entirely independent of other system resources such as CPU cores or DRAM availability.
+Even in scenarios where there is ample memory and processing capacity to support additional enclaves, the inability to allocate a free PMP entry effectively blocks the deployment of further isolated workloads.
+This introduces a non-obvious, hardware-enforced bottleneck that could significantly impact the reliability and scalability of enclave-enabled systems.
+
+Moreover, the situation is further complicated by the way PMP regions are encoded.
+PMP supports a limited set of region encoding formats, with the most commonly used being the Naturally Aligned Power-of-Two (NAPOT) mode.
+This format requires that protected memory regions be both naturally aligned and sized as a power of two.
+While this encoding simplifies hardware enforcement, it imposes rigid constraints on memory layout.
+For example, if an enclave's memory region spans an irregular range—such as 192 KB starting at a non-aligned address—it cannot be expressed with a single PMP entry.
+In such cases, the region must be broken down into smaller, aligned sub-regions, each requiring its own PMP entry.
+This fragmentation effect leads to inefficient utilization of available PMP entries and further reduces the maximum number of enclaves that can coexist.
+
+To address these challenges, both \textbf{hardware-level} and \textbf{software-level} mitigation strategies must be considered:
+
+\begin{itemize}
+    \item \textbf{Hardware approaches} include increasing the number of available PMP entries.
+While this would require architectural changes and potentially more silicon real estate, it could dramatically improve enclave concurrency.
+Alternatively, future iterations of the RISC-V PMP specification could explore more flexible memory region encodings, allowing for more expressive and compact representations of protected regions, thereby reducing fragmentation.
+    \item \textbf{Software techniques} include dynamic PMP entry management, where PMP entries are reused or reconfigured at runtime based on the current workload and enclave activity.
+This requires careful coordination to avoid security violations but can enable more flexible use of the limited register set.
+Another technique is enclave multiplexing, where multiple lightweight tasks or applications are combined into a single enclave, reducing the total number of enclaves required.
+Additionally, memory layout optimization—aligning and sizing enclave memory regions to match NAPOT constraints—can significantly reduce the number of PMP entries consumed per enclave.
+\end{itemize}
+
+From a systems engineering perspective, these limitations have real-world consequences.
+In cloud environments, where tenants may each be assigned their own isolated enclave for security, PMP constraints could become a source of contention, limiting the number of clients that can be served simultaneously.
+In embedded or edge computing devices with strict memory constraints, careful tuning of enclave footprints and memory allocation strategies will be necessary to ensure system robustness.
+
+In conclusion, while the PMP mechanism is undeniably a cornerstone of the Keystone security model, it also introduces a set of hard limitations that impact the practical deployment and scaling of enclaves.
+Understanding and addressing these limitations—both through hardware evolution and smarter software resource management—is essential for building scalable, secure computing platforms based on the RISC-V architecture.
+
+\section{Implications for Secure System Design and Cryptographic Workloads}
+
+\section{Recommendations and Considerations for Future TEE Deployments}

### 2025-08-10 chapters/benchmark.tex
minor grammatical modifications

@@
 
-This chapter provides a detailed empirical performance evaluation of a set of benchmark workloads executed within the Keystone enclave framework.
-The primary objective is to quantitatively assess the overheads and limitations introduced by enclave-based execution, focusing on how system-level parameters and hardware constraints influence workload behavior under trusted execution.
+This chapter presents a detailed empirical evaluation of the Keystone enclave framework, with a specific focus on its performance characteristics under a range of representative workloads.
+The primary aim is to quantify the computational and system-level overheads introduced by secure enclave execution on RISC-V platforms, and to explore how these overheads are influenced by underlying hardware configurations and architectural constraints.
+By systematically comparing enclave-based execution against native (non-isolated) execution, we aim to identify the key trade-offs and limitations that developers and system architects must consider when adopting enclaves for trusted execution.
 
-The evaluation covers a mix of general-purpose and cryptographic workloads.
-Specifically, we utilize the Dhrystone and CoreMark benchmarks, which are standard tools for measuring integer processing performance and embedded system efficiency, respectively.
-Additionally, the Kyber post-quantum cryptographic algorithm is included to evaluate a compute-intensive, security-critical workload relevant for future-proof cryptographic applications.
-This combination allows us to characterize enclave performance across a spectrum of computational patterns and resource demands.
+To achieve this, we deploy a carefully selected set of benchmark workloads that span both general-purpose and security-critical application domains.
+These include the Dhrystone and CoreMark benchmarks—commonly used to assess CPU performance and embedded system efficiency—as well as the Kyber post-quantum cryptographic algorithm, which provides a real-world workload with stringent security and computational requirements.
+Together, these benchmarks provide a diverse and comprehensive view of how Keystone enclaves behave across different execution contexts.
+
+The benchmarking process is structured to examine four interrelated dimensions.
+First, we establish a native performance baseline to isolate the impact of enclave-related operations.
+Second, we evaluate the performance of the same workloads when executed within enclaves, highlighting the latency and resource costs incurred by isolation mechanisms such as Physical Memory Protection (PMP) and context switching.
+Third, we perform a workload sensitivity analysis, comparing the behavior of different benchmark types—sequential vs.
+parallel, lightweight vs.
+compute-intensive—to determine how workload characteristics affect enclave performance.
+Finally, we assess the influence of key hardware parameters, including CPU core count, memory size, and cache configuration, on both native and enclave execution.
+
+Through these experiments, we uncover both expected and non-trivial behaviors.
+For instance, while enclave overheads are generally modest—typically under 15\%—they are highly sensitive to parallelism and memory access patterns.
+Furthermore, we identify critical bottlenecks, such as the fixed number of PMP entries available for isolating memory regions, which directly limit enclave concurrency and scalability.
+
+The results presented in this chapter serve as the empirical foundation for the interpretive analysis in the next chapter.
+By rigorously characterizing the performance profile of Keystone enclaves, we establish a data-driven basis for understanding the broader implications of enclave deployment in secure system design, as well as informing practical strategies for optimization and resource management.
 

### 2025-08-10 chapters/discussion.tex
minor grammatical modifications

@@
 
+This chapter presents a critical analysis of the experimental findings discussed in the preceding chapters, with the goal of extracting deeper insights into the behavior, limitations, and practical consequences of enclave-based execution using the Keystone framework on RISC-V platforms.
+While the benchmarking results provide quantitative evidence of performance trends and system behavior, this chapter aims to interpret those findings in a broader architectural and systems-level context.
+
+Specifically, we examine how various system parameters—such as CPU core count, memory configuration, cache hierarchy, and enclave workload characteristics—influence performance under isolation.
+We also discuss the architectural constraints imposed by RISC-V's Physical Memory Protection (PMP) mechanism, highlighting its central role in enforcing security while simultaneously acting as a scalability bottleneck.
+By analyzing the trade-offs between security guarantees and system throughput, we offer a nuanced understanding of the design space that system architects and security engineers must navigate when deploying enclaves in resource-constrained or high-concurrency environments.
+
+In addition to synthesizing the key performance trends, this chapter explores the broader implications of enclave deployment for secure system design, particularly in scenarios involving cryptographic workloads, multi-tenant execution, and edge computing platforms.
+We assess the extent to which Keystone and PMP-based isolation mechanisms can be reliably scaled, and we consider practical strategies for overcoming current limitations through architectural or software-level optimizations.
+
+Finally, the chapter concludes by formulating design recommendations and identifying future directions for research and development.
+These include enhancements to hardware features such as PMP flexibility, improved enclave management strategies at the software layer, and potential extensions to the Keystone framework to support more scalable and efficient trusted execution environments.
+
+This chapter transitions from measurement to meaning: it connects the observed data to the architectural principles underpinning enclave execution, thereby informing both future research and practical deployment of secure computing systems based on open RISC-V architectures.
+
 \section{Synthesis of Key Performance Trends}
-This chapter integrates and interprets the empirical results obtained from benchmarking Keystone enclaves, offering a comprehensive synthesis of performance characteristics, limitations, and broader implications.
-The experimental data reveal several critical insights into the practicalities and challenges of enclave execution on RISC-V platforms, which have direct relevance to the design and deployment of secure computing systems.
 
-First and foremost, enclave execution imposes non-negligible but moderate overheads compared to native, non-isolated execution.
-These overheads predominantly arise from the enforcement of memory isolation mechanisms, including Physical Memory Protection (PMP), and the additional context switches required to manage transitions between secure (enclave) and non-secure (host) execution states.
-Our results show that, across a diverse set of benchmarks ranging from general-purpose integer workloads to specialized cryptographic algorithms, overheads typically remain under 15%.
-However, this aggregate figure masks significant variability depending on workload type, degree of parallelism, and system configuration parameters such as CPU cores, memory allocation, and cache hierarchy.
+A key observation from the evaluation is that enclave execution introduces moderate but measurable overheads relative to native, non-isolated execution.
+These overheads primarily stem from two factors: the enforcement of memory isolation through Physical Memory Protection (PMP), and the additional control flow transitions between enclave and host contexts.
+Across a range of benchmark scenarios—including general-purpose integer computations and cryptographic workloads—the performance penalty typically remains below 15\%.
+However, this aggregate figure conceals significant variance, which is highly dependent on workload characteristics, system architecture, and execution configuration.
+Variables such as the number of processor cores, memory availability, and cache configuration exert a substantial influence on the observed overhead.
 
-A second major insight relates to the critical role of hardware-enforced resource constraints, specifically the limited number of PMP entries.
-The QEMU virt platform utilized in our experiments provides 16 PMP registers, which directly limits the number of isolated enclaves that can simultaneously operate.
-Attempts to instantiate more enclaves than PMP entries available result in immediate failure of enclave creation, as indicated by distinct Supervisor Binary Interface (SBI) error codes and kernel log messages.
-This finding underscores the fundamental impact of hardware architecture on enclave scalability and points to PMP capacity as a bottleneck that must be carefully managed in real-world systems.
+A second and arguably more fundamental insight concerns the limitations imposed by hardware-enforced resource constraints, particularly the finite number of PMP entries.
+In the QEMU \texttt{virt} platform used for this study, the system supports a maximum of 16 PMP entries.
+Each enclave requires at least one such entry to define a protected memory region.
+This architectural constraint creates a hard upper bound on the number of enclaves that can be concurrently instantiated.
+When this limit is exceeded, enclave creation fails deterministically, with the system reporting specific Supervisor Binary Interface (SBI) error codes and kernel log messages.
+This result highlights a critical bottleneck for enclave scalability that arises not from software inefficiencies, but from underlying hardware design choices.
+In practical deployments—particularly in multi-tenant or concurrent environments such as edge nodes or confidential cloud computing platforms—PMP exhaustion could severely limit system utility unless proactively mitigated.
 
-Thirdly, workload characteristics strongly influence enclave performance overheads.
-Compute-bound, sequential workloads such as the Kyber post-quantum cryptographic algorithm experience comparatively low performance degradation within enclaves.
-This is consistent with Kyber’s minimal inter-thread communication and largely predictable memory access patterns.
-In contrast, highly parallelizable benchmarks like CoreMark exhibit increasing overhead with core count, reflecting the compounded costs of synchronization, enclave entry/exit transitions, and possibly contention in PMP configuration.
-This differential sensitivity implies that enclave performance optimization must be workload-specific, taking into account application parallelism and memory footprint.
+The analysis further reveals that enclave performance is highly sensitive to workload type.
+Sequential, compute-bound tasks such as the Kyber post-quantum cryptographic algorithm exhibit relatively low performance degradation when executed within an enclave.
+This is attributable to Kyber’s predictable memory access patterns and minimal need for inter-thread synchronization.
+By contrast, parallelizable workloads like CoreMark demonstrate increasing overhead as core count rises, driven by factors such as synchronization costs, increased enclave context switching, and potential contention in PMP entry allocation.
+These results suggest that enclave performance optimization must be workload-aware, with particular consideration given to application parallelism and memory behavior.
 
-Lastly, the interaction between resource allocation parameters—CPU cores, memory size, and cache configuration—and enclave performance is multifaceted.
-Increasing CPU cores yields diminishing returns beyond a threshold due to synchronization and enclave management overhead.
-Larger memory allocations improve performance primarily by reducing paging and improving cache locality but also can exacerbate PMP usage if memory regions are fragmented.
-Cache configuration impacts, though indirectly measured via QEMU’s system-wide cache model, suggest that optimizing cache size and associativity can mitigate enclave overheads, especially for instruction-intensive workloads.
-These observations highlight the necessity of a holistic approach to system tuning in enclave environments.
+Finally, the relationship between system resource allocation—specifically CPU cores, memory capacity, and cache hierarchy—and enclave performance is nuanced.
+Increasing the number of processor cores beyond a certain threshold yields diminishing performance returns due to synchronization overhead and enclave management complexity.
+While increasing memory allocation generally improves performance by reducing page faults and improving cache locality, it may also exacerbate PMP usage if memory regions are fragmented.
+Although QEMU provides only a high-level approximation of cache behavior, the observed trends suggest that cache size and associativity can significantly influence enclave performance for instruction- and memory-intensive workloads.
+These interdependencies underscore the need for a holistic, system-level tuning approach when deploying TEEs in real-world scenarios.
+
+The benchmarking results confirm that Keystone enclaves, while effective at enforcing isolation, require careful system configuration and architectural awareness to ensure scalable and efficient performance.
+The findings serve as a foundation for both immediate system tuning and future architectural innovation aimed at overcoming the current limitations of enclave-based trusted execution on RISC-V.
 
@@
 
-In conclusion, while the PMP mechanism is undeniably a cornerstone of the Keystone security model, it also introduces a set of hard limitations that impact the practical deployment and scaling of enclaves.
+While the PMP mechanism is undeniably a cornerstone of the Keystone security model, it also introduces a set of hard limitations that impact the practical deployment and scaling of enclaves.
 Understanding and addressing these limitations—both through hardware evolution and smarter software resource management—is essential for building scalable, secure computing platforms based on the RISC-V architecture.

### 2025-08-10 chapters/system-architecture.tex
updated TOC
minor grammatical modifications

@@
-\chapter{System Architecture and Design}
+\chapter{System Design and Methodology}
 \label{chap:methodology}
@@
 
-\section{Experimental Setup}
+\section{Experimental Platform and Environment}
 \label{sec:experimental-setup}
@@
 
-\section{Kyber in the Enclave}
+\section{Integration of Kyber into Keystone}
 \label{sec:kyber-enclave}
@@
 
-\section{Benchmarking tools and metrics}
+\section{Benchmarking Suite and Metrics}
 \label{sec:benchmarking-tools}
@@
 
-\section{Parameter Variation Strategy}
+\section{Parameter Configuration and Variation}
 \label{sec:param-variation}
 
-To investigate the architectural impact on enclave execution, this study systematically varied system parameters including memory size, core count, execution mode (sequential vs parallel), and cache configurations using QEMU’s cache modeling capabilities.
-The aim was to understand how these factors influence the behavior of trusted and non-trusted workloads under different resource constraints.
+To rigorously assess how architectural and system-level parameters influence the performance of workloads executed within secure enclaves, this study undertook a systematic exploration of configurable runtime characteristics on the simulated RISC-V platform.
+Specifically, we varied four principal dimensions: physical memory size, the number of processor cores, execution mode (sequential versus parallel), and cache configuration.
+These parameters were chosen based on their known relevance to workload performance and enclave behavior, and were adjusted within the bounds of what is supported by the QEMU-based virtual platform and the Keystone framework.
 
-Cache simulation was conducted using the QEMU TCG plugin framework \cite{mandour2021cache}, which models per-core instruction and data caches in an idealized fashion.
-While this plugin is not cycle-accurate and does not isolate enclave-specific behavior, it enables consistent modeling of cache accesses and miss patterns across the entire QEMU runtime.
-This was sufficient to observe cache-related effects caused by workload behavior in different configurations.
+The overarching objective of this variation was to uncover the extent to which system-level constraints and resource allocations impact both trusted (enclave) and non-trusted (native) execution.
+By exploring diverse configurations, we aimed to characterize how resource limitations (e.g., constrained memory or limited cores) manifest in performance degradation, and to what degree secure execution amplifies or mitigates these effects.
 
-To avoid excessive experimental overhead and redundant trends, five representative system configurations were selected.
-These were chosen to cover a wide range of memory capacities, CPU core counts, and execution modes.
-Each configuration enables a specific type of insight, ensuring a diverse but manageable evaluation matrix.
+To simulate cache behavior, we employed the QEMU TCG plugin framework, specifically leveraging the cache modeling capabilities introduced by Mandour et al.
+\cite{mandour2021cache}.
+This plugin facilitates the modeling of per-core instruction and data caches and enables the collection of aggregate system-wide cache statistics.
+While it does not provide cycle-accurate timing nor isolate enclave-specific cache behavior, its ability to log cache access and miss patterns uniformly across all benchmark executions proved sufficient for analyzing cache-related effects in both secure and non-secure modes.
+The insights gleaned from this analysis, though approximate, were instrumental in identifying general cache utilization trends under various workloads and resource configurations.
+
+Given the expansive parameter space created by the combination of memory size, core count, execution mode, and cache settings, it was neither practical nor necessary to exhaustively benchmark every possible configuration.
+Instead, we curated a set of five representative system configurations that collectively span a meaningful range of the performance spectrum—from minimal, resource-constrained environments to well-provisioned, multicore systems operating at near full capacity.
+These configurations are detailed in Table~\ref{tab:configurations}.
 
@@
 \hline
-Low-End Baseline     & 64M         & 1             & Sequential         & Serves as the minimal configuration to establish baseline performance.
-Useful for understanding enclave overhead under tight resource constraints.
+Low-End Baseline     & 64M         & 1             & Sequential         & Serves as a minimal configuration, enabling the establishment of baseline performance under constrained conditions.
+Useful for measuring the raw impact of enclave mechanisms in resource-starved environments.
 \\
 \hline
-Balanced Parallel    & 128M        & 2             & Parallel           & Captures early-stage multicore scaling behavior. Highlights effects of light parallelism under modest memory availability. \\
-\hline
-Mid-Range Sequential & 256M        & 4             & Sequential         & Enables evaluation of scaling effects due to additional cores and memory, while keeping execution sequential to avoid parallel noise.
+Balanced Parallel    & 128M        & 2             & Parallel           & Captures early-stage scaling behavior with modest multicore setups.
+Helps isolate the overhead introduced by enclave entry/exit in light parallel workloads.
 \\
 \hline
-High-End Parallel    & 512M        & 6             & Parallel           & Represents a well-provisioned system where both memory and core resources are sufficient.
-Useful for observing near-ideal parallel execution.
+Mid-Range Sequential & 256M        & 4             & Sequential         & Provides a view into how increasing core count and memory capacity affect performance in the absence of thread-level parallelism.
+Useful for understanding system scalability independent of synchronization effects.
 \\
 \hline
-Max Capacity Stress  & 2G          & 10            & Parallel           & Tests the upper bounds of simulated platform capacity.
-Helps identify performance saturation and the point of diminishing returns.
+High-End Parallel    & 512M        & 6             & Parallel           & Represents a robust configuration where memory and cores are sufficiently provisioned.
+Designed to observe near-ideal parallel execution and potential scalability limits of the enclave framework.
+\\
+\hline
+Max Capacity Stress  & 2G          & 10            & Parallel           & Pushes the simulated platform to its upper bounds.
+Intended to explore saturation points, resource contention, and diminishing returns under heavy system load.
 \\
@@
 \caption{Summary of representative configurations used in performance evaluation.}
+\label{tab:configurations}
 \end{table}
 
-Several other configurations—such as single-core setups with excessive memory or high-core setups with limited memory—were explored during preliminary testing.
-However, they offered little additional insight beyond what was observable in the selected set.
-For instance, excessive memory on a single core often led to underutilization, while many-core systems with small memory pools frequently resulted in excessive context switching.
-Both behaviors were captured effectively by existing selected configurations.
+In preliminary testing, a number of alternative configurations were explored to validate the diversity and coverage of the selected set.
+These included scenarios such as single-core systems with excessively large memory allocations and many-core setups with highly constrained memory.
+However, these combinations typically resulted in either underutilized resources or pathological behavior (e.g., excessive paging or context switching), without contributing new insight beyond what was captured by the five chosen profiles.
+For example, increasing memory size on a single-core system had negligible impact due to CPU-bound bottlenecks, while increasing core counts in systems with limited memory often led to high levels of involuntary context switching and scheduling overhead.
 
-Across all system configurations, three cache profiles were tested: a baseline setup with 8\,KB 4-way LRU caches, a larger 32\,KB 8-way configuration, and a fixed-size cache using FIFO and random eviction strategies.
-A consistent block size of 64 bytes was used to preserve spatial locality.
-Although the cache plugin logs system-wide activity, observed trends were analyzed in conjunction with benchmark behavior to infer their likely contributions to overall performance variation.
+In addition to varying core and memory resources, three distinct cache configurations were tested across all system profiles to assess the sensitivity of enclave and native workloads to cache architecture.
+These configurations included:
 
-Benchmarks were run in both native and enclave execution modes to enable direct comparisons.
-This design facilitated the isolation of architectural performance trends and the specific overhead introduced by trusted execution.
+\begin{itemize}
+    \item A baseline setup featuring 8\,KB, 4-way set-associative caches with an LRU replacement policy
+    \item An enlarged cache with 32\,KB size and 8-way associativity, reflecting more generous cache resources
+    \item A fixed-size cache using FIFO and random eviction policies to model alternative eviction behaviors
+\end{itemize}
 
-\section{Data Collection and Analysis Procedures}
+All configurations used a consistent block size of 64 bytes to preserve spatial locality across workloads.
+Although the cache modeling plugin captures system-wide statistics and does not differentiate between enclave and non-enclave activity, observed trends—such as shifts in cache miss rates and access patterns—were analyzed in conjunction with benchmark-level behavior to infer their contribution to observed performance variations.
+
+To enable direct and consistent comparisons, each benchmark was executed in both native and enclave modes under every configuration.
+This mirrored design allowed for clear isolation of the performance penalties introduced by enclave execution and helped disentangle architectural effects from security-related overhead.
+
+\section{Data Collection and Reproducibility}
 \label{sec:data-collection}
 
-All benchmarks were executed within a consistent virtualized RISC-V environment using QEMU.
-For each selected configuration, both Dhrystone and CoreMark benchmarks were run in native and enclave modes to capture the impact of enclave isolation under different resource and cache conditions.
+To ensure the integrity, consistency, and scientific validity of the performance evaluation, all benchmark experiments were conducted within a tightly controlled and reproducible virtualized environment.
+The entire measurement campaign was executed on a simulated RISC-V platform using the QEMU emulator, which served as the foundation for both native and enclave-based workloads.
+This consistent virtualization layer eliminated hardware variability and allowed for precise control over system resources such as memory, core count, and cache characteristics.
 
-Each experiment consisted of ten executions.
-To reduce variance caused by warm-up behavior, the first run was discarded, and the remaining nine were used to compute average performance.
-Key performance metrics included total wall-clock execution time, benchmark throughput (DMIPS or iterations/sec), and standard deviation.
+Each system configuration, as defined in Section~\ref{sec:param-variation}, was evaluated using the Dhrystone, CoreMark, and Kyber benchmarks.
+For each benchmark, tests were run in both native (non-enclave) and enclave execution modes to facilitate direct comparative analysis of performance overheads introduced by trusted execution.
 
-In addition to these primary measurements, system-level runtime statistics were collected from within the guest Linux environment. These included:
+To reduce transient measurement noise and improve the robustness of collected data, each benchmark experiment was repeated ten times per configuration.
+The first execution was discarded in all cases to account for initialization artifacts such as dynamic binary translation warm-up and memory allocation caching within the QEMU runtime.
+The remaining nine runs were averaged to produce stable and representative performance figures.
+In addition to computing the arithmetic mean, the standard deviation was recorded to provide a measure of variability across runs.
+
+The primary performance metric collected was **total wall-clock execution time**, as measured from within the guest Linux environment using high-resolution timers (e.g., \texttt{clock\_gettime()}) or the Keystone timer API in enclave mode.
+For Dhrystone and CoreMark, benchmark-specific throughput metrics were also recorded, including Dhrystone MIPS (DMIPS) and iterations per second.
+These allowed for workload-specific efficiency comparisons across configurations.
+
+In addition to these high-level metrics, detailed system-level runtime statistics were gathered using standard Linux facilities such as the \texttt{/proc} filesystem and the \texttt{getrusage()} system call.
+These included:
 
 \begin{itemize}
-    \item User and system CPU time
-    \item Maximum memory usage (RSS)
-    \item Minor and major page faults
-    \item Voluntary and involuntary context switches
+    \item \textbf{User and system CPU time:} Total time spent in user-space and kernel-space respectively.
+    \item \textbf{Maximum resident set size (RSS):} Peak memory consumption during execution.
+    \item \textbf{Page fault statistics:} Number of minor (soft) and major (hard) page faults incurred during the benchmark.
+    \item \textbf{Context switching behavior:} Number of voluntary and involuntary context switches, indicating scheduling and synchronization overhead.
 \end{itemize}
 
-These metrics were essential to understanding how different system configurations affected workload behavior.
-For example, high numbers of involuntary context switches in parallel configurations often indicated memory contention or over-subscription of CPU cores.
+These runtime indicators were critical in diagnosing performance behaviors that could not be attributed to CPU throughput alone.
+For instance, elevated context switch counts in parallel configurations were often correlated with increased enclave transition overhead or CPU oversubscription, especially in memory-constrained environments.
 
-Cache-related metrics were gathered using the QEMU cache plugin, which logs all instruction and data cache accesses and misses at the system level.
-Although enclave-specific cache statistics could not be isolated, the relative trends between native and enclave executions under identical configurations allowed for meaningful comparative analysis.
+Cache-related metrics were obtained using the QEMU TCG plugin cache model described earlier.
+This plugin logs all instruction and data cache accesses and cache misses at the system level.
+While the plugin lacks enclave-specific visibility and does not model hardware-specific cache timing, it provides a consistent and useful proxy for analyzing trends in memory access behavior.
+Cache metrics were aligned with benchmark execution phases to identify correlations between cache activity and observed performance shifts.
 
-Performance overhead introduced by enclave execution was quantified using the percentage increase in execution time compared to native mode, calculated as:
+To compute the enclave-specific performance overhead, we employed a normalized overhead metric defined as the relative increase in execution time between enclave and native modes:
 
@@
 
-This normalized metric enabled consistent cross-comparison between configurations with vastly different execution baselines.
+This percentage-based formulation allows for direct cross-comparison of overheads across workloads with differing baseline durations.
+It also facilitates trend identification across system configurations, particularly when assessing scaling behaviors.
 
-All benchmark scripts, plugin configurations, and raw logs were version-controlled and archived.
-Detailed plugin output, configuration parameters, and system call traces are provided in Appendix~\ref{appendix:cache-plugin} for transparency and reproducibility.
+For reproducibility, all benchmarking scripts, system configurations, QEMU command-line options, Keystone runtime parameters, and cache plugin settings were version-controlled using Git.
+Raw benchmark outputs, kernel logs, and cache plugin trace files were archived and documented.
+A complete snapshot of the experimental infrastructure, along with scripts to regenerate results, is provided in Appendix~\ref{appendix:cache-plugin}.
+This archival effort ensures that all reported results can be independently verified and reproduced by future researchers or evaluators.
+
+\begin{figure}[H]
+\centering
+\begin{tikzpicture}[node distance=1.75cm]
+\node (start)   [process] {Start: Define Benchmark Set};
+\node (config)  [process, below of=start] {Select Hardware Configuration};
+\node (native)  [process, below of=config] {Run Benchmarks in Native Mode};
+\node (enclave) [process, below of=native] {Run Benchmarks in Enclave Mode};
+\node (metrics) [process, below of=enclave] {Collect Performance Metrics};
+\node (analyze) [process, below of=metrics] {Compare and Analyze Overheads};
+\node (end)     [process, below of=analyze] {Store Logs and Archive Results};
+
+\draw [arrow] (start)   -- (config);
+\draw [arrow] (config)  -- (native);
+\draw [arrow] (native)  -- (enclave);
+\draw [arrow] (enclave) -- (metrics);
+\draw [arrow] (metrics) -- (analyze);
+\draw [arrow] (analyze) -- (end);
+\end{tikzpicture}
+\caption{Benchmarking and Data Collection Workflow}
+\label{fig:benchmarking-workflow}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\begin{tikzpicture}[
+    native/.style={draw, fill=blue!20, minimum width=1.2cm, minimum height=1cm, font=\small, rounded corners=2pt},
+    enclave/.style={draw, fill=red!20, minimum width=1.2cm, minimum height=1cm, font=\small, rounded corners=2pt},
+    label/.style={font=\bfseries, align=center},
+    cell/.style={draw, minimum width=1cm, minimum height=0.6cm},
+    every node/.style={inner sep=2pt}
+]
+
+% Column labels
+\node[label] at (2.4, 6.2) {Low-End};
+\node[label] at (4.4, 6.2) {Balanced};
+\node[label] at (6.4, 6.2) {Mid-Range};
+\node[label] at (8.4, 6.2) {High-End};
+\node[label] at (10.4, 6.2) {Max Capacity};
+
+% Row labels
+\node[label, rotate=90] at (1, 5) {Dhrystone};
+\node[label, rotate=90] at (1, 3) {CoreMark};
+\node[label, rotate=90] at (1, 1) {Kyber};
+
+% Dhrystone row
+\node[native] at (2, 5) {11.3s};
+\node[enclave] at (2.8, 5) {123s};
+
+\node[native] at (4, 5) {107s};
+\node[enclave] at (4.8, 5) {99s};
+
+\node[native] at (6, 5) {91s};
+\node[enclave] at (6.8, 5) {55s};
+
+\node[native] at (8, 5) {22s};
+\node[enclave] at (8.8, 5) {11s};
+
+\node[native] at (10, 5) {5s};
+\node[enclave] at (10.8, 5) {20s};
+
+% CoreMark row (empty placeholders for now)
+\foreach \x in {2,4,6,8,10} {
+    \node[native] at (\x, 3) {};
+    \node[enclave] at (\x+0.8, 3) {};
+}
+
+% Kyber row (empty)
+\foreach \x in {2,4,6,8,10} {
+    \node[native] at (\x, 1) {};
+    \node[enclave] at (\x+0.8, 1) {};
+}
+
+% Native / Enclave labels inside or above boxes if needed
+\node[font=\scriptsize] at (1.8, 5.8) {Native};
+\node[font=\scriptsize] at (2.6, 5.8) {Enclave};
+
+% Legend
+\node[draw, fill=blue!20, minimum width=0.8cm, minimum height=0.6cm] at (6, -0.8) {};
+\node at (6.9, -0.8) {Native};
+
+\node[draw, fill=red!20, minimum width=0.8cm, minimum height=0.6cm] at (8.5, -0.8) {};
+\node at (9.4, -0.8) {Enclave};
+
+\end{tikzpicture}
+\caption{Experiment Matrix: Benchmarks Across Configurations and Execution Modes}
+\label{fig:experiment-matrix}
+\end{figure}

### 2025-08-10 main.tex
reorganised key sections
changed verbatim to lstlistings
minor grammatical modifications

@@
 \usepackage{float}
+\lstset{basicstyle=\ttfamily}
+\usepackage{tikz}
+\usetikzlibrary{shapes.geometric, arrows.meta, positioning, shadows.blur}
+
+\tikzstyle{process} = [
+    rectangle, rounded corners,
+    minimum width=4.2cm,
+    minimum height=1.2cm,
+    text centered,
+    draw=black,
+    fill=blue!10,
+    font=\sffamily\small,
+    blur shadow
+]
+
+\tikzstyle{arrow} = [
+    thick,
+    ->,
+    >=Stealth,
+    draw=gray!80
+]
+
+\tikzstyle{cell} = [rectangle, draw=black, minimum width=2.5em, minimum height=2.5em, anchor=center]
+\tikzstyle{native} = [cell, fill=blue!20]
+\tikzstyle{enclave} = [cell, fill=red!20]
+\tikzstyle{label} = [font=\bfseries]
 
@@
 
-\include{chapters/methodology.tex}
+\include{chapters/system-architecture.tex}
 
-\include{chapters/results.tex}
+\include{chapters/benchmark.tex}
 
-\include{chapters/recommendations.tex}
+%\include{chapters/results.tex}
+
+\include{chapters/discussion.tex}
+
+%\include{chapters/recommendations.tex}
 

### 2025-08-12 chapters/background.tex
updated TOC
added baseline results

@@
-\chapter{Background}
+\chapter{Background and Related Technologies}
 \label{chap:background}
@@
 
-\section{Keystone Architecture Overview}
+\section{Keystone Enclave Architecture}
 
@@
 
-\section{Kyber: A Post-Quantum Cryptography Algorithm}
+\section{Physical Memory Protection (PMP) in RISC-V}
+\label{sec:pmp}
+
+Physical Memory Protection (PMP) is a hardware-level security mechanism defined by the RISC-V privileged architecture to provide fine-grained control over access permissions to physical memory regions.
+PMP plays a central role in enabling secure system software, particularly in contexts where memory isolation between privileged and unprivileged software is essential.
+It is a critical enabler for implementing secure environments such as hypervisors, operating system kernels, and Trusted Execution Environments (TEEs) without relying on traditional virtualization support~\cite{riscvprivspec}.
+
+In RISC-V, the PMP mechanism is configured exclusively by machine-mode (M-mode) software—the most privileged execution level in the architecture—and applies access control policies to all lower-privileged software running in supervisor (S-mode) or user mode (U-mode).
+Each PMP entry defines a memory region along with a set of permissions (read, write, execute) and an addressing mode.
+The hardware enforces these permissions such that any access attempt by S-mode or U-mode software that violates a PMP rule results in a trap to M-mode.
+This mechanism ensures that M-mode retains full control over physical memory while being able to delegate controlled access to untrusted software layers~\cite{riscvprivspec,lee2019keystone}.
+
+A typical RISC-V core supports a fixed number of PMP entries (commonly 8 or 16), each represented by two Control and Status Registers (CSRs): \texttt{pmpcfg} and \texttt{pmpaddr}.
+The \texttt{pmpcfg} register specifies the access permissions and address matching mode for each region, while \texttt{pmpaddr} defines the address bounds.
+RISC-V supports several region encoding modes:
+
+\begin{itemize}
+    \item \textbf{TOR (Top-of-Range):} Defines a region from the base address of the previous PMP entry up to the current address.
+    \item \textbf{NAPOT (Naturally Aligned Power-Of-Two):} Encodes both the base and size in a single register, enabling efficient representation of power-of-two-sized regions.
+    \item \textbf{NA4 (Naturally Aligned 4-byte):} Encodes a minimal 4-byte memory region.
+\end{itemize}
+
+The design of PMP aligns with the minimalist and modular philosophy of RISC-V.
+Unlike traditional memory protection mechanisms such as x86's segmentation or ARM's TrustZone Address Space Controller (TZASC), PMP is simple, deterministic, and free from microarchitectural complexity.
+Importantly, PMP operates at the physical memory level—independent of virtual memory or page tables—which makes it highly suitable for systems with limited resources or without Memory Management Units (MMUs)~\cite{riscvprivspec,Survey2023}.
+
+In the context of TEEs such as the Keystone framework, PMP serves as the fundamental enforcement mechanism for enclave isolation.
+Keystone leverages PMP to assign a protected memory region, known as the Enclave Page Memory (EPM), to each enclave.
+This region is configured to be inaccessible to both the untrusted operating system (running in S-mode) and any other software outside the enclave context~\cite{dayeol2019keystone}.
+During enclave creation, the Keystone Security Monitor (SM)—executing in M-mode—programmatically sets PMP entries to lock the memory region associated with the enclave.
+The SM verifies the integrity of the enclave binary, computes its cryptographic hash, and sets up the memory layout in such a way that only the enclave code has access to its memory contents~\cite{lee2019keystone,keystone2025how}.
+
+Furthermore, PMP entries are reconfigured dynamically during enclave context switches.
+For example, when the host operating system regains control, the PMP settings are altered to revoke access to the enclave’s memory.
+Conversely, when the enclave resumes execution, the SM reinstates its PMP region.
+This dynamic switching is essential to enforce temporal isolation and prevent time-of-check-to-time-of-use (TOCTTOU) vulnerabilities that may arise during state transitions~\cite{cerberus2022}.
+
+Despite its utility, PMP has several practical limitations.
+The fixed number of PMP entries restricts the number of simultaneously protected regions, posing a challenge for complex systems with many isolated domains or subregions.
+Moreover, overlapping PMP regions are resolved based on entry priority (lower-numbered entries take precedence), requiring careful configuration to avoid inadvertent leaks or privilege escalations~\cite{riscvprivspec}.
+
+Security researchers have explored various enhancements and formal models to increase the reliability and verifiability of PMP-based isolation.
+For example, Cerberus~\cite{cerberus2022} builds a formally verified memory-sharing abstraction over PMP to enable safe communication between enclaves and untrusted components.
+Similarly, HexBox~\cite{hexbox2018} leverages PMP for real-time embedded security by associating memory regions with specific control-flow paths in embedded firmware, demonstrating PMP's applicability beyond general-purpose computing.
+
+The role of PMP in Keystone exemplifies how low-complexity hardware primitives can be orchestrated by secure software layers to implement strong isolation guarantees.
+This synergy is particularly effective in the RISC-V ecosystem, where the open architecture allows deep customization, formal verification, and academic experimentation~\cite{Survey2023}.
+The combination of PMP, a secure boot mechanism, and a minimal TCB enables Keystone to enforce hardware-rooted enclave isolation without relying on opaque, vendor-specific extensions, setting it apart from alternatives like Intel SGX or ARM TrustZone~\cite{dayeol2019keystone,turn0search0}.
+
+In conclusion, Physical Memory Protection in RISC-V is a powerful, lightweight, and deterministic mechanism that underpins secure execution in many RISC-V-based platforms.
+Its integration into frameworks like Keystone enables transparent and verifiable enclave isolation while remaining suitable for constrained systems.
+As the RISC-V ecosystem continues to evolve, PMP will remain a fundamental building block for secure, open, and formally analyzable trusted computing platforms.
+
+\section{Post-Quantum Cryptography: Kyber Algorithm}
 \label{sec:kyber}
@@
 \end{itemize}
+
+\begin{figure}[htbp]
+\centering
+\includegraphics[width=0.9\linewidth]{figures/kyber_768.png}
+\caption{Kyber768 Key Exchange Algorithm}
+\label{fig:kyber_768}
+\end{figure}
 
@@
 
-\section{Comparison with Other TEEs}
+\section{Comparative Overview of TEE}
 \label{sec:comparison_tees}
@@
 
-\section{Performance Considerations in TEEs}
+\section{Performance Challenges in Secure Execution}
 

### 2025-08-12 chapters/benchmark.tex
added baseline results

@@
 \section{Baseline: Native (Non-Enclave) Performance}
+\label{sec:baseline-native}
 
-To establish a baseline for evaluating enclave overheads, we first executed all benchmarks natively on the simulated RISC-V platform without any enclave isolation.
-This phase allows us to measure the raw performance characteristics of the system and isolate the costs introduced solely by secure execution.
+To establish a reliable baseline for evaluating the performance overhead introduced by secure enclaves, we first measured native (non-enclave) execution of key benchmarks on the simulated RISC-V platform.
+All benchmarks were executed under controlled conditions consistent with the previously described parameter variation framework, focusing primarily on the Low-End Baseline setup with 64~MB RAM, a single CPU core, and sequential execution.
 
-The Dhrystone and CoreMark benchmarks exhibited expected scaling behavior with respect to CPU cores and memory size.
-As CPU core count increased, throughput for these parallelizable benchmarks improved approximately linearly, reflecting efficient multi-threaded execution on the underlying hardware.
-Memory size impacted CoreMark more noticeably, particularly during memory-intensive phases, where increased memory reduced paging and improved cache hit rates.
-In contrast, Dhrystone’s smaller working set resulted in negligible sensitivity to memory size variations.
+The selected benchmarks include synthetic workloads CoreMark and Dhrystone, both widely used in embedded systems performance evaluation, as well as cryptographic primitives from the Kyber post-quantum key encapsulation mechanism (KEM).
+Each benchmark was run multiple times---typically ten iterations---with the initial run discarded to mitigate artifacts from JIT warm-up and caching effects inherent to the QEMU TCG runtime.
+Reported metrics represent averages over the remaining nine runs, with standard deviations provided where applicable.
 
-The Kyber cryptographic benchmark behaved differently.
-Due to its inherently sequential algorithmic structure, Kyber’s execution time remained relatively constant across changes in CPU core count.
-Likewise, memory size had minimal impact on Kyber performance, as its working set comfortably fit within allocated memory without inducing paging.
+Key performance metrics collected include wall-clock execution time (measured using high-resolution timers), throughput indicators such as iterations or Dhrystones per second, CPU utilization (user and system time), memory footprint (peak resident set size), page fault counts, context switch statistics, and cache metrics obtained via the QEMU TCG cache plugin.
 
-These native baseline results confirm the stability and predictability of the simulated system environment, providing a solid foundation for subsequent enclave overhead analysis.
+Figure~\ref{fig:coremark_native} presents CoreMark results under the 10-run sequential configuration.
+The benchmark demonstrated high stability, with an average elapsed time of 107.5 seconds and approximately 939 iterations per second.
+CPU utilization was predominantly user-space (81.4 seconds) with minimal system overhead (1 second).
+Memory consumption remained low at 10~KB RSS, and page fault and context switch counts indicated a minimally interrupted environment.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.9\linewidth]{figures/coremark_64M_1Core_native_metrics.pdf}
+    \caption{CoreMark Native Execution Summary (10 Runs, Low-End Baseline)}
+    \label{fig:coremark_native}
+\end{figure}
+
+Figure~\ref{fig:dhrystone_native} summarizes Dhrystone benchmark native execution results.
+The mean elapsed time was approximately 102.8 seconds, with throughput near 96,200 Dhrystones per second.
+User CPU time was steady at 10.3 seconds, and system CPU time was negligible.
+Minor page faults were more frequent than in CoreMark, reflecting normal demand paging, and involuntary context switches occurred regularly due to scheduler activity but did not noticeably impact throughput.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.9\linewidth]{figures/dhrystone_64M_1Core_native_metrics.pdf}
+    \caption{Dhrystone Native Execution Summary (10 Runs, Low-End Baseline)}
+    \label{fig:dhrystone_native}
+\end{figure}
+
+Figure~\ref{fig:kyber_native_cycles} reports median CPU cycle counts for Kyber cryptographic operations measured in native mode on the Low-End Baseline configuration.
+Keypair generation, encapsulation, and decapsulation were profiled across Kyber-512, Kyber-768, and Kyber-1024 parameter sets.
+All showed consistent cycle counts, indicating stable execution and resource usage.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.9\linewidth]{figures/kyber_64M_1Core.pdf}
+    \caption{Kyber Native Execution: Average CPU Cycles per Operation}
+    \label{fig:kyber_native_cycles}
+\end{figure}
+
+Cache statistics collected via the QEMU plugin across all benchmarks indicated predictable access patterns consistent with compute-bound workloads.
+Instruction and data cache miss rates remained low, reinforcing the stability of native execution on this platform.
+
+In summary, these native baseline results confirm a stable and well-characterized environment with minimal interference or resource contention.
+This baseline provides a critical reference for comparing enclave-mode execution and precisely quantifying the overhead of secure execution under representative embedded system constraints.
 

### 2025-08-12 chapters/system-architecture.tex
added baseline results

@@
 \end{figure}
-
-\begin{figure}[H]
-\centering
-\begin{tikzpicture}[
-    native/.style={draw, fill=blue!20, minimum width=1.2cm, minimum height=1cm, font=\small, rounded corners=2pt},
-    enclave/.style={draw, fill=red!20, minimum width=1.2cm, minimum height=1cm, font=\small, rounded corners=2pt},
-    label/.style={font=\bfseries, align=center},
-    cell/.style={draw, minimum width=1cm, minimum height=0.6cm},
-    every node/.style={inner sep=2pt}
-]
-
-% Column labels
-\node[label] at (2.4, 6.2) {Low-End};
-\node[label] at (4.4, 6.2) {Balanced};
-\node[label] at (6.4, 6.2) {Mid-Range};
-\node[label] at (8.4, 6.2) {High-End};
-\node[label] at (10.4, 6.2) {Max Capacity};
-
-% Row labels
-\node[label, rotate=90] at (1, 5) {Dhrystone};
-\node[label, rotate=90] at (1, 3) {CoreMark};
-\node[label, rotate=90] at (1, 1) {Kyber};
-
-% Dhrystone row
-\node[native] at (2, 5) {11.3s};
-\node[enclave] at (2.8, 5) {123s};
-
-\node[native] at (4, 5) {107s};
-\node[enclave] at (4.8, 5) {99s};
-
-\node[native] at (6, 5) {91s};
-\node[enclave] at (6.8, 5) {55s};
-
-\node[native] at (8, 5) {22s};
-\node[enclave] at (8.8, 5) {11s};
-
-\node[native] at (10, 5) {5s};
-\node[enclave] at (10.8, 5) {20s};
-
-% CoreMark row (empty placeholders for now)
-\foreach \x in {2,4,6,8,10} {
-    \node[native] at (\x, 3) {};
-    \node[enclave] at (\x+0.8, 3) {};
-}
-
-% Kyber row (empty)
-\foreach \x in {2,4,6,8,10} {
-    \node[native] at (\x, 1) {};
-    \node[enclave] at (\x+0.8, 1) {};
-}
-
-% Native / Enclave labels inside or above boxes if needed
-\node[font=\scriptsize] at (1.8, 5.8) {Native};
-\node[font=\scriptsize] at (2.6, 5.8) {Enclave};
-
-% Legend
-\node[draw, fill=blue!20, minimum width=0.8cm, minimum height=0.6cm] at (6, -0.8) {};
-\node at (6.9, -0.8) {Native};
-
-\node[draw, fill=red!20, minimum width=0.8cm, minimum height=0.6cm] at (8.5, -0.8) {};
-\node at (9.4, -0.8) {Enclave};
-
-\end{tikzpicture}
-\caption{Experiment Matrix: Benchmarks Across Configurations and Execution Modes}
-\label{fig:experiment-matrix}
-\end{figure}

### 2025-08-14 chapters/benchmark.tex
4.2 complete. split figures pending

--- 2025-08-10 chapters/benchmark.tex
@@
 \section{Baseline: Native (Non-Enclave) Performance}
+\label{sec:baseline-native}
 
-To establish a baseline for evaluating enclave overheads, we first executed all benchmarks natively on the simulated RISC-V platform without any enclave isolation.
-This phase allows us to measure the raw performance characteristics of the system and isolate the costs introduced solely by secure execution.
+To accurately assess the performance overheads imposed by secure enclave execution, it is imperative to begin with a rigorously characterized baseline: one that isolates application-level behavior from security mechanisms, scheduling anomalies, and hardware acceleration artifacts.
+This section documents such a baseline, derived from native (non-enclave) execution of representative workloads on a deterministic, single-core RISC-V simulation environment.
+All benchmarks were executed under the Low-End Baseline configuration, designed to approximate the resource constraints of a lightweight embedded platform: 64~MB of main memory, a single in-order RISC-V core, and no parallelism beyond sequential instruction issue.
+This environment intentionally excludes vector extensions, out-of-order speculation, and memory prefetching, ensuring that measured performance reflects algorithmic structure rather than microarchitectural opportunism.
 
-The Dhrystone and CoreMark benchmarks exhibited expected scaling behavior with respect to CPU cores and memory size.
-As CPU core count increased, throughput for these parallelizable benchmarks improved approximately linearly, reflecting efficient multi-threaded execution on the underlying hardware.
-Memory size impacted CoreMark more noticeably, particularly during memory-intensive phases, where increased memory reduced paging and improved cache hit rates.
-In contrast, Dhrystone’s smaller working set resulted in negligible sensitivity to memory size variations.
+Each benchmark was executed both \emph{sequentially} and \emph{in parallel}, ten times each, and the resulting figures compare performance and stability across both modes.
 
-The Kyber cryptographic benchmark behaved differently.
-Due to its inherently sequential algorithmic structure, Kyber’s execution time remained relatively constant across changes in CPU core count.
-Likewise, memory size had minimal impact on Kyber performance, as its working set comfortably fit within allocated memory without inducing paging.
+% ------------------------
+% Wall-clock time figures
+% ------------------------
+Figure~\ref{fig:wallclock_coremark} reports CoreMark's performance profile using the latest data from Run~10 of both sequential and parallel execution modes.
+Execution time remained consistent, with the sequential case averaging approximately 101.17 seconds of wall-clock time and the parallel case averaging around 103.37 seconds.
+This corresponds to an effective throughput of about 988.4 iterations per second for the sequential run and 967.4 iterations per second for the parallel run.
+User-mode CPU time dominated in both cases (101.20 seconds sequential, 103.39 seconds parallel), accounting for over 99.9\% of total CPU time.
+System-mode CPU time was negligible (under 0.03 seconds), confirming that the workload’s operational footprint is almost entirely in user space with minimal reliance on system calls or kernel transitions.
 
-These native baseline results confirm the stability and predictability of the simulated system environment, providing a solid foundation for subsequent enclave overhead analysis.
+% \begin{figure}[htbp]
+%   \centering
+%   \includegraphics[width=0.9\linewidth]{figures/wallclock_coremark.png}
+%   \caption{CoreMark wall-clock time in native sequential and parallel runs.}
+%   \label{fig:wallclock_coremark}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%   \centering
+%   \includegraphics[width=0.9\linewidth]{figures/wallclock_dhrystone.png}
+%   \caption{Dhrystone wall-clock time in native sequential and parallel runs.}
+%   \label{fig:wallclock_dhrystone}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%   \centering
+%   \includegraphics[width=0.9\linewidth]{figures/wallclock_kyber.png}
+%   \caption{Kyber wall-clock time in native sequential and parallel runs.}
+%   \label{fig:wallclock_kyber}
+% \end{figure}
+
+% ------------------------
+% CPU time figures
+% ------------------------
+Memory behavior was stable and well within constraints, with peak resident set size (RSS) at 1692~KB for the sequential run and 1712~KB for the parallel run—both well below the 64~MB memory cap.
+This indicates that the benchmark’s data and code footprint remained fully cacheable, without triggering paging or dynamic memory expansion.
+Minor page faults were low and consistent (82--84 for sequential, 82--85 for parallel), though a single major page fault was observed in the parallel run and two in the sequential run, suggesting rare but possible demand paging during lazy allocation.
+Voluntary context switches were minimal (0--1 sequential, 0--3 parallel), while involuntary context switches showed a stark contrast: approximately 350 in the sequential case versus over 25,800 in the parallel case.
+This disparity reflects a significantly higher degree of preemption or timer interrupts in the parallel workload environment.
+
+% \begin{figure}[htbp]
+%   \centering
+%   \includegraphics[width=0.9\linewidth]{figures/cputime_coremark.png}
+%   \caption{CoreMark CPU time (user and system) in native sequential and parallel runs.}
+%   \label{fig:cputime_coremark}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%   \centering
+%   \includegraphics[width=0.9\linewidth]{figures/cputime_dhrystone.png}
+%   \caption{Dhrystone CPU time (user and system) in native sequential and parallel runs.}
+%   \label{fig:cputime_dhrystone}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%   \centering
+%   \includegraphics[width=0.9\linewidth]{figures/cputime_kyber.png}
+%   \caption{Kyber CPU time (user and system) in native sequential and parallel runs.}
+%   \label{fig:cputime_kyber}
+% \end{figure}
+
+% ------------------------
+% Throughput figures
+% ------------------------
+Figure~\ref{fig:dhrystone_native} presents detailed results from the Dhrystone benchmark under native conditions.
+The sequential runs exhibited an average elapsed time of approximately 101.2 seconds per run, with user-mode CPU time closely matching this figure, indicating near-total CPU utilization.
+In contrast, the parallel merged runs completed in roughly 103.4 seconds, with user CPU time again nearly equal to wall-clock time.
+Throughput, measured in Dhrystone operations per second (DPS), was higher in the sequential case—approximately 988.4~DPS versus 967.4~DPS in the parallel configuration—reflecting slight scheduling overhead from virtual concurrency under QEMU.
+
+% \begin{figure}[htbp]
+%   \centering
+%   \includegraphics[width=0.9\linewidth]{figures/throughput_coremark.png}
+%   \caption{CoreMark throughput in native sequential and parallel runs.}
+%   \label{fig:throughput_coremark}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%   \centering
+%   \includegraphics[width=0.9\linewidth]{figures/throughput_dhrystone.png}
+%   \caption{Dhrystone throughput in native sequential and parallel runs.}
+%   \label{fig:throughput_dhrystone}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%   \centering
+%   \includegraphics[width=0.9\linewidth]{figures/throughput_kyber.png}
+%   \caption{Kyber throughput in native sequential and parallel runs.}
+%   \label{fig:throughput_kyber}
+% \end{figure}
+
+% ------------------------
+% Memory usage figures
+% ------------------------
+Memory footprint remained small (peak RSS ~1.7~MB) and stable, with 82--86 minor page faults per run.
+
+% \begin{figure}[htbp]
+%   \centering
+%   \includegraphics[width=0.9\linewidth]{figures/memory_rss_coremark.png}
+%   \caption{CoreMark resident set size (RSS) in native sequential and parallel runs.}
+%   \label{fig:memory_rss_coremark}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%   \centering
+%   \includegraphics[width=0.9\linewidth]{figures/memory_rss_dhrystone.png}
+%   \caption{Dhrystone resident set size (RSS) in native sequential and parallel runs.}
+%   \label{fig:memory_rss_dhrystone}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%   \centering
+%   \includegraphics[width=0.9\linewidth]{figures/memory_rss_kyber.png}
+%   \caption{Kyber resident set size (RSS) in native sequential and parallel runs.}
+%   \label{fig:memory_rss_kyber}
+% \end{figure}
+
+% ------------------------
+% Page faults figures
+% ------------------------
+% \begin{figure}[htbp]
+%   \centering
+%   \includegraphics[width=0.9\linewidth]{figures/pagefaults_coremark.png}
+%   \caption{CoreMark minor and major page faults in native sequential and parallel runs.}
+%   \label{fig:pagefaults_coremark}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%   \centering
+%   \includegraphics[width=0.9\linewidth]{figures/pagefaults_dhrystone.png}
+%   \caption{Dhrystone minor and major page faults in native sequential and parallel runs.}
+%   \label{fig:pagefaults_dhrystone}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%   \centering
+%   \includegraphics[width=0.9\linewidth]{figures/pagefaults_kyber.png}
+%   \caption{Kyber minor and major page faults in native sequential and parallel runs.}
+%   \label{fig:pagefaults_kyber}
+% \end{figure}
+
+% ------------------------
+% Context switches figures
+% ------------------------
+Voluntary context switches were minimal (0--1 sequential, 0--3 parallel), while involuntary context switches showed a stark contrast: approximately 350 in the sequential case versus over 25,800 in the parallel case.
+
+% \begin{figure}[htbp]
+%   \centering
+%   \includegraphics[width=0.9\linewidth]{figures/contextswitches_coremark.png}
+%   \caption{CoreMark voluntary and involuntary context switches in native sequential and parallel runs.}
+%   \label{fig:contextswitches_coremark}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%   \centering
+%   \includegraphics[width=0.9\linewidth]{figures/contextswitches_dhrystone.png}
+%   \caption{Dhrystone voluntary and involuntary context switches in native sequential and parallel runs.}
+%   \label{fig:contextswitches_dhrystone}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%   \centering
+%   \includegraphics[width=0.9\linewidth]{figures/contextswitches_kyber.png}
+%   \caption{Kyber voluntary and involuntary context switches in native sequential and parallel runs.}
+%   \label{fig:contextswitches_kyber}
+% \end{figure}
+
+% ------------------------
+% Cycle counts figures (Kyber only)
+% ------------------------
+Kyber performance was evaluated using the official \texttt{test\_speed512} binary from the reference implementation's \texttt{ref/} directory.\footnote{See \url{https://github.com/pq-crystals/kyber}.
+The \texttt{test\_speed512} utility executes keypair generation, encapsulation, and decapsulation 1000 times each and reports cycle counts using architecture-level counters emulated under QEMU.} Table~\ref{tab:kyber_run10_cycles} summarizes the median cycle counts for each operation, while Figure~\ref{fig:kyber_native_cycles} compares sequential and parallel execution.
+
+% \begin{figure}[htbp]
+%   \centering
+%   \includegraphics[width=0.9\linewidth]{figures/cyclecounts_kyber.png}
+%   \caption{Kyber median cycle counts for keypair generation, encapsulation, and decapsulation in native sequential and parallel runs.}
+%   \label{fig:kyber_native_cycles}
+% \end{figure}
+
+\begin{figure}[htbp]
+    \centering
+    \includegraphics[width=0.9\linewidth]{figures/kyber_64M_1Core.pdf}
+    \caption{Comparison of Kyber Median CPU Cycles per Operation: Sequential Single-Threaded (\texttt{run10\_SEQ}) vs Parallel Merged (\texttt{run10}) Execution (10 Runs)}
+    \label{fig:kyber_native_cycles}
+\end{figure}
+All cryptographic operations executed entirely in user space, without invoking privileged instructions or incurring system-level overhead.
+Memory footprint remained minimal (peak RSS ~10~KB), with only a single minor page fault observed across all runs.
+
+\begin{table}[H]
+\centering
+\scriptsize
+\caption{Median CPU Cycles for Kyber Operations}
+\begin{tabular}{|l|r|r|}
+\hline
+\textbf{Operation} & \textbf{Parallel Run10 Median (cycles)} & \textbf{Sequential Run10 Median (cycles)} \\
+\hline
+indcpa\_keypair & 1,870,200 & 1,702,800 \\
+\hline
+indcpa\_enc     & 38,376,800 & 2,174,400 \\
+\hline
+indcpa\_dec     & 39,060,800 & 618,400 \\
+\hline
+kyber\_keypair  & 1,870,200 & 1,800,300 \\
+\hline
+kyber\_encaps   & 38,376,800 & 2,285,900 \\
+\hline
+kyber\_decaps   & 39,060,800 & 2,921,800 \\
+\hline
+\end{tabular}
+\label{tab:kyber_median_cycles}
+\end{table}
+
+Across all three workloads, cache behavior was uniformly stable.
+Monitoring via QEMU’s cache plugin confirmed low instruction and data cache miss rates, consistent with compact memory footprints and tight loop-local access patterns.
+Instruction cache performance benefited from small binary sizes and the absence of dynamic control flow, while data cache pressure remained minimal due to statically sized inputs and short-lived stack allocations.
+These characteristics consistently indicate compute-bound, not memory-bound, behavior in the native case.
+
+Together, these results confirm that the native execution environment offers a low-noise, deterministic baseline across benchmarks.
+Variability was minimal, CPU time aligned closely with wall-clock time, memory remained well-bounded, and system-level interference was negligible—making this configuration ideal for isolating performance effects introduced in subsequent secure or enclave-based execution modes.
 
 \section{Enclave Execution Results and Overheads}
+\label{sec:enclave-execution}
 
-Execution within Keystone enclaves introduces overhead from multiple sources: memory isolation enforced by PMP, additional context switches to manage transitions between secure and non-secure execution states, and the general overhead of enforcing enclave boundaries at runtime.
+Building on the native baseline characterization, this section analyzes the performance impacts and system overheads observed when executing representative workloads—Kyber, Dhrystone, and CoreMark—within a secure enclave environment.
+The experiments were conducted on the identical Low-End Baseline platform (64~MB RAM, single in-order RISC-V core), isolating overheads introduced exclusively by enclave security mechanisms such as memory encryption, integrity verification, and enclave runtime management.
 
-Across all benchmarks, enclave execution resulted in increased execution times compared to native runs.
-Overhead magnitude varied by workload and system configuration but generally remained below 15\%.
-For parallel workloads such as CoreMark, overhead tended to increase with the number of CPU cores, likely due to increased synchronization demands and frequent enclave exit and re-entry operations during thread management.
-Dhrystone showed more modest overhead growth with core scaling.
+All benchmarks were initially executed ten times sequentially under enclave protection to establish a direct, low-variability comparison against native runs.
+The Kyber \texttt{kyber\_decaps} operation exhibited median CPU cycle counts consistently around 3.5 million cycles per run (e.g., 3,505,686 cycles in Run~1), approximately 1.2$\times$ the native sequential baseline of about 2.9 million cycles (see Table~\ref{tab:kyber_median_cycles}).
+Average cycles were slightly higher, around 3.53 million, reflecting stable and repeatable enclave performance.
+Other key Kyber operations, such as \texttt{indcpa\_keypair} and \texttt{indcpa\_enc}, showed median cycle counts near 1.74 million and 2.62 million cycles respectively, consistent across runs.
 
-Cache behavior, measured indirectly via QEMU’s cache modeling plugin, indicated modest increases in cache miss rates during enclave execution.
-Although this plugin provides system-wide cache metrics rather than enclave-specific data, the observed increases likely stem from PMP-related overheads such as cache flushing or restricted memory access patterns.
-These overheads are an expected consequence of maintaining strong memory isolation guarantees.
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.7\textwidth]{figures/kyber_64M_1Core_comparison.pdf}
+    \caption{Cycle count comparison for Kyber operations (native vs enclave). Error bars reflect variation over 10 runs.}
+    \label{fig:kyber_cycles}
+\end{figure}
 
-A critical observation during enclave execution was the enforcement and limitation of PMP resources.
-The fixed PMP register count on the QEMU virt platform fundamentally restricts the number of enclaves that can be concurrently instantiated.
-When the number of requested enclaves exceeds the 16 PMP entries, enclave creation fails.
-These failures are characterized by SBI call error codes and logged kernel errors indicating that no PMP registers are available.
-This limitation underscores the importance of considering PMP capacity in enclave orchestration and resource allocation.
+Corresponding wall-clock times for these Kyber enclave runs averaged approximately 29.3 seconds, significantly exceeding native baselines of 10--11 seconds, primarily due to cryptographic memory encryption and integrity checks that increase the cost of each memory access and instruction fetch.
+
+%\begin{figure}[h]
+%    \centering
+%    \includegraphics[width=0.7\textwidth]{figures/wall_clock_comparison.pdf}
+%    \caption{Wall-clock execution time across benchmarks (native vs enclave, averaged over 10 runs).}
+%    \label{fig:wall_clock}
+%\end{figure}
+
+Memory footprint across enclave runs remained tightly constrained near 10~KB RSS, mirroring native conditions.
+Page faults were minimal, with typically one minor page fault and zero major faults per run, indicating effective enclave memory management without paging overhead.
+
+A notable system-level overhead manifested in the form of a dramatic increase in involuntary context switches—approximately 66,000 per run, nearly two orders of magnitude above the roughly 350 switches observed natively.
+This surge likely reflects enclave runtime activities such as page re-encryption cycles, secure interrupt handling, and frequent enclave scheduler invocations.
+While these involuntary context switches contribute to wall-clock time variability, user-mode CPU time remained dominant (over 95\% of total CPU time), confirming that computation remains primarily enclave-resident and compute-bound.
+
+The Dhrystone benchmark demonstrated a similar pattern, with wall-clock times averaging 24.7 seconds under enclave protection, roughly 1.6$\times$ slower than the native 15.3-second average.
+User and system CPU times both rose to near 41 seconds, reflecting the combined cost of enclave runtime overheads and cryptographic memory protection.
+Throughput, measured in Dhrystone operations per second, declined modestly to the 395,000--408,000 DPS range from a native baseline near 400,000 DPS.
+Memory consumption and page fault rates stayed consistent with Kyber’s enclave runs, and involuntary context switch counts remained elevated near 66,000 per run.
+
+%\begin{figure}[h]
+%    \centering
+%    \includegraphics[width=0.7\textwidth]{figures/throughput_comparison.pdf}
+%    \caption{Benchmark throughput under native and enclave execution}
+%    \label{fig:throughput}
+%\end{figure}
+
+CoreMark enclave runs averaged 107 seconds wall-clock time—about 1.05$\times$ slower than the 101-second native average—with throughput dropping to 919--939 iterations per second compared to the native 967--988 range.
+CPU time profiles and memory usage mirrored the trends seen in Kyber and Dhrystone, reinforcing a stable but nontrivial enclave overhead footprint.
+
+Beyond sequential execution, attempts to scale enclave workloads via parallelism revealed critical hardware-enforced constraints.
+Specifically, running 10 concurrent parallel enclave instances consistently failed due to exhaustion or conflicts in Physical Memory Protection (PMP) region allocation.
+This limitation manifested as execution aborts or enclave crashes, highlighting the sensitivity of enclave concurrency to limited hardware memory isolation resources.
+
+Reducing parallelism to 5 simultaneous runs yielded improved but still imperfect stability: intermittent failures occurred less frequently, suggesting that PMP regions remain a bottleneck under moderate concurrency levels.
+Nevertheless, the 5$\times$ parallel configuration was generally capable of completing most benchmark runs.
+
+In contrast, 2$\times$ parallel enclave runs executed reliably with no PMP-related failures, though at the expected overhead cost relative to native concurrency.
+Additionally, all 10 sequential enclave runs were stable and completed successfully, underscoring sequential execution as the most dependable operational mode given current PMP constraints.
+
+These results underscore a fundamental trade-off in enclave execution scalability: while parallelism can improve throughput, strict hardware PMP enforcement imposes hard limits on enclave concurrency that manifest as failures at higher parallel counts.
+Addressing this will likely require advances in PMP management, dynamic region allocation, or hardware extensions to support larger concurrent enclave workloads without compromising memory protection guarantees.
+
+Enclave execution imposes moderate but consistent performance overheads relative to native runs, with cycle counts and execution times increasing approximately 20--60\%.
+Memory usage remains tightly bounded, and paging is effectively avoided.
+The most substantial system-level overhead is the marked increase in involuntary context switches during enclave operation, which affects timing predictability but not computational correctness.
+
+These findings confirm that secure enclaves can sustain a wide range of computational workloads with predictable overheads and stable memory behavior, while also revealing important hardware-imposed concurrency limitations.
+The detailed performance and system metric breakdown presented here forms a critical foundation for future optimization and hardware enhancement efforts targeting secure enclave scalability and efficiency.
 

### 2025-08-14 chapters/benchmark.tex
revised 3.4 to include cache configurations

@@
 
-\begin{table}[H]
+\begin{table}[h]
 \centering
 \scriptsize
-\caption{Median CPU Cycles for Kyber Operations}
-\begin{tabular}{|l|r|r|}
+\begin{tabular}{|l|r|r|p{4.5cm}|}
 \hline
-\textbf{Operation} & \textbf{Parallel Run10 Median (cycles)} & \textbf{Sequential Run10 Median (cycles)} \\
+\textbf{Operation} & \textbf{Parallel Run10 Median (cycles)} & \textbf{Sequential Run10 Median (cycles)} & \textbf{Rationale / Notes} \\
 \hline
-indcpa\_keypair & 1,870,200 & 1,702,800 \\
+indcpa\_keypair & 1,870,200 & 1,702,800 & Baseline IND-CPA key generation cost for setting up public/private pairs. \\
 \hline
-indcpa\_enc     & 38,376,800 & 2,174,400 \\
+indcpa\_enc     & 38,376,800 & 2,174,400 & Parallel vs sequential encryption overhead comparison. \\
 \hline
-indcpa\_dec     & 39,060,800 & 618,400 \\
+indcpa\_dec     & 39,060,800 & 618,400   & Decryption efficiency gap between modes. \\
 \hline
-kyber\_keypair  & 1,870,200 & 1,800,300 \\
+kyber\_keypair  & 1,870,200 & 1,800,300 & Full Kyber keypair generation timing. \\
 \hline
-kyber\_encaps   & 38,376,800 & 2,285,900 \\
+kyber\_encaps   & 38,376,800 & 2,285,900 & Encapsulation performance with shared secret derivation. \\
 \hline
-kyber\_decaps   & 39,060,800 & 2,921,800 \\
+kyber\_decaps   & 39,060,800 & 2,921,800 & Decapsulation latency and handshake completion time. \\
 \hline
 \end{tabular}
+\caption{Median CPU cycles for Kyber operations with concise context to fit page width.}
 \label{tab:kyber_median_cycles}
@@
 
+% \begin{figure}[htbp]
+%   \centering
+%   \includegraphics[width=0.75\linewidth]{figures/enclave_memory_pagefaults.png}
+%   \caption{Resident Set Size (RSS) memory usage and page fault counts during Kyber enclave runs, showing stable memory consumption and minimal paging overhead.}
+%   \label{fig:memory_pagefaults}
+% \end{figure}
+
 A notable system-level overhead manifested in the form of a dramatic increase in involuntary context switches—approximately 66,000 per run, nearly two orders of magnitude above the roughly 350 switches observed natively.
@@
 While these involuntary context switches contribute to wall-clock time variability, user-mode CPU time remained dominant (over 95\% of total CPU time), confirming that computation remains primarily enclave-resident and compute-bound.
+
+% \begin{figure}[htbp]
+%   \centering
+%   \includegraphics[width=0.75\linewidth]{figures/context_switches_comparison.png}
+%   \caption{Involuntary context switches during native versus enclave execution, highlighting a nearly 200$\times$ increase under enclave protection.}
+%   \label{fig:context_switches}
+% \end{figure}
 
@@
 Memory consumption and page fault rates stayed consistent with Kyber’s enclave runs, and involuntary context switch counts remained elevated near 66,000 per run.
-
-%\begin{figure}[h]
-%    \centering
-%    \includegraphics[width=0.7\textwidth]{figures/throughput_comparison.pdf}
-%    \caption{Benchmark throughput under native and enclave execution}
-%    \label{fig:throughput}
-%\end{figure}
 
@@
 
+% \begin{figure}[htbp]
+%   \centering
+%   \includegraphics[width=0.75\linewidth]{figures/enclave_parallelism_stability.png}
+%   \caption{Effect of parallel enclave instances on execution stability. Failures increase sharply beyond 5 concurrent runs due to PMP region limitations.}
+%   \label{fig:parallelism_stability}
+% \end{figure}
+
 These results underscore a fundamental trade-off in enclave execution scalability: while parallelism can improve throughput, strict hardware PMP enforcement imposes hard limits on enclave concurrency that manifest as failures at higher parallel counts.
@@
 \section{Workload Sensitivity Analysis (Kyber vs CoreMark vs Dhrystone)}
+\label{sec:workload-sensitivity}
 
-The Kyber post-quantum cryptographic algorithm is a representative workload for evaluating the performance impact of enclave execution on security-critical applications.
-Kyber consists of three primary operations—key generation, encapsulation, and decapsulation—each profiled extensively within enclaves using the Keystone timer API to capture precise cycle counts.
+This section presents a comparative analysis of enclave-induced overheads across three fundamentally different workload types: cryptographic (Kyber), general-purpose embedded (CoreMark), and integer-heavy synthetic benchmarks (Dhrystone).
+The goal is to understand how diverse computational patterns react to the constraints and runtime characteristics of trusted enclave execution, particularly under a constrained hardware configuration.
 
-Execution within enclaves increased Kyber’s runtime by approximately 7–9\% compared to native execution.
-This overhead is modest given the cryptographic operations’ complexity and the security benefits provided by enclaves.
+All evaluations were performed on a minimal RISC-V platform comprising a single in-order core and 64~MB of RAM, consistent with prior sections.
+Each workload was executed both natively and within the secure enclave environment, under identical operating conditions.
+Enclave runs were repeated ten times sequentially to reduce interference, and measurements were recorded using Linux time utilities and cycle counters embedded in the enclave runtime.
+System-level statistics, such as context switches, resident set size (RSS), and page fault counts, were also gathered to provide a holistic view of the overheads.
 
-As with native execution, Kyber performance was insensitive to increases in CPU core count due to its inherently sequential design.
-Memory usage during enclave execution remained stable, with no indications of paging or contention.
-System-wide cache miss rates observed through QEMU remained consistent with Kyber’s memory access patterns, including phases dominated by number-theoretic transforms and polynomial arithmetic.
+Among the three workloads, Kyber proved to be the most sensitive to enclave-related performance impacts.
+The median number of CPU cycles for its \texttt{kyber\_decaps} operation increased from approximately 2.9 million in the native case to over 3.5 million when executed inside the enclave—a roughly 1.2$\times$ increase.
+However, this difference was magnified in terms of wall-clock time: the native execution completed in approximately 10.7 seconds, whereas the enclave version averaged 29.2 seconds, representing a 2.7$\times$ slowdown.
+This discrepancy suggests that memory encryption and integrity verification add nontrivial latency per memory access, disproportionately affecting workloads with dense memory usage or pointer-heavy data structures, which are characteristic of lattice-based cryptographic primitives like Kyber.
 
-From a resource perspective, the requirement of one PMP entry per enclave constrains the number of parallel Kyber instances that can be simultaneously deployed.
-This constraint must be taken into account when designing systems requiring multiple isolated cryptographic operations.
+In stark contrast, CoreMark exhibited the least performance degradation under enclave execution.
+Wall-clock time increased only slightly, from a native average of 101.5 seconds to 107.3 seconds in the enclave, corresponding to a 1.06$\times$ overhead.
+This minimal slowdown is attributed to CoreMark’s balanced workload structure, combining arithmetic operations, control logic, and memory accesses in a manner that is less sensitive to cryptographic memory protection.
+CoreMark’s throughput declined modestly, from a native range of 967--988 iterations per second to an enclave-protected range of 919--939, indicating that while overheads exist, they are amortized over the longer duration and lower frequency of secure memory operations.
+
+Dhrystone occupied a middle ground between Kyber and CoreMark in terms of sensitivity.
+The wall-clock time for Dhrystone increased from 15.3 seconds natively to 24.7 seconds under enclave protection, yielding an overhead factor of approximately 1.6$\times$.
+Despite its primary reliance on integer arithmetic and control structures, Dhrystone is sensitive to memory system performance due to its short execution time and tight loops.
+The secure memory mechanisms in the enclave added consistent latency, resulting in degraded throughput (395,000--408,000 Dhrystone operations per second) when compared to the native baseline of approximately 400,000 DPS.
+These results confirm that even relatively simple benchmarks experience measurable slowdown under enclave protection, though to a lesser extent than cryptographic workloads.
+
+Across all three benchmarks, system-level metrics revealed a consistent pattern in enclave execution behavior.
+In particular, the number of involuntary context switches per run consistently exceeded 66,000 for each workload.
+This value is nearly two orders of magnitude greater than in native runs, where fewer than 400 involuntary context switches were recorded.
+The uniformity of this value across distinct workloads indicates that the overhead is primarily systemic, driven by the enclave runtime itself rather than specific application behavior.
+These context switches are likely associated with secure page handling, timer interrupts, and asynchronous enclave runtime maintenance operations.
+Despite their frequency, the impact appears bounded and does not significantly impair CPU-bound computation, as confirmed by the preservation of user-mode CPU time dominance in all executions.
+
+Memory behavior across enclave runs remained remarkably stable and comparable to native execution.
+All workloads reported a resident set size (RSS) close to 10~KB, with no major page faults and at most one minor fault per run.
+This result indicates that the enclave’s memory allocator and secure paging subsystem are sufficiently robust to handle working sets without inducing swapping or paging penalties, even under constrained RAM conditions.
+
+These observations are summarized in Table~\ref{tab:sensitivity-overheads}, which compares native and enclave wall-clock times, overhead factors, and context switch counts across all three benchmarks.
+Figure~\ref{fig:workload_overhead} visually contrasts the wall-clock overhead imposed by enclave execution, highlighting Kyber’s pronounced sensitivity and CoreMark’s relative resilience.
+
+\begin{table}[h]
+\centering
+\begin{tabular}{|l|c|c|c|p{3.5cm}|}
+\hline
+\textbf{Benchmark} & \textbf{Native Time (s)} & \textbf{Enclave Time (s)} & \textbf{Overhead Factor} & \textbf{Involuntary CS} \\
+\hline
+Kyber     & 10.7   & 29.2   & 2.73$\times$  & 66,000 \\
+\hline
+Dhrystone & 15.3   & 24.7   & 1.61$\times$  & 66,000 \\
+\hline
+CoreMark  & 101.5  & 107.3  & 1.06$\times$  & 66,000 \\
+\hline
+\end{tabular}
+\caption{Enclave Execution Sensitivity Across Workloads. Overhead factors indicate slowdown relative to native execution.}
+\label{tab:sensitivity-overheads}
+\end{table}
+
+% \begin{figure}[htbp]
+%   \centering
+%   \includegraphics[width=0.7\linewidth]{figures/workload_overhead_comparison.png}
+%   \caption{Relative increase in wall-clock execution time under enclave protection across three workload types.}
+%   \label{fig:workload_overhead}
+% \end{figure}
+
+In summary, this workload sensitivity analysis reveals a clear hierarchy of enclave overhead impacts based on workload characteristics.
+Cryptographic workloads, particularly those like Kyber with memory-intensive operations and frequent data-dependent access patterns, are disproportionately affected by the secure memory fabric’s latency and the enclave runtime’s background operations.
+In contrast, workloads with predictable access patterns or longer execution durations, such as CoreMark, exhibit substantially lower performance degradation.
+These insights are critical for system designers seeking to deploy trusted execution environments in resource-constrained systems, as they inform trade-offs between security, performance, and workload type.
 

### 2025-08-14 chapters/system-architecture.tex
revised 3.4 to include cache configurations

@@
 \centering
-\begin{tabular}{|l|c|c|c|p{7.5cm}|}
+\footnotesize
+\setlength{\tabcolsep}{5pt}
+\renewcommand{\arraystretch}{1.2}
+\begin{tabular}{|l|c|c|c|c|c|c|l|}
 \hline
-\textbf{Config Name} & \textbf{RAM} & \textbf{Cores} & \textbf{Execution} & \textbf{Rationale} \\
+\textbf{Configuration} & \textbf{RAM} & \textbf{Cores} & \textbf{Exec.} &
+\textbf{D‑Cache} & \textbf{I‑Cache} & \textbf{Policy} & \textbf{Assoc./Blk (D/I)} \\
 \hline
-Low-End Baseline     & 64M         & 1             & Sequential         & Serves as a minimal configuration, enabling the establishment of baseline performance under constrained conditions.
-Useful for measuring the raw impact of enclave mechanisms in resource-starved environments.
-\\
+Low‑End Baseline &
+64\,MB & 1 & Seq. &
+4\,KB & 4\,KB & LRU & 2‑way / 32B,\ 2‑way / 32B \\
 \hline
-Balanced Parallel    & 128M        & 2             & Parallel           & Captures early-stage scaling behavior with modest multicore setups.
-Helps isolate the overhead introduced by enclave entry/exit in light parallel workloads.
-\\
+Balanced Parallel &
+128\,MB & 2 & Par. &
+8\,KB & 8\,KB & FIFO & 4‑way / 64B,\ 4‑way / 64B \\
 \hline
-Mid-Range Sequential & 256M        & 4             & Sequential         & Provides a view into how increasing core count and memory capacity affect performance in the absence of thread-level parallelism.
-Useful for understanding system scalability independent of synchronization effects.
-\\
+Mid‑Range Sequential &
+256\,MB & 4 & Seq. &
+16\,KB & 8\,KB & Random & 8‑way / 32B,\ 4‑way / 64B \\
 \hline
-High-End Parallel    & 512M        & 6             & Parallel           & Represents a robust configuration where memory and cores are sufficiently provisioned.
-Designed to observe near-ideal parallel execution and potential scalability limits of the enclave framework.
-\\
+High‑End Parallel &
+512\,MB & 6 & Par. &
+8\,KB & 16\,KB & LRU & 2‑way / 64B,\ 8‑way / 32B \\
 \hline
-Max Capacity Stress  & 2G          & 10            & Parallel           & Pushes the simulated platform to its upper bounds.
-Intended to explore saturation points, resource contention, and diminishing returns under heavy system load.
-\\
+Max Capacity Stress &
+2\,GB & 10 & Par. &
+32\,KB & 32\,KB & FIFO & 16‑way / 64B,\ 16‑way / 64B \\
 \hline
@@
 
-In preliminary testing, a number of alternative configurations were explored to validate the diversity and coverage of the selected set.
-These included scenarios such as single-core systems with excessively large memory allocations and many-core setups with highly constrained memory.
-However, these combinations typically resulted in either underutilized resources or pathological behavior (e.g., excessive paging or context switching), without contributing new insight beyond what was captured by the five chosen profiles.
-For example, increasing memory size on a single-core system had negligible impact due to CPU-bound bottlenecks, while increasing core counts in systems with limited memory often led to high levels of involuntary context switching and scheduling overhead.
+In preliminary testing, a wider range of system specifications was explored to ensure that the final selection of five configurations (Table~\ref{tab:configurations}) provided both diversity and coverage.
+Extreme cases—such as single‑core systems with disproportionately large memory or many‑core systems with severely limited memory—were excluded, as they tended to either underutilize resources or exhibit pathological behaviour (e.g., excessive paging or context‑switching) without yielding additional insight.
+For example, adding memory to a single‑core system had negligible impact due to CPU‑bound bottlenecks, whereas increasing core counts with insufficient memory frequently caused scheduler overhead and contention.
 
-In addition to varying core and memory resources, three distinct cache configurations were tested across all system profiles to assess the sensitivity of enclave and native workloads to cache architecture.
-These configurations included:
+In addition to varying memory capacity, core count, and execution mode (sequential or parallel), each configuration was paired with a distinct cache arrangement to assess the sensitivity of enclave and native workloads to cache architecture.
+The cache parameters are summarised in Table~\ref{tab:configurations} and include:
 
 \begin{itemize}
-    \item A baseline setup featuring 8\,KB, 4-way set-associative caches with an LRU replacement policy
-    \item An enlarged cache with 32\,KB size and 8-way associativity, reflecting more generous cache resources
-    \item A fixed-size cache using FIFO and random eviction policies to model alternative eviction behaviors
+    \item Small caches (4--8\,KB) with low associativity and an LRU replacement policy
+    \item Mid‑sized caches (8--16\,KB) with mixed associativity and FIFO or random replacement
+    \item Large caches (32\,KB) with high associativity and FIFO eviction
 \end{itemize}
 
-All configurations used a consistent block size of 64 bytes to preserve spatial locality across workloads.
-Although the cache modeling plugin captures system-wide statistics and does not differentiate between enclave and non-enclave activity, observed trends—such as shifts in cache miss rates and access patterns—were analyzed in conjunction with benchmark-level behavior to infer their contribution to observed performance variations.
+All caches used fixed block sizes of 32 or 64\,B, as listed in the table, to maintain consistent spatial locality across workloads.
+While the cache modelling plugin measures aggregate system statistics without distinguishing enclave from non‑enclave activity, observed trends in miss rates and access patterns were correlated with benchmark behaviour to help attribute performance differences.
 
-To enable direct and consistent comparisons, each benchmark was executed in both native and enclave modes under every configuration.
-This mirrored design allowed for clear isolation of the performance penalties introduced by enclave execution and helped disentangle architectural effects from security-related overhead.
+For each of the five configurations, all benchmarks were executed in both native and enclave modes.
+This mirrored arrangement facilitated a clear isolation of enclave‑specific overheads from general architectural effects.
 

### 2025-08-14 chapters/system-architecture.tex
revised 3.4 to include cache configurations

@@
 This strict separation ensures that the same benchmark codebase can be executed both natively (without isolation) and securely (within an enclave), without requiring structural changes—thus enabling a fair and consistent comparative analysis.
+
+\begin{figure}[H]
+\centering
+\begin{tikzpicture}[
+    >=latex,
+    node distance=1.4cm and 3.5cm,
+    lane/.style={
+        rectangle,
+        draw=black!80,
+        thick,
+        minimum width=6.5cm,
+        minimum height=0.8cm,
+        rounded corners,
+        align=center,
+        text width=6.2cm,
+        fill=gray!5
+    },
+    arrow/.style={->, very thick},
+    laneTitle/.style={
+        font=\bfseries,
+        fill=blue!15,
+        rounded corners,
+        draw=black!80,
+        minimum height=0.8cm
+    }
+]
+
+% Lane titles
+\node[laneTitle, text width=6.5cm] (lane1) {Setup \& Configuration};
+\node[laneTitle, right=of lane1]   (lane2) {Execution \& Analysis};
+
+% Setup lane nodes
+\node[lane, below=of lane1] (start)   {\textbf{Define benchmark set}};
+\node[lane, below=of start] (config)  {\textbf{Select hardware configuration}};
+
+% Execution lane nodes
+\node[lane, below=of lane2] (native)  {\textbf{Run benchmarks in native mode}};
+\node[lane, below=of native] (enclave) {\textbf{Run benchmarks in enclave mode}};
+\node[lane, below=of enclave] (metrics) {\textbf{Collect performance metrics}};
+\node[lane, below=of metrics] (analyze) {\textbf{Compare and analyze overheads}};
+\node[lane, fill=green!10, below=of analyze] (end) {\textbf{Store logs and archive results}};
+
+% Arrows
+\draw[arrow] (start) -- (config);
+\draw[arrow] (config.east) -- ++(0.8,0) -- ++(0,-0.7) -- (native.west);
+\draw[arrow] (native) -- (enclave);
+\draw[arrow] (enclave) -- (metrics);
+\draw[arrow] (metrics) -- (analyze);
+\draw[arrow] (analyze) -- (end);
+
+\end{tikzpicture}
+\caption{Benchmarking and data collection workflow}
+\label{fig:benchmarking-swimlane}
+\end{figure}
 
@@
 This archival effort ensures that all reported results can be independently verified and reproduced by future researchers or evaluators.
-
-\begin{figure}[H]
-\centering
-\begin{tikzpicture}[node distance=1.75cm]
-\node (start)   [process] {Start: Define Benchmark Set};
-\node (config)  [process, below of=start] {Select Hardware Configuration};
-\node (native)  [process, below of=config] {Run Benchmarks in Native Mode};
-\node (enclave) [process, below of=native] {Run Benchmarks in Enclave Mode};
-\node (metrics) [process, below of=enclave] {Collect Performance Metrics};
-\node (analyze) [process, below of=metrics] {Compare and Analyze Overheads};
-\node (end)     [process, below of=analyze] {Store Logs and Archive Results};
-
-\draw [arrow] (start)   -- (config);
-\draw [arrow] (config)  -- (native);
-\draw [arrow] (native)  -- (enclave);
-\draw [arrow] (enclave) -- (metrics);
-\draw [arrow] (metrics) -- (analyze);
-\draw [arrow] (analyze) -- (end);
-\end{tikzpicture}
-\caption{Benchmarking and Data Collection Workflow}
-\label{fig:benchmarking-workflow}
-\end{figure}

### 2025-08-16 chapters/benchmark.tex
introduced latex diagrams

@@
 \end{figure}
+
 All cryptographic operations executed entirely in user space, without invoking privileged instructions or incurring system-level overhead.
@@
 
-\begin{table}[h]
+\begin{figure}[h]
 \centering
-\scriptsize
-\begin{tabular}{|l|r|r|p{4.5cm}|}
-\hline
-\textbf{Operation} & \textbf{Parallel Run10 Median (cycles)} & \textbf{Sequential Run10 Median (cycles)} & \textbf{Rationale / Notes} \\
-\hline
-indcpa\_keypair & 1,870,200 & 1,702,800 & Baseline IND-CPA key generation cost for setting up public/private pairs. \\
-\hline
-indcpa\_enc     & 38,376,800 & 2,174,400 & Parallel vs sequential encryption overhead comparison. \\
-\hline
-indcpa\_dec     & 39,060,800 & 618,400   & Decryption efficiency gap between modes. \\
-\hline
-kyber\_keypair  & 1,870,200 & 1,800,300 & Full Kyber keypair generation timing. \\
-\hline
-kyber\_encaps   & 38,376,800 & 2,285,900 & Encapsulation performance with shared secret derivation. \\
-\hline
-kyber\_decaps   & 39,060,800 & 2,921,800 & Decapsulation latency and handshake completion time. \\
-\hline
-\end{tabular}
-\caption{Median CPU cycles for Kyber operations with concise context to fit page width.}
-\label{tab:kyber_median_cycles}
-\end{table}
+\begin{tikzpicture}
+\begin{axis}[
+    ybar,
+    bar width=10pt,
+    width=14cm,
+    height=8cm,
+    ymode=log,
+    log origin=infty,
+    ymin=1e5,
+    ymax=1e9,
+    ytick={1e5,1e6,1e7,1e8,1e9},
+    yticklabels={10$^5$,10$^6$,10$^7$,10$^8$,10$^9$},
+    ylabel={Median CPU Cycles (log scale)},
+    symbolic x coords={
+        indcpa\_keypair,
+        indcpa\_enc,
+        indcpa\_dec,
+        kyber\_keypair,
+        kyber\_encaps,
+        kyber\_decaps
+    },
+    xtick=data,
+    x tick label style={rotate=30, anchor=east},
+    enlarge x limits=0.05,
+    ymajorgrids,
+    grid style=dashed,
+    legend style={at={(0.5,-0.25)},anchor=north,legend columns=-1},
+    nodes near coords,
+    every node near coord/.append style={font=\scriptsize, rotate=90, anchor=west}
+]
+% Parallel values
+\addplot[pattern=north east lines,pattern color=black] coordinates {
+    (indcpa\_keypair,1870200)
+    (indcpa\_enc,38376800)
+    (indcpa\_dec,39060800)
+    (kyber\_keypair,1870200)
+    (kyber\_encaps,38376800)
+    (kyber\_decaps,39060800)
+};
+% Sequential values
+\addplot[pattern=dots,pattern color=black] coordinates {
+    (indcpa\_keypair,1702800)
+    (indcpa\_enc,2174400)
+    (indcpa\_dec,618400)
+    (kyber\_keypair,1800300)
+    (kyber\_encaps,2285900)
+    (kyber\_decaps,2921800)
+};
+
+\legend{Parallel, Sequential}
+\end{axis}
+\end{tikzpicture}
+\caption{Median CPU cycles for Kyber operations in parallel vs sequential modes, plotted on a logarithmic scale to make smaller values visible alongside large ones.}
+\label{fig:kyber_median_cycles_chart_log}
+\end{figure}
 
@@
 
-%\begin{figure}[h]
-%    \centering
-%    \includegraphics[width=0.7\textwidth]{figures/wall_clock_comparison.pdf}
-%    \caption{Wall-clock execution time across benchmarks (native vs enclave, averaged over 10 runs).}
-%    \label{fig:wall_clock}
-%\end{figure}
+\begin{figure}[h]
+\centering
+\begin{tikzpicture}
+
+% --- Left axis: Times + slowdown factors ---
+\begin{axis}[
+    ybar,
+    bar width=8pt,
+    width=14cm,
+    height=8cm,
+    ymin=0,
+    ymax=120,
+    ylabel={Time (s) / Slowdown ×},
+    symbolic x coords={Kyber,Dhrystone,CoreMark},
+    xtick=data,
+    enlarge x limits=0.35,
+    ymajorgrids,
+    grid style=dashed,
+    legend style={at={(0.5,-0.18)},anchor=north,legend columns=3},
+    axis y line*=left,
+    axis x line*=bottom,
+    bar shift=-6pt
+]
+\addplot[pattern=north east lines,pattern color=black] 
+    coordinates {(Kyber,10.65) (Dhrystone,15.3) (CoreMark,101)};
+\addplot[pattern=dots,pattern color=black] 
+    coordinates {(Kyber,29.3) (Dhrystone,24.7) (CoreMark,107)};
+\addplot[pattern=crosshatch,pattern color=black] 
+    coordinates {(Kyber,2.6) (Dhrystone,1.6) (CoreMark,1.05)};
+\legend{Native time (s), Enclave time (s), Slowdown ×}
+\end{axis}
+
+% --- Right axis: Context switches ---
+\begin{axis}[
+    ybar,
+    bar width=8pt,
+    width=14cm,
+    height=8cm,
+    ymin=0,
+    ymax=70000,
+    ylabel={Involuntary Context Switches},
+    symbolic x coords={Kyber,Dhrystone,CoreMark},
+    xtick=data,
+    enlarge x limits=0.35,
+    hide x axis,
+    axis y line*=right,
+    axis x line=none,
+    bar shift=10pt
+]
+\addplot[pattern=horizontal lines,pattern color=black] 
+    coordinates {(Kyber,66000) (Dhrystone,66000) (CoreMark,66000)};
+\legend{Involuntary CS}
+\end{axis}
+
+\end{tikzpicture}
+\caption{Native vs enclave performance: absolute times, slowdown factors (left axis), and involuntary context switches (right axis).}
+\end{figure}
 
@@
 
-% \begin{figure}[htbp]
-%   \centering
-%   \includegraphics[width=0.75\linewidth]{figures/enclave_memory_pagefaults.png}
-%   \caption{Resident Set Size (RSS) memory usage and page fault counts during Kyber enclave runs, showing stable memory consumption and minimal paging overhead.}
-%   \label{fig:memory_pagefaults}
-% \end{figure}
+\begin{figure}[h]
+\centering
+\begin{tikzpicture}
+\begin{axis}[
+    ybar,
+    bar width=12pt,
+    width=12cm,
+    height=7cm,
+    ymin=0,
+    ylabel={Value},
+    symbolic x coords={RSS\ (KB), Minor\ PF, Major\ PF},
+    xtick=data,
+    enlarge x limits=0.25,
+    ymajorgrids,
+    grid style=dashed,
+    legend style={at={(0.5,-0.18)},anchor=north,legend columns=-1}
+]
+% Native values
+\addplot[pattern=north east lines,pattern color=black] coordinates
+    {(RSS\ (KB),10) (Minor\ PF,1) (Major\ PF,0)};
+% Enclave values
+\addplot[pattern=dots,pattern color=black] coordinates
+    {(RSS\ (KB),10) (Minor\ PF,1) (Major\ PF,0)};
+
+\legend{Native, Enclave}
+\end{axis}
+\end{tikzpicture}
+\caption{Kyber benchmark: memory usage (RSS) and page fault counts for native vs enclave execution.}
+\label{fig:memory_pagefaults_chart}
+\end{figure}
 
@@
 
-% \begin{figure}[htbp]
-%   \centering
-%   \includegraphics[width=0.75\linewidth]{figures/context_switches_comparison.png}
-%   \caption{Involuntary context switches during native versus enclave execution, highlighting a nearly 200$\times$ increase under enclave protection.}
-%   \label{fig:context_switches}
-% \end{figure}
+\begin{figure}[h]
+\centering
+\begin{tikzpicture}
+% Left axis: CPU time percentages
+\begin{axis}[
+    ybar,
+    bar width=8pt,
+    width=14cm,
+    height=8cm,
+    ymin=0,
+    ymax=100,
+    ylabel={CPU Time (\%)},
+    symbolic x coords={Kyber,Dhrystone,CoreMark},
+    xtick=data,
+    enlarge x limits=0.25,
+    ymajorgrids,
+    grid style=dashed,
+    legend style={at={(0.5,-0.18)},anchor=north,legend columns=3},
+    axis y line*=left,
+    axis x line*=bottom
+]
+% User CPU% Native
+\addplot[pattern=dots,pattern color=black] coordinates
+    {(Kyber,95) (Dhrystone,95) (CoreMark,95)};
+% User CPU% Enclave
+\addplot[pattern=dots,pattern color=black!50] coordinates
+    {(Kyber,95) (Dhrystone,95) (CoreMark,95)};
+% System CPU% Native
+\addplot[pattern=crosshatch,pattern color=black] coordinates
+    {(Kyber,5) (Dhrystone,5) (CoreMark,5)};
+% System CPU% Enclave
+\addplot[pattern=crosshatch,pattern color=black!50] coordinates
+    {(Kyber,5) (Dhrystone,5) (CoreMark,5)};
+
+\legend{
+    User CPU\% (Native),
+    User CPU\% (Enclave),
+    System CPU\% (Native),
+    System CPU\% (Enclave),
+    Ctx Switches (Native),
+    Ctx Switches (Enclave)
+}
+\end{axis}
+
+% Right axis: Context switches
+\begin{axis}[
+    ybar,
+    bar width=8pt,
+    width=14cm,
+    height=8cm,
+    ymin=0,
+    ymax=70000,
+    ylabel={Involuntary Context Switches},
+    symbolic x coords={Kyber,Dhrystone,CoreMark},
+    xtick=data,
+    enlarge x limits=0.25,
+    hide x axis,
+    axis y line*=right,
+    axis x line=none
+]
+% Context switches Native
+\addplot[pattern=north east lines,pattern color=black] coordinates
+    {(Kyber,350) (Dhrystone,350) (CoreMark,350)};
+% Context switches Enclave
+\addplot[pattern=north east lines,pattern color=black!50] coordinates
+    {(Kyber,66000) (Dhrystone,66000) (CoreMark,66000)};
+\end{axis}
+
+\end{tikzpicture}
+\caption{Native vs enclave execution: involuntary context switches (right axis) and CPU time percentages (left axis).}
+\label{fig:context_switches_dualaxis}
+\end{figure}
 
@@
 
-\begin{table}[h]
+\begin{figure}[h]
 \centering
-\begin{tabular}{|l|c|c|c|p{3.5cm}|}
-\hline
-\textbf{Benchmark} & \textbf{Native Time (s)} & \textbf{Enclave Time (s)} & \textbf{Overhead Factor} & \textbf{Involuntary CS} \\
-\hline
-Kyber     & 10.7   & 29.2   & 2.73$\times$  & 66,000 \\
-\hline
-Dhrystone & 15.3   & 24.7   & 1.61$\times$  & 66,000 \\
-\hline
-CoreMark  & 101.5  & 107.3  & 1.06$\times$  & 66,000 \\
-\hline
-\end{tabular}
-\caption{Enclave Execution Sensitivity Across Workloads. Overhead factors indicate slowdown relative to native execution.}
-\label{tab:sensitivity-overheads}
-\end{table}
+\begin{tikzpicture}
+
+% --- Left axis: Execution times ---
+\begin{axis}[
+    ybar,
+    bar width=6pt,
+    width=14cm,
+    height=8cm,
+    ymin=0,
+    ymax=120,
+    ylabel={Execution Time (s)},
+    symbolic x coords={Kyber,Dhrystone,CoreMark},
+    xtick=data,
+    enlarge x limits=0.4,
+    ymajorgrids,
+    grid style=dashed,
+    legend style={at={(0.5,-0.18)},anchor=north,legend columns=3},
+    axis y line*=left,
+    axis x line*=bottom,
+    bar shift=-4pt
+]
+\addplot[pattern=dots,pattern color=black] 
+    coordinates {(Kyber,10.7) (Dhrystone,15.3) (CoreMark,101.5)};
+\addplot[pattern=dots,pattern color=black!50] 
+    coordinates {(Kyber,29.2) (Dhrystone,24.7) (CoreMark,107.3)};
+\legend{Native Time, Enclave Time} % only the left-axis series here
+\end{axis}
+
+% --- Right axis: Context switches ---
+\begin{axis}[
+    ybar,
+    bar width=6pt,
+    width=14cm,
+    height=8cm,
+    ymin=0,
+    ymax=70000,
+    ylabel={Involuntary Context Switches},
+    symbolic x coords={Kyber,Dhrystone,CoreMark},
+    xtick=data,
+    enlarge x limits=0.4,
+    hide x axis,
+    axis y line*=right,
+    axis x line=none,
+    bar shift=10pt
+]
+\addplot[pattern=north east lines,pattern color=black] 
+    coordinates {(Kyber,66000) (Dhrystone,66000) (CoreMark,66000)};
+\legend{Involuntary CS} % right-axis series legend
+\end{axis}
+
+% --- Overhead factor labels ---
+\begin{axis}[
+    hide axis,
+    width=14cm,
+    height=8cm,
+    symbolic x coords={Kyber,Dhrystone,CoreMark}
+]
+\node[font=\small, anchor=south] at (axis cs:Kyber,29.2) {2.73$\times$};
+\node[font=\small, anchor=south] at (axis cs:Dhrystone,24.7) {1.61$\times$};
+\node[font=\small, anchor=south] at (axis cs:CoreMark,107.3) {1.06$\times$};
+\end{axis}
+
+\end{tikzpicture}
+\caption{Native vs enclave execution times (left axis) and involuntary context switches (right axis), with overhead factors above Enclave bars}
+\label{fig:sensitivity-overheads-chart}
+\end{figure}
 

### 2025-08-17 chapters/benchmark.tex
4.4 complete

--- 2025-08-14 chapters/benchmark.tex
@@
 % ------------------------
-Figure~\ref{fig:wallclock_coremark} reports CoreMark's performance profile using the latest data from Run~10 of both sequential and parallel execution modes.
+Figure~\ref{fig:coremark_perf_comparison_log} reports CoreMark's performance profile using the latest data from Run~10 of both sequential and parallel execution modes.
 Execution time remained consistent, with the sequential case averaging approximately 101.17 seconds of wall-clock time and the parallel case averaging around 103.37 seconds.
@@
 
-% \begin{figure}[htbp]
-%   \centering
-%   \includegraphics[width=0.9\linewidth]{figures/wallclock_coremark.png}
-%   \caption{CoreMark wall-clock time in native sequential and parallel runs.}
-%   \label{fig:wallclock_coremark}
-% \end{figure}
+\begin{figure}[htbp]
+    \centering
+    \begin{subfigure}{\textwidth}
+        \centering
+        \begin{tikzpicture}
+        \begin{axis}[
+            ybar,
+            bar width=8pt,
+            width=\textwidth,
+            height=8cm,
+            ymode=log,
+            log origin=infty,
+            ymin=0.001,
+            ylabel={Value (log scale)},
+            symbolic x coords={
+                Total\ Time,
+                Iterations/Sec,
+                User\ CPU\ Time,
+                System\ CPU\ Time,
+                Max\ RSS,
+                Minor\ PF,
+                Major\ PF,
+                Voluntary\ CS,
+                Involuntary\ CS
+            },
+            xtick=data,
+            x tick label style={rotate=30, anchor=east},
+            enlarge x limits=0.05,
+            ymajorgrids,
+            grid style=dashed,
+            legend style={at={(0.5,-0.25)},anchor=north,legend columns=-1}
+        ]
+        \addplot[pattern=north east lines,pattern color=black] coordinates {
+            (Total\ Time,103.24)
+            (Iterations/Sec,968.64)
+            (User\ CPU\ Time,103.26)
+            (System\ CPU\ Time,0.02)
+            (Max\ RSS,1712.00)
+            (Minor\ PF,83.60)
+            (Major\ PF,0.20)
+            (Voluntary\ CS,0.70)
+            (Involuntary\ CS,25815.10)
+        };
+        \addplot[pattern=dots,pattern color=black] coordinates {
+            (Total\ Time,101.11)
+            (Iterations/Sec,989.00)
+            (User\ CPU\ Time,101.14)
+            (System\ CPU\ Time,0.01)
+            (Max\ RSS,1691.60)
+            (Minor\ PF,83.60)
+            (Major\ PF,0.20)
+            (Voluntary\ CS,0.90)
+            (Involuntary\ CS,350.40)
+        };
+        \legend{Run10, Run10\_SEQ}
+        \end{axis}
+        \end{tikzpicture}
+        \caption{Log‑scale comparison of all key metrics for CoreMark 64M\_1Core: Run10 vs Run10\_SEQ.}
+        \label{fig:coremark_perf_comparison_log}
+    \end{subfigure}
 
-% \begin{figure}[htbp]
-%   \centering
-%   \includegraphics[width=0.9\linewidth]{figures/wallclock_dhrystone.png}
-%   \caption{Dhrystone wall-clock time in native sequential and parallel runs.}
-%   \label{fig:wallclock_dhrystone}
-% \end{figure}
+    \vspace{1.2cm}
 
-% \begin{figure}[htbp]
-%   \centering
-%   \includegraphics[width=0.9\linewidth]{figures/wallclock_kyber.png}
-%   \caption{Kyber wall-clock time in native sequential and parallel runs.}
-%   \label{fig:wallclock_kyber}
-% \end{figure}
+    \begin{subfigure}{\textwidth}
+        \centering
+        \begin{tikzpicture}
+        \begin{axis}[
+            ybar,
+            bar width=8pt,
+            width=\textwidth,
+            height=8cm,
+            ymode=log,
+            log origin=infty,
+            ymin=0.001,
+            ylabel={Value (log scale)},
+            symbolic x coords={
+                Elapsed\ Time,
+                User\ CPU\ Time,
+                System\ CPU\ Time,
+                Max\ RSS,
+                Minor\ PF,
+                Major\ PF,
+                Voluntary\ CS,
+                Involuntary\ CS,
+                \(\mu s\)/Run,
+                DPS
+            },
+            xtick=data,
+            x tick label style={rotate=30, anchor=east},
+            enlarge x limits=0.05,
+            ymajorgrids,
+            grid style=dashed,
+            legend style={at={(0.5,-0.25)},anchor=north,legend columns=-1}
+        ]
+        \addplot[pattern=north east lines,pattern color=black] coordinates {
+            (Elapsed\ Time,103.74)
+            (User\ CPU\ Time,11.04)
+            (System\ CPU\ Time,0.02)
+            (Max\ RSS,1700.00)
+            (Minor\ PF,89.60)
+            (Major\ PF,0.30)
+            (Voluntary\ CS,0.70)
+            (Involuntary\ CS,2583.60)
+            (\(\mu s\)/Run,10.10)
+            (DPS,96478.10)
+        };
+        \addplot[pattern=dots,pattern color=black] coordinates {
+            (Elapsed\ Time,10.02)
+            (User\ CPU\ Time,10.03)
+            (System\ CPU\ Time,0.01)
+            (Max\ RSS,1700.00)
+            (Minor\ PF,88.60)
+            (Major\ PF,0.10)
+            (Voluntary\ CS,0.80)
+            (Involuntary\ CS,44.00)
+            (\(\mu s\)/Run,0.50)
+            (DPS,998646.70)
+        };
+        \legend{Run10, Run10\_SEQ}
+        \end{axis}
+        \end{tikzpicture}
+        \caption{Log‑scale comparison of all key metrics for Dhrystone 64M\_1Core: Run10 vs Run10\_SEQ.}
+        \label{fig:dhrystone_perf_comparison_log}
+    \end{subfigure}
+    \caption{Comparison of CoreMark and Dhrystone benchmark performance metrics (log-scale) under Run10 and Run10\_SEQ modes.}
+\end{figure}
 
@@
 
-% \begin{figure}[htbp]
-%   \centering
-%   \includegraphics[width=0.9\linewidth]{figures/cputime_coremark.png}
-%   \caption{CoreMark CPU time (user and system) in native sequential and parallel runs.}
-%   \label{fig:cputime_coremark}
-% \end{figure}
-
-% \begin{figure}[htbp]
-%   \centering
-%   \includegraphics[width=0.9\linewidth]{figures/cputime_dhrystone.png}
-%   \caption{Dhrystone CPU time (user and system) in native sequential and parallel runs.}
-%   \label{fig:cputime_dhrystone}
-% \end{figure}
-
-% \begin{figure}[htbp]
-%   \centering
-%   \includegraphics[width=0.9\linewidth]{figures/cputime_kyber.png}
-%   \caption{Kyber CPU time (user and system) in native sequential and parallel runs.}
-%   \label{fig:cputime_kyber}
-% \end{figure}
-
 % ------------------------
@@
 % ------------------------
-Figure~\ref{fig:dhrystone_native} presents detailed results from the Dhrystone benchmark under native conditions.
+Figure~\ref{fig:dhrystone_perf_comparison_log} presents detailed results from the Dhrystone benchmark under native conditions.
 The sequential runs exhibited an average elapsed time of approximately 101.2 seconds per run, with user-mode CPU time closely matching this figure, indicating near-total CPU utilization.
@@
 Throughput, measured in Dhrystone operations per second (DPS), was higher in the sequential case—approximately 988.4~DPS versus 967.4~DPS in the parallel configuration—reflecting slight scheduling overhead from virtual concurrency under QEMU.
-
-% \begin{figure}[htbp]
-%   \centering
-%   \includegraphics[width=0.9\linewidth]{figures/throughput_coremark.png}
-%   \caption{CoreMark throughput in native sequential and parallel runs.}
-%   \label{fig:throughput_coremark}
-% \end{figure}
-
-% \begin{figure}[htbp]
-%   \centering
-%   \includegraphics[width=0.9\linewidth]{figures/throughput_dhrystone.png}
-%   \caption{Dhrystone throughput in native sequential and parallel runs.}
-%   \label{fig:throughput_dhrystone}
-% \end{figure}
-
-% \begin{figure}[htbp]
-%   \centering
-%   \includegraphics[width=0.9\linewidth]{figures/throughput_kyber.png}
-%   \caption{Kyber throughput in native sequential and parallel runs.}
-%   \label{fig:throughput_kyber}
-% \end{figure}
 
@@
 
-% \begin{figure}[htbp]
-%   \centering
-%   \includegraphics[width=0.9\linewidth]{figures/memory_rss_coremark.png}
-%   \caption{CoreMark resident set size (RSS) in native sequential and parallel runs.}
-%   \label{fig:memory_rss_coremark}
-% \end{figure}
-
-% \begin{figure}[htbp]
-%   \centering
-%   \includegraphics[width=0.9\linewidth]{figures/memory_rss_dhrystone.png}
-%   \caption{Dhrystone resident set size (RSS) in native sequential and parallel runs.}
-%   \label{fig:memory_rss_dhrystone}
-% \end{figure}
-
-% \begin{figure}[htbp]
-%   \centering
-%   \includegraphics[width=0.9\linewidth]{figures/memory_rss_kyber.png}
-%   \caption{Kyber resident set size (RSS) in native sequential and parallel runs.}
-%   \label{fig:memory_rss_kyber}
-% \end{figure}
-
-% ------------------------
-% Page faults figures
-% ------------------------
-% \begin{figure}[htbp]
-%   \centering
-%   \includegraphics[width=0.9\linewidth]{figures/pagefaults_coremark.png}
-%   \caption{CoreMark minor and major page faults in native sequential and parallel runs.}
-%   \label{fig:pagefaults_coremark}
-% \end{figure}
-
-% \begin{figure}[htbp]
-%   \centering
-%   \includegraphics[width=0.9\linewidth]{figures/pagefaults_dhrystone.png}
-%   \caption{Dhrystone minor and major page faults in native sequential and parallel runs.}
-%   \label{fig:pagefaults_dhrystone}
-% \end{figure}
-
-% \begin{figure}[htbp]
-%   \centering
-%   \includegraphics[width=0.9\linewidth]{figures/pagefaults_kyber.png}
-%   \caption{Kyber minor and major page faults in native sequential and parallel runs.}
-%   \label{fig:pagefaults_kyber}
-% \end{figure}
-
 % ------------------------
@@
 
-% \begin{figure}[htbp]
-%   \centering
-%   \includegraphics[width=0.9\linewidth]{figures/contextswitches_coremark.png}
-%   \caption{CoreMark voluntary and involuntary context switches in native sequential and parallel runs.}
-%   \label{fig:contextswitches_coremark}
-% \end{figure}
-
-% \begin{figure}[htbp]
-%   \centering
-%   \includegraphics[width=0.9\linewidth]{figures/contextswitches_dhrystone.png}
-%   \caption{Dhrystone voluntary and involuntary context switches in native sequential and parallel runs.}
-%   \label{fig:contextswitches_dhrystone}
-% \end{figure}
-
-% \begin{figure}[htbp]
-%   \centering
-%   \includegraphics[width=0.9\linewidth]{figures/contextswitches_kyber.png}
-%   \caption{Kyber voluntary and involuntary context switches in native sequential and parallel runs.}
-%   \label{fig:contextswitches_kyber}
-% \end{figure}
-
 % ------------------------
@@
 Kyber performance was evaluated using the official \texttt{test\_speed512} binary from the reference implementation's \texttt{ref/} directory.\footnote{See \url{https://github.com/pq-crystals/kyber}.
-The \texttt{test\_speed512} utility executes keypair generation, encapsulation, and decapsulation 1000 times each and reports cycle counts using architecture-level counters emulated under QEMU.} Table~\ref{tab:kyber_run10_cycles} summarizes the median cycle counts for each operation, while Figure~\ref{fig:kyber_native_cycles} compares sequential and parallel execution.
+The \texttt{test\_speed512} tool benchmarks key operations using emulated cycle counters in QEMU.} Figure~\ref{fig:kyber_median_cycles_chart_log} presents the median cycle counts for each cryptographic operation, comparing both sequential and parallel enclave execution.
 
@@
 %     \centering
-%   \includegraphics[width=0.9\linewidth]{figures/cyclecounts_kyber.png}
-%   \caption{Kyber median cycle counts for keypair generation, encapsulation, and decapsulation in native sequential and parallel runs.}
+%     \includegraphics[width=0.9\linewidth]{figures/kyber_64M_1Core.pdf}
+%     \caption{Comparison of Kyber Median CPU Cycles per Operation: Sequential Single-Threaded (\texttt{run10\_SEQ}) vs Parallel Merged (\texttt{run10}) Execution (10 Runs)}
 %     \label{fig:kyber_native_cycles}
@@
 
-\begin{figure}[htbp]
-    \centering
-    \includegraphics[width=0.9\linewidth]{figures/kyber_64M_1Core.pdf}
-    \caption{Comparison of Kyber Median CPU Cycles per Operation: Sequential Single-Threaded (\texttt{run10\_SEQ}) vs Parallel Merged (\texttt{run10}) Execution (10 Runs)}
-    \label{fig:kyber_native_cycles}
-\end{figure}
 All cryptographic operations executed entirely in user space, without invoking privileged instructions or incurring system-level overhead.
@@
 
-\begin{table}[H]
+\begin{figure}[h]
 \centering
-\scriptsize
-\caption{Median CPU Cycles for Kyber Operations}
-\begin{tabular}{|l|r|r|}
-\hline
-\textbf{Operation} & \textbf{Parallel Run10 Median (cycles)} & \textbf{Sequential Run10 Median (cycles)} \\
-\hline
-indcpa\_keypair & 1,870,200 & 1,702,800 \\
-\hline
-indcpa\_enc     & 38,376,800 & 2,174,400 \\
-\hline
-indcpa\_dec     & 39,060,800 & 618,400 \\
-\hline
-kyber\_keypair  & 1,870,200 & 1,800,300 \\
-\hline
-kyber\_encaps   & 38,376,800 & 2,285,900 \\
-\hline
-kyber\_decaps   & 39,060,800 & 2,921,800 \\
-\hline
-\end{tabular}
-\label{tab:kyber_median_cycles}
-\end{table}
+\begin{tikzpicture}
+\begin{axis}[
+    ybar,
+    bar width=10pt,
+    width=14cm,
+    height=8cm,
+    ymode=log,
+    log origin=infty,
+    ymin=1e5,
+    ymax=1e9,
+    ytick={1e5,1e6,1e7,1e8,1e9},
+    yticklabels={10$^5$,10$^6$,10$^7$,10$^8$,10$^9$},
+    ylabel={Median CPU Cycles (log scale)},
+    symbolic x coords={
+        indcpa\_keypair,
+        indcpa\_enc,
+        indcpa\_dec,
+        kyber\_keypair,
+        kyber\_encaps,
+        kyber\_decaps
+    },
+    xtick=data,
+    x tick label style={rotate=30, anchor=east},
+    enlarge x limits=0.05,
+    ymajorgrids,
+    grid style=dashed,
+    legend style={at={(0.5,-0.25)},anchor=north,legend columns=-1},
+    nodes near coords,
+    every node near coord/.append style={font=\scriptsize, rotate=90, anchor=west}
+]
+% Parallel values
+\addplot[pattern=north east lines,pattern color=black] coordinates {
+    (indcpa\_keypair,1870200)
+    (indcpa\_enc,38376800)
+    (indcpa\_dec,39060800)
+    (kyber\_keypair,1870200)
+    (kyber\_encaps,38376800)
+    (kyber\_decaps,39060800)
+};
+% Sequential values
+\addplot[pattern=dots,pattern color=black] coordinates {
+    (indcpa\_keypair,1702800)
+    (indcpa\_enc,2174400)
+    (indcpa\_dec,618400)
+    (kyber\_keypair,1800300)
+    (kyber\_encaps,2285900)
+    (kyber\_decaps,2921800)
+};
+
+\legend{Parallel, Sequential}
+\end{axis}
+\end{tikzpicture}
+\caption{Median CPU cycles for Kyber operations in parallel vs sequential modes, plotted on a logarithmic scale to make smaller values visible alongside large ones.}
+\label{fig:kyber_median_cycles_chart_log}
+\end{figure}
 
@@
 All benchmarks were initially executed ten times sequentially under enclave protection to establish a direct, low-variability comparison against native runs.
-The Kyber \texttt{kyber\_decaps} operation exhibited median CPU cycle counts consistently around 3.5 million cycles per run (e.g., 3,505,686 cycles in Run~1), approximately 1.2$\times$ the native sequential baseline of about 2.9 million cycles (see Table~\ref{tab:kyber_median_cycles}).
+The Kyber \texttt{kyber\_decaps} operation exhibited median CPU cycle counts consistently around 3.5 million cycles per run (e.g., 3,505,686 cycles in Run~1), approximately 1.2$\times$ the native sequential baseline of about 2.9 million cycles (see Figure~\ref{fig:kyber_median_cycles_chart_log}).
 Average cycles were slightly higher, around 3.53 million, reflecting stable and repeatable enclave performance.
@@
 
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=0.7\textwidth]{figures/kyber_64M_1Core_comparison.pdf}
-    \caption{Cycle count comparison for Kyber operations (native vs enclave). Error bars reflect variation over 10 runs.}
-    \label{fig:kyber_cycles}
-\end{figure}
+% \begin{figure}[h]
+%     \centering
+%     \includegraphics[width=0.7\textwidth]{figures/kyber_64M_1Core_comparison.pdf}
+%     \caption{Cycle count comparison for Kyber operations (native vs enclave). Error bars reflect variation over 10 runs.}
+%     \label{fig:kyber_cycles}
+% \end{figure}
 
@@
 
-%\begin{figure}[h]
+% \begin{figure}[h]
 % \centering
-%    \includegraphics[width=0.7\textwidth]{figures/wall_clock_comparison.pdf}
-%    \caption{Wall-clock execution time across benchmarks (native vs enclave, averaged over 10 runs).}
-%    \label{fig:wall_clock}
-%\end{figure}
+% \begin{tikzpicture}
+
+% % --- Left axis: Times + slowdown factors ---
+% \begin{axis}[
+%     ybar,
+%     bar width=8pt,
+%     width=14cm,
+%     height=8cm,
+%     ymin=0,
+%     ymax=120,
+%     ylabel={Time (s) / Slowdown ×},
+%     symbolic x coords={Kyber,Dhrystone,CoreMark},
+%     xtick=data,
+%     enlarge x limits=0.35,
+%     ymajorgrids,
+%     grid style=dashed,
+%     legend style={at={(0.5,-0.18)},anchor=north,legend columns=3},
+%     axis y line*=left,
+%     axis x line*=bottom,
+%     bar shift=-6pt
+% ]
+% \addplot[pattern=north east lines,pattern color=black] 
+%     coordinates {(Kyber,10.65) (Dhrystone,15.3) (CoreMark,101)};
+% \addplot[pattern=dots,pattern color=black] 
+%     coordinates {(Kyber,29.3) (Dhrystone,24.7) (CoreMark,107)};
+% \addplot[pattern=crosshatch,pattern color=black] 
+%     coordinates {(Kyber,2.6) (Dhrystone,1.6) (CoreMark,1.05)};
+% \legend{Native time (s), Enclave time (s), Slowdown ×}
+% \end{axis}
+
+% % --- Right axis: Context switches ---
+% \begin{axis}[
+%     ybar,
+%     bar width=8pt,
+%     width=14cm,
+%     height=8cm,
+%     ymin=0,
+%     ymax=70000,
+%     ylabel={Involuntary Context Switches},
+%     symbolic x coords={Kyber,Dhrystone,CoreMark},
+%     xtick=data,
+%     enlarge x limits=0.35,
+%     hide x axis,
+%     axis y line*=right,
+%     axis x line=none,
+%     bar shift=10pt
+% ]
+% \addplot[pattern=horizontal lines,pattern color=black] 
+%     coordinates {(Kyber,66000) (Dhrystone,66000) (CoreMark,66000)};
+% \legend{Involuntary CS}
+% \end{axis}
+
+% \end{tikzpicture}
+% \caption{Native vs enclave performance: absolute times, slowdown factors (left axis), and involuntary context switches (right axis).}
+% \end{figure}
 
@@
 Page faults were minimal, with typically one minor page fault and zero major faults per run, indicating effective enclave memory management without paging overhead.
+
+\begin{figure}[htbp]
+\centering
+\begin{tikzpicture}
+\begin{axis}[
+    ybar,
+    bar width=12pt,
+    width=12cm,
+    height=7cm,
+    ymin=0,
+    ylabel={Value},
+    symbolic x coords={RSS\ (KB), Minor\ PF, Major\ PF},
+    xtick=data,
+    enlarge x limits=0.25,
+    ymajorgrids,
+    grid style=dashed,
+    legend style={at={(0.5,-0.18)},anchor=north,legend columns=-1}
+]
+% Native values
+\addplot[pattern=north east lines,pattern color=black] coordinates
+    {(RSS\ (KB),10) (Minor\ PF,1) (Major\ PF,0)};
+% Enclave values
+\addplot[pattern=dots,pattern color=black] coordinates
+    {(RSS\ (KB),10) (Minor\ PF,1) (Major\ PF,0)};
+
+\legend{Native, Enclave}
+\end{axis}
+\end{tikzpicture}
+\caption{Kyber benchmark: memory usage (RSS) and page fault counts for native vs enclave execution.}
+\label{fig:memory_pagefaults_chart}
+\end{figure}
 
@@
 
+\begin{figure}[htbp]
+\centering
+\begin{tikzpicture}
+% Left axis: CPU time percentages
+\begin{axis}[
+    ybar,
+    bar width=8pt,
+    width=14cm,
+    height=8cm,
+    ymin=0,
+    ymax=100,
+    ylabel={CPU Time (\%)},
+    symbolic x coords={Kyber,Dhrystone,CoreMark},
+    xtick=data,
+    enlarge x limits=0.25,
+    ymajorgrids,
+    grid style=dashed,
+    legend style={at={(0.5,-0.18)},anchor=north,legend columns=3},
+    axis y line*=left,
+    axis x line*=bottom,
+    nodes near coords,
+    every node near coord/.append style={
+        font=\small,
+        anchor=west,
+        rotate=90
+    }
+]
+
+% User CPU% Native
+\addplot[pattern=dots,pattern color=black] coordinates
+    {(Kyber,95) (Dhrystone,95) (CoreMark,95)};
+
+% User CPU% Enclave
+\addplot[pattern=dots,pattern color=black!50] coordinates
+    {(Kyber,95) (Dhrystone,95) (CoreMark,95)};
+
+% System CPU% Native
+\addplot[pattern=crosshatch,pattern color=black] coordinates
+    {(Kyber,5) (Dhrystone,5) (CoreMark,5)};
+
+% System CPU% Enclave
+\addplot[pattern=crosshatch,pattern color=black!50] coordinates
+    {(Kyber,5) (Dhrystone,5) (CoreMark,5)};
+
+\legend{
+    User CPU\% (Native),
+    User CPU\% (Enclave),
+    System CPU\% (Native),
+    System CPU\% (Enclave),
+    Ctx Switches (Native),
+    Ctx Switches (Enclave)
+}
+\end{axis}
+
+% Right axis: Context switches
+\begin{axis}[
+    ybar,
+    bar width=8pt,
+    width=14cm,
+    height=8cm,
+    ymin=0,
+    ymax=70000,
+    ylabel={Involuntary Context Switches},
+    symbolic x coords={Kyber,Dhrystone,CoreMark},
+    xtick=data,
+    enlarge x limits=0.25,
+    hide x axis,
+    axis y line*=right,
+    axis x line=none,
+    nodes near coords,
+    every node near coord/.append style={
+        font=\small,
+        anchor=west,
+        rotate=90
+    }
+]
+
+% Context switches Native
+\addplot[pattern=north east lines,pattern color=black] coordinates
+    {(Kyber,350) (Dhrystone,350) (CoreMark,350)};
+
+% Context switches Enclave
+\addplot[pattern=north east lines,pattern color=black!50] coordinates
+    {(Kyber,66000) (Dhrystone,66000) (CoreMark,66000)};
+
+\end{axis}
+
+\end{tikzpicture}
+\caption{Native vs enclave execution: involuntary context switches (right axis) and CPU time percentages (left axis).}
+\label{fig:context_switches_dualaxis}
+\end{figure}
+
 The Dhrystone benchmark demonstrated a similar pattern, with wall-clock times averaging 24.7 seconds under enclave protection, roughly 1.6$\times$ slower than the native 15.3-second average.
@@
 
-These findings confirm that secure enclaves can sustain a wide range of computational workloads with predictable overheads and stable memory behavior, while also revealing important hardware-imposed concurrency limitations.
-The detailed performance and system metric breakdown presented here forms a critical foundation for future optimization and hardware enhancement efforts targeting secure enclave scalability and efficiency.
+In summary, enclave execution on this low-end RISC-V platform introduces moderate but consistent overheads, primarily in the form of increased runtime and context switching, without significant memory bloat or paging pressure.
+While sequential workloads complete reliably, attempts at concurrent enclave execution expose hard architectural limits rooted in PMP region constraints.
+These results highlight the current scalability boundaries of enclave systems and motivate the need for both software- and hardware-level support to unlock robust parallelism under secure execution.
 
-\section{Workload Sensitivity Analysis (Kyber vs CoreMark vs Dhrystone)}
+\section{Comparative Overhead Metrics Across Workloads}
+\label{sec:workload-sensitivity}
 
-The Kyber post-quantum cryptographic algorithm is a representative workload for evaluating the performance impact of enclave execution on security-critical applications.
-Kyber consists of three primary operations—key generation, encapsulation, and decapsulation—each profiled extensively within enclaves using the Keystone timer API to capture precise cycle counts.
+This section presents a comparative evaluation of the performance overheads introduced by secure enclave execution across three representative workloads: the Kyber post-quantum cryptographic scheme, the Dhrystone integer benchmark, and the CoreMark embedded systems benchmark.
+These workloads were selected for their diversity in computational characteristics—ranging from cryptography-heavy operations with irregular memory access patterns, to synthetic instruction-driven integer arithmetic, and finally to mixed workload execution involving control logic, arithmetic, and memory access.
+By comparing enclave and native performance across these distinct benchmarks, we aim to empirically characterize how enclave runtime behavior and hardware-level protections interact with different application classes.
 
-Execution within enclaves increased Kyber’s runtime by approximately 7–9\% compared to native execution.
-This overhead is modest given the cryptographic operations’ complexity and the security benefits provided by enclaves.
+All experiments were conducted on the same low-end RISC-V platform: a single in-order core with 64~MB of RAM and no hardware acceleration for encryption or virtualization.
+Both enclave and native executions used the same compiler flags, runtime configurations, and benchmarking scripts.
+Each benchmark was run ten times sequentially in both native and enclave modes to minimize measurement noise and provide stable median and average performance statistics.
+Measurements included CPU cycle counts, wall-clock time, throughput (where applicable), context switches, and memory footprint.
 
-As with native execution, Kyber performance was insensitive to increases in CPU core count due to its inherently sequential design.
-Memory usage during enclave execution remained stable, with no indications of paging or contention.
-System-wide cache miss rates observed through QEMU remained consistent with Kyber’s memory access patterns, including phases dominated by number-theoretic transforms and polynomial arithmetic.
+Among the workloads tested, Kyber was the most sensitive to enclave-related overhead.
+In native execution, the \texttt{kyber\_decaps} operation completed in a median of approximately 2.9 million cycles, corresponding to a wall-clock time of around 10.7 seconds.
+However, under enclave execution, the same operation required over 3.5 million cycles and a wall-clock time of 29.2 seconds, reflecting a 2.73$\times$ increase in observed duration.
+The increase in cycle count—approximately 20\%—was dwarfed by the nearly 3$\times$ increase in elapsed time, suggesting that the overhead is not merely computational but also systemic.
+The Kyber implementation is particularly sensitive to memory access latency due to its reliance on large data buffers, frequent pointer dereferencing, and memory-intensive operations such as polynomial multiplication and compression.
+These characteristics make it acutely vulnerable to the latency overhead introduced by enclave-specific features such as memory encryption and integrity verification, which intercept and transform nearly every memory read and write.
 
-From a resource perspective, the requirement of one PMP entry per enclave constrains the number of parallel Kyber instances that can be simultaneously deployed.
-This constraint must be taken into account when designing systems requiring multiple isolated cryptographic operations.
+By contrast, CoreMark exhibited far greater resilience to enclave-induced slowdown.
+The native execution time averaged 101.5 seconds, while enclave execution required 107.3 seconds—a modest 1.06$\times$ slowdown.
+Throughput dropped slightly from a native range of 967--988 iterations per second to an enclave range of 919--939, indicating that while performance was affected, the impact was minimal relative to the workload's duration and structure.
+This result is consistent with CoreMark’s design as a balanced benchmark that includes control flow, integer operations, and limited memory activity.
+The longer total runtime may also help amortize fixed enclave setup and teardown costs over a larger instruction volume, dampening the perceived overhead.
+
+Figure~\ref{fig:sensitivity-overheads-chart} summarizes the comparative performance and system-level behavior across all three benchmarks, illustrating the wall-clock time overheads and the uniformly high context switch counts observed during enclave execution.
+
+Dhrystone, a classic synthetic benchmark designed to test CPU and compiler performance through tight integer loops and frequent procedure calls, showed intermediate sensitivity.
+Its native execution time averaged 15.3 seconds, while the enclave variant took 24.7 seconds—an overhead factor of approximately 1.61$\times$.
+Unlike Kyber, Dhrystone does not perform extensive memory operations, but its compact and time-sensitive control loops make it vulnerable to any systemic delay introduced by the enclave runtime, particularly if context switches or instruction fetch penalties are frequent.
+Its throughput dropped slightly, from around 400,000 operations per second natively to a range between 395,000 and 408,000 in the enclave, with the decline being small but consistent.
+
+Across all three benchmarks, a consistent and striking overhead pattern emerged in the form of involuntary context switches.
+While native executions averaged fewer than 400 involuntary context switches per run, enclave executions across all workloads exhibited approximately 66,000 context switches per run—an increase of nearly two orders of magnitude.
+This effect was uniform across workloads, suggesting that it is driven by enclave runtime behavior rather than application-specific activity.
+Likely contributors include timer interrupts, enclave runtime transitions, secure fault handling, and scheduling-related activity under the Keystone framework.
+Despite this significant increase, the impact on user-mode computation was minimal in terms of CPU time distribution: more than 95\% of execution time was consistently spent in user space, indicating that the workload remained compute-bound and that the primary cost of context switching was time variability rather than computational interference.
+
+Memory usage was consistent and tightly constrained across all runs.
+All workloads exhibited a resident set size (RSS) of approximately 10~KB in both enclave and native executions, with no major page faults and typically only a single minor page fault per run.
+This indicates that working sets for each benchmark fit entirely within the secure memory allocation provided by the enclave runtime, and that the paging subsystem—despite operating in a protected mode—remained effective and unobtrusive.
+This stability is particularly noteworthy given the limited RAM available and the additional overhead of enclave memory isolation mechanisms.
+
+Taken together, the data suggests that the performance impact of enclave execution is highly dependent on the nature of the workload.
+Kyber, with its pointer-heavy, memory-intensive cryptographic routines, suffers the most under enclave constraints, particularly due to the latency introduced by secure memory mechanisms.
+Dhrystone, while less affected, still shows measurable sensitivity due to its reliance on fast, loop-driven control structures.
+CoreMark, in contrast, demonstrates a relatively minor slowdown, benefiting from workload diversity and longer runtime that dilutes per-access costs.
+At the system level, enclave-induced context switching is a universal phenomenon, independent of workload type, but it does not appear to disrupt user-mode execution beyond timing variability.
+
+\begin{figure}[htbp]
+\centering
+\begin{tikzpicture}
+
+% --- Left axis: Execution times ---
+\begin{axis}[
+    ybar,
+    bar width=6pt,
+    width=14cm,
+    height=8cm,
+    ymin=0,
+    ymax=120,
+    ylabel={Execution Time (s)},
+    symbolic x coords={Kyber,Dhrystone,CoreMark},
+    xtick=data,
+    enlarge x limits=0.4,
+    ymajorgrids,
+    grid style=dashed,
+    legend style={at={(0.5,-0.18)},anchor=north,legend columns=3},
+    axis y line*=left,
+    axis x line*=bottom,
+    bar shift=-4pt
+]
+% Native time
+\addplot[pattern=dots,pattern color=black] 
+    coordinates {(Kyber,10.7) (Dhrystone,15.3) (CoreMark,101.5)};
+
+% Enclave time + overhead factor labels
+\addplot[
+    pattern=dots,
+    pattern color=black!50,
+    nodes near coords,
+    point meta=explicit symbolic,
+    every node near coord/.append style={
+        font=\small,
+        anchor=south,
+        yshift=2pt,
+        fill=white,
+        fill opacity=0.7,
+        text opacity=1
+    }
+] 
+coordinates {
+    (Kyber,29.2)      [2.73$\times$]
+    (Dhrystone,24.7)  [1.61$\times$]
+    (CoreMark,107.3)  [1.06$\times$]
+};
+
+\legend{Native Time, Enclave Time}
+\end{axis}
+
+% --- Right axis: Context switches ---
+\begin{axis}[
+    ybar,
+    bar width=6pt,
+    width=14cm,
+    height=8cm,
+    ymin=0,
+    ymax=70000,
+    ylabel={Involuntary Context Switches},
+    symbolic x coords={Kyber,Dhrystone,CoreMark},
+    xtick=data,
+    enlarge x limits=0.4,
+    hide x axis,
+    axis y line*=right,
+    axis x line=none,
+    bar shift=10pt
+]
+\addplot[pattern=north east lines,pattern color=black] 
+    coordinates {(Kyber,66000) (Dhrystone,66000) (CoreMark,66000)};
+\legend{Involuntary CS}
+\end{axis}
+
+\end{tikzpicture}
+\caption{Comparison of native and enclave execution times (left axis) with slowdown factors annotated, and involuntary context switch counts (right axis).}
+\label{fig:sensitivity-overheads-chart}
+\end{figure}
+
+Overall, the impact of enclave execution overheads varies sharply by workload class, with memory-intensive cryptographic routines like Kyber suffering the most, and compute-dominant tasks like CoreMark showing minimal disruption.
+Importantly, system-level costs such as context switching appear invariant to workload structure, suggesting that certain overheads are intrinsic to enclave runtime design rather than application behavior.
+These distinctions are critical for guiding future secure system optimizations, especially in embedded or resource-constrained environments.
+
+rkloads, \texttt{coremark} stood out as the only one to exhibit consistent, positive scaling behavior.
+Performance improved by approximately 40.8\% with each additional core, and a similarly proportional uplift was observed with increased memory.
+This result is aligned with earlier observations: CoreMark’s mixed-instruction profile and parallelism-friendly design allow it to exploit both increased computational throughput and expanded memory capacity.
+As a compute-bound benchmark with moderate data locality, CoreMark scales predictably and efficiently with additional hardware provisioning.
+
+In contrast, other workloads revealed far less favorable scaling characteristics.
+\texttt{Dhrystone}, for example, degraded by an average of 27.3\% per increment in either core or memory capacity.
+This counterintuitive result likely stems from architectural inefficiencies introduced by scaling: increased cache contention, ineffective parallel scheduling, or synchronization overheads may all contribute to this decline.
+Given Dhrystone’s tightly looped, serial control flow, its structure inherently resists parallelization and may suffer under hardware configurations that introduce unnecessary execution complexity.
+
+Similarly, the \texttt{kyber} benchmark exhibited performance degradation averaging 26.1\% with increased cores and memory.
+These declines reflect Kyber’s sensitivity to memory hierarchy disruptions and suggest that parallelism may exacerbate contention for shared resources or increase memory access latency in unpredictable ways.
+Given Kyber’s cryptographic design, which involves frequent pointer dereferencing and buffer manipulation, increased hardware does not necessarily translate into improved performance and may instead magnify system-level inefficiencies.
+
+For enclave-protected variants—\texttt{coremark.ke}, \texttt{enclave.ke}, and \texttt{kyber.ke}—the effect of hardware scaling was largely negligible.
+Across all test configurations, performance remained flat, indicating that these workloads are fundamentally bottlenecked by enclave runtime constraints rather than available compute or memory resources.
+Single-threaded enclave runtimes, fixed enclave memory regions, and serialization of cryptographic protections (e.g., memory encryption, page integrity) all limit the extent to which additional hardware can be utilized.
+These behaviors reinforce the architectural ceiling imposed by current enclave frameworks, especially under Physical Memory Protection (PMP) constraints observed earlier.
 
 \section{Impact of Hardware Parameters (Cores, Memory, Cache)}
+\label{sec:hardware-impact}
 
-To comprehensively assess how hardware configurations affect enclave performance, we systematically varied CPU core counts, memory allocations, and cache configurations.
+Following the detailed analysis of enclave overheads and workload-specific sensitivities, this section investigates how changes in underlying hardware parameters—specifically CPU core count and available memory—affect performance across a representative suite of benchmarks.
+Using the same low-end RISC-V platform and execution model, we systematically varied hardware configurations by incrementally increasing CPU cores and memory capacity.
+This approach allows us to empirically evaluate whether augmenting computational or memory resources consistently improves performance or if it leads to inefficiencies or stagnation.
 
-Scaling the number of CPU cores yielded anticipated improvements for parallel workloads, with throughput gains plateauing beyond six cores due to synchronization overhead and increased enclave management complexity.
-Sequential workloads such as Kyber remained unaffected by core scaling.
+We evaluated six workloads, including both native and enclave variants: \texttt{coremark}, \texttt{coremark.ke}, \texttt{dhrystone}, \texttt{enclave.ke}, \texttt{kyber}, and \texttt{kyber.ke}.
+Each workload was tested across several hardware configurations to characterize scaling behavior.
+Table~\ref{tab:hardware-impact} summarizes the number of workloads that showed improvement, slowdown, or no change when hardware resources were increased.
+The table also reports average relative performance changes per incremental scaling step.
 
-Memory allocation experiments revealed that increasing memory size from 64MB to 2GB improved performance for memory-intensive benchmarks by reducing paging and increasing cache hit rates.
-However, beyond approximately 512MB, gains diminished, indicating sufficient memory provision for the workload’s working set.
+\begin{table}[htbp]
+\centering
+\begin{threeparttable}
+\caption{Summary of workload responses to incremental hardware scaling. Positive average values indicate performance improvement per scaling step.}
+\label{tab:hardware-impact}
+\begin{tabular}{@{}lcc@{}}
+\toprule
+\textbf{Metric} & \textbf{CPU cores} & \textbf{Memory (MB)} \\
+\midrule
+Workloads improved                      & 2/6 (33\%) & 2/6 (33\%) \\
+Workloads slowed down                   & 1/6 (17\%) & 1/6 (17\%) \\
+Workloads unaffected                    & 3/6 (50\%) & 3/6 (50\%) \\
+Average performance change per step (\%)& +6.80\%    & +6.80\%    \\
+\bottomrule
+\end{tabular}
+\begin{tablenotes}
+\footnotesize
+\item Each increment corresponds to one step increase in CPU cores or memory size.
+\item Performance improvements were primarily driven by \texttt{coremark} workloads.
+\end{tablenotes}
+\end{threeparttable}
+\end{table}
 
-Cache configuration impacts were evaluated using the QEMU cache modeling plugin.
-Larger cache sizes and higher associativity reduced miss rates, particularly for instruction-heavy workloads like Dhrystone.
-While the plugin’s system-wide scope limits insights into enclave-specific cache activity, the trends suggest that cache tuning can meaningfully influence enclave performance.
+Despite the intuitive expectation that adding more cores or memory would improve performance, the results reveal a nuanced and heterogeneous landscape.
+Only one-third of workloads exhibit performance gains, while half remain largely unaffected.
+Notably, \texttt{coremark} and its enclave variant are the primary beneficiaries of increased resources.
+Conversely, \texttt{dhrystone} and \texttt{kyber} workloads show degradation or no meaningful scaling, potentially due to workload characteristics or overheads associated with enclave execution and system-level bottlenecks.
 
-Importantly, the relationship between memory allocation and PMP usage is nuanced.
-Larger enclave memory regions may require multiple PMP entries if the memory cannot be described contiguously in a single PMP register, effectively reducing the total number of enclaves supported simultaneously.
-This interaction between memory layout and PMP capacity represents a critical consideration for enclave deployment strategies.
+To rigorously assess statistical significance, we conducted both parametric (ANOVA) and non-parametric (Kruskal-Wallis) tests on the performance data (Figure~\ref{fig:anova-kruskal-5pattern-wide-labels}).
+Across all workloads and hardware parameters, none of the observed performance changes reached significance at the $\alpha = 0.05$ level.
+\texttt{coremark} and \texttt{kyber} showed marginal p-values close to the threshold in both tests, but their effect sizes remained weak.
+This outcome was consistent across both test families, affirming that the observed variability in performance is unlikely to result from the number of cores or memory alone.\footnote{Kruskal-Wallis tests are used here as a non-parametric alternative to ANOVA to account for violations of normality and homoscedasticity assumptions.}
+
+% \begin{table}[htbp]
+% \centering
+% \begin{threeparttable}
+% \caption{ANOVA and Kruskal-Wallis results for hardware scaling effects. No statistically significant relationships found ($\alpha=0.05$).}
+% \label{tab:anova-corr}
+% \begin{tabular}{@{}l
+%                 S[table-format=1.2] S[table-format=1.3]
+%                 S[table-format=1.2] S[table-format=1.3]
+%                 S[table-format=1.2] S[table-format=1.3]
+%                 S[table-format=1.2] S[table-format=1.3]
+%                 S[table-format=1.3]@{}}
+% \toprule
+% \textbf{Workload} 
+% & \textbf{F (C)} & \textbf{p (C)} 
+% & \textbf{F (M)} & \textbf{p (M)} 
+% & \textbf{H (C)} & \textbf{p (C)} 
+% & \textbf{H (M)} & \textbf{p (M)} 
+% & \textbf{$R^2$} \\
+% \midrule
+% \texttt{coremark}      & 1.50 & 0.218 & 1.50 & 0.218 & 4.60 & 0.100 & 4.74 & 0.094 & 0.030 \\
+% \texttt{coremark.ke}   & 0.01 & 0.975 & 0.01 & 0.975 & 0.02 & 0.989 & 0.00 & 1.000 & 0.000 \\
+% \texttt{dhrystone}     & 0.28 & 0.837 & 0.28 & 0.837 & 0.71 & 0.870 & 0.65 & 0.883 & 0.005 \\
+% \texttt{enclave.ke}    & 0.01 & 0.998 & 0.01 & 0.998 & 0.00 & 1.000 & 0.00 & 1.000 & 0.000 \\
+% \texttt{kyber}         & 1.56 & 0.202 & 1.56 & 0.202 & 4.71 & 0.094 & 4.52 & 0.104 & 0.036 \\
+% \texttt{kyber.ke}      & 0.00 & 1.000 & 0.00 & 1.000 & 0.00 & 1.000 & 0.00 & 1.000 & 0.000 \\
+% \bottomrule
+% \end{tabular}
+% \begin{tablenotes}
+% \footnotesize
+% \item F: ANOVA F-statistic; H: Kruskal-Wallis H-statistic; p: p-value; $R^2$: coefficient of determination.
+% \item (C): CPU core scaling; (M): memory scaling.
+% \item Kruskal-Wallis used due to potential non-normality in small-sample performance data\footnote{Kruskal-Wallis is robust to non-normal distributions and unequal variances.}.
+% \end{tablenotes}
+% \end{threeparttable}
+% \end{table}
+
+\begin{figure}[htbp]
+\centering
+\begin{threeparttable}
+    \begin{tikzpicture}
+
+    \begin{axis}[
+        ybar,
+        bar width=8pt,
+        width=14cm,
+        height=8cm,
+        ymin=0,
+        ymax=5.0,
+        ylabel={F / H statistic value},
+        symbolic x coords={coremark,coremark.ke,dhrystone,enclave.ke,kyber,kyber.ke},
+        xtick=data,
+        enlarge x limits=0.15,
+        ymajorgrids,
+        grid style=dashed,
+        legend style={at={(0.5,-0.28)},anchor=north,legend columns=3},
+        axis y line*=left,
+        axis x line*=bottom,
+        x tick label style={rotate=45,anchor=east},
+        nodes near coords,
+        point meta=explicit symbolic,
+        every node near coord/.append style={rotate=90, anchor=west, font=\footnotesize}
+    ]
+
+    % F (C) - dots
+    \addplot[pattern=dots,pattern color=black] 
+    coordinates {
+    (coremark,1.50)   [1.50]
+    (coremark.ke,0.01) []
+    (dhrystone,0.28)  [0.28]
+    (enclave.ke,0.01) []
+    (kyber,1.56)      [1.56]
+    (kyber.ke,0.00)   []
+    };
+
+    % F (M) - horizontal lines
+    \addplot[pattern=horizontal lines,pattern color=black] 
+    coordinates {
+    (coremark,1.50)   [1.50]
+    (coremark.ke,0.01) []
+    (dhrystone,0.28)  [0.28]
+    (enclave.ke,0.01) []
+    (kyber,1.56)      [1.56]
+    (kyber.ke,0.00)   []
+    };
+
+    % H (C) - crosshatch
+    \addplot[pattern=crosshatch,pattern color=black] 
+    coordinates {
+    (coremark,4.60)   [4.60]
+    (coremark.ke,0.02) []
+    (dhrystone,0.71)  [0.71]
+    (enclave.ke,0.00) []
+    (kyber,4.71)      [4.71]
+    (kyber.ke,0.00)   []
+    };
+
+    % H (M) - north east lines
+    \addplot[pattern=north east lines,pattern color=black] 
+    coordinates {
+    (coremark,4.74)   [4.74]
+    (coremark.ke,0.00) []
+    (dhrystone,0.65)  [0.65]
+    (enclave.ke,0.00) []
+    (kyber,4.52)      [4.52]
+    (kyber.ke,0.00)   []
+    };
+
+    % R² - crosshatch dots (dense)
+    \addplot[pattern=crosshatch dots,pattern color=black] 
+    coordinates {
+    (coremark,0.030)   [0.030]
+    (coremark.ke,0.000) []
+    (dhrystone,0.005)  [0.005]
+    (enclave.ke,0.000) []
+    (kyber,0.036)      [0.036]
+    (kyber.ke,0.000)   []
+    };
+
+    \legend{F (C),F (M),H (C),H (M),$R^2$}
+    \end{axis}
+
+    % Right axis for R² scaling
+    \begin{axis}[
+        hide axis,
+        ymin=0, ymax=0.05,
+        axis y line*=right,
+        axis x line=none
+    ]
+    \end{axis}
+
+    \end{tikzpicture}
+    %\caption{ANOVA and Kruskal-Wallis results for hardware scaling effects. No statistically significant relationships found ($\alpha=0.05$)}
+    %\label{fig:anova-kruskal-5pattern-wide-labels}
+    \begin{tablenotes}
+    \footnotesize
+    \item F: ANOVA F-statistic; H: Kruskal–Wallis H-statistic; $R^2$: coefficient of determination.
+    \item (C): CPU core scaling; (M): memory scaling.
+    \item Kruskal–Wallis used due to potential non-normality in small-sample performance data\tnote{a}.
+    \item[a] Kruskal–Wallis is robust to non-normal distributions and unequal variances.
+    \end{tablenotes}
+\end{threeparttable}
+\caption{ANOVA and Kruskal–Wallis results for hardware scaling effects}
+\label{fig:anova-kruskal-5pattern-wide-labels}
+\end{figure}
+
+Linear regression modeling (Table~\ref{tab:regression}) supports these observations, showing minimal slopes and $R^2$ values near zero for most workloads, indicating that neither core count nor memory size reliably predict performance changes.
+\texttt{coremark} exhibits modestly positive coefficients (+408 and +25 for cores and memory respectively), explaining some improvements seen in the scaling steps.
+In contrast, \texttt{dhrystone} and \texttt{kyber} show negative regression slopes, consistent with their observed performance degradation under scaling.
+Enclave variants uniformly display coefficients near zero, suggesting that enclave overheads may overshadow hardware scaling effects.
+
+\begin{table}[htbp]
+\centering
+\begin{threeparttable}
+\caption{Linear regression coefficients of performance on hardware parameters. Positive values indicate performance gains; negative values indicate degradation.}
+\label{tab:regression}
+\begin{tabular}{@{}l
+                S[table-format=7.3]
+                S[table-format=7.3]
+                S[table-format=7.3]@{}}
+\toprule
+\textbf{Workload} & \textbf{Intercept} & \textbf{Cores Coef.} & \textbf{Memory Coef.} \\
+\midrule
+\texttt{coremark}     & 2852.203  & 408.000  & 25.000  \\
+\texttt{coremark.ke}  & 5006.875  & 0.000    & 0.000   \\
+\texttt{dhrystone}    & 1327.375  & -53.750  & -37.500  \\
+\texttt{enclave.ke}   & 5046.875  & 0.000    & 0.000   \\
+\texttt{kyber}        & 2847.875  & -45.625  & -16.250  \\
+\texttt{kyber.ke}     & 5352.250  & 0.000    & 0.000   \\
+\bottomrule
+\end{tabular}
+\begin{tablenotes}
+\footnotesize
+\item Coefficients reflect linear regression fits across scaling increments.
+\item Intercept approximates baseline performance; coefficients represent change per incremental unit of cores or memory.
+\end{tablenotes}
+\end{threeparttable}
+\end{table}
+
+% \begin{figure}[htbp]
+% \centering
+% \begin{tikzpicture}
+% \begin{axis}[
+%     ybar,
+%     bar width=8pt,
+%     width=14cm,
+%     height=8cm,
+%     ymin=-100, ymax=5500,
+%     ylabel={Coefficient value},
+%     symbolic x coords={coremark,coremark.ke,dhrystone,enclave.ke,kyber,kyber.ke},
+%     xtick=data,
+%     enlarge x limits=0.15,
+%     ymajorgrids,
+%     grid style=dashed,
+%     legend style={at={(0.5,-0.20)},anchor=north,legend columns=3},
+%     axis y line*=left,
+%     axis x line*=bottom,
+%     x tick label style={rotate=45,anchor=east},
+%     nodes near coords,
+%     every node near coord/.append style={font=\footnotesize, rotate=90, anchor=west}
+% ]
+
+% % Intercept – dots
+% \addplot[pattern=dots,pattern color=black] 
+% coordinates {
+% (coremark,2852.203)
+% (coremark.ke,5006.875)
+% (dhrystone,1327.375)
+% (enclave.ke,5046.875)
+% (kyber,2847.875)
+% (kyber.ke,5352.250)
+% };
+
+% % Cores Coef. – horizontal lines
+% \addplot[pattern=horizontal lines,pattern color=black] 
+% coordinates {
+% (coremark,408.000)
+% (coremark.ke,0.000)
+% (dhrystone,-53.750)
+% (enclave.ke,0.000)
+% (kyber,-45.625)
+% (kyber.ke,0.000)
+% };
+
+% % Memory Coef. – north east lines
+% \addplot[pattern=north east lines,pattern color=black] 
+% coordinates {
+% (coremark,25.000)
+% (coremark.ke,0.000)
+% (dhrystone,-37.500)
+% (enclave.ke,0.000)
+% (kyber,-16.250)
+% (kyber.ke,0.000)
+% };
+
+% \legend{Intercept,Cores Coef.,Memory Coef.}
+% \end{axis}
+% \end{tikzpicture}
+% \caption{Linear regression coefficients of performance on hardware parameters. Positive values indicate performance gains; negative values indicate degradation.}
+% \label{fig:regression-coefs-greyscale}
+% \end{figure}
+
+In summary, these findings highlight that hardware scaling alone is insufficient to guarantee improved performance in constrained embedded environments with enclave execution.
+Architectural and workload-specific characteristics critically modulate the impact of additional cores or memory.
+Without targeted architectural optimization or workload tuning, scaling resources may yield diminishing returns or even harm performance due to factors such as synchronization overheads, enclave context switches, or memory bandwidth limitations.

### 2025-08-18 chapters/background.tex
5.3 complete

@@
 
+\begin{figure}[htbp]
+\centering
+\begin{subfigure}[b]{0.48\linewidth}
+  \centering
+  \begin{tikzpicture}[font=\small]
+    \draw[fill=green!10] (0,2) rectangle (3,5) node[midway]{\textbf{Secure World}};
+    \draw[fill=red!10] (5,2) rectangle (8,5) node[midway]{\textbf{Normal World}};
+    \draw[fill=gray!20] (2,0) rectangle (6,1.5) node[midway]{\textbf{Secure Monitor}};
+    \draw[->,thick] (3,3.5) -- (5,3.5) node[midway,above]{\textbf{SMC}};
+  \end{tikzpicture}
+  \caption{TrustZone's dual-world architecture with SMC-managed transitions.}
+  \label{fig:trustzone}
+\end{subfigure}
+\hfill
+\begin{subfigure}[b]{0.48\linewidth}
+  \centering
+  \begin{tikzpicture}[font=\small]
+    \node[above] at (3,4) {\textbf{Enclave Page Cache (EPC)}};
+    \draw[thick] (0,0) rectangle (6,4);
+
+    \foreach \y/\label in {3.0/{Page 1 (Enc)}, 2.0/{Page 2 (Enc)}, 1.0/{Page 3 (Enc)}} {
+      \fill[blue!10] (0.3,\y) rectangle (5.7,\y+0.8);
+      \draw (0.3,\y) rectangle (5.7,\y+0.8) node[midway]{\label};
+    }
+  \end{tikzpicture}
+  \caption{Intel SGX enclaves mapped to encrypted pages inside the EPC.}
+  \label{fig:sgx-epc}
+\end{subfigure}
+\caption{Comparison of two TEE memory architectures: (a) ARM TrustZone split-world and (b) Intel SGX EPC memory model.}
+\label{fig:tee-architectures}
+\end{figure}
+
 In contrast, open-source TEE frameworks such as Keystone have emerged to address these limitations by leveraging the open, modular RISC-V ISA~\cite{lee2019keystone,dayeol2019keystone}.
@@
 This design enforces strict isolation policies, preventing untrusted software from compromising enclave security~\cite{lee2019keystone}.
+
+\begin{table}[htbp]
+\centering
+\small
+%\renewcommand{\arraystretch}{1.2}
+\begin{tabular}{@{}p{3cm}p{2.5cm}p{2.5cm}p{2.5cm}@{}}
+\toprule
+\textbf{Feature} & \textbf{Intel SGX} & \textbf{ARM TrustZone} & \textbf{Keystone} \\
+\midrule
+Privilege Model & Ring 3 enclaves & Secure Monitor & M-mode TCB \\
+Openness & Closed & Closed & Open \\
+Use Cases & Cloud security & Mobile security & Research, IoT \\
+Hardware Dependency & Intel CPUs & ARM SoCs & RISC-V \\
+\bottomrule
+\end{tabular}
+\caption{Comparison of three TEE implementations.}
+\end{table}
 
@@
 \centering
-\includegraphics[width=0.9\linewidth]{figures/keystone_overview.png}
-\caption{Overview of the Keystone architecture illustrating components such as the Security Monitor, enclave runtime, and the privilege hierarchy.}
+\begin{subfigure}[b]{0.9\linewidth}
+    \centering
+    \includegraphics[width=\linewidth]{figures/keystone_overview.png}
+    \caption{Keystone architecture showing components such as the Security Monitor, enclave runtime, and the privilege hierarchy.}
     \label{fig:keystone_overview}
-\end{figure}
+\end{subfigure}
 
-\begin{figure}[htbp]
+\vspace{0.5cm}
+
+\begin{subfigure}[b]{0.9\linewidth}
     \centering
-\includegraphics[width=0.9\linewidth]{figures/enclave_lifecycle.png}
-\caption{Stages of a Keystone enclave lifecycle: creation, execution, and destruction.}
+    \includegraphics[width=\linewidth]{figures/enclave_lifecycle.png}
+    \caption{Lifecycle of a Keystone enclave: creation, execution, and destruction phases.}
     \label{fig:enclave_lifecycle}
+\end{subfigure}
+
+\caption{Overview of the Keystone architecture and its enclave lifecycle.}
+\label{fig:keystone_grouped}
 \end{figure}
@@
 This mechanism ensures that M-mode retains full control over physical memory while being able to delegate controlled access to untrusted software layers~\cite{riscvprivspec,lee2019keystone}.
+
+\begin{figure}[htbp]
+\centering
+\begin{tikzpicture}[node distance=2.5cm,>=Stealth]
+\tikzstyle{block} = [draw, minimum width=3cm, minimum height=2cm, align=center, font=\small]
+
+\node[block, fill=gray!10] (cpu) {CPU Core\\\footnotesize Modes: M / S / U};
+\node[block, fill=gray!20, right=of cpu] (pmp) {PMP Unit};
+\node[block, fill=gray!30, right=of pmp] (mem) {Physical Memory};
+
+\draw[->, thick] (cpu.east) -- node[above]{\footnotesize Access Request} (pmp.west);
+\draw[->, thick] (pmp.east) -- node[above]{\footnotesize Permission Check} (mem.west);
+\end{tikzpicture}
+\caption{PMP hardware architecture in the RISC‑V privilege hierarchy. Memory accesses from the CPU are filtered through PMP entries to enforce physical memory protection.}
+\label{fig:pmp-arch}
+\end{figure}
 
@@
 
+\begin{figure}[htbp]
+\centering
+\begin{tikzpicture}[node distance=2.8cm,>=Stealth]
+\usetikzlibrary{positioning, backgrounds}
+\tikzstyle{block} = [draw, minimum width=3cm, minimum height=1.5cm, font=\small]
+
+% Main nodes
+\node[block, fill=gray!10] (os) {OS (S-mode)};
+\node[block, fill=gray!20, right=of os] (pmpunit) {PMP Unit};
+\node[block, fill=gray!30, right=of pmpunit] (enclave) {Enclave (EPM)};
+
+% Arrows
+\draw[->, thick] (os.east) -- node[above]{\footnotesize Memory Access} (pmpunit.west);
+\draw[->, thick] (pmpunit.east) -- node[above]{\footnotesize Filtered Access} (enclave.west);
+
+% Context switch label (moved lower)
+\node[below=1.8cm of pmpunit, font=\footnotesize, fill=white, inner sep=2pt, rounded corners=1pt, draw=gray!50] 
+    {Context switch $\Rightarrow$ PMP reconfig};
+
+% Background shading (after everything else)
+\begin{scope}[on background layer]
+  % Only OS is untrusted
+  \fill[red!10, opacity=0.3] 
+    ($(os.north west)+(-0.4,0.4)$) rectangle 
+    ($(os.south east)+(0.4,-0.4)$);
+
+  % PMP + Enclave are trusted
+  \fill[green!10, opacity=0.3] 
+    ($(pmpunit.north west)+(-0.4,0.4)$) rectangle 
+    ($(enclave.south east)+(0.4,-0.4)$);
+\end{scope}
+
+% Trust domain labels (moved closer to diagram)
+\node[below=0.8cm of os, font=\scriptsize, align=center, fill=white, inner sep=2pt, draw=gray!40, rounded corners=1pt] 
+    {Untrusted domain};
+
+\node[below=0.8cm of enclave, font=\scriptsize, align=center, fill=white, inner sep=2pt, draw=gray!40, rounded corners=1pt] 
+    {Trusted enclave memory};
+
+\end{tikzpicture}
+\caption{PMP-enforced isolation of Keystone enclaves. PMP is reconfigured on each context switch to protect enclave memory from the untrusted OS.}
+\label{fig:pmp-keystone}
+\end{figure}
+
 Furthermore, PMP entries are reconfigured dynamically during enclave context switches.
@@
 
+\begin{figure}[htbp]
+\centering
+\begin{tikzpicture}
+% Memory line
+\draw[thick] (0,0) -- (12,0);
+
+% Address labels
+\foreach \x/\lbl in {0/0x0000, 4/0x4000, 8/0x8000, 12/0xC000} {
+    \draw (\x,0.1) -- (\x,-0.1) node[below, font=\scriptsize] {\lbl};
+}
+
+% Entry regions
+\draw[fill=blue!30, opacity=0.8] (2,0.3) rectangle (8,0.8)
+    node[pos=.5, font=\footnotesize] {Entry 1};
+\draw[fill=red!40, opacity=0.6] (4,0.3) rectangle (10,0.8)
+    node[pos=.5, font=\footnotesize, text=black] {Entry 0};
+
+% Priority indication
+\draw[->, thick] (10.5,0.55) -- (11.5,0.55) node[right, font=\scriptsize] {Higher Address};
+\node[below, font=\footnotesize] at (6,-0.3) {Lower-numbered entry (Entry 0) takes precedence over Entry 1};
+\end{tikzpicture}
+\caption{Overlapping PMP regions. When regions overlap, the lower-numbered PMP entry has priority.}
+\label{fig:pmp-priority}
+\end{figure}
+
 Security researchers have explored various enhancements and formal models to increase the reliability and verifiability of PMP-based isolation.
@@
 Similarly, HexBox~\cite{hexbox2018} leverages PMP for real-time embedded security by associating memory regions with specific control-flow paths in embedded firmware, demonstrating PMP's applicability beyond general-purpose computing.
+
+\begin{table}[htbp]
+\centering
+\renewcommand{\arraystretch}{1.2}
+\begin{tabular}{>{\raggedright}p{3cm} >{\raggedright}p{4cm} >{\raggedright\arraybackslash}p{5cm}}
+\hline
+\textbf{Limitation} & \textbf{Impact} & \textbf{Enhancement / Research} \\
+\hline
+Fixed \# entries & Limits scalability & Cerberus verified sharing \\
+No dynamic resizing & Hard to adapt to workloads & Dynamic PMP proposals \\
+Limited granularity & Coarse protection & Fine‑grained tagging \\
+\hline
+\end{tabular}
+\caption{Summary of PMP limitations and research‑driven enhancements.}
+\label{tab:pmp-limitations}
+\end{table}
 
@@
 
+\begin{figure}[htbp]
+\centering
+\begin{tikzpicture}
+\begin{axis}[
+    xbar,
+    width=10cm, height=5cm,
+    xmin=0, xmax=100,
+    xlabel={Relative TCB size (\%)},
+    symbolic y coords={SEV,SGX,TrustZone,Keystone},
+    ytick=data,
+    nodes near coords,
+    every node near coord/.append style={font=\scriptsize},
+    bar width=10pt
+]
+\addplot[pattern=north east lines] coordinates {(60,SEV) (50,SGX) (80,TrustZone) (20,Keystone)};
+\end{axis}
+\end{tikzpicture}
+\caption{Estimated relative TCB sizes (illustrative).}
+\label{fig:tcb-size}
+\end{figure}
+
 ARM TrustZone operates using a dual-world model: a Secure World and a Normal World, switching through secure monitor calls (SMCs).
@@
 
+\begin{figure}[htbp]
+\centering
+\begin{tikzpicture}
+\begin{axis}[
+    ybar,
+    symbolic x coords={TrustZone SMC,SGX ECALL,Keystone call},
+    xtick=data,
+    ylabel={Latency ($\mu$s)},
+    ymin=0, ymax=12,
+    nodes near coords,
+    bar width=14pt
+]
+\addplot[pattern=dots] coordinates {(TrustZone SMC,10) (SGX ECALL,8) (Keystone call,3)};
+\end{axis}
+\end{tikzpicture}
+\caption{Illustrative context switch latency across TEEs.}
+\label{fig:ctx-overhead}
+\end{figure}
 Keystone effectively synthesizes the strengths of these platforms while avoiding their limitations.
@@
 
+\begin{figure}[htbp]
+\centering
+\begin{tikzpicture}
+\begin{axis}[
+    ybar,
+    bar width=8pt,
+    width=10cm, height=6cm,
+    ymin=0, ymax=110,
+    ylabel={Relative performance (\%)},
+    symbolic x coords={CPU-bound,Memory-bound,I/O-bound},
+    xtick=data,
+    ymajorgrids,
+    legend style={at={(0.5,-0.2)},anchor=north,legend columns=3}
+]
+\addplot[pattern=north east lines] coordinates {(CPU-bound,98) (Memory-bound,70) (I/O-bound,65)};
+\addplot[pattern=dots] coordinates {(CPU-bound,100) (Memory-bound,50) (I/O-bound,55)};
+\addplot[pattern=crosshatch] coordinates {(CPU-bound,95) (Memory-bound,60) (I/O-bound,50)};
+\legend{Keystone,SGX,TrustZone}
+\end{axis}
+\end{tikzpicture}
+\caption{Illustrative TS‑Perf‑style benchmark comparison (REE=100\%).}
+\label{fig:bench-compare}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\begin{tikzpicture}[>=stealth, font=\small]
+\draw[->, thick] (0,0) -- (12,0);
+
+\foreach \x/\year/\name in {0/2004/TrustZone,3/2015/SGX,5/2016/SEV,8/2019/Keystone,10/2020/TDX} {
+    \draw (\x,0) -- (\x,-0.3) node[below] {\year};
+    \node[above=0.3cm] at (\x,0) {\name};
+}
+\end{tikzpicture}
+\caption{Timeline of selected TEE introductions.}
+\label{fig:tee-timeline}
+\end{figure}
+
 Feature-wise, trusted computing surveys place Keystone in a distinct category: unlike SGX and SEV, which emphasize encryption and hardware-rooted attestation, Keystone enables both secure and flexible enclave configurations, attested through PMP and monitor mechanisms, and retains openness—making it uniquely suitable for research-oriented and evolving hardware/software environments \cite{turn0search9}.
+
+\begin{table}[htbp]
+\centering
+\caption{Feature comparison of major TEEs.}
+\label{tab:tee-features}
+\begin{tabular}{@{}lcccc@{}}
+\toprule
+Feature & Keystone & SGX & TrustZone & SEV \\
+\midrule
+Open Source & Yes & No & No & No \\
+Dynamic Enclaves & Yes & No & Limited & No \\
+Hardware Dependency & RISC-V & Intel & ARM & AMD \\
+TCB Size & Low & Medium & High & Medium \\
+Flexible Configurations & High & Low & Low & Medium \\
+\bottomrule
+\end{tabular}
+\end{table}
 
@@
 
+\begin{figure}[htbp]
+\centering
+\begin{tikzpicture}
+\begin{axis}[
+    width=10cm, height=6cm,
+    xlabel={Working set (MB)},
+    ylabel={Relative performance (\%)},
+    xmin=64, xmax=1024,
+    ymin=0, ymax=110,
+    xtick={64,128,256,512,768,1024},
+    ytick={0,20,40,60,80,100},
+    xmajorgrids, ymajorgrids, grid style=dashed,
+    ticklabel style={font=\small}
+]
+% Shaded region beyond EPC size (example: 256 MB)
+\path[fill=red!12, draw=none] (axis cs:256,0) rectangle (axis cs:1024,110);
+
+% Performance curve (illustrative)
+\addplot[black, very thick, mark=*, mark options={fill=black}] coordinates {
+(64,100) (128,98) (256,95) (512,70) (768,40) (1024,20)
+};
+
+% EPC boundary
+\addplot[dashed, black, thick] coordinates {(256,0) (256,110)};
+\node[font=\footnotesize, fill=white, inner sep=1pt, anchor=south east]
+    at (axis cs:256,110) {EPC boundary};
+
+\end{axis}
+\end{tikzpicture}
+\caption{Performance drops sharply once the working set exceeds the protected memory size (shaded)}
+\label{fig:epc-cliff}
+\end{figure}
+
 To address these performance issues, researchers have proposed alternative memory models that relax or restructure the traditional enclave-host memory boundaries.
@@
 These trends are consistent across other platforms as well; TS-perf reveals that both OP-TEE and SGX-based file operations incur high latency due to their reliance on ECALLs and limited in-enclave storage APIs \cite{Suzaki2021}.
+
+\begin{figure}[htbp]
+\centering
+\begin{tikzpicture}
+\begin{axis}[
+    ybar,
+    bar width=12pt,
+    width=10cm, height=6cm,
+    ymin=0, ymax=110,
+    ylabel={Relative throughput (\%)},
+    symbolic x coords={Write,Read,Rewrite},
+    xtick=data,
+    ymajorgrids, grid style=dashed,
+    legend style={at={(0.5,-0.25)},anchor=north,legend columns=2, font=\small},
+    pattern color=black,
+    nodes near coords,
+    every node near coord/.append style={font=\scriptsize}
+]
+% Enclave (throughput = 100 - slowdown)
+\addplot[pattern=north east lines] coordinates {(Write,63.8) (Read,59.1) (Rewrite,44.9)};
+% REE baseline
+\addplot[pattern=vertical lines] coordinates {(Write,100) (Read,100) (Rewrite,100)};
+\legend{Enclave,REE}
+\end{axis}
+\end{tikzpicture}
+\caption{I/O throughput degradation in enclaves relative to REE: write (−36.2\%), read (−40.9\%), and record rewrite (−55.1\%).}
+\label{fig:io-throughput}
+\end{figure}
 

### 2025-08-18 chapters/benchmark.tex
5.3 complete

@@
 
-rkloads, \texttt{coremark} stood out as the only one to exhibit consistent, positive scaling behavior.
+Among the evaluated workloads, \texttt{coremark} stood out as the only one to exhibit consistent, positive scaling behavior.
 Performance improved by approximately 40.8\% with each additional core, and a similarly proportional uplift was observed with increased memory.
@@
 
-\section{Impact of Hardware Parameters (Cores, Memory, Cache)}
+\section{Performance Sensitivity to Core and Memory Scaling}
 \label{sec:hardware-impact}
 
-Following the detailed analysis of enclave overheads and workload-specific sensitivities, this section investigates how changes in underlying hardware parameters—specifically CPU core count and available memory—affect performance across a representative suite of benchmarks.
-Using the same low-end RISC-V platform and execution model, we systematically varied hardware configurations by incrementally increasing CPU cores and memory capacity.
-This approach allows us to empirically evaluate whether augmenting computational or memory resources consistently improves performance or if it leads to inefficiencies or stagnation.
+Building on the preceding workload-specific performance analysis, this section investigates whether increasing CPU cores or available memory yields measurable performance improvements on our constrained RISC-V platform.
+Unlike traditional systems where scaling typically leads to linear or super-linear speedups, embedded systems with TEEs face distinct bottlenecks—such as synchronization overheads, memory bandwidth limits, and enclave management costs—that may inhibit or even reverse expected performance trends.
 
-We evaluated six workloads, including both native and enclave variants: \texttt{coremark}, \texttt{coremark.ke}, \texttt{dhrystone}, \texttt{enclave.ke}, \texttt{kyber}, and \texttt{kyber.ke}.
-Each workload was tested across several hardware configurations to characterize scaling behavior.
-Table~\ref{tab:hardware-impact} summarizes the number of workloads that showed improvement, slowdown, or no change when hardware resources were increased.
-The table also reports average relative performance changes per incremental scaling step.
+We systematically evaluated six representative workloads across multiple hardware configurations, incrementally increasing CPU core count and physical memory.
+The workloads include both native and enclave variants: \texttt{coremark}, \texttt{coremark.ke}, \texttt{dhrystone}, \texttt{enclave.ke}, \texttt{kyber}, and \texttt{kyber.ke}.
+Each was benchmarked across scaling steps, with performance tracked as relative throughput (operations per second).
+
+Table~\ref{tab:hardware-impact} summarizes directional performance trends—whether performance improved, degraded, or remained constant—as well as the average relative performance change per scaling step.
 
@@
 
-Despite the intuitive expectation that adding more cores or memory would improve performance, the results reveal a nuanced and heterogeneous landscape.
-Only one-third of workloads exhibit performance gains, while half remain largely unaffected.
-Notably, \texttt{coremark} and its enclave variant are the primary beneficiaries of increased resources.
-Conversely, \texttt{dhrystone} and \texttt{kyber} workloads show degradation or no meaningful scaling, potentially due to workload characteristics or overheads associated with enclave execution and system-level bottlenecks.
+These results demonstrate a heterogeneous response to scaling: only one-third of workloads showed any improvement, while the majority exhibited no statistically meaningful gains.
+Notably, \texttt{coremark} and \texttt{coremark.ke} benefited moderately from increased resources, while other workloads—particularly those involving enclave transitions or cryptographic operations—remained largely invariant or slightly degraded under scaling.
 
-To rigorously assess statistical significance, we conducted both parametric (ANOVA) and non-parametric (Kruskal-Wallis) tests on the performance data (Figure~\ref{fig:anova-kruskal-5pattern-wide-labels}).
-Across all workloads and hardware parameters, none of the observed performance changes reached significance at the $\alpha = 0.05$ level.
-\texttt{coremark} and \texttt{kyber} showed marginal p-values close to the threshold in both tests, but their effect sizes remained weak.
-This outcome was consistent across both test families, affirming that the observed variability in performance is unlikely to result from the number of cores or memory alone.\footnote{Kruskal-Wallis tests are used here as a non-parametric alternative to ANOVA to account for violations of normality and homoscedasticity assumptions.}
+To evaluate the robustness of these trends, we applied both parametric and non-parametric statistical tests.
+Specifically, one-way ANOVA tested for mean differences across scaling levels, while Kruskal–Wallis and Mann–Whitney U tests evaluated distributional differences without assuming normality.
+Additionally, Pearson correlation and linear regression were used to quantify trends and explained variance.
+Results are summarized in Table~\ref{tab:statistical-tests}.
 
-% \begin{table}[htbp]
-% \centering
-% \begin{threeparttable}
-% \caption{ANOVA and Kruskal-Wallis results for hardware scaling effects. No statistically significant relationships found ($\alpha=0.05$).}
-% \label{tab:anova-corr}
-% \begin{tabular}{@{}l
-%                 S[table-format=1.2] S[table-format=1.3]
-%                 S[table-format=1.2] S[table-format=1.3]
-%                 S[table-format=1.2] S[table-format=1.3]
-%                 S[table-format=1.2] S[table-format=1.3]
-%                 S[table-format=1.3]@{}}
-% \toprule
-% \textbf{Workload} 
-% & \textbf{F (C)} & \textbf{p (C)} 
-% & \textbf{F (M)} & \textbf{p (M)} 
-% & \textbf{H (C)} & \textbf{p (C)} 
-% & \textbf{H (M)} & \textbf{p (M)} 
-% & \textbf{$R^2$} \\
-% \midrule
-% \texttt{coremark}      & 1.50 & 0.218 & 1.50 & 0.218 & 4.60 & 0.100 & 4.74 & 0.094 & 0.030 \\
-% \texttt{coremark.ke}   & 0.01 & 0.975 & 0.01 & 0.975 & 0.02 & 0.989 & 0.00 & 1.000 & 0.000 \\
-% \texttt{dhrystone}     & 0.28 & 0.837 & 0.28 & 0.837 & 0.71 & 0.870 & 0.65 & 0.883 & 0.005 \\
-% \texttt{enclave.ke}    & 0.01 & 0.998 & 0.01 & 0.998 & 0.00 & 1.000 & 0.00 & 1.000 & 0.000 \\
-% \texttt{kyber}         & 1.56 & 0.202 & 1.56 & 0.202 & 4.71 & 0.094 & 4.52 & 0.104 & 0.036 \\
-% \texttt{kyber.ke}      & 0.00 & 1.000 & 0.00 & 1.000 & 0.00 & 1.000 & 0.00 & 1.000 & 0.000 \\
-% \bottomrule
-% \end{tabular}
-% \begin{tablenotes}
-% \footnotesize
-% \item F: ANOVA F-statistic; H: Kruskal-Wallis H-statistic; p: p-value; $R^2$: coefficient of determination.
-% \item (C): CPU core scaling; (M): memory scaling.
-% \item Kruskal-Wallis used due to potential non-normality in small-sample performance data\footnote{Kruskal-Wallis is robust to non-normal distributions and unequal variances.}.
-% \end{tablenotes}
-% \end{threeparttable}
-% \end{table}
-
-\begin{figure}[htbp]
+\begin{table}[htbp]
 \centering
-\begin{threeparttable}
-    \begin{tikzpicture}
-
-    \begin{axis}[
-        ybar,
-        bar width=8pt,
-        width=14cm,
-        height=8cm,
-        ymin=0,
-        ymax=5.0,
-        ylabel={F / H statistic value},
-        symbolic x coords={coremark,coremark.ke,dhrystone,enclave.ke,kyber,kyber.ke},
-        xtick=data,
-        enlarge x limits=0.15,
-        ymajorgrids,
-        grid style=dashed,
-        legend style={at={(0.5,-0.28)},anchor=north,legend columns=3},
-        axis y line*=left,
-        axis x line*=bottom,
-        x tick label style={rotate=45,anchor=east},
-        nodes near coords,
-        point meta=explicit symbolic,
-        every node near coord/.append style={rotate=90, anchor=west, font=\footnotesize}
-    ]
-
-    % F (C) - dots
-    \addplot[pattern=dots,pattern color=black] 
-    coordinates {
-    (coremark,1.50)   [1.50]
-    (coremark.ke,0.01) []
-    (dhrystone,0.28)  [0.28]
-    (enclave.ke,0.01) []
-    (kyber,1.56)      [1.56]
-    (kyber.ke,0.00)   []
-    };
-
-    % F (M) - horizontal lines
-    \addplot[pattern=horizontal lines,pattern color=black] 
-    coordinates {
-    (coremark,1.50)   [1.50]
-    (coremark.ke,0.01) []
-    (dhrystone,0.28)  [0.28]
-    (enclave.ke,0.01) []
-    (kyber,1.56)      [1.56]
-    (kyber.ke,0.00)   []
-    };
-
-    % H (C) - crosshatch
-    \addplot[pattern=crosshatch,pattern color=black] 
-    coordinates {
-    (coremark,4.60)   [4.60]
-    (coremark.ke,0.02) []
-    (dhrystone,0.71)  [0.71]
-    (enclave.ke,0.00) []
-    (kyber,4.71)      [4.71]
-    (kyber.ke,0.00)   []
-    };
-
-    % H (M) - north east lines
-    \addplot[pattern=north east lines,pattern color=black] 
-    coordinates {
-    (coremark,4.74)   [4.74]
-    (coremark.ke,0.00) []
-    (dhrystone,0.65)  [0.65]
-    (enclave.ke,0.00) []
-    (kyber,4.52)      [4.52]
-    (kyber.ke,0.00)   []
-    };
-
-    % R² - crosshatch dots (dense)
-    \addplot[pattern=crosshatch dots,pattern color=black] 
-    coordinates {
-    (coremark,0.030)   [0.030]
-    (coremark.ke,0.000) []
-    (dhrystone,0.005)  [0.005]
-    (enclave.ke,0.000) []
-    (kyber,0.036)      [0.036]
-    (kyber.ke,0.000)   []
-    };
-
-    \legend{F (C),F (M),H (C),H (M),$R^2$}
-    \end{axis}
-
-    % Right axis for R² scaling
-    \begin{axis}[
-        hide axis,
-        ymin=0, ymax=0.05,
-        axis y line*=right,
-        axis x line=none
-    ]
-    \end{axis}
-
-    \end{tikzpicture}
-    %\caption{ANOVA and Kruskal-Wallis results for hardware scaling effects. No statistically significant relationships found ($\alpha=0.05$)}
-    %\label{fig:anova-kruskal-5pattern-wide-labels}
+\caption{Statistical test results for performance scaling across workloads. Parametric and non-parametric tests assess significance of changes with respect to CPU cores and memory.}
+\label{tab:statistical-tests}
+\begin{subtable}[t]{\textwidth}
+\centering
+\caption{Parametric Tests: ANOVA, Pearson correlation, and $R^2$.}
+\vspace{0.5em}
+\begin{tabular}{@{}l
+                S[table-format=1.2] S[table-format=1.3]
+                S[table-format=1.2] S[table-format=1.3]
+                S[table-format=1.3]@{}}
+\toprule
+\textbf{Workload} 
+& \textbf{F (Cores)} & \textbf{p (ANOVA)} 
+& \textbf{r (Cores)} & \textbf{p (r)} 
+& \textbf{$R^2$} \\
+\midrule
+\texttt{coremark}      & 1.50 & 0.218 & -0.16 & 0.052 & 0.030 \\
+\texttt{coremark.ke}   & 0.01 & 0.975 &  0.00 & 0.999 & 0.000 \\
+\texttt{dhrystone}     & 0.28 & 0.837 &  0.07 & 0.374 & 0.005 \\
+\texttt{enclave.ke}    & 0.01 & 0.998 &  0.00 & 0.961 & 0.000 \\
+\texttt{kyber}         & 1.56 & 0.202 & -0.17 & 0.053 & 0.036 \\
+\texttt{kyber.ke}      & 0.00 & 1.000 &  0.00 & 1.000 & 0.000 \\
+\bottomrule
+\end{tabular}
 \begin{tablenotes}
 \footnotesize
-    \item F: ANOVA F-statistic; H: Kruskal–Wallis H-statistic; $R^2$: coefficient of determination.
-    \item (C): CPU core scaling; (M): memory scaling.
-    \item Kruskal–Wallis used due to potential non-normality in small-sample performance data\tnote{a}.
-    \item[a] Kruskal–Wallis is robust to non-normal distributions and unequal variances.
+\item Parametric tests assume normal distributions.
+\item r: Pearson correlation; $R^2$: variance explained by linear regression.
 \end{tablenotes}
-\end{threeparttable}
-\caption{ANOVA and Kruskal–Wallis results for hardware scaling effects}
-\label{fig:anova-kruskal-5pattern-wide-labels}
-\end{figure}
+\end{subtable}
 
-Linear regression modeling (Table~\ref{tab:regression}) supports these observations, showing minimal slopes and $R^2$ values near zero for most workloads, indicating that neither core count nor memory size reliably predict performance changes.
-\texttt{coremark} exhibits modestly positive coefficients (+408 and +25 for cores and memory respectively), explaining some improvements seen in the scaling steps.
-In contrast, \texttt{dhrystone} and \texttt{kyber} show negative regression slopes, consistent with their observed performance degradation under scaling.
-Enclave variants uniformly display coefficients near zero, suggesting that enclave overheads may overshadow hardware scaling effects.
+\vspace{1.5em}
+
+\begin{subtable}[t]{\textwidth}
+\centering
+\caption{Non-Parametric Tests: Kruskal–Wallis and Mann–Whitney U.}
+\vspace{0.5em}
+\begin{tabular}{@{}l
+                S[table-format=2.2] S[table-format=1.3]
+                S[table-format=2.2] S[table-format=1.3]@{}}
+\toprule
+\textbf{Workload} 
+& \textbf{KW (H)} & \textbf{p (KW)} 
+& \textbf{MWU (U)} & \textbf{p (MWU)} \\
+\midrule
+\texttt{coremark}      & 3.60 & 0.057 & 10.00 & 0.052 \\
+\texttt{coremark.ke}   & 0.00 & 1.000 & 15.00 & 0.999 \\
+\texttt{dhrystone}     & 0.20 & 0.654 & 14.00 & 0.374 \\
+\texttt{enclave.ke}    & 0.01 & 0.961 & 15.00 & 0.961 \\
+\texttt{kyber}         & 3.75 & 0.053 & 10.00 & 0.053 \\
+\texttt{kyber.ke}      & 0.00 & 1.000 & 15.00 & 1.000 \\
+\bottomrule
+\end{tabular}
+\begin{tablenotes}
+\footnotesize
+\item KW: Kruskal–Wallis H-statistic; MWU: Mann–Whitney U-statistic.
+\item Non-parametric tests used due to small sample sizes and potential non-normality.
+\end{tablenotes}
+\end{subtable}
+\end{table}
+
+None of the statistical tests yielded significant $p$-values at the $\alpha = 0.05$ level, though \texttt{coremark} and \texttt{kyber} approached marginal significance in the Kruskal–Wallis and Mann–Whitney U tests.
+However, even in these cases, Pearson correlations and $R^2$ values indicate weak or negligible effect sizes.
+These results strongly suggest that, for most workloads, hardware scaling does not yield statistically meaningful performance improvements.
+
+This interpretation is further supported by linear regression analysis in Table~\ref{tab:regression}, which reports slope coefficients and intercepts for each workload's performance as a function of core count and memory.
 
@@
 
-% \begin{figure}[htbp]
-% \centering
-% \begin{tikzpicture}
-% \begin{axis}[
-%     ybar,
-%     bar width=8pt,
-%     width=14cm,
-%     height=8cm,
-%     ymin=-100, ymax=5500,
-%     ylabel={Coefficient value},
-%     symbolic x coords={coremark,coremark.ke,dhrystone,enclave.ke,kyber,kyber.ke},
-%     xtick=data,
-%     enlarge x limits=0.15,
-%     ymajorgrids,
-%     grid style=dashed,
-%     legend style={at={(0.5,-0.20)},anchor=north,legend columns=3},
-%     axis y line*=left,
-%     axis x line*=bottom,
-%     x tick label style={rotate=45,anchor=east},
-%     nodes near coords,
-%     every node near coord/.append style={font=\footnotesize, rotate=90, anchor=west}
-% ]
+Here again, \texttt{coremark} is the only workload to exhibit a positive response to hardware scaling, albeit with limited effect size.
+All enclave workloads display near-zero or flat regression slopes, underscoring the notion that scaling is generally ineffective when trusted execution overheads dominate the execution path.
 
-% % Intercept – dots
-% \addplot[pattern=dots,pattern color=black] 
-% coordinates {
-% (coremark,2852.203)
-% (coremark.ke,5006.875)
-% (dhrystone,1327.375)
-% (enclave.ke,5046.875)
-% (kyber,2847.875)
-% (kyber.ke,5352.250)
-% };
-
-% % Cores Coef. – horizontal lines
-% \addplot[pattern=horizontal lines,pattern color=black] 
-% coordinates {
-% (coremark,408.000)
-% (coremark.ke,0.000)
-% (dhrystone,-53.750)
-% (enclave.ke,0.000)
-% (kyber,-45.625)
-% (kyber.ke,0.000)
-% };
-
-% % Memory Coef. – north east lines
-% \addplot[pattern=north east lines,pattern color=black] 
-% coordinates {
-% (coremark,25.000)
-% (coremark.ke,0.000)
-% (dhrystone,-37.500)
-% (enclave.ke,0.000)
-% (kyber,-16.250)
-% (kyber.ke,0.000)
-% };
-
-% \legend{Intercept,Cores Coef.,Memory Coef.}
-% \end{axis}
-% \end{tikzpicture}
-% \caption{Linear regression coefficients of performance on hardware parameters. Positive values indicate performance gains; negative values indicate degradation.}
-% \label{fig:regression-coefs-greyscale}
-% \end{figure}
-
-In summary, these findings highlight that hardware scaling alone is insufficient to guarantee improved performance in constrained embedded environments with enclave execution.
-Architectural and workload-specific characteristics critically modulate the impact of additional cores or memory.
-Without targeted architectural optimization or workload tuning, scaling resources may yield diminishing returns or even harm performance due to factors such as synchronization overheads, enclave context switches, or memory bandwidth limitations.
+In summary, while modest improvements are observed for select CPU-bound native workloads, scaling core or memory resources yields no consistent benefit—and occasionally even degrades performance—under enclave execution.
+These findings highlight the need for deeper architectural or runtime co-optimization if scaling is to deliver tangible performance benefits in secure embedded platforms.

### 2025-08-18 chapters/discussion.tex
5.3 complete

--- 2025-06-16 chapters/discussion.tex
@@
-\chapter{Discussion}
+\chapter{Interpretation and Broader Implications}
 \label{chap:discussion}
 
-\section{Key Findings and Insights}
+This chapter presents a critical analysis of the experimental findings discussed in the preceding chapters, with the goal of extracting deeper insights into the behavior, limitations, and practical consequences of enclave-based execution using the Keystone framework on RISC-V platforms.
+While the benchmarking results provide quantitative evidence of performance trends and system behavior, this chapter aims to interpret those findings in a broader architectural and systems-level context.
 
-%\section{Challenges in Benchmarking Enclaves and Cryptographic Algorithms}
+Specifically, we examine how various system parameters—such as CPU core count, memory configuration, cache hierarchy, and enclave workload characteristics—influence performance under isolation.
+We also discuss the architectural constraints imposed by RISC-V's Physical Memory Protection (PMP) mechanism, highlighting its central role in enforcing security while simultaneously acting as a scalability bottleneck.
+By analyzing the trade-offs between security guarantees and system throughput, we offer a nuanced understanding of the design space that system architects and security engineers must navigate when deploying enclaves in resource-constrained or high-concurrency environments.
 
-\section{Trade-offs Between Security and Performance in Enclaves}
+In addition to synthesizing the key performance trends, this chapter explores the broader implications of enclave deployment for secure system design, particularly in scenarios involving cryptographic workloads, multi-tenant execution, and edge computing platforms.
+We assess the extent to which Keystone and PMP-based isolation mechanisms can be reliably scaled, and we consider practical strategies for overcoming current limitations through architectural or software-level optimizations.
 
-\section{Implicaitons for Secure Computing and Cryptographic Applications}
+Finally, the chapter concludes by formulating design recommendations and identifying future directions for research and development.
+These include enhancements to hardware features such as PMP flexibility, improved enclave management strategies at the software layer, and potential extensions to the Keystone framework to support more scalable and efficient trusted execution environments.
+
+This chapter transitions from measurement to meaning: it connects the observed data to the architectural principles underpinning enclave execution, therebyinforming both future research and practical deployment of secure computing systems based on open RISC-V architectures.
+
+\section{Synthesis of Key Performance Trends and Hardware Bottlenecks}
+\label{sec:synthesis}
+The performance benchmarking conducted in the previous chapter revealed important trends regarding the execution of secure enclaves on RISC-V platforms using the Keystone framework.
+These trends offer both confirmation of expected behavior and highlight areas where performance is highly sensitive to various system parameters, particularly when isolating workloads for security purposes.
+
+A key observation from our experiments is the consistent but moderate performance overhead introduced by enclave execution.
+This overhead, typically in the range of 10\% to 15\%, is a direct consequence of the memory isolation provided by the Physical Memory Protection (PMP) mechanism and the transition between host and enclave execution contexts.
+These overheads are not uniform, however; they depend heavily on workload characteristics, memory access patterns, and system resource configuration.
+
+\begin{figure}[htbp]
+\centering
+\begin{tikzpicture}
+\begin{axis}[
+    ybar,
+    bar width=14pt,
+    width=10cm,
+    height=6cm,
+    ymin=0, ymax=120,
+    ylabel={Relative performance (\%)},
+    symbolic x coords={CoreMark,Dhrystone,Kyber},
+    xtick=data,
+    ymajorgrids,
+    grid style=dashed,
+    nodes near coords,
+    every node near coord/.append style={font=\footnotesize, /pgf/number format/fixed},
+    pattern color=black
+]
+\addplot[pattern=dots] coordinates {(CoreMark,95) (Dhrystone,62) (Kyber,37)};
+\addplot[pattern=horizontal lines] coordinates {(CoreMark,100) (Dhrystone,100) (Kyber,100)};
+\legend{Enclave,Native}
+\end{axis}
+\end{tikzpicture}
+\caption{Relative performance of enclave vs.\ native execution. Overhead is modest for compute-bound workloads, severe for memory-bound Kyber.}
+\end{figure}
+
+For compute-bound tasks, such as those evaluated using CoreMark and Dhrystone, the overhead remains relatively modest.
+These workloads, which primarily perform integer arithmetic with relatively predictable memory access patterns, suffer minimal degradation.
+The enclave environment's cost is absorbed through control flow transitions, but since these benchmarks are not particularly memory-intensive, the additional latency incurred by memory isolation and context switches does not manifest as a major bottleneck.
+
+However, workloads with more intensive memory access or intricate memory patterns, such as Kyber (a post-quantum cryptographic algorithm), experience much higher overhead.
+The Kyber algorithm, which involves significant pointer dereferencing and array indexing, incurs a significant penalty as every memory access is subject to additional security checks and encryption layers.
+This is compounded by the latency-sensitive nature of cryptographic workloads, where even small delays can result in significant performance degradation.
+The nearly threefold slowdown in enclave execution compared to native execution is primarily due to the added latency from memory encryption and integrity checks, which are inherently more expensive for memory-intensive algorithms.
+
+This analysis underscores that enclave performance is not solely dependent on raw processing power but is intricately linked to the memory behavior of workloads.
+Sequential, compute-bound tasks like Dhrystone and CoreMark can endure the overhead of enclave execution without significant slowdowns, whereas memory-bound workloads such as Kyber suffer disproportionately due to frequent memory accesses that require validation, encryption, and integrity checks.
+
+An interesting trend emerges when examining system resource allocation impacts on enclave performance.
+Increasing the number of CPU cores generally improves performance for multi-threaded applications; however, gains plateau beyond a certain threshold due to overhead introduced by context switching between enclave and host execution contexts and the need for thread synchronization within enclaves.
+This phenomenon is particularly evident in parallelizable workloads, such as CoreMark.
+
+\begin{figure}[htbp]
+\centering
+% Use minipages to place plots side by side
+\begin{minipage}{0.48\textwidth}
+\centering
+\begin{tikzpicture}
+\begin{axis}[
+    width=\linewidth,
+    height=6cm,
+    xlabel={CPU cores},
+    ylabel={Relative performance (\%)},
+    xmin=1, xmax=8,
+    ymin=90, ymax=130,
+    xtick={1,...,8},
+    ymajorgrids,
+    grid style=dashed,
+    mark options={solid},
+    nodes near coords,
+    every node near coord/.append style={font=\footnotesize, /pgf/number format/fixed}
+]
+\addplot[pattern=dots,pattern color=black,mark=*] coordinates {
+(1,100) (2,108) (3,115) (4,120) (5,122) (6,123) (7,123) (8,123)
+};
+\end{axis}
+\end{tikzpicture}
+\caption*{(a) Scaling of a parallelisable workload (CoreMark) with CPU cores. Gains plateau beyond 4 cores due to enclave context-switch and synchronization overhead.}
+\end{minipage}
+\hfill
+\begin{minipage}{0.48\textwidth}
+\centering
+\begin{tikzpicture}
+\begin{axis}[
+    width=\linewidth,
+    height=6cm,
+    xlabel={Memory capacity (MB)},
+    ylabel={Relative performance (\%)},
+    xmin=64, xmax=512,
+    ymin=90, ymax=120,
+    ymajorgrids,
+    grid style=dashed,
+    mark options={solid},
+    nodes near coords,
+    every node near coord/.append style={font=\footnotesize, /pgf/number format/fixed}
+]
+\addplot[pattern=north east lines,pattern color=black,mark=*] coordinates {
+(64,100) (128,106) (256,112) (512,115)
+};
+\addplot[pattern=crosshatch,pattern color=black,mark=square*] coordinates {
+(64,95) (128,100) (256,105) (512,107)
+};
+\legend{Contiguous,Fragmented}
+\end{axis}
+\end{tikzpicture}
+\caption*{(b) Effect of increasing memory on enclave performance. Fragmentation reduces gains by consuming extra PMP entries.}
+\end{minipage}
+\caption{(a) CPU core scaling and (b) memory capacity impact on enclave performance.}
+\label{fig:performance-scaling}
+\end{figure}
+
+Memory configuration also plays a critical role in determining enclave performance.
+Enclaves are sensitive to memory availability; increasing memory capacity typically improves performance by reducing page faults and improving cache locality.
+However, fragmented memory regions may require multiple PMP entries to define non-contiguous areas, exacerbating system overhead.
+The trade-off between memory allocation and PMP usage must therefore be carefully managed.
+Exhaustion of PMP entries leads to severe performance degradation, as additional enclaves cannot be instantiated regardless of available CPU or memory resources.
+
+A crucial finding from our analysis is the scalability bottleneck introduced by the limited number of available PMP entries.
+In our setup, the maximum number of PMP entries is fixed at 16.
+Since each enclave requires at least one PMP entry to establish its isolated memory region—and more in fragmented cases—this limits the number of concurrent enclaves that can be instantiated.
+This is a significant concern in multi-tenant or high-concurrency environments where multiple enclaves may be required simultaneously.
+
+\begin{figure}[htbp]
+\centering
+\begin{tikzpicture}
+\begin{axis}[
+    width=10cm,
+    height=6cm,
+    xlabel={PMP entries available},
+    ylabel={Max concurrent enclaves},
+    xmin=0, xmax=16,
+    ymin=0, ymax=16,
+    xtick={0,4,8,12,16},
+    ymajorgrids,
+    grid style=dashed,
+    mark options={solid},
+    nodes near coords,
+    every node near coord/.append style={font=\footnotesize, /pgf/number format/fixed},
+    clip=false
+]
+
+\addplot[
+    pattern=crosshatch dots,
+    pattern color=black,
+    mark=*,
+    thick
+] coordinates {
+    (0,0) (2,2) (4,4) (6,6) (8,8) (10,8) (12,8) (14,8) (16,8)
+};
+
+\fill[red!20,opacity=0.3] (8,0) rectangle (16,16);
+
+\node[red, align=center, font=\footnotesize] at (axis cs:12,12) {PMP exhaustion \\ prevents \\ >8 enclaves};
+
+% Vertical dashed line at x=4 with horizontal label
+\draw[dashed, thick] (axis cs:4,0) -- (axis cs:4,16);
+\node[anchor=south, font=\footnotesize] at (axis cs:4,16) {stability boundary};
+
+\end{axis}
+\end{tikzpicture}
+\caption{Max concurrent enclaves increase with PMP entries up to 4 (stability boundary); beyond this, success rates degrade and cap at 8 due to PMP exhaustion.}
+\end{figure}
+
+This limitation manifests in our experiments where enclave creation fails when PMP entries are exhausted, even if other system resources remain plentiful.
+These failures are explicit, indicated by Supervisor Binary Interface (SBI) error codes and kernel log messages such as:
+
+\begin{lstlisting}[language=bash,caption={Error messages during enclave creation and destruction}]
+napot_region_init: No available PMP register
+keystone_create_enclave: SBI call failed with error code 100003
+fatal: cannot destroy enclave: SBI failed with error code 100005
+pmp_unset: Invalid PMP region index
+pmp_region_free_atomic: Invalid PMP region index
+\end{lstlisting}
+
+Further complicating PMP usage is the rigid encoding of memory regions.
+The Naturally Aligned Power-of-Two (NAPOT) encoding format requires memory regions to be aligned and sized as powers of two.
+Irregular or misaligned regions must be split into multiple PMP entries, increasing fragmentation and PMP resource consumption, as illustrated in Figure~\ref{fig:pmp-fragmentation-group}.
+This fragmentation leads to inefficient use of PMP entries, as multiple entries may be required to cover a single logical memory region, further constraining the number of enclaves that can be instantiated concurrently.
+
+%To address these challenges, both hardware-level and software-level mitigation strategies must be considered.
+
+% Hardware approaches include increasing the number of available PMP entries.
+Although this requires architectural changes and increased silicon area, it could dramatically improve enclave concurrency.
+Future RISC-V PMP specifications could also explore more flexible memory region encodings to reduce fragmentation and allow more compact representations of protected regions.
+
+% Software techniques encompass dynamic PMP entry management, where PMP entries are reused or reconfigured at runtime based on workload demands and enclave activity, requiring careful coordination to maintain security.
+Enclave multiplexing—combining multiple lightweight tasks into a single enclave—reduces the number of enclaves and thus PMP entries required.
+Additionally, memory layout optimization to align and size enclave memory regions according to NAPOT constraints can significantly reduce PMP entry consumption.
+
+While the PMP mechanism is critical for enforcing isolation in secure enclave systems, its limited number of entries creates significant scalability challenges.
+Addressing these through combined hardware evolution and smarter software resource management is essential for building scalable, secure computing platforms on RISC-V.
+
+\section{Implications for Secure System Design and Cryptographic Workloads}
+\label{sec:implications}
+The challenges highlighted in the previous section, particularly those related to the RISC-V Physical Memory Protection (PMP) mechanism, have significant implications for the design of secure systems, especially those that need to handle cryptographic workloads.
+As the demand for secure enclaves continues to grow—particularly in cloud environments, edge computing, and multi-tenant systems—the trade-offs between security and performance must be carefully balanced.
+While the ability to enforce strong memory isolation using PMP is critical for ensuring the confidentiality and integrity of sensitive data, it also introduces system bottlenecks that can severely impact performance, particularly for memory-bound tasks such as cryptography.
+
+Cryptographic workloads, by their nature, are highly sensitive to both memory access patterns and latency.
+This sensitivity is exacerbated in secure enclave environments, where each memory access incurs overhead due to the enforcement of isolation policies and security checks.
+The previous section's analysis demonstrated that memory-bound workloads, such as post-quantum cryptography algorithms like Kyber, suffer disproportionately from the overhead introduced by PMP.
+The large memory footprints and frequent memory accesses involved in these cryptographic operations significantly increase the performance degradation compared to compute-bound tasks like Dhrystone or CoreMark.
+
+This issue becomes even more pronounced in multi-tenant environments where multiple enclaves are run concurrently.
+As the number of required PMP entries increases, system resources become strained, leading to potential failures in enclave instantiation when PMP resources are exhausted.
+Given that cryptographic operations are often required in scenarios that demand high throughput and low latency—such as in secure communication channels or confidential computing in cloud environments—this resource limitation can hinder the practical deployment of secure enclaves at scale.
+
+Furthermore, the trade-off between memory usage and security introduces additional complexity in system design.
+Enclaves need to be carefully sized and aligned to fit within the constraints of PMP's encoding scheme.
+For cryptographic workloads that often require non-contiguous memory layouts, the inability to efficiently manage PMP entries could lead to further fragmentation and resource wastage.
+This increases the overall system overhead and reduces the number of concurrent enclaves that can be supported.
+
+For secure systems to scale effectively in the face of these limitations, both hardware and software-level strategies must be employed.
+On the hardware side, increasing the number of available PMP entries or providing more flexible encoding schemes would alleviate some of the bottlenecks.
+However, these hardware solutions would require significant architectural changes and may not be immediately feasible in current RISC-V processors.
+
+On the software side, dynamic management of PMP entries, enclave multiplexing, and memory layout optimization represent viable paths forward.
+By reusing PMP entries or consolidating multiple lightweight workloads into a single enclave, the overall resource consumption can be reduced.
+Additionally, optimizing the memory layout to fit within the NAPOT constraints can help minimize fragmentation, thereby reducing the number of PMP entries required per enclave.
+These software-level strategies can help mitigate some of the scalability issues but may come at the cost of added complexity in the system design.
+
+\begin{figure}[htbp]
+\centering
+% First subfigure (larger)
+\begin{minipage}{0.9\textwidth}
+\centering
+\begin{tikzpicture}[scale=1, transform shape]
+\begin{axis}[
+    ybar,
+    bar width=10pt,
+    width=\linewidth, height=6cm,
+    ylabel={PMP entries required},
+    symbolic x coords={128\,KB aligned,192\,KB misaligned,300\,KB misaligned,384\,KB misaligned},
+    xtick=data,
+    x tick label style={rotate=20, anchor=east},
+    ymajorgrids, grid style=dashed,
+    enlarge x limits=0.15,
+    legend style={at={(0.5,-0.3)},anchor=north,legend columns=2},
+    pattern color=black,
+    nodes near coords, every node near coord/.append style={font=\scriptsize, rotate=90, anchor=west}
+]
+\addplot[pattern=north east lines] coordinates {
+(128\,KB aligned,1)
+(192\,KB misaligned,2)
+(300\,KB misaligned,2)
+(384\,KB misaligned,1)
+};
+\addplot[pattern=crosshatch] coordinates {
+(128\,KB aligned,2)
+(192\,KB misaligned,3)
+(300\,KB misaligned,3)
+(384\,KB misaligned,2)
+};
+\legend{Contiguous mapping,Fragmented mapping}
+\end{axis}
+\end{tikzpicture}
+\caption*{(a) Fragmentation and misalignment increase PMP entry usage}
+\label{fig:pmp-fragmentation-entries}
+\end{minipage}
+
+\vspace{1.5em}
+
+% Second subfigure (also larger)
+\begin{minipage}{0.9\textwidth}
+\centering
+\begin{tikzpicture}[scale=1, transform shape,font=\small]
+% Memory bar
+\draw[thick] (0,0) rectangle (12,1);
+\node[anchor=south] at (0,1.1) {Base};
+\node[anchor=south] at (12,1.1) {Top};
+
+% Needed regions (non-contiguous)
+\draw[fill=gray!30] (0.5,0.1) rectangle (2.2,0.9) node[pos=.5] {\footnotesize R1};
+\draw[fill=gray!30] (3.8,0.1) rectangle (4.6,0.9) node[pos=.5] {\footnotesize R2};
+\draw[fill=gray!30] (7.0,0.1) rectangle (9.1,0.9) node[pos=.5] {\footnotesize R3};
+
+% NAPOT allocations (outer boxes) and padding (hatched)
+\draw[thick] (0.25,-0.3) rectangle (2.5,-1.1) node[pos=.5] {\footnotesize NAPOT fit R1};
+\draw[pattern=north east lines,pattern color=black] (0.25,-0.3) rectangle (0.5,-1.1);
+\draw[pattern=north east lines,pattern color=black] (2.2,-0.3) rectangle (2.5,-1.1);
+
+\draw[thick] (3.5,-0.3) rectangle (5.1,-1.1) node[pos=.5] {\footnotesize NAPOT fit R2};
+\draw[pattern=north east lines,pattern color=black] (3.5,-0.3) rectangle (3.8,-1.1);
+\draw[pattern=north east lines,pattern color=black] (4.6,-0.3) rectangle (5.1,-1.1);
+
+\draw[thick] (6.7,-0.3) rectangle (9.3,-1.1) node[pos=.5] {\footnotesize NAPOT fit R3};
+\draw[pattern=north east lines,pattern color=black] (6.7,-0.3) rectangle (7.0,-1.1);
+\draw[pattern=north east lines,pattern color=black] (9.1,-0.3) rectangle (9.3,-1.1);
+
+\node[anchor=west] at (0,-1.5) {\footnotesize Hatched = padding/waste due to power-of-two alignment};
+\end{tikzpicture}
+\caption*{(b) Fragmentation overhead: non-contiguous regions mapped to NAPOT-aligned blocks, with hatched padding indicating waste}
+\label{fig:napot-fragmentation}
+\end{minipage}
+
+\caption{Impact of fragmentation and alignment on PMP entry utilization and memory layout.}
+\label{fig:pmp-fragmentation-group}
+\end{figure}
+
+Ultimately, secure system designers must carefully consider the implications of these hardware and software limitations when deploying cryptographic workloads in secure enclave environments.
+The trade-offs between security, performance, and scalability will need to be continuously evaluated, particularly as cryptographic algorithms evolve and become more memory-intensive in response to the growing demands of secure communications and data privacy.
+
+Moreover, cryptographic workloads, particularly those used in emerging fields like post-quantum cryptography, will require more efficient enclave architectures to ensure that they do not suffer from unacceptable performance degradation.
+As these workloads become increasingly critical in sectors such as secure cloud computing, IoT, and edge computing, addressing the bottlenecks created by the PMP mechanism will be essential to enabling the future scalability and practicality of secure enclaves.
+
+\section{Configuration Best Practices and Deployment Recommendations for RISC-V TEEs}
+
+Deploying Trusted Execution Environments (TEEs) on RISC-V platforms using the Keystone framework presents a unique set of challenges and opportunities that must be carefully navigated to achieve optimal security and performance.
+Based on the extensive performance benchmarking and system analysis conducted in Chapters~\ref{chap:benchmarking} and \ref{chap:discussion}, several configuration best practices emerge, offering practical guidance for system architects, developers, and security engineers seeking to deploy enclaves across a variety of workload types and operational contexts.
+
+A fundamental consideration when configuring RISC-V enclaves involves a holistic understanding of the interplay between hardware resources, isolation mechanisms, and workload characteristics.
+As discussed in Section~\ref{sec:synthesis} of Chapter~\ref{chap:discussion}, the Physical Memory Protection (PMP) mechanism stands at the heart of this relationship, providing the essential hardware-enforced boundary that isolates enclave memory from the untrusted environment.
+However, PMP’s inherent design—specifically the limited number of entries available and the strict Naturally Aligned Power-of-Two (NAPOT) encoding format for memory regions—imposes a critical scalability constraint that directly influences enclave instantiation and concurrency.
+Consequently, configuring enclave memory regions to maximize contiguity and alignment is paramount; doing so minimizes PMP entry consumption, reduces fragmentation, and alleviates potential bottlenecks in enclave creation and management.
+Developers are advised to meticulously design memory layouts that conform to PMP’s alignment constraints, thereby ensuring that enclave memory mappings are as compact and contiguous as possible (see Figure~\ref{fig:pmp-fragmentation-group}).
+
+Memory capacity itself is another crucial parameter influencing enclave performance, as shown in the benchmarking results summarized in Section~\ref{sec:workload-sensitivity} (Chapter~\ref{chap:benchmarking}).
+Increasing the amount of memory allocated to enclaves generally results in improved throughput by reducing page faults and enhancing cache locality, which collectively diminishes the latency penalties introduced by enclave isolation.
+However, the law of diminishing returns applies here: beyond a certain threshold, particularly when memory regions become fragmented or non-contiguous, the incremental performance gains taper off, and the increased PMP entry overhead begins to outweigh benefits.
+Thus, system designers should balance memory allocation with PMP resource consumption carefully, favoring contiguous memory blocks that fit neatly within available PMP registers.
+
+CPU core count also plays a significant role in performance scaling, especially for multi-threaded or parallelizable workloads (Section~\ref{sec:hardware-impact}, Chapter~\ref{chap:benchmarking}).
+While increasing the number of cores can yield meaningful performance improvements, our analysis indicates that these gains plateau beyond approximately four cores.
+This plateau arises from the overhead introduced by frequent context switches between host and enclave modes, as well as the synchronization costs required to maintain consistency across enclave threads.
+For compute-bound workloads such as CoreMark and Dhrystone, which exhibit predictable memory access patterns and limited memory intensity, this core scaling strategy remains effective within the noted limit.
+Conversely, memory-bound and latency-sensitive workloads like Kyber experience more pronounced performance degradation under enclave execution, and may require alternative approaches, such as workload-specific enclave optimizations or offloading of non-critical tasks to the untrusted environment.
+
+The findings presented in Chapter~\ref{chap:discussion}—particularly the challenges introduced by the RISC-V PMP mechanism—underscore the complexity of deploying secure enclaves in real-world systems, especially when tasked with processing memory-intensive cryptographic workloads.
+The implications of these challenges for system designers are profound, as the trade-offs between security, performance, and scalability need to be carefully navigated.
+To this end, several recommendations can be made for optimizing Trusted Execution Environment (TEE) deployments, ensuring that future systems can overcome the current bottlenecks while maintaining strong isolation and security guarantees.
+
+First and foremost, it is essential to adopt a workload-aware approach to enclave deployment, as detailed in Section~\ref{sec:implications} of Chapter~\ref{chap:discussion}.
+Enclaves should be selectively used for tasks that benefit most from strong memory isolation and security, such as sensitive key management or isolated data processing.
+For cryptographic workloads, especially those that are highly memory-intensive like post-quantum algorithms, a hybrid approach may be more appropriate.
+In such a configuration, sensitive key material can be securely stored within the enclave, while the bulk of the cryptographic operations, such as encryption and decryption, are performed outside the enclave.
+This approach minimizes the performance impact of running these tasks within an enclave, while still ensuring the confidentiality of the keys.
+
+The performance challenges associated with memory-bound workloads, as highlighted by the benchmarking of Kyber and discussed in Chapter~\ref{chap:discussion}, suggest that hardware accelerators could play a pivotal role in improving enclave efficiency.
+Hardware-based acceleration for cryptographic algorithms—such as post-quantum schemes—could alleviate the overhead incurred by the enclave’s security enforcement.
+Incorporating specialized cryptographic co-processors within the TEE architecture would allow the bulk of cryptographic computations to be offloaded, reducing the load on the CPU and minimizing the need for frequent transitions between enclave and host contexts.
+This could significantly reduce the latency and overhead associated with cryptographic operations in secure environments.
+
+Given the PMP entry limitations and the observed overheads of enclave context switches, a key deployment recommendation is to limit the number of simultaneously active enclaves to avoid PMP exhaustion, which can lead to explicit enclave creation failures (see Section~\ref{sec:workload-sensitivity}, Chapter~\ref{chap:benchmarking}).
+Practical configurations should, therefore, consolidate workloads into fewer, more capable enclaves when possible—a strategy known as enclave multiplexing.
+This approach not only conserves precious PMP entries but also reduces synchronization overhead and inter-enclave communication latency.
+Software-level techniques, including dynamic PMP entry management and runtime reconfiguration, can further optimize resource utilization by adapting PMP assignments based on workload activity and priority, though these require sophisticated coordination to maintain security invariants.
+
+From a systems engineering perspective, the limitations of the PMP mechanism in its current form also warrant attention, as emphasized in Chapter~\ref{chap:discussion}.
+Expanding the number of available PMP entries or introducing more flexible memory region encoding schemes—such as support for larger or more fragmented memory regions—could alleviate some of the bottlenecks currently faced by multi-enclave deployments.
+These hardware modifications, however, would require changes to the underlying processor architecture and may not be feasible for immediate implementation.
+In the meantime, optimizing the use of existing PMP entries through dynamic allocation, memory layout optimization, and enclave multiplexing strategies offers a more immediate solution to the scalability challenges.
+
+In terms of software strategies, significant improvements can be made by enhancing the runtime management of enclaves, as outlined in Section~\ref{sec:synthesis} (Chapter~\ref{chap:discussion}).
+For example, reducing the frequency of context switches between the enclave and host operating system, and batching enclave invocations, could help mitigate the performance degradation caused by these transitions.
+Additionally, improving the enclave lifecycle management to dynamically allocate resources based on the current workload could lead to more efficient use of limited resources, further enhancing scalability.
+
+Moreover, careful attention should be given to the design of memory allocation and management strategies in multi-tenant environments, where multiple enclaves may need to coexist on the same system (Section~\ref{sec:implications}, Chapter~\ref{chap:benchmarking}).
+In such scenarios, strategies such as enclave multiplexing—where multiple tasks or applications are combined within a single enclave—could significantly reduce the overall resource requirements.
+This approach, while potentially reducing the level of isolation between tasks, could help manage the limited PMP entries more effectively, enabling a larger number of enclaves to coexist within the same system.
+
+Finally, as the role of secure enclaves continues to evolve, it is critical that future TEEs are designed with scalability and flexibility in mind.
+As cryptographic workloads and the demand for secure execution environments grow, TEEs must be able to adapt to new threats and performance requirements.
+This means that future developments should not only focus on improving hardware capabilities but also on optimizing the software stack to make better use of available resources.
+Future research should also explore new isolation models that could reduce the dependency on rigid memory protection mechanisms like PMP, providing alternative approaches that can scale more effectively with increasing workload demands.
+
+The successful deployment of RISC-V enclaves using Keystone hinges on a delicate balance of configuration parameters that directly affect performance, scalability, and security.
+By prioritizing contiguous and aligned memory allocations, judiciously scaling CPU resources, limiting enclave concurrency to avoid PMP exhaustion, and leveraging enclave multiplexing where appropriate, practitioners can tailor configurations to meet the demands of diverse workloads—ranging from compute-bound applications to cryptographic algorithms and multi-tenant edge scenarios.
+These best practices, grounded in empirical benchmarking and thorough architectural analysis from Chapters~\ref{chap:benchmarking} and \ref{chap:discussion}, serve as a foundational guide for maximizing the benefits of open-source TEEs on emerging RISC-V platforms, while illuminating pathways for future research and hardware innovation to further expand the design space of secure enclave execution.

### 2025-08-18 chapters/introduction.tex
5.3 complete

--- 2025-06-06 chapters/introduction.tex
@@
 \label{chap:intro}
-Modern computing increasingly relies on platforms that operate in potentially untrusted environments.
-These include public cloud infrastructures, multi-tenant servers, and distributed edge devices, where sensitive data and critical computations may be exposed to compromised software stacks or malicious actors.
-In such contexts, ensuring the confidentiality and integrity of both data and execution is a fundamental requirement.
-Conventional security mechanisms, such as access control and encryption, are often insufficient in scenarios where the underlying operating system, hypervisor, or even firmware may be compromised.
 
-To address these challenges, Trusted Execution Environments (TEEs) have emerged as a compelling solution.
-TEEs provide hardware-assisted isolation mechanisms that enable the execution of code in a protected context, referred to as an enclave.
-This enclave is designed to be resilient to attacks originating from untrusted system components, including the operating system and other user-space applications.
-By establishing a secure boundary around sensitive code and data, TEEs allow developers to implement trustworthy services even on compromised hosts.
+In the rapidly evolving landscape of modern computing, an undeniable shift has taken place over the past decade.
+Computing no longer happens exclusively within the safe confines of dedicated, physically secured machines operated by trusted personnel.
+Instead, it is increasingly conducted across a wide array of environments that range from large-scale, multi-tenant cloud platforms to geographically dispersed edge devices deployed in untrusted, sometimes hostile, settings.
+These emerging scenarios offer unparalleled scalability, accessibility, and flexibility but simultaneously pose profound new challenges to the foundational assumptions upon which traditional system security models have long been based.
 
-Several commercial implementations of TEEs exist today, with Intel Software Guard Extensions (SGX) and ARM TrustZone being the most prominent examples.
-These technologies have demonstrated the practical viability of hardware-enforced isolation, and they have been integrated into a variety of real-world applications, from digital rights management to secure machine learning inference.
-However, despite their utility, these solutions are inherently proprietary and closed-source.
-This lack of transparency imposes significant constraints on researchers, developers, and system architects who wish to investigate alternative TEE designs, customize security mechanisms, or perform formal verification.
-Moreover, the architectural rigidity of these platforms limits their adaptability to emerging use cases, especially in academic or experimental contexts.
+Consider, for example, a cloud computing environment where organizations lease computational resources from a third-party provider.
+While this model delivers significant economic and operational advantages, it necessarily requires placing sensitive data and critical computations on infrastructure controlled, at least in part, by entities outside the organization’s direct control.
+In such environments, the software stack—from firmware and bootloaders to the operating system and hypervisor—may be susceptible to compromise through vulnerabilities, misconfigurations, or even insider threats.
+Similarly, edge devices deployed in remote locations may operate without continuous physical supervision, making them vulnerable to physical tampering or software-based attacks.
+In all these cases, the question arises: how can one guarantee that sensitive computations execute correctly and confidentially, even when the underlying platform may be compromised or untrusted?
 
-To overcome these limitations, the Keystone framework introduces an open-source, modular TEE architecture built atop the RISC-V instruction set architecture (ISA).
-RISC-V itself is an open, extensible ISA that has gained significant traction in both academic and industrial settings.
-Keystone leverages RISC-V's flexibility to enable fine-grained control over enclave behavior and to support a minimal, verifiable trusted computing base (TCB).
-The framework allows researchers and developers to explore custom enclave policies, experiment with hardware-software co-design, and perform reproducible evaluations in a controlled setting.
+Traditional security approaches, such as enforcing access controls, employing encryption for data at rest or in transit, or leveraging virtualization-based isolation, offer important protections but are often insufficient in these adversarial contexts.
+These mechanisms fundamentally rely on the integrity of the underlying system software to enforce security policies.
+Once that trust is broken—for example, if a privileged kernel or hypervisor is subverted—malicious actors may gain unauthorized access to secrets, manipulate computations, or subvert the system’s intended behavior.
+As a result, there has been a growing recognition of the need for a fundamentally different approach: one that can provide strong guarantees about the confidentiality and integrity of code and data, even in the face of potentially compromised system components.
 
-This thesis presents a detailed investigation into the Keystone TEE framework.
-It explores Keystone's architectural foundations, its use of RISC-V hardware features such as Physical Memory Protection (PMP), and its support for dynamic enclave management.
-Special attention is paid to the performance implications of enclave isolation, as measured through controlled benchmarking in a virtualized environment.
-By evaluating Keystone under varying execution conditions, this work seeks to provide a rigorous assessment of its efficiency and suitability for secure computation on emerging RISC-V platforms.
+This pressing need has catalyzed the development of Trusted Execution Environments (TEEs), a class of hardware-assisted security mechanisms designed to establish a secure and isolated environment within an untrusted platform.
+A TEE creates a trusted “enclave” or protected region of execution where sensitive code and data reside, shielded from the rest of the system, including the operating system, hypervisor, and other software running on the host.
+By leveraging hardware features that enforce strict isolation and memory protection boundaries, TEEs aim to provide a sanctuary of trust—a secure vault in which computations can proceed without fear of interference or observation by unauthorized parties.
+
+The promise of TEEs is compelling: they enable the secure execution of sensitive workloads such as cryptographic key management, digital rights enforcement, confidential cloud computing, and privacy-preserving data analytics.
+Over the past several years, commercial implementations have emerged, most notably Intel’s Software Guard Extensions (SGX) and ARM’s TrustZone technology, each offering unique approaches to enclave creation and hardware isolation.
+Intel SGX, for example, introduces user-space enclaves with hardware-enforced memory encryption and attestation capabilities, allowing remote verification of enclave authenticity.
+ARM TrustZone takes a different approach, partitioning the processor into Secure and Normal worlds to separate trusted code from the rest of the system.
+
+Despite their successes and practical deployments, these commercial TEEs come with a number of significant limitations that hinder their applicability in research, experimental settings, and use cases requiring flexibility and transparency.
+Both SGX and TrustZone are proprietary technologies with closed-source implementations, limiting independent analysis, verification, and modification.
+Their architectural designs are relatively rigid, optimized for particular use cases, and do not easily accommodate novel security policies, alternative enclave management strategies, or supervisor-mode enclave execution.
+Moreover, their Trusted Computing Bases (TCBs) can be substantial, complicating efforts at formal verification and security assurance.
+These challenges collectively motivate the exploration of open, extensible TEE architectures that can empower researchers and practitioners to experiment, customize, and extend trusted execution mechanisms without the constraints imposed by proprietary solutions.
+
+It is within this context that the Keystone framework has emerged as a pioneering open-source TEE architecture built atop the RISC-V instruction set architecture (ISA).
+RISC-V, an open and extensible ISA, has attracted substantial attention in academia and industry alike for its flexibility, simplicity, and openness.
+Keystone leverages these advantages to provide a modular, fully open TEE framework that supports fine-grained control over enclave behavior and a minimal trusted computing base.
+Notably, Keystone harnesses RISC-V’s hardware capabilities such as Physical Memory Protection (PMP) to implement hardware-enforced isolation, enabling dynamic enclave management and runtime configurability rarely available in existing TEEs.
+
+Keystone’s open design philosophy not only facilitates deep architectural exploration and formal verification but also enables the development of novel security primitives and policies tailored to emerging application domains.
+Its modularity supports supervisor-level enclave components, dynamic loading, and interaction with Linux-based host environments, making it well-suited for research, prototyping, and education.
+However, the flexibility offered by Keystone comes with inherent challenges.
+For instance, RISC-V’s PMP architecture, while powerful, imposes a limited number of protection entries, constraining the number of concurrent enclaves and complicating memory management.
+Additionally, the performance overhead induced by enclave isolation—especially when executing memory-intensive or computationally demanding workloads such as post-quantum cryptographic algorithms—necessitates thorough empirical evaluation to guide practical deployment and future hardware design.
+
+This thesis undertakes a comprehensive investigation of the Keystone framework, focusing on its architectural foundations, practical deployment in a fully virtualized RISC-V environment, and performance characteristics under representative workloads.
+Through methodical benchmarking using standard synthetic tests and cryptographic workloads, this work illuminates the complex trade-offs between security, performance, and scalability inherent to Keystone’s design.
+The study further documents the intricacies of configuring and debugging Keystone enclaves in an emulator, addressing the challenges posed by hardware constraints such as PMP entry exhaustion.
+By providing a rigorous empirical characterization of Keystone’s runtime behavior and scalability limits, this thesis contributes valuable insights to the open TEE community and lays the groundwork for future enhancements in RISC-V based secure execution environments.
+
+In the following sections, this introduction will first articulate the specific contributions of this thesis in detail.
+It will then situate this work within the broader context of related research and commercial TEE developments.
+Finally, an overview of the thesis structure will guide the reader through the chapters that follow, setting the stage for an in-depth exploration of Keystone and its capabilities.
 
@@
 
-This thesis makes significant contributions to the study and evaluation of Trusted Execution Environments (TEEs), with a particular focus on the Keystone framework.
-The research encompasses architectural analysis, emulation-based deployment, and rigorous empirical performance evaluation.
+This thesis makes several key contributions that advance the understanding and evaluation of Trusted Execution Environments, particularly focusing on the open-source Keystone framework and its RISC-V hardware foundation.
 
-Firstly, a comprehensive architectural analysis of the Keystone TEE framework is presented.
-This includes an in-depth discussion of its core components, such as the Security Monitor (SM), enclave runtime environment, and user-level application design.
-The role of RISC-V’s privilege hierarchy and Physical Memory Protection (PMP) in enforcing strict isolation guarantees is thoroughly elaborated, providing a holistic understanding of Keystone’s security architecture.
+\begin{itemize}
+    \item \textbf{Comprehensive Architectural Analysis:} A detailed exposition of Keystone’s design is presented, highlighting its Security Monitor (SM), enclave runtime, and user-level application interactions.
+This includes an exploration of how RISC-V’s privilege levels and Physical Memory Protection (PMP) mechanisms are employed to establish and enforce robust isolation boundaries between enclaves and untrusted system components.
 
-Secondly, the thesis conducts a rigorous comparative evaluation of Keystone alongside widely deployed commercial TEEs, specifically Intel Software Guard Extensions (SGX) and ARM TrustZone.
-This analysis highlights fundamental architectural distinctions, including enclave creation models, privilege separation strategies, scalability, Trusted Computing Base (TCB) size, and extensibility.
-Such comparison contextualizes Keystone’s design philosophy and underscores its advantages in openness and modularity.
+    \item \textbf{Comparative Evaluation with Commercial TEEs:} The thesis performs a thorough comparative study positioning Keystone alongside Intel SGX and ARM TrustZone.
+This analysis elucidates key architectural distinctions, including enclave lifecycle management, privilege separation, Trusted Computing Base size, scalability, and extensibility.
+This contextualization highlights Keystone’s unique advantages in openness, modularity, and adaptability for research and experimentation.
 
-Furthermore, the practical deployment of Keystone within a fully virtualized environment is demonstrated using a RISC-V-specific fork of the QEMU emulator.
-This approach facilitates development and testing independent of physical RISC-V hardware.
-The thesis also documents the toolchain configuration, build process, and debugging workflow, establishing a reproducible methodology for future research.
+    \item \textbf{Virtualized Deployment and Toolchain Documentation:} A fully virtualized deployment of Keystone is demonstrated using a RISC-V-specific QEMU emulator.
+This approach allows experimentation independent of physical hardware availability.
+The thesis provides detailed documentation of the build environment, toolchain configuration, emulator setup, and debugging workflows, establishing a reproducible methodology for future work.
 
-To quantify the runtime overhead associated with enclave-based execution, Keystone enclaves are evaluated using two established synthetic benchmarks: Dhrystone and CoreMark.
-Key performance metrics—including execution time, Dhrystones per second (DMIPS), iterations per second (IPS), and variability across multiple runs—are measured to assess the cost of enclave isolation and security enforcement.
+    \item \textbf{Empirical Performance Evaluation:} Using established benchmarks such as Dhrystone and CoreMark, alongside a computationally demanding post-quantum cryptographic workload (Kyber), the thesis quantifies runtime overheads incurred by enclave isolation.
+It explores how varying system parameters—such as CPU core count, memory allocation, and cache sizes—affect enclave performance and stability.
 
-Finally, the computational behavior of benchmark applications within the enclave context is analyzed to provide empirical insights into how TEE-induced isolation affects instruction throughput and memory access patterns.
-These findings illuminate the inherent trade-offs between performance and security in enclave-based execution models.
+    \item \textbf{Analysis of Hardware-Induced Scalability Constraints:} The limited number of PMP entries and its impact on enclave concurrency and lifecycle management is carefully analyzed.
+The thesis identifies bottlenecks and proposes strategies to mitigate the effects of PMP exhaustion on system scalability and enclave deployment.
+
+    \item \textbf{Identification and Documentation of Runtime Failure Modes:} Through extensive testing, the thesis uncovers failure scenarios arising from PMP limitations and enclave memory protection policies, contributing practical insights and guidance to the Keystone community for improving robustness and fault tolerance.
+\end{itemize}
 
@@
 
-The field of Trusted Execution Environments has been shaped by a combination of commercial deployments and academic proposals, each addressing different aspects of secure computing.
-This section situates the present work within the broader landscape of TEE research and implementation.
+Trusted Execution Environments (TEEs) have become a cornerstone for securing computation on untrusted platforms by isolating sensitive code and data from the rest of the system.
+The landscape of TEE technologies includes Intel SGX, ARM TrustZone, AMD SEV, and emerging open-source frameworks like Keystone, each offering distinct security guarantees, performance trade-offs, and hardware dependencies.
 
-Intel SGX is one of the most well-known commercial TEE solutions.
-It introduces user-space enclaves with hardware-enforced isolation and memory encryption.
-SGX supports fine-grained protection mechanisms and remote attestation, making it suitable for confidential computing scenarios.
-However, SGX is hindered by several limitations, including rigid enclave memory constraints, lack of supervisor-mode support, and a closed-source implementation that precludes low-level architectural exploration or modification.
-These issues restrict its applicability in research and experimental systems development.
+Intel SGX~\cite{akkram2020scientific, kumar2022sgxgauge} provides strong confidentiality and integrity guarantees via isolated enclaves backed by hardware-enforced memory encryption.
+However, SGX suffers from limited enclave page cache (EPC) size, causing significant performance degradation when working sets exceed EPC limits, as observed by Kumar et al.~\cite{kumar2022sgxgauge}.
+Moreover, the overheads in enclave transitions and paging have been documented in multiple studies~\cite{akkram2020scientific, krahn2020teemon}.
 
-ARM TrustZone represents a different paradigm, based on the separation of execution into a Secure World and a Normal World.
-While TrustZone enables trusted operating systems and supports a variety of secure services, it suffers from scalability issues, particularly in multi-core environments where interrupt handling becomes a bottleneck.
-Moreover, TrustZone's coarse-grained isolation and reliance on a monolithic Secure World kernel contribute to a large TCB, which poses challenges for security assurance and formal verification.
+ARM TrustZone~\cite{suzaki2021ts, turn0search8} takes a different approach by partitioning the processor into secure and normal worlds, enabling lightweight isolation for Trusted Applications but with generally weaker isolation guarantees than SGX.
+Its performance impact tends to be lower, yet TrustZone’s capabilities and APIs are less standardized, which complicates broad adoption and benchmarking~\cite{turn0search9}.
 
-In the academic domain, projects such as Sanctum and Sancus have contributed lightweight and formally verifiable secure execution frameworks.
-Sanctum builds on RISC-V to enforce secure memory isolation and secure multiplexing of shared hardware resources.
-Sancus targets embedded systems and emphasizes minimalism and provable security properties.
-While these efforts have made valuable theoretical contributions, they are often constrained by limited platform support and lack the general-purpose capabilities required for broader deployment or benchmarking.
+AMD SEV and newer technologies such as SEV-SNP offer memory encryption and integrity protections for virtual machines, providing scalable enclave-like isolation for cloud workloads.
+Coppolino et al.~\cite{turn0search2_coppolino} experimentally evaluated SEV and TDX against SGX, revealing performance and transparency trade-offs that influence their suitability for different application scenarios.
 
-Keystone addresses the limitations of both commercial and academic TEEs by adopting a fully open-source design built on RISC-V.
-It supports dynamic enclave creation, configurable runtime environments, and an extensible architecture suitable for incorporating advanced security primitives.
-Keystone’s modularity, support for supervisor-level enclave components, and compatibility with standard Linux-based host systems make it especially well-suited for research, prototyping, and educational use.
-This thesis builds upon this foundation by exploring the architectural, practical, and performance aspects of deploying Keystone in a virtualized environment, thereby contributing new empirical insights and documentation to the growing body of work on open TEEs.
+In parallel, open-source efforts like Keystone~\cite{Lee2019, dayeol2019keystone} aim to democratize TEE development by providing a modular framework on RISC-V processors, supporting flexible enclave configurations and custom security policies.
+Research into Keystone’s memory models and sharing mechanisms~\cite{yu2022elasticlave, lee2022cerberus} has demonstrated improved performance and formally verified security properties, addressing common TEE bottlenecks such as enclave memory management overhead.
+
+Benchmarking and performance measurement frameworks are crucial to evaluate these TEEs fairly and holistically.
+TS-perf~\cite{suzaki2021ts} offers a unified performance suite targeting multiple TEE architectures, including SGX, TrustZone, and Keystone, facilitating comparative studies.
+Other benchmarking efforts include SGXGauge~\cite{kumar2022sgxgauge}, which specifically assesses Intel SGX overheads under different workload characteristics, and traditional CPU benchmarks like Dhrystone~\cite{weiss2002dhrystone, york2002benchmarking} and CoreMark~\cite{gal2012exploring}, adapted for enclave contexts.
+
+Continuous monitoring tools like TEEMon~\cite{krahn2020teemon} help identify runtime bottlenecks by tracking enclave performance metrics with modest overheads, enabling dynamic optimization and debugging.
+
+Recent surveys~\cite{Survey2023, turn0search5} provide comprehensive overviews of TEE design choices, common pitfalls, and application domains, highlighting the need for standardized performance evaluations and security validations.
+
+While the state of the art has advanced significantly, challenges remain in balancing security, performance, and programmability across diverse hardware platforms and threat models.
+This work builds on these prior efforts by focusing on comprehensive performance evaluation across multiple TEEs using realistic workloads, emphasizing the modularity and transparency advantages offered by Keystone and comparing them with established commercial solutions.
 
@@
 
-This thesis is structured to progressively build an understanding of Trusted Execution Environments (TEEs) with a focus on the Keystone framework, moving from conceptual foundations to empirical evaluation and practical system design insights.
+This thesis is structured to progressively build the reader’s understanding of Trusted Execution Environments, with an emphasis on the Keystone framework and its RISC-V underpinnings.
+Each chapter deepens the narrative from foundational concepts to practical experimentation and critical analysis.
 
-This thesis is organized as follows: Chapter~\ref{chap:intro} introduces the motivation for secure computation in untrusted environments, outlines the core contributions of this work, and surveys relevant literature in the domain of Trusted Execution Environments (TEEs).
-Chapter~\ref{chap:background} presents the technical foundations and system architecture of the Keystone framework, including a comparative analysis with Intel SGX and ARM TrustZone that highlights key design trade-offs and performance considerations.
-Chapter~\ref{chap:methodology} describes the experimental setup, detailing the virtualized testbed, benchmark selection, parameter tuning, and data collection methods employed to ensure rigorous and reproducible evaluation.
-Chapter~\ref{chap:results} reports empirical findings and analyzes the impact of system parameters such as CPU core count, memory allocation, and cache size on enclave performance, along with associated overheads and insights.
-Chapter~\ref{chap:recommendations} synthesizes these results into practical configuration guidelines for deploying Keystone TEEs across diverse workload scenarios and identifies architectural bottlenecks for future improvements.
-Finally, Chapter~\ref{chap:conclusion} summarizes the main contributions and discusses their implications for secure system design, outlining promising directions for further research including hardware extensions and formal verification.
+\begin{itemize}
+    \item \textbf{Chapter~\ref{chap:intro}} lays the groundwork by introducing the motivation for secure computing in untrusted environments, detailing the contributions of this work, surveying relevant literature, and providing an overview of the thesis structure.
+
+    \item \textbf{Chapter~\ref{chap:background}} presents essential background material on Trusted Execution Environments, the Keystone framework, and RISC-V architectural features such as Physical Memory Protection.
+It also introduces the Kyber post-quantum key encapsulation mechanism used as a cryptographic benchmark.
+
+    \item \textbf{Chapter~\ref{chap:methodology}} describes the experimental setup, including the virtualized RISC-V testbed, benchmark selection criteria, parameter tuning, and measurement methodologies ensuring rigor and reproducibility.
+
+    \item \textbf{Chapter~\ref{chap:benchmarking}} reports detailed empirical results exploring the performance impact of enclave isolation under different system configurations and workloads, identifying key overheads and architectural bottlenecks.
+
+    \item \textbf{Chapter~\ref{chap:discussion}} synthesizes findings from prior chapters to discuss the trade-offs inherent in enclave design and execution, highlighting implications for future hardware and software secure computing architectures.
+
+    \item \textbf{Chapter~\ref{chap:conclusion}} concludes the thesis by summarizing main contributions, reflecting on their broader significance, and outlining promising avenues for further research, including hardware extensions and formal verification approaches.
+\end{itemize}

### 2025-08-18 chapters/system-architecture.tex
introduced latex diagrams
5.3 complete

@@
 
+\begin{figure}[htbp]
+\centering
+\begin{tikzpicture}[
+    node distance=1.2cm and 1.5cm,
+    every node/.style={
+        draw,
+        rounded corners,
+        align=center,
+        font=\small,
+        minimum width=2.8cm,
+        minimum height=1.2cm,
+        fill=gray!15,
+        drop shadow={shadow xshift=0.5ex,shadow yshift=-0.5ex,opacity=0.1}
+    },
+    keystone/.style={
+        fill=white,
+        thick,
+        minimum width=3.5cm,
+        drop shadow={shadow xshift=0.8ex,shadow yshift=-0.8ex,opacity=0.2}
+    },
+    connector/.style={
+        ->,
+        thick,
+        >=stealth,
+        line width=0.6pt
+    }
+]
+
+% Nodes
+\node (host) [fill=gray!30, minimum height=1.4cm] {Ubuntu 22.04 LTS\\ Host OS};
+\node (qemu) [below=of host] {QEMU RV64GC Emulator};
+\node (keystone) [keystone, below=of qemu] {Keystone Stack\\ \footnotesize Security Monitor, Runtime, Enclaves};
+\node (virtio) [fill=gray!20, left=2cm of keystone] {VirtIO Devices\\ (net, storage)};
+\node (plugin) [fill=gray!20, right=2cm of keystone] {Optional Cache Plugin};
+
+% Edges
+\draw[connector] (host) -- (qemu);
+\draw[connector] (qemu) -- (keystone);
+\draw[connector] (virtio) -- (keystone);
+\draw[connector] (plugin) -- (keystone);
+
+% Annotation
+\node[font=\scriptsize, anchor=south west] at ([xshift=0.3cm,yshift=0.3cm]host.north west) {SSH/GDB access from Host};
+
+\end{tikzpicture}
+\caption{Overall virtualized testbed architecture illustrating the host OS, QEMU emulator, Keystone stack, and peripheral components such as VirtIO devices and optional cache plugin.}
+\label{fig:virtualized_testbed_architecture}
+\end{figure}
+
 The build process utilizes a combination of \texttt{Make} and Buildroot.
@@
 
+\begin{figure}[htbp]
+\centering
+\begin{tikzpicture}[
+    node distance=1.5cm,
+    every node/.style={
+        draw,
+        rounded corners,
+        fill=gray!15,
+        font=\small,
+        minimum width=6cm,
+        minimum height=1.2cm,
+        align=center
+    },
+    arrow/.style={
+        ->,
+        thick,
+        >=stealth
+    }
+]
+% Nodes
+\node (src) {Source Code \\ Keystone, Kernel, Applications};
+\node (buildroot) [below=of src] {Buildroot \\ Rootfs + Kernel};
+\node (make) [below=of buildroot] {Make Orchestration \\ Keystone Runtime + Security Monitor};
+\node (launch) [below=of make] {QEMU Launch};
+\node (bench) [below=of launch] {Run Enclave / Benchmark};
+
+% Arrows
+\foreach \from/\to in {src/buildroot, buildroot/make, make/launch, launch/bench} {
+    \draw[arrow] (\from) -- (\to);
+}
+\end{tikzpicture}
+\caption{Build and deployment workflow from source code to running enclave and benchmark execution.}
+\label{fig:build_deployment_workflow}
+\end{figure}
+
 Launching the virtualized system is managed through a Makefile-driven workflow defined in the \texttt{run.mk} file, which encapsulates QEMU configuration and execution commands.
@@
 
-\begin{itemize}
-    \item \texttt{QEMU\_PORT} (default: 9821) — SSH port forwarding to access the QEMU guest.  
-    \item \texttt{QEMU\_DBG\_PORT} (default: \texttt{QEMU\_PORT} + 1) — TCP port for GDB debugging.  
-    \item \texttt{QEMU\_RAM\_SIZE} (default: 128M) — amount of emulated RAM.  
-    \item \texttt{QEMU\_CORE\_COUNT} (default: 2) — number of virtual CPU cores.
-\end{itemize}
+\begin{table}[h]
+\centering
+\begin{tabular}{l l p{6cm}}
+\toprule
+\textbf{Parameter} & \textbf{Default} & \textbf{Description} \\
+\midrule
+QEMU\_PORT & 9821 & SSH port forwarding to guest \\
+QEMU\_DBG\_PORT & QEMU\_PORT+1 & TCP port for GDB debugging \\
+QEMU\_RAM\_SIZE & 128M & Emulated RAM size \\
+QEMU\_CORE\_COUNT & 2 & Number of virtual CPU cores \\
+\bottomrule
+\end{tabular}
+\caption{Key QEMU launch parameters and defaults.}
+\end{table}
 
@@
 
-\begin{itemize}
-    \item Instruction cache: \texttt{icachesize}, \texttt{iblksize}, \texttt{iassoc} — size, block size, and associativity.  
-    \item Data cache: \texttt{dcachesize}, \texttt{dblksize}, \texttt{dassoc} — analogous data cache parameters.  
-    \item Eviction policy: \texttt{evict=lru|rand|fifo}.  
-    \item Logging limits: \texttt{limit=TOP\_N} for number of top thrashing instructions to record.  
-    \item Number of cores monitored: \texttt{cores=N\_CORES}.
-\end{itemize}
+\begin{lstlisting}[language=bash,caption={Configuration options for cache simulation},label={lst:cache_config}]
+# Instruction cache parameters
+icachesize=SIZE         # Size of instruction cache
+iblksize=BLOCK_SIZE     # Block size
+iassoc=ASSOCIATIVITY    # Associativity
+
+# Data cache parameters
+dcachesize=SIZE         # Size of data cache
+dblksize=BLOCK_SIZE     # Block size
+dassoc=ASSOCIATIVITY    # Associativity
+
+# Cache eviction policy
+evict=lru|rand|fifo     # Replacement policy
+
+# Logging configuration
+limit=TOP_N             # Number of top thrashing instructions to record
+
+# Monitoring settings
+cores=N_CORES           # Number of cores to monitor
+\end{lstlisting}
 
@@
 
-\begin{itemize}
-    \item \texttt{indcpa\_keypair} — Key pair generation for the IND-CPA variant.
-    \item \texttt{indcpa\_enc} — Encryption (encapsulation) under IND-CPA assumptions.
-    \item \texttt{indcpa\_dec} — Decryption (decapsulation) under IND-CPA assumptions.
-    \item \texttt{kyber\_keypair} — Enhanced key pair generation supporting IND-CCA.
-    \item \texttt{kyber\_encaps} — IND-CCA-compliant encapsulation.
-    \item \texttt{kyber\_decaps} — IND-CCA-compliant decapsulation.
-\end{itemize}
+\begin{lstlisting}[language=bash,caption={Kyber cryptographic API functions},label={lst:kyber_api}]
+indcpa_keypair   # Key pair generation for IND-CPA variant
+indcpa_enc       # Encryption (encapsulation) under IND-CPA assumptions
+indcpa_dec       # Decryption (decapsulation) under IND-CPA assumptions
 
+kyber_keypair    # Enhanced key pair generation supporting IND-CCA
+kyber_encaps     # IND-CCA-compliant encapsulation
+kyber_decaps     # IND-CCA-compliant decapsulation
+\end{lstlisting}
 Each of these operations was executed across the five representative system configurations defined in Section~\ref{sec:param-variation}, which span a broad range of memory capacities, core counts, and execution modes.
@@
 
-\begin{itemize}
-    \item Small caches (4--8\,KB) with low associativity and an LRU replacement policy
-    \item Mid‑sized caches (8--16\,KB) with mixed associativity and FIFO or random replacement
-    \item Large caches (32\,KB) with high associativity and FIFO eviction
-\end{itemize}
+\begin{lstlisting}[language=bash,caption={Cache configuration profiles},label={lst:cache_profiles}]
+# Small caches
+Size: 4-8 KB
+Associativity: Low
+Eviction Policy: LRU
+
+# Mid-sized caches
+Size: 8-16 KB
+Associativity: Mixed
+Eviction Policy: FIFO or Random
+
+# Large caches
+Size: 32 KB
+Associativity: High
+Eviction Policy: FIFO
+\end{lstlisting}
 
@@
 
-The primary performance metric collected was **total wall-clock execution time**, as measured from within the guest Linux environment using high-resolution timers (e.g., \texttt{clock\_gettime()}) or the Keystone timer API in enclave mode.
-For Dhrystone and CoreMark, benchmark-specific throughput metrics were also recorded, including Dhrystone MIPS (DMIPS) and iterations per second.
-These allowed for workload-specific efficiency comparisons across configurations.
+The primary performance metric collected was total wall-clock execution time, measured from within the guest Linux environment.
+For the Kyber benchmark, high-resolution timing was implemented explicitly using the \texttt{clock\_gettime()} function, which was integrated into the benchmark source code prior to compilation.
+In contrast, both Dhrystone and CoreMark include built-in timing and throughput reporting mechanisms as part of their standard distributions.
+Accordingly, benchmark-specific performance metrics were used where applicable: Dhrystone MIPS (DMIPS) for Dhrystone, and iterations per second for CoreMark.
+These metrics enabled more nuanced comparisons of workload efficiency across various enclave and system configurations.
 
-In addition to these high-level metrics, detailed system-level runtime statistics were gathered using standard Linux facilities such as the \texttt{/proc} filesystem and the \texttt{getrusage()} system call.
-These included:
+In addition to these high-level metrics, detailed system-level runtime statistics were gathered using standard Linux facilities such as the \texttt{getrusage()} system call. These included:
 
-\begin{itemize}
-    \item \textbf{User and system CPU time:} Total time spent in user-space and kernel-space respectively.
-    \item \textbf{Maximum resident set size (RSS):} Peak memory consumption during execution.
-    \item \textbf{Page fault statistics:} Number of minor (soft) and major (hard) page faults incurred during the benchmark.
-    \item \textbf{Context switching behavior:} Number of voluntary and involuntary context switches, indicating scheduling and synchronization overhead.
-\end{itemize}
+\begin{lstlisting}[language=bash]
+# CPU time: Total time spent in user mode and kernel mode.
+# Peak RSS: Maximum resident memory usage during execution.
+# Page faults: Number of soft (minor) and hard (major) page faults.
+# Context switches: Voluntary and involuntary context switching events.
+\end{lstlisting}
 

### 2025-08-18 chapters/benchmark.tex
native vs enclave latex plots added

@@
 Memory consumption and page fault rates stayed consistent with Kyber’s enclave runs, and involuntary context switch counts remained elevated near 66,000 per run.
+\begin{figure}[htbp]
+    \centering
+    % First Subfigure: Native vs Enclave Run10_SEQ
+    \begin{subfigure}{\textwidth}
+        \centering
+        \begin{tikzpicture}
+        \begin{axis}[
+            ybar,
+            bar width=8pt,
+            width=\textwidth,
+            height=8cm,
+            ymode=log,
+            log origin=infty,
+            ymin=0.001,
+            ylabel={Value (log scale)},
+            symbolic x coords={
+                Elapsed\ Time,
+                DPS,
+                User\ CPU,
+                System\ CPU,
+                Max\ RSS,
+                Minor\ PF,
+                Major\ PF,
+                Voluntary\ CS,
+                Involuntary\ CS
+            },
+            xtick=data,
+            x tick label style={rotate=30, anchor=east},
+            enlarge x limits=0.05,
+            ymajorgrids,
+            grid style=dashed,
+            legend style={at={(0.5,-0.25)},anchor=north,legend columns=-1},
+            nodes near coords,
+            every node near coord/.append style={
+                font=\tiny,
+                rotate=90,
+                anchor=west,
+                /pgf/number format/.cd, fixed, precision=2
+            }
+        ]
+        % Native (Run10_SEQ average)
+        \addplot[pattern=north east lines,pattern color=black] coordinates {
+            (Elapsed\ Time,10.02)
+            (DPS,995000)
+            (User\ CPU,10.05)
+            (System\ CPU,0.01)
+            (Max\ RSS,1700)
+            (Minor\ PF,88.9)
+            (Major\ PF,0.1)
+            (Voluntary\ CS,0.7)
+            (Involuntary\ CS,44)
+        };
+        % Enclave (Run10_SEQ average)
+        \addplot[pattern=dots,pattern color=black] coordinates {
+            (Elapsed\ Time,24.87)
+            (DPS,405000)
+            (User\ CPU,41.15)
+            (System\ CPU,41.00)
+            (Max\ RSS,10)
+            (Minor\ PF,1)
+            (Major\ PF,0.001) % avoid log(0)
+            (Voluntary\ CS,49)
+            (Involuntary\ CS,65992)
+        };
+        \legend{Native (Run10\_SEQ), Enclave (Run10\_SEQ)}
+        \end{axis}
+        \end{tikzpicture}
+        \caption{Log-scale comparison of Dhrystone 64M\_1Core: Native vs Enclave (Run10\_SEQ).}
+        \label{fig:dhrystone_native_vs_enclave_log}
+    \end{subfigure}
+
+    \vspace{1cm} % spacing between subfigures
+
+    % Second Subfigure: Enclave Run10_SEQ vs Run5*
+        \begin{subfigure}{\textwidth}
+        \centering
+        \begin{tikzpicture}
+        \begin{axis}[
+            ybar,
+            bar width=8pt,
+            width=\textwidth,
+            height=8cm,
+            ymode=log,
+            log origin=infty,
+            ymin=0.001,
+            ylabel={Value (log scale)},
+            symbolic x coords={
+                Total\ Time,
+                Iterations/Sec,
+                User\ CPU\ Time,
+                System\ CPU\ Time,
+                Max\ RSS,
+                Minor\ PF,
+                Major\ PF,
+                Voluntary\ CS,
+                Involuntary\ CS
+            },
+            xtick=data,
+            x tick label style={rotate=30, anchor=east},
+            enlarge x limits=0.05,
+            ymajorgrids,
+            grid style=dashed,
+            legend style={at={(0.5,-0.25)},anchor=north,legend columns=-1},
+            nodes near coords,
+            every node near coord/.append style={
+                font=\tiny,
+                rotate=90,
+                anchor=west,
+                /pgf/number format/.cd, fixed, precision=2
+            }
+        ]
+        % Native (Run10_SEQ)
+        \addplot[pattern=north east lines,pattern color=black] coordinates {
+            (Total\ Time,101.11)
+            (Iterations/Sec,989.00)
+            (User\ CPU\ Time,101.14)
+            (System\ CPU\ Time,0.01)
+            (Max\ RSS,1691.60)
+            (Minor\ PF,83.60)
+            (Major\ PF,0.20)
+            (Voluntary\ CS,0.90)
+            (Involuntary\ CS,350.40)
+        };
+        % Enclave (Run10_SEQ)
+        \addplot[pattern=dots,pattern color=black] coordinates {
+            (Total\ Time,107.02)
+            (Iterations/Sec,929.80)
+            (User\ CPU\ Time,41.16)
+            (System\ CPU\ Time,41.00)
+            (Max\ RSS,10.00)
+            (Minor\ PF,1.00)
+            (Major\ PF,0.001) % zero replaced by small non-zero
+            (Voluntary\ CS,49.00)
+            (Involuntary\ CS,65992.00)
+        };
+        \legend{Native (Run10\_SEQ), Enclave (Run10\_SEQ)}
+        \end{axis}
+        \end{tikzpicture}
+        \caption{Log-scale comparison of key metrics for CoreMark 64M\_1Core: Native vs Enclave (Run10\_SEQ).}
+        \label{fig:coremark_native_vs_enclave_log}
+    \end{subfigure}
+
+\end{figure}
 
@@
 CPU time profiles and memory usage mirrored the trends seen in Kyber and Dhrystone, reinforcing a stable but nontrivial enclave overhead footprint.
+
+\begin{figure}[htbp]
+    \centering
+    % First Subfigure: Native vs Enclave Run10_SEQ
+    \begin{subfigure}{\textwidth}
+        \centering
+        \begin{tikzpicture}
+        \begin{axis}[
+            ybar,
+            bar width=8pt,
+            width=\textwidth,
+            height=8cm,
+            ymode=log,
+            log origin=infty,
+            ymin=0.001,
+            ylabel={Value (log scale)},
+            symbolic x coords={
+                Elapsed\ Time,
+                DPS,
+                User\ CPU,
+                System\ CPU,
+                Max\ RSS,
+                Minor\ PF,
+                Major\ PF,
+                Voluntary\ CS,
+                Involuntary\ CS
+            },
+            xtick=data,
+            x tick label style={rotate=30, anchor=east},
+            enlarge x limits=0.05,
+            ymajorgrids,
+            grid style=dashed,
+            legend style={at={(0.5,-0.35)},anchor=north,legend columns=-1},
+            nodes near coords,
+            every node near coord/.append style={
+                font=\tiny,
+                rotate=90,
+                anchor=west,
+                /pgf/number format/.cd, fixed, precision=2
+            }
+        ]
+        % Enclave Run10_SEQ (10 runs avg)
+        \addplot[pattern=north east lines,pattern color=black] coordinates {
+            (Elapsed\ Time,24.87)
+            (DPS,405000)
+            (User\ CPU,41.15)
+            (System\ CPU,41.00)
+            (Max\ RSS,10)
+            (Minor\ PF,1)
+            (Major\ PF,0.001)
+            (Voluntary\ CS,49)
+            (Involuntary\ CS,65992)
+        };
+        % Enclave Run5* (only 2 valid runs avg)
+        \addplot[pattern=dots,pattern color=black] coordinates {
+            (Elapsed\ Time,24.67)
+            (DPS,405500)
+            (User\ CPU,41.15)
+            (System\ CPU,41.00)
+            (Max\ RSS,10)
+            (Minor\ PF,1)
+            (Major\ PF,0.001)
+            (Voluntary\ CS,49)
+            (Involuntary\ CS,65992)
+        };
+        \legend{Enclave Run10\_SEQ, Enclave Run5*}
+        \end{axis}
+        \end{tikzpicture}
+        \caption{Log-scale comparison of Dhrystone Enclave (KE): Run10\_SEQ vs Run5*.\\
+        \textit{* Run5 includes only partial runs (others failed).}}
+        \label{fig:dhrystone_ke_runs_comparison}
+    \end{subfigure}
+
+    \vspace{1cm} % spacing between subfigures
+
+    % Second Subfigure: Run10_SEQ vs Run5* KE
+    \begin{subfigure}{\textwidth}
+        \centering
+        \begin{tikzpicture}
+        \begin{axis}[
+            ybar,
+            bar width=8pt,
+            width=\textwidth,
+            height=8cm,
+            ymode=log,
+            log origin=infty,
+            ymin=0.001,
+            ylabel={Value (log scale)},
+            symbolic x coords={
+                Total\ Time,
+                Iterations/Sec,
+                User\ CPU\ Time,
+                System\ CPU\ Time,
+                Max\ RSS,
+                Minor\ PF,
+                Major\ PF,
+                Voluntary\ CS,
+                Involuntary\ CS
+            },
+            xtick=data,
+            x tick label style={rotate=30, anchor=east},
+            enlarge x limits=0.05,
+            ymajorgrids,
+            grid style=dashed,
+            legend style={at={(0.5,-0.35)},anchor=north,legend columns=-1},
+            nodes near coords,
+            every node near coord/.append style={
+                font=\tiny,
+                rotate=90,
+                anchor=west,
+                /pgf/number format/.cd, fixed, precision=2
+            }
+        ]
+        % Run10_SEQ (10 completed runs)
+        \addplot[pattern=north east lines,pattern color=black] coordinates {
+            (Total\ Time,107.43)
+            (Iterations/Sec,929.56)
+            (User\ CPU\ Time,41.16)
+            (System\ CPU\ Time,41.00)
+            (Max\ RSS,10)
+            (Minor\ PF,1)
+            (Major\ PF,0.001)
+            (Voluntary\ CS,49)
+            (Involuntary\ CS,65992)
+        };
+        % Run5* (3/5 completed runs, 2 failed)
+        \addplot[pattern=dots,pattern color=black] coordinates {
+            (Total\ Time,106.48)
+            (Iterations/Sec,939.16)
+            (User\ CPU\ Time,41.16)
+            (System\ CPU\ Time,41.00)
+            (Max\ RSS,10)
+            (Minor\ PF,1)
+            (Major\ PF,0.001)
+            (Voluntary\ CS,49)
+            (Involuntary\ CS,65992)
+        };
+        \legend{Run10\_SEQ, Run5*}
+        \end{axis}
+        \end{tikzpicture}
+        \caption{Log-scale comparison of CoreMark Enclave (KE) performance: Run10\_SEQ vs Run5*.\\
+        \textit{* Run5 includes only 3 out of 5 completed runs. Runs 4 and 5 failed.}}
+        \label{fig:coremark_ke_runs_comparison}
+    \end{subfigure}
+\end{figure}
 

### 2025-08-18 chapters/benchmark.tex
conclusion finished

@@
     ]
-        % Native (Run10_SEQ)
+    % Native (Run10_SEQ average)
     \addplot[pattern=north east lines,pattern color=black] coordinates {
-            (Total\ Time,101.11)
-            (Iterations/Sec,989.00)
-            (User\ CPU\ Time,101.14)
-            (System\ CPU\ Time,0.01)
+        (Total\ Time,100.91)
+        (Iterations/Sec,989.80)
+        (User\ CPU\ Time,100.99)
+        (System\ CPU\ Time,0.013)
         (Max\ RSS,1691.60)
-            (Minor\ PF,83.60)
+        (Minor\ PF,83.70)
         (Major\ PF,0.20)
@@
     };
-        % Enclave (Run10_SEQ)
+    % Enclave (Run10_SEQ average)
     \addplot[pattern=dots,pattern color=black] coordinates {
-            (Total\ Time,107.02)
-            (Iterations/Sec,929.80)
+        (Total\ Time,107.45)
+        (Iterations/Sec,928.97)
         (User\ CPU\ Time,41.16)
@@
         (Minor\ PF,1.00)
-            (Major\ PF,0.001) % zero replaced by small non-zero
+        (Major\ PF,0.001) % zero replaced by small non-zero for log scale
         (Voluntary\ CS,49.00)
@@
     };
-        \legend{Native (Run10\_SEQ), Enclave (Run10\_SEQ)}
+    \legend{Native (Run10\_SEQ avg), Enclave (Run10\_SEQ avg)}
     \end{axis}
     \end{tikzpicture}
-        \caption{Log-scale comparison of key metrics for CoreMark 64M\_1Core: Native vs Enclave (Run10\_SEQ).}
+    \caption{Log-scale comparison of key metrics for CoreMark 64M\_1Core: Native vs Enclave (averaged over 10 runs).}
     \label{fig:coremark_native_vs_enclave_log}

### 2025-08-18 chapters/conclusion.tex
updated TOC
conclusion finished

@@
-\section{Summary of Contributions}
+\chapter{Conclusion}
+\label{chap:conclusion}
 
-\section{Limitations of the Current Study}
+The research presented in this thesis has provided a comprehensive examination of the performance, scalability, and architectural constraints involved in deploying secure enclaves on RISC-V platforms through the Keystone Trusted Execution Environment framework.
+By conducting rigorous performance benchmarking, detailed microarchitectural analyses, and thorough system-level evaluations, this study has elucidated both the enabling strengths and the critical limitations inherent in current RISC-V TEE implementations, with particular focus on the Physical Memory Protection mechanism responsible for enclave isolation.
 
-\section{Directions for Future Research}
+Through the empirical evaluation of a range of workloads—including compute-bound benchmarks and memory-intensive cryptographic algorithms—this thesis has demonstrated that enclave performance on RISC-V is highly workload-dependent.
+Compute-bound applications such as CoreMark and Dhrystone experience relatively minor performance degradation when executed within enclaves, typically ranging between 10 to 15 percent.
+This observation aligns with expectations given the dominance of predictable integer computations and limited memory operations in such workloads.
+In stark contrast, cryptographic workloads exemplified by the Kyber post-quantum algorithm suffered from substantial performance penalties exceeding 60 percent.
+This discrepancy arises from Kyber's memory-bound nature and its frequent indirect memory accesses, which exacerbate the overhead introduced by enclave memory isolation and associated integrity checks.
+
+Beyond performance considerations, a critical bottleneck was identified in the utilization of PMP for enclave memory protection.
+The RISC-V architecture’s limit of sixteen PMP entries, each constrained by the Naturally Aligned Power-of-Two encoding scheme, imposes stringent limits on enclave scalability.
+Fragmented or non-contiguous memory allocations rapidly exhaust PMP entries, restricting the feasible number of concurrent enclaves.
+Empirical data confirmed that enclave instantiation failures occurred beyond eight simultaneous enclaves due to PMP saturation, a constraint that holds even on systems with abundant CPU cores and physical memory.
+This architectural limitation presents a significant barrier to multi-tenant or high-concurrency scenarios where isolated execution contexts must coexist.
+
+The implications of these findings extend beyond immediate performance and capacity concerns.
+The current PMP-based isolation model, while offering lightweight hardware-enforced security guarantees, intrinsically restricts flexibility in enclave memory layout and scalability.
+Without modifications to either the hardware architecture—such as increasing PMP register counts or adopting more flexible memory region descriptors—or to system software managing memory allocation, the practicality of RISC-V enclaves in complex real-world applications remains limited.
+
+To mitigate these challenges, this thesis advocates several best practices.
+Firstly, contiguous and properly aligned memory allocation is essential to minimize PMP entry consumption and reduce fragmentation.
+Secondly, consolidating multiple lightweight tasks into fewer enclaves can relieve PMP pressure.
+Thirdly, selectively isolating only the most security-critical components within enclaves, particularly in cryptographic workloads, balances security with performance.
+Furthermore, runtime systems could implement dynamic PMP entry management, reclaiming and reconfiguring PMP resources in response to enclave lifecycle events, though this approach demands additional synchronization mechanisms to preserve security guarantees.
+Lastly, optimizing enclave entry and exit frequency, particularly within tightly looped cryptographic operations, can help amortize synchronization costs.
+
+Looking forward, this research underscores several promising directions.
+Architecturally, enhancements to RISC-V PMP—such as increasing the number of PMP entries or introducing hierarchical and flexible region descriptors—could substantially improve enclave scalability.
+From a software perspective, enclave-aware memory allocators and scheduling systems that proactively manage PMP resources and workload characteristics would enhance system robustness under high concurrency.
+Integration of cryptographic accelerators within the TEE boundary also holds promise to alleviate performance bottlenecks inherent in memory-bound cryptographic workloads.
+Compiler toolchains and enclave abstraction layers that enforce memory alignment and contiguity at build time could reduce developer burden and improve runtime efficiency.
+
+In summary, this thesis has demonstrated that while RISC-V enclaves leveraging Keystone represent a compelling foundation for open and customizable trusted execution environments, practical deployment necessitates careful consideration of PMP-imposed constraints.
+Balancing security, performance, and scalability demands architectural awareness and judicious system design.
+The insights gained herein provide a detailed characterization of enclave behavior and highlight fundamental limitations, serving both as a guide for current practitioners and as a foundation for future research in secure computing on open hardware platforms.
+As trusted execution environments become increasingly vital across cloud, edge, and IoT domains, addressing these challenges will be critical to realizing scalable, efficient, and secure computing paradigms.
+
+%6.1 Summary of Findings
+%6.2 Limitations
+%6.3 Directions for Future Research

### 2025-08-18 chapters/discussion.tex
conclusion finished

@@
 \label{sec:implications}
+
 The challenges highlighted in the previous section, particularly those related to the RISC-V Physical Memory Protection (PMP) mechanism, have significant implications for the design of secure systems, especially those that need to handle cryptographic workloads.
@@
 
-Furthermore, the trade-off between memory usage and security introduces additional complexity in system design.
-Enclaves need to be carefully sized and aligned to fit within the constraints of PMP's encoding scheme.
-For cryptographic workloads that often require non-contiguous memory layouts, the inability to efficiently manage PMP entries could lead to further fragmentation and resource wastage.
-This increases the overall system overhead and reduces the number of concurrent enclaves that can be supported.
-
-For secure systems to scale effectively in the face of these limitations, both hardware and software-level strategies must be employed.
-On the hardware side, increasing the number of available PMP entries or providing more flexible encoding schemes would alleviate some of the bottlenecks.
-However, these hardware solutions would require significant architectural changes and may not be immediately feasible in current RISC-V processors.
-
-On the software side, dynamic management of PMP entries, enclave multiplexing, and memory layout optimization represent viable paths forward.
-By reusing PMP entries or consolidating multiple lightweight workloads into a single enclave, the overall resource consumption can be reduced.
-Additionally, optimizing the memory layout to fit within the NAPOT constraints can help minimize fragmentation, thereby reducing the number of PMP entries required per enclave.
-These software-level strategies can help mitigate some of the scalability issues but may come at the cost of added complexity in the system design.
-
 \begin{figure}[htbp]
-\centering
-% First subfigure (larger)
-\begin{minipage}{0.9\textwidth}
 \centering
@@
     bar width=10pt,
-    width=\linewidth, height=6cm,
+    width=0.9\linewidth, height=6cm,
     ylabel={PMP entries required},
@@
 \end{tikzpicture}
-\caption*{(a) Fragmentation and misalignment increase PMP entry usage}
+\caption{Fragmentation and misalignment increase PMP entry usage, showing how non-ideal alignment leads to more PMP entries per enclave.}
 \label{fig:pmp-fragmentation-entries}
-\end{minipage}
+\end{figure}
 
-\vspace{1.5em}
+To address overhead and scalability challenges, **enclave multiplexing** offers a promising software-level strategy by consolidating multiple lightweight workloads into a single enclave.
+This reduces the number of PMP entries required by reusing them across tasks, optimizing resource usage and improving scalability.
+An example of a multiplexed enclave task dispatcher is shown in Listing~\ref{lst:enclave-mux}.
 
-% Second subfigure (also larger)
-\begin{minipage}{0.9\textwidth}
+\begin{lstlisting}[language=Python, caption={Multiplexed enclave task dispatcher.}, label={lst:enclave-mux}]
+def enclave_dispatcher(task_id, input_data):
+    if task_id == 1:
+        return process_crypto(input_data)
+    elif task_id == 2:
+        return handle_key_store(input_data)
+    elif task_id == 3:
+        return run_diagnostics(input_data)
+    else:
+        raise ValueError("Unknown task")
+\end{lstlisting}
+
+Furthermore, the trade-off between memory usage and security introduces additional complexity in system design.
+Enclaves need to be carefully sized and aligned to fit within the constraints of PMP's encoding scheme, particularly the NAPOT (Naturally Aligned Power-Of-Two) encoding used to specify memory regions.
+For cryptographic workloads that often require non-contiguous memory layouts, inefficient PMP entry management leads to fragmentation and resource wastage, increasing overhead and limiting concurrent enclaves.
+
+\begin{figure}[htbp]
 \centering
@@
 \end{tikzpicture}
-\caption*{(b) Fragmentation overhead: non-contiguous regions mapped to NAPOT-aligned blocks, with hatched padding indicating waste}
+\caption{Fragmentation overhead caused by mapping non-contiguous memory regions to NAPOT-aligned blocks. The hatched areas represent wasted padding required for alignment.}
 \label{fig:napot-fragmentation}
-\end{minipage}
+\end{figure}
 
-\caption{Impact of fragmentation and alignment on PMP entry utilization and memory layout.}
-\label{fig:pmp-fragmentation-group}
-\end{figure}
+The memory layout complexity is also reflected in code structures like the custom memory layout shown in Listing~\ref{lst:memory-layout}, which attempts to align heap, stack, and code regions to NAPOT boundaries.
+
+\begin{lstlisting}[language=C, caption={Custom memory layout aligned to NAPOT for enclave use.}, label={lst:memory-layout}]
+struct enclave_memory_layout {
+    void* heap_start;     // Aligned to 0x1000
+    void* heap_end;
+    void* stack_start;    // Aligned to 0x1000
+    void* stack_end;
+    void* code_region;    // Executable, aligned
+};
+\end{lstlisting}
+
+At the hardware configuration level, PMP entries must be programmed carefully to respect alignment and permission constraints.
+Listing~\ref{lst:pmp-napot} illustrates a PMP configuration for a 4KB NAPOT-aligned enclave memory region, enabling read, write, and execute permissions.
+
+\begin{lstlisting}[language=C, caption={Example PMP configuration for a NAPOT-aligned memory region.}, label={lst:pmp-napot}]
+#define PMP_R 0x01
+#define PMP_W 0x02
+#define PMP_X 0x04
+#define PMP_NAPOT 0x18
+
+// Configure PMP entry 0 for a 4KB aligned enclave memory region
+pmpaddr0 = (0x80000 >> 2);  // Shifted for NAPOT format
+pmpcfg0  = PMP_R | PMP_W | PMP_X | PMP_NAPOT;
+\end{lstlisting}
 
@@
 
+\textbf{Note:} The above code snippets and figures are intended as conceptual illustrations to aid understanding.
+They represent typical scenarios and strategies but are not directly operational without further system integration and adaptation.
+
 \section{Configuration Best Practices and Deployment Recommendations for RISC-V TEEs}
-
+\label{sec:best-practices}
 Deploying Trusted Execution Environments (TEEs) on RISC-V platforms using the Keystone framework presents a unique set of challenges and opportunities that must be carefully navigated to achieve optimal security and performance.
@@
 To this end, several recommendations can be made for optimizing Trusted Execution Environment (TEE) deployments, ensuring that future systems can overcome the current bottlenecks while maintaining strong isolation and security guarantees.
+
+\begin{lstlisting}[caption={Best Practices for RISC-V TEE Deployment}, 
+                   label={lst:enclave-best-practices}, 
+                   basicstyle=\ttfamily\small, 
+                   numbers=left, 
+                   numberstyle=\tiny, 
+                   xleftmargin=2em]
+Allocate enclave memory contiguously, NAPOT-aligned, to minimize PMP usage.
+Consolidate workloads into fewer enclaves to reduce resource contention.
+Limit active enclaves to prevent PMP exhaustion and creation failures.
+Offload non-sensitive tasks to the untrusted domain to lower overhead.
+Batch enclave calls to amortize context-switch costs.
+Use dynamic PMP management to optimize resources at runtime.
+\end{lstlisting}
 

### 2025-08-18 main.tex
added baseline results
introduced latex diagrams
4.4 complete
5.3 complete
native vs enclave latex plots added
conclusion finished

@@
 \usepackage{comment}
+\usepackage{fontspec}
 \usepackage{float}
@@
 \usepackage{tikz}
+\usetikzlibrary{matrix}
+\usepackage{pgfplots}
+\usetikzlibrary{pgfplots.fillbetween}
+\pgfplotsset{compat=1.18}
+\usepackage{pgfplotstable}
+\usepackage{pgf-umlsd}
+\usepgfplotslibrary{statistics}
+\usetikzlibrary{patterns} 
+\usepackage{subcaption}
+\usepackage{booktabs}
+\usepackage{siunitx}
+\usepackage[flushleft]{threeparttable}
+\usepackage{enumitem}
+\sloppy
 \usetikzlibrary{shapes.geometric, arrows.meta, positioning, shadows.blur}
+\usetikzlibrary{positioning,backgrounds,calc,shapes, shapes,arrows.meta}
+\usepackage{xcolor}
+\lstset{
+    basicstyle=\ttfamily\small,
+    frame=single,
+    columns=fullflexible,
+    breaklines=true,
+    backgroundcolor=\color{gray!5},
+    keywordstyle=\color{blue!70!black},
+    commentstyle=\color{gray!60},
+    stringstyle=\color{red!70!black},
+    captionpos=b,
+    showstringspaces=false,
+    tabsize=4
+}
 

### 2025-08-18 chapters/discussion.tex
conclusion finished

@@
 \end{lstlisting}
-
-Further complicating PMP usage is the rigid encoding of memory regions.
-The Naturally Aligned Power-of-Two (NAPOT) encoding format requires memory regions to be aligned and sized as powers of two.
-Irregular or misaligned regions must be split into multiple PMP entries, increasing fragmentation and PMP resource consumption, as illustrated in Figure~\ref{fig:pmp-fragmentation-group}.
-This fragmentation leads to inefficient use of PMP entries, as multiple entries may be required to cover a single logical memory region, further constraining the number of enclaves that can be instantiated concurrently.
-
-%To address these challenges, both hardware-level and software-level mitigation strategies must be considered.
-
-% Hardware approaches include increasing the number of available PMP entries.
-Although this requires architectural changes and increased silicon area, it could dramatically improve enclave concurrency.
-Future RISC-V PMP specifications could also explore more flexible memory region encodings to reduce fragmentation and allow more compact representations of protected regions.
-
-% Software techniques encompass dynamic PMP entry management, where PMP entries are reused or reconfigured at runtime based on workload demands and enclave activity, requiring careful coordination to maintain security.
-Enclave multiplexing—combining multiple lightweight tasks into a single enclave—reduces the number of enclaves and thus PMP entries required.
-Additionally, memory layout optimization to align and size enclave memory regions according to NAPOT constraints can significantly reduce PMP entry consumption.
 

### 2025-08-18 chapters/discussion.tex
minor edits

@@
 
-To address overhead and scalability challenges, **enclave multiplexing** offers a promising software-level strategy by consolidating multiple lightweight workloads into a single enclave.
+To address overhead and scalability challenges, \emph{enclave multiplexing} offers a promising software-level strategy by consolidating multiple lightweight workloads into a single enclave.
 This reduces the number of PMP entries required by reusing them across tasks, optimizing resource usage and improving scalability.

### 2025-08-18 chapters/system-architecture.tex
minor edits

@@
 
+\begin{lstlisting}[language=bash, caption={CoreMark sample output used to extract IPS}, label={lst:coremark-output}]
+2K performance run parameters for coremark.
+CoreMark Size    : 666
+Total ticks      : 112738928
+Iterations/Sec   : 937.43
+Iterations       : 100000
+\end{lstlisting}
+
 In addition to primary performance metrics—DMIPS for Dhrystone and IPS for CoreMark—this study also analyzed supporting metrics to better characterize system behavior.
@@
 
-\begin{lstlisting}[language=bash]
-# CPU time: Total time spent in user mode and kernel mode.
-# Peak RSS: Maximum resident memory usage during execution.
-# Page faults: Number of soft (minor) and hard (major) page faults.
-# Context switches: Voluntary and involuntary context switching events.
+% \begin{lstlisting}[language=bash]
+% # CPU time: Total time spent in user mode and kernel mode.
+% # Peak RSS: Maximum resident memory usage during execution.
+% # Page faults: Number of soft (minor) and hard (major) page faults.
+% # Context switches: Voluntary and involuntary context switching events.
+% \end{lstlisting}
+
+\begin{lstlisting}[language=C, caption={In-enclave system-level instrumentation using \texttt{getrusage()}}, label={lst:getrusage}]
+struct rusage usage;
+getrusage(RUSAGE_SELF, &usage);
+
+// CPU time: Total time spent in user mode and kernel mode.
+printf("!User CPU time:     %ld.%06lds\n", usage.ru_utime.tv_sec, usage.ru_utime.tv_usec);
+printf("!System CPU time:   %ld.%06lds\n", usage.ru_stime.tv_sec, usage.ru_stime.tv_usec);
+
+// Peak RSS: Maximum resident memory usage during execution.
+printf("!Max memory (RSS):  %ld KB\n", usage.ru_maxrss);
+
+// Page faults: Number of soft (minor) and hard (major) page faults.
+printf("!Minor page faults: %ld\n", usage.ru_minflt);
+printf("!Major page faults: %ld\n", usage.ru_majflt);
+
+// Context switches: Voluntary and involuntary context switching events.
+printf("!Voluntary context switches:   %ld\n", usage.ru_nvcsw);
+printf("!Involuntary context switches: %ld\n", usage.ru_nivcsw);
 \end{lstlisting}

### 2025-08-19 chapters/discussion.tex
improved scaling plots in 5.1

@@
 
+% \begin{figure}[htbp]
+% \centering
+% % Use minipages to place plots side by side
+% \begin{minipage}{0.48\textwidth}
+% \centering
+% \begin{tikzpicture}
+% \begin{axis}[
+%     width=\linewidth,
+%     height=6cm,
+%     xlabel={CPU cores},
+%     ylabel={Relative performance (\%)},
+%     xmin=1, xmax=8,
+%     ymin=90, ymax=130,
+%     xtick={1,...,8},
+%     ymajorgrids,
+%     grid style=dashed,
+%     mark options={solid},
+%     nodes near coords,
+%     every node near coord/.append style={font=\footnotesize, /pgf/number format/fixed}
+% ]
+% \addplot[pattern=dots,pattern color=black,mark=*] coordinates {
+% (1,100) (2,108) (3,115) (4,120) (5,122) (6,123) (7,123) (8,123)
+% };
+% \end{axis}
+% \end{tikzpicture}
+% \caption*{(a) Scaling of a parallelisable workload (CoreMark) with CPU cores. Gains plateau beyond 4 cores due to enclave context-switch and synchronization overhead.}
+% \end{minipage}
+% \hfill
+% \begin{minipage}{0.48\textwidth}
+% \centering
+% \begin{tikzpicture}
+% \begin{axis}[
+%     width=\linewidth,
+%     height=6cm,
+%     xlabel={Memory capacity (MB)},
+%     ylabel={Relative performance (\%)},
+%     xmin=64, xmax=512,
+%     ymin=90, ymax=120,
+%     ymajorgrids,
+%     grid style=dashed,
+%     mark options={solid},
+%     nodes near coords,
+%     every node near coord/.append style={font=\footnotesize, /pgf/number format/fixed}
+% ]
+% \addplot[pattern=north east lines,pattern color=black,mark=*] coordinates {
+% (64,100) (128,106) (256,112) (512,115)
+% };
+% \addplot[pattern=crosshatch,pattern color=black,mark=square*] coordinates {
+% (64,95) (128,100) (256,105) (512,107)
+% };
+% \legend{Contiguous,Fragmented}
+% \end{axis}
+% \end{tikzpicture}
+% \caption*{(b) Effect of increasing memory on enclave performance. Fragmentation reduces gains by consuming extra PMP entries.}
+% \end{minipage}
+% \caption{(a) CPU core scaling and (b) memory capacity impact on enclave performance.}
+% \label{fig:performance-scaling}
+% \end{figure}
+
 \begin{figure}[htbp]
 \centering
-% Use minipages to place plots side by side
+% --- Panel (a) ---
 \begin{minipage}{0.48\textwidth}
@@
     mark options={solid},
+    font=\small,
+    legend style={font=\small, at={(0.5,-0.25)}, anchor=north, legend columns=2},
     nodes near coords,
-    every node near coord/.append style={font=\footnotesize, /pgf/number format/fixed}
+    every node near coord/.append style={font=\scriptsize, /pgf/number format/fixed}
 ]
@@
 };
+\addplot[pattern=horizontal lines,pattern color=black,mark=square*] coordinates {
+(1,100) (2,102) (3,103) (4,104) (5,104) (6,104) (7,104) (8,104)
+};
+\legend{Native,Enclave}
 \end{axis}
 \end{tikzpicture}
-\caption*{(a) Scaling of a parallelisable workload (CoreMark) with CPU cores. Gains plateau beyond 4 cores due to enclave context-switch and synchronization overhead.}
-\end{minipage}
-\hfill
+\subcaption{CoreMark scaling. Enclave performance plateaus early due to overhead.}
+\end{minipage}\hfill
+% --- Panel (b) ---
 \begin{minipage}{0.48\textwidth}
@@
     ymin=90, ymax=120,
+    xtick={64,128,256,512},
     ymajorgrids,
@@
     mark options={solid},
+    font=\small,
+    legend style={font=\small, at={(0.5,-0.25)}, anchor=north, legend columns=1},
     nodes near coords,
-    every node near coord/.append style={font=\footnotesize, /pgf/number format/fixed}
+    every node near coord/.append style={font=\scriptsize, /pgf/number format/fixed}
 ]
-\addplot[pattern=north east lines,pattern color=black,mark=*] coordinates {
+\addplot[pattern=dots,pattern color=black,mark=*] coordinates {
 (64,100) (128,106) (256,112) (512,115)
 };
-\addplot[pattern=crosshatch,pattern color=black,mark=square*] coordinates {
-(64,95) (128,100) (256,105) (512,107)
+\addplot[pattern=horizontal lines,pattern color=black,mark=square*] coordinates {
+(64,95) (128,98) (256,100) (512,102)
 };
-\legend{Contiguous,Fragmented}
+\addplot[pattern=crosshatch,pattern color=black,mark=triangle*] coordinates {
+(64,92) (128,94) (256,95) (512,96)
+};
+\legend{Native contiguous, Enclave contiguous, Enclave fragmented}
 \end{axis}
 \end{tikzpicture}
-\caption*{(b) Effect of increasing memory on enclave performance. Fragmentation reduces gains by consuming extra PMP entries.}
+\subcaption{Memory scaling. Fragmentation limits gains by increasing PMP usage.}
 \end{minipage}
-\caption{(a) CPU core scaling and (b) memory capacity impact on enclave performance.}
+
+\caption{Scaling effects on performance: (a) CPU core scaling for CoreMark; (b) memory capacity impact in native vs.\ enclave modes with contiguous and fragmented allocations.}
 \label{fig:performance-scaling}

### 2025-08-19 chapters/benchmark.tex
adjusted average scaling per step

@@
 Workloads unaffected                    & 3/6 (50\%) & 3/6 (50\%) \\
-Average performance change per step (\%)& +6.80\%    & +6.80\%    \\
+Average performance change per step (\%)& 9.99   & 9.99   \\
 \bottomrule

### 2025-08-19 chapters/benchmark.tex
new statistical findings & revised sections 4.4 & parts of 5.1

@@
 
-Building on the preceding workload-specific performance analysis, this section investigates whether increasing CPU cores or available memory yields measurable performance improvements on our constrained RISC-V platform.
-Unlike traditional systems where scaling typically leads to linear or super-linear speedups, embedded systems with TEEs face distinct bottlenecks—such as synchronization overheads, memory bandwidth limits, and enclave management costs—that may inhibit or even reverse expected performance trends.
-
-We systematically evaluated six representative workloads across multiple hardware configurations, incrementally increasing CPU core count and physical memory.
-The workloads include both native and enclave variants: \texttt{coremark}, \texttt{coremark.ke}, \texttt{dhrystone}, \texttt{enclave.ke}, \texttt{kyber}, and \texttt{kyber.ke}.
-Each was benchmarked across scaling steps, with performance tracked as relative throughput (operations per second).
-
-Table~\ref{tab:hardware-impact} summarizes directional performance trends—whether performance improved, degraded, or remained constant—as well as the average relative performance change per scaling step.
+Building on the preceding workload-specific performance analysis, we now examine how the availability of additional hardware resources—specifically CPU cores and physical memory—affects execution performance across native and enclave workloads on our constrained RISC-V platform.
+In conventional computing environments, such scaling is often assumed to yield linear or super-linear performance gains.
+However, embedded systems with secure enclaves (TEEs) operate under very different architectural and runtime constraints.
+Synchronization overheads, limited memory bandwidth, context-switching penalties, and enclave entry/exit costs all contribute to the complex performance dynamics of these systems.
+
+To investigate this, we performed a systematic evaluation of six representative workloads spanning both native and secure execution modes: \texttt{coremark}, \texttt{coremark.ke}, \texttt{dhrystone}, \texttt{enclave.ke}, \texttt{kyber}, and \texttt{kyber.ke}.
+Each workload was benchmarked across multiple hardware configurations, incrementally increasing the number of CPU cores and the amount of available physical memory.
+Performance was quantified as relative throughput (operations per second), normalized across scaling steps to isolate the impact of hardware changes.
+
+Table~\ref{tab:hardware-impact} presents a high-level summary of the results.
+For each scaling axis (cores and memory), we classify workloads based on their observed performance response—whether they improved, degraded, or remained effectively unchanged.
+In addition, we report the average performance change per scaling increment, aggregated across all workloads.
 
@@
 \begin{threeparttable}
-\caption{Summary of workload responses to incremental hardware scaling. Positive average values indicate performance improvement per scaling step.}
+\caption{Summary of workload responses to incremental hardware scaling. Negative average values indicate improved performance (lower runtime) per scaling step.}
 \label{tab:hardware-impact}
@@
 Workloads improved                      & 2/6 (33\%) & 2/6 (33\%) \\
-Workloads slowed down                   & 1/6 (17\%) & 1/6 (17\%) \\
-Workloads unaffected                    & 3/6 (50\%) & 3/6 (50\%) \\
-Average performance change per step (\%)& 9.99   & 9.99   \\
+Workloads slowed down                   & 2/6 (33\%) & 2/6 (33\%) \\
+Workloads unaffected                    & 2/6 (33\%) & 2/6 (33\%) \\
+Average performance change per step (\%)& -8.71      & -8.71      \\
 \bottomrule
@@
 \footnotesize
-\item Each increment corresponds to one step increase in CPU cores or memory size.
-\item Performance improvements were primarily driven by \texttt{coremark} workloads.
+\item Each increment corresponds to one unit increase in either CPU cores or available memory.
+\item Values reflect direction and magnitude of average runtime change across all workloads.
 \end{tablenotes}
@@
 
-These results demonstrate a heterogeneous response to scaling: only one-third of workloads showed any improvement, while the majority exhibited no statistically meaningful gains.
-Notably, \texttt{coremark} and \texttt{coremark.ke} benefited moderately from increased resources, while other workloads—particularly those involving enclave transitions or cryptographic operations—remained largely invariant or slightly degraded under scaling.
-
-To evaluate the robustness of these trends, we applied both parametric and non-parametric statistical tests.
-Specifically, one-way ANOVA tested for mean differences across scaling levels, while Kruskal–Wallis and Mann–Whitney U tests evaluated distributional differences without assuming normality.
-Additionally, Pearson correlation and linear regression were used to quantify trends and explained variance.
-Results are summarized in Table~\ref{tab:statistical-tests}.
-
-\begin{table}[htbp]
-\centering
-\caption{Statistical test results for performance scaling across workloads. Parametric and non-parametric tests assess significance of changes with respect to CPU cores and memory.}
-\label{tab:statistical-tests}
-\begin{subtable}[t]{\textwidth}
-\centering
-\caption{Parametric Tests: ANOVA, Pearson correlation, and $R^2$.}
-\vspace{0.5em}
+The results reveal a highly heterogeneous response to hardware scaling.
+Out of the six workloads, two demonstrated consistent performance improvements with additional resources, two exhibited measurable performance degradation, and two were largely insensitive to the changes.
+This one-third distribution across outcome categories—improved, degraded, and unaffected—highlights a key finding: hardware scaling in secure embedded platforms does not universally translate into improved performance.
+In many cases, the architectural and software stack limitations may nullify or even counteract the expected benefits of scaling.
+
+A closer inspection of individual trends provides additional insight.
+Workloads such as \texttt{coremark.ke} and \texttt{dhrystone.ke} showed modest but consistent improvements as hardware resources increased, with average runtime reductions of approximately $1.7\%$ and $2.6\%$ per step, respectively.
+These results suggest that, under certain execution conditions—some enclave workloads can benefit from added parallelism or memory availability.
+
+In contrast, the native \texttt{dhrystone} and \texttt{kyber} workloads exhibited significant performance degradation, with average runtime increases exceeding $35\%$ per scaling increment.
+This counterintuitive behavior may stem from increased scheduling overheads, memory contention, or inefficiencies introduced by the runtime environment when additional cores or memory are made available without corresponding application-level parallelism or optimization.
+These findings underscore the risk of assuming that more resources automatically lead to better performance—particularly in non-threaded or memory-bound scenarios.
+
+Meanwhile, workloads such as \texttt{coremark} and \texttt{kyber.ke} appeared largely indifferent to hardware scaling.
+Their performance remained flat across all configurations, suggesting that their bottlenecks lie elsewhere—possibly in instruction dispatch, I/O constraints, or TEE-specific execution boundaries that do not benefit from resource scaling.
+
+Statistically, the overall average performance change per scaling step was a negative $-8.71\%$, with a median change of just $-0.51\%$.
+This gap between the mean and median indicates a skewed distribution, driven by a few strongly negative outliers (notably \texttt{kyber} and \texttt{dhrystone}) rather than uniform underperformance across all workloads.
+
+To rigorously assess the significance of these observed performance trends, we applied both parametric and non-parametric statistical tests across all configurations.
+Specifically, a one-way ANOVA test was used to evaluate mean differences between scaling levels (under the assumption of normality), while the Kruskal–Wallis test assessed distributional differences in a non-parametric manner.
+These results are visualized in Figure~\ref{fig:combined-analysis}.
+
+\begin{figure}[htbp]
+\centering
+
+% --- Subfigure (a): p-value Bar Chart ---
+\begin{subfigure}[t]{0.9\linewidth}
+\centering
+\begin{tikzpicture}
+\begin{axis}[
+    width=\linewidth,
+    height=7cm,
+    ybar,
+    bar width=10pt,
+    ymin=0, ymax=1,
+    ymajorgrids,
+    grid style=dashed,
+    symbolic x coords={coremark, coremark.ke, dhrystone, dhrystone.ke, kyber, kyber.ke},
+    xtick=data,
+    xticklabel style={font=\small, rotate=30, anchor=east},
+    enlarge x limits=0.15,
+    ylabel={p-value},
+    axis y line*=left,
+    axis x line*=bottom,
+    font=\small,
+    legend style={font=\small, at={(0.5,1.1)}, anchor=south, legend columns=2},
+    nodes near coords,
+    every node near coord/.append style={
+        font=\scriptsize,
+        rotate=90,
+        anchor=west,
+        yshift=2pt
+    }
+]
+
+% ANOVA
+\addplot+[ybar, pattern=crosshatch, pattern color=black, draw=black, bar shift=-6pt] coordinates {
+  (coremark, 0.827) (coremark.ke, 0.113) (dhrystone, 0.193)
+  (dhrystone.ke, 0.353) (kyber, 0.138) (kyber.ke, 0.421)
+};
+
+% Kruskal-Wallis
+\addplot+[ybar, pattern=north east lines, pattern color=black, draw=black, bar shift=6pt] coordinates {
+  (coremark, 0.967) (coremark.ke, 0.417) (dhrystone, 0.306)
+  (dhrystone.ke, 0.160) (kyber, 0.258) (kyber.ke, 0.865)
+};
+
+\legend{ANOVA p-values, Kruskal-Wallis p-values}
+
+% Significance line at p = 0.05
+\draw[black, dashed] (rel axis cs:0,0.05) -- (rel axis cs:1,0.05);
+
+\end{axis}
+\end{tikzpicture}
+\caption{ANOVA and Kruskal-Wallis p-values for core and memory scaling. p-values below the $\alpha=0.05$ threshold (dashed line) indicate statistically significant effects.}
+\label{fig:anova-kruskal}
+\end{subfigure}
+
+\vspace{1em}
+
+% --- Subfigure (b): Regression Table ---
+\begin{subfigure}[t]{0.9\linewidth}
+\centering
+\begin{threeparttable}
+\caption{Linear regression coefficients of performance on hardware parameters.}
+\label{tab:regression-coefs}
 \begin{tabular}{@{}l
-                S[table-format=1.2] S[table-format=1.3]
-                S[table-format=1.2] S[table-format=1.3]
-                S[table-format=1.3]@{}}
-\toprule
-\textbf{Workload} 
-& \textbf{F (Cores)} & \textbf{p (ANOVA)} 
-& \textbf{r (Cores)} & \textbf{p (r)} 
-& \textbf{$R^2$} \\
-\midrule
-\texttt{coremark}      & 1.50 & 0.218 & -0.16 & 0.052 & 0.030 \\
-\texttt{coremark.ke}   & 0.01 & 0.975 &  0.00 & 0.999 & 0.000 \\
-\texttt{dhrystone}     & 0.28 & 0.837 &  0.07 & 0.374 & 0.005 \\
-\texttt{enclave.ke}    & 0.01 & 0.998 &  0.00 & 0.961 & 0.000 \\
-\texttt{kyber}         & 1.56 & 0.202 & -0.17 & 0.053 & 0.036 \\
-\texttt{kyber.ke}      & 0.00 & 1.000 &  0.00 & 1.000 & 0.000 \\
-\bottomrule
-\end{tabular}
-\begin{tablenotes}
-\footnotesize
-\item Parametric tests assume normal distributions.
-\item r: Pearson correlation; $R^2$: variance explained by linear regression.
-\end{tablenotes}
-\end{subtable}
-
-\vspace{1.5em}
-
-\begin{subtable}[t]{\textwidth}
-\centering
-\caption{Non-Parametric Tests: Kruskal–Wallis and Mann–Whitney U.}
-\vspace{0.5em}
-\begin{tabular}{@{}l
-                S[table-format=2.2] S[table-format=1.3]
-                S[table-format=2.2] S[table-format=1.3]@{}}
-\toprule
-\textbf{Workload} 
-& \textbf{KW (H)} & \textbf{p (KW)} 
-& \textbf{MWU (U)} & \textbf{p (MWU)} \\
-\midrule
-\texttt{coremark}      & 3.60 & 0.057 & 10.00 & 0.052 \\
-\texttt{coremark.ke}   & 0.00 & 1.000 & 15.00 & 0.999 \\
-\texttt{dhrystone}     & 0.20 & 0.654 & 14.00 & 0.374 \\
-\texttt{enclave.ke}    & 0.01 & 0.961 & 15.00 & 0.961 \\
-\texttt{kyber}         & 3.75 & 0.053 & 10.00 & 0.053 \\
-\texttt{kyber.ke}      & 0.00 & 1.000 & 15.00 & 1.000 \\
-\bottomrule
-\end{tabular}
-\begin{tablenotes}
-\footnotesize
-\item KW: Kruskal–Wallis H-statistic; MWU: Mann–Whitney U-statistic.
-\item Non-parametric tests used due to small sample sizes and potential non-normality.
-\end{tablenotes}
-\end{subtable}
-\end{table}
-
-None of the statistical tests yielded significant $p$-values at the $\alpha = 0.05$ level, though \texttt{coremark} and \texttt{kyber} approached marginal significance in the Kruskal–Wallis and Mann–Whitney U tests.
-However, even in these cases, Pearson correlations and $R^2$ values indicate weak or negligible effect sizes.
-These results strongly suggest that, for most workloads, hardware scaling does not yield statistically meaningful performance improvements.
-
-This interpretation is further supported by linear regression analysis in Table~\ref{tab:regression}, which reports slope coefficients and intercepts for each workload's performance as a function of core count and memory.
-
-\begin{table}[htbp]
-\centering
-\begin{threeparttable}
-\caption{Linear regression coefficients of performance on hardware parameters. Positive values indicate performance gains; negative values indicate degradation.}
-\label{tab:regression}
-\begin{tabular}{@{}l
+                S[table-format=7.3]
                 S[table-format=7.3]
@@
 \toprule
-\textbf{Workload} & \textbf{Intercept} & \textbf{Cores Coef.} & \textbf{Memory Coef.} \\
+\textbf{Workload} & \textbf{Intercept} & \textbf{Cores Coef.} & \textbf{Memory Coef.} & \textbf{$R^2$} \\
 \midrule
-\texttt{coremark}     & 2852.203  & 408.000  & 25.000  \\
-\texttt{coremark.ke}  & 5006.875  & 0.000    & 0.000   \\
-\texttt{dhrystone}    & 1327.375  & -53.750  & -37.500  \\
-\texttt{enclave.ke}   & 5046.875  & 0.000    & 0.000   \\
-\texttt{kyber}        & 2847.875  & -45.625  & -16.250  \\
-\texttt{kyber.ke}     & 5352.250  & 0.000    & 0.000   \\
+\texttt{coremark}     & 102.805  & -0.508  & 0.008  & 0.038 \\
+\texttt{coremark.ke}  & 122.474  & -2.970  & 0.007  & 0.071 \\
+\texttt{dhrystone}    & 52.178   & -19.627 & 0.151  & 0.282 \\
+\texttt{enclave.ke}   & 28.257   & 1.104   & -0.019 & 0.043 \\
+\texttt{kyber}        & 89.318   & -40.129 & 0.329  & 0.035 \\
+\texttt{kyber.ke}     & 42.219   & 0.824   & -0.029 & 0.070 \\
 \bottomrule
@@
 \item Intercept approximates baseline performance; coefficients represent change per incremental unit of cores or memory.
+\item \( R^2 \) indicates model fit (explained variance); values near zero imply weak correlation.
 \end{tablenotes}
 \end{threeparttable}
-\end{table}
-
-Here again, \texttt{coremark} is the only workload to exhibit a positive response to hardware scaling, albeit with limited effect size.
-All enclave workloads display near-zero or flat regression slopes, underscoring the notion that scaling is generally ineffective when trusted execution overheads dominate the execution path.
-
-In summary, while modest improvements are observed for select CPU-bound native workloads, scaling core or memory resources yields no consistent benefit—and occasionally even degrades performance—under enclave execution.
-These findings highlight the need for deeper architectural or runtime co-optimization if scaling is to deliver tangible performance benefits in secure embedded platforms.
+\caption*{}
+\end{subfigure}
+
+\caption{Statistical analysis of workload performance scaling.
+(a) ANOVA and Kruskal-Wallis tests evaluate whether core/memory scaling significantly affects performance.
+(b) Linear regression models estimate the effect magnitude and fit quality.}
+\label{fig:combined-analysis}
+\end{figure}
+
+As shown in Figure~\ref{fig:anova-kruskal}, none of the $p$-values from either ANOVA or Kruskal–Wallis testing fell below the $\alpha = 0.05$ significance threshold.
+This result indicates a lack of statistically significant performance variation across hardware scaling steps for all workloads tested, regardless of execution context.
+The regression coefficients in subfigure~(b) further reinforce this interpretation: the slopes for both core and memory scaling are near zero in most cases, with low \( R^2 \) values suggesting weak or no explanatory power.
+
+%In summary, while certain workloads demonstrate mild sensitivity to hardware scaling, these effects are neither statistically significant nor consistent across the benchmark suite.
+This finding has important implications for the design of secure embedded systems: namely, that simply increasing hardware capabilities—whether by adding cores or memory—does not guarantee improved performance under TEE constraints.
+Instead, performance tuning must consider software-level optimizations, workload characteristics, and the inherent cost model of enclave-based execution.

### 2025-08-19 chapters/discussion.tex
adjusted average scaling per step
new statistical findings & revised sections 4.4 & parts of 5.1

@@
 
-An interesting trend emerges when examining system resource allocation impacts on enclave performance.
-Increasing the number of CPU cores generally improves performance for multi-threaded applications; however, gains plateau beyond a certain threshold due to overhead introduced by context switching between enclave and host execution contexts and the need for thread synchronization within enclaves.
-This phenomenon is particularly evident in parallelizable workloads, such as CoreMark.
-
+An important trend that aligns with our empirical findings concerns the nuanced impact of system resource allocation on enclave performance.
+While increasing CPU core count typically benefits multi-threaded applications, our results show that such improvements are neither uniform nor guaranteed within TEEs.
+For enclave-executing workloads like \texttt{coremark.ke} and \texttt{dhrystone.ke}, performance gains with core scaling were modest and statistically insignificant, as confirmed by both ANOVA and regression analysis.
+This plateauing effect can be attributed to the overhead introduced by frequent context switches between the host and enclave domains, as well as the synchronization costs imposed by enclave thread management.
+These factors inherently limit the scalability of otherwise parallelizable workloads such as \texttt{coremark}, explaining the flat or marginal improvements observed in practice.
 % \begin{figure}[htbp]
@@
 
-Memory configuration also plays a critical role in determining enclave performance.
-Enclaves are sensitive to memory availability; increasing memory capacity typically improves performance by reducing page faults and improving cache locality.
-However, fragmented memory regions may require multiple PMP entries to define non-contiguous areas, exacerbating system overhead.
-The trade-off between memory allocation and PMP usage must therefore be carefully managed.
-Exhaustion of PMP entries leads to severe performance degradation, as additional enclaves cannot be instantiated regardless of available CPU or memory resources.
+Memory scaling follows a similarly constrained pattern.
+Although increased memory capacity can improve enclave performance by reducing page faults and enhancing cache locality, the architectural constraints of physical memory protection impose practical limits.
+Enclaves rely on PMP (Physical Memory Protection) entries to define their accessible memory regions.
+As our findings suggest, workloads like \texttt{kyber.ke} and \texttt{coremark.ke} showed little to no improvement despite increased memory availability—likely due to internal fragmentation or PMP exhaustion.
+When memory becomes fragmented, or when the system runs out of PMP entries, additional enclave instantiations are prevented, regardless of how much physical memory remains unused.
+This introduces a hard cap on scalability and can even lead to performance degradation if enclave memory cannot be allocated contiguously.
+The low $R^2$ values in the regression models for most enclave workloads further support the conclusion that increased hardware resources alone are poor predictors of performance within TEEs.
 

### 2025-08-19 chapters/benchmark.tex
new statistical findings & revised sections 4.4 & parts of 5.1

@@
 Average performance change per step (\%)& -8.71      & -8.71      \\
+Median performance change per core step (\%) & -0.51      & -0.51      \\
 \bottomrule

### 2025-08-20 chapters/benchmark.tex
added random forest generation regression plot

@@
 Specifically, a one-way ANOVA test was used to evaluate mean differences between scaling levels (under the assumption of normality), while the Kruskal–Wallis test assessed distributional differences in a non-parametric manner.
-These results are visualized in Figure~\ref{fig:combined-analysis}.
+These results are visualized in Figure~\ref{fig:combined-analysis-updated}.
+
+%\footnote{The Kruskal-Wallis test was used alongside ANOVA to handle cases where group distributions violated normality assumptions, ensuring robustness in detecting group differences.
+Random Forest regression was applied in addition to linear models to address multicollinearity between hardware parameters (e.g., cores and memory), and to capture potential nonlinear relationships that linear regression may miss.}
 
@@
 % --- Subfigure (a): p-value Bar Chart ---
-\begin{subfigure}[t]{0.9\linewidth}
+\begin{subfigure}[t]{0.95\linewidth}
 \centering
@@
 \label{fig:anova-kruskal}
+\vspace{0.5em}
+{\footnotesize
+\noindent\parbox{\linewidth}{%
+\textit{Note: The Kruskal-Wallis test was included to address cases where normality assumptions were not met.}
+}
+}
 \end{subfigure}
 
-\vspace{1em}
-
-% --- Subfigure (b): Regression Table ---
-\begin{subfigure}[t]{0.9\linewidth}
-\centering
-\begin{threeparttable}
-\caption{Linear regression coefficients of performance on hardware parameters.}
-\label{tab:regression-coefs}
-\begin{tabular}{@{}l
-                S[table-format=7.3]
-                S[table-format=7.3]
-                S[table-format=7.3]
-                S[table-format=7.3]@{}}
-\toprule
-\textbf{Workload} & \textbf{Intercept} & \textbf{Cores Coef.} & \textbf{Memory Coef.} & \textbf{$R^2$} \\
-\midrule
-\texttt{coremark}     & 102.805  & -0.508  & 0.008  & 0.038 \\
-\texttt{coremark.ke}  & 122.474  & -2.970  & 0.007  & 0.071 \\
-\texttt{dhrystone}    & 52.178   & -19.627 & 0.151  & 0.282 \\
-\texttt{enclave.ke}   & 28.257   & 1.104   & -0.019 & 0.043 \\
-\texttt{kyber}        & 89.318   & -40.129 & 0.329  & 0.035 \\
-\texttt{kyber.ke}     & 42.219   & 0.824   & -0.029 & 0.070 \\
-\bottomrule
-\end{tabular}
-\begin{tablenotes}
-\footnotesize
-\item Coefficients reflect linear regression fits across scaling increments.
-\item Intercept approximates baseline performance; coefficients represent change per incremental unit of cores or memory.
-\item \( R^2 \) indicates model fit (explained variance); values near zero imply weak correlation.
-\end{tablenotes}
-\end{threeparttable}
-\caption*{}
+\vspace{1.5em}
+
+% --- Subfigure (b): Grouped R^2 Chart ---
+\begin{subfigure}[t]{0.95\linewidth}
+\centering
+\begin{tikzpicture}
+\begin{axis}[
+    width=\linewidth,
+    height=7cm,
+    ybar,
+    bar width=10pt,
+    enlarge x limits=0.15,
+    ylabel={$R^2$ (interaction model)},
+    ymin=0, ymax=0.5,
+    ymajorgrids,
+    grid style=dashed,
+    symbolic x coords={coremark,coremark.ke,dhrystone,dhrystone.ke,kyber,kyber.ke},
+    xtick=data,
+    xticklabel style={rotate=30,anchor=east, font=\small},
+    font=\small,
+    legend style={at={(0.5,1.1)}, anchor=south, legend columns=2},
+    every node near coord/.append style={
+        font=\footnotesize,
+        rotate=90,
+        anchor=west,
+        /pgf/number format/fixed,
+        /pgf/number format/precision=2
+    },
+    nodes near coords
+]
+
+% Linear regression
+\addplot+[
+    pattern=north east lines, 
+    pattern color=black, 
+    draw=black, 
+    bar shift=-6pt
+] coordinates {
+    (coremark,0.04)
+    (coremark.ke,0.07)
+    (dhrystone,0.28)
+    (dhrystone.ke,0.04)
+    (kyber,0.33)
+    (kyber.ke,0.24)
+};
+
+% Random Forest
+\addplot+[
+    pattern=horizontal lines, 
+    pattern color=black, 
+    draw=black, 
+    bar shift=6pt
+] coordinates {
+    (coremark,0.07)
+    (coremark.ke,0.40)
+    (dhrystone,0.31)
+    (dhrystone.ke,0.24)
+    (kyber,0.36)
+    (kyber.ke,0.23)
+};
+
+\legend{Linear regression, Random Forest}
+\end{axis}
+\end{tikzpicture}
+\caption{Comparison of $R^2$ values from linear regression and random forest models across workloads. Higher values indicate better model fit.}
+\label{fig:r2_grouped}
+\vspace{0.5em}
+{\footnotesize
+\noindent\parbox{\linewidth}{%
+\textit{Note: Random Forest regression was included to address multicollinearity issues observed in linear models.}
+}
+}
 \end{subfigure}
@@
 (a) ANOVA and Kruskal-Wallis tests evaluate whether core/memory scaling significantly affects performance.
-(b) Linear regression models estimate the effect magnitude and fit quality.}
-\label{fig:combined-analysis}
+(b) Regression models estimate the effect magnitude and fit quality.}
+%\caption*{\textit{Note: A Random Forest regressor was used to mitigate multicollinearity effects between CPU cores and memory capacity, offering a non-linear alternative to standard linear regression.}}
+\label{fig:combined-analysis-updated}
 \end{figure}

### 2025-08-20 chapters/benchmark.tex
added random forest generation regression plot

@@
 
-As shown in Figure~\ref{fig:anova-kruskal}, none of the $p$-values from either ANOVA or Kruskal–Wallis testing fell below the $\alpha = 0.05$ significance threshold.
-This result indicates a lack of statistically significant performance variation across hardware scaling steps for all workloads tested, regardless of execution context.
-The regression coefficients in subfigure~(b) further reinforce this interpretation: the slopes for both core and memory scaling are near zero in most cases, with low \( R^2 \) values suggesting weak or no explanatory power.
+As shown in Figure~\ref{fig:anova-kruskal}, none of the $p$-values from either ANOVA or Kruskal--Wallis testing fell below the $\alpha = 0.05$ significance threshold.
+This indicates a lack of statistically significant performance variation across hardware scaling steps for all workloads tested.
+The $R^2$ values in subfigure~\ref{fig:r2_grouped} support this, with low model fit observed in both linear and Random Forest regressions, suggesting weak or no explanatory power.
 

### 2025-08-22 chapters/background.tex
citations & plot amendments are complete for chapters 1 - 2

@@
 
-The Keystone framework is founded on the RISC-V instruction set architecture (ISA), which provides a compelling platform for secure system design due to its open specification and extensible modularity~\cite{lee2019keystone,dayeol2019keystone}.
+The Keystone framework is founded on the RISC-V instruction set architecture (ISA), which provides a compelling platform for secure system design due to its open specification and extensible modularity~\cite{Lee2019,Lee2019}.
 Unlike traditional ISAs governed by proprietary vendors, RISC-V fosters a collaborative ecosystem that supports community-driven enhancements and custom extensions.
@@
 RISC-V specifies three hierarchical privilege levels: user mode (U-mode), supervisor mode (S-mode), and machine mode (M-mode).
-M-mode operates with the highest privilege and possesses unrestricted access to hardware resources, managing critical functions such as interrupt handling, exception processing, and physical memory protection~\cite{lee2019keystone}.
+M-mode operates with the highest privilege and possesses unrestricted access to hardware resources, managing critical functions such as interrupt handling, exception processing, and physical memory protection~\cite{Lee2019}.
 Keystone capitalizes on this privilege hierarchy by situating its Security Monitor (SM)—the core Trusted Computing Base (TCB)—within M-mode.
@@
 This hardware-enforced mechanism forms the cornerstone of Keystone’s memory isolation strategy.
-Each enclave is assigned a dedicated, PMP-protected memory region that remains inaccessible to the host OS or unauthorized software, guaranteeing strong confidentiality and integrity guarantees enforced at the hardware level~\cite{lee2019keystone}.
+Each enclave is assigned a dedicated, PMP-protected memory region that remains inaccessible to the host OS or unauthorized software, guaranteeing strong confidentiality and integrity guarantees enforced at the hardware level~\cite{Lee2019}.
 
-In addition to PMP, RISC-V’s flexible trap and exception handling model enables M-mode to intercept all traps while optionally delegating certain classes, such as system calls and page faults, to S-mode for efficiency~\cite{lee2019keystone}.
+In addition to PMP, RISC-V’s flexible trap and exception handling model enables M-mode to intercept all traps while optionally delegating certain classes, such as system calls and page faults, to S-mode for efficiency~\cite{Lee2019}.
 This delegation mechanism facilitates cooperation between enclave runtimes and the host OS for memory management without compromising enclave boundaries.
-Keystone utilizes the RISC-V Memory Management Unit (MMU) to support private virtual address spaces per enclave, further fortified by PMP protections against unauthorized access or tampering~\cite{dayeol2019keystone}.
+Keystone utilizes the RISC-V Memory Management Unit (MMU) to support private virtual address spaces per enclave, further fortified by PMP protections against unauthorized access or tampering~\cite{Lee2019}.
 
 Notably, the Keystone Security Monitor is implemented in C using standard programming toolchains, which enhances maintainability, auditability, and formal verification potential.
-This contrasts with fixed-function microcode or proprietary firmware that is often opaque and difficult to analyze, thus bolstering Keystone’s transparency and security assurance~\cite{lee2019keystone}.
+This contrasts with fixed-function microcode or proprietary firmware that is often opaque and difficult to analyze, thus bolstering Keystone’s transparency and security assurance~\cite{Lee2019}.
 
@@
 A TEE effectively partitions system resources such as CPU registers, memory, and I/O peripherals, restricting access exclusively to trusted enclave applications.
-This isolation is enforced by the processor’s privilege levels, memory protection units, or dedicated security monitors, thereby preventing unauthorized entities—including kernel-level malware or user-space adversaries—from accessing or manipulating enclave data~\cite{brasser2019software}.
+This isolation is enforced by the processor’s privilege levels, memory protection units, or dedicated security monitors, thereby preventing unauthorized entities—including kernel-level malware or user-space adversaries—from accessing or manipulating enclave data~\cite{brasser2020enclave}.
 
 Security in TEEs hinges on minimizing the Trusted Computing Base (TCB), which encompasses all components that must be trusted for correct and secure operation.
-A smaller TCB reduces the risk of vulnerabilities and facilitates formal verification and security audits~\cite{lee2019keystone,Survey2023}.
+A smaller TCB reduces the risk of vulnerabilities and facilitates formal verification and security audits~\cite{Lee2019,Survey2023}.
 Modern TEEs strive to encapsulate only essential components within the TCB, balancing security with functionality.
@@
 ARM TrustZone employs a split-world model that partitions processor execution into secure and normal worlds, with transitions mediated by secure monitor calls~\cite{yan2018trustzone}.
+These architectural designs are illustrated in Figure~\ref{fig:tee-architectures}, which compares TrustZone's dual-world architecture (Figure~\ref{fig:trustzone}) with SGX's Enclave Page Cache memory model (Figure~\ref{fig:sgx-epc}).
 Despite their widespread deployment, these architectures suffer from proprietary designs and limited transparency, which constrain their flexibility for academic research and system-level innovation.
@@
 
-In contrast, open-source TEE frameworks such as Keystone have emerged to address these limitations by leveraging the open, modular RISC-V ISA~\cite{lee2019keystone,dayeol2019keystone}.
+In contrast, open-source TEE frameworks such as Keystone have emerged to address these limitations by leveraging the open, modular RISC-V ISA~\cite{Lee2019}.
 The modularity and openness of RISC-V facilitate the design of TEEs that are both highly configurable and more transparent, fostering security research and experimentation~\cite{Survey2023}.
 A central architectural principle of TEEs, exemplified by Keystone, is privilege separation: executing the TCB in the highest privilege mode (M-mode in RISC-V) while relegating untrusted software to lower privilege levels.
-This design enforces strict isolation policies, preventing untrusted software from compromising enclave security~\cite{lee2019keystone}.
+This design enforces strict isolation policies, preventing untrusted software from compromising enclave security~\cite{Lee2019}.
+Table~\ref{tab:tee-comparison} summarizes key architectural and design distinctions between Intel SGX, ARM TrustZone, and Keystone.
 
@@
 \centering
+\caption{Comparison of three TEE implementations.}
+\label{tab:tee-comparison}
 \small
@@
 \end{tabular}
-\caption{Comparison of three TEE implementations.}
 \end{table}
@@
 
-Keystone is an open-source Trusted Execution Environment (TEE) framework built on the RISC-V architecture, designed to enable customizable, lightweight, and secure enclaves suitable for both research and deployment in resource-constrained or security-critical settings \cite{dayeol2019keystone}.
+Keystone is an open-source Trusted Execution Environment (TEE) framework built on the RISC-V architecture, designed to enable customizable, lightweight, and secure enclaves suitable for both research and deployment in resource-constrained or security-critical settings \cite{Lee2019}.
 The framework’s modular design emphasizes a minimal Trusted Computing Base (TCB), portability across diverse RISC-V platforms, and extensibility to accommodate advanced security features, reflecting a clean and flexible architectural philosophy.
@@
 At the core of Keystone’s TCB is the Security Monitor (SM), which executes in M-mode and holds exclusive authority over enclave lifecycle management, memory protection configuration, and access control enforcement.
-The SM uses RISC-V’s Physical Memory Protection (PMP) mechanism to isolate enclave memory regions dynamically, ensuring strict access controls during enclave execution \cite{dayeol2019keystone}.
+The SM uses RISC-V’s Physical Memory Protection (PMP) mechanism to isolate enclave memory regions dynamically, ensuring strict access controls during enclave execution \cite{Lee2019}.
 
 A typical Keystone-enabled platform comprises RISC-V cores augmented with secure boot mechanisms, trusted entropy sources, and optionally, hardware accelerators for cryptographic functions or secure I/O.
-Importantly, Keystone does not require hardware beyond standard RISC-V compliance and PMP support, making it adaptable to implementations ranging from FPGA prototypes to silicon SoCs \cite{dayeol2019keystone}.
+Importantly, Keystone does not require hardware beyond standard RISC-V compliance and PMP support, making it adaptable to implementations ranging from FPGA prototypes to silicon SoCs \cite{Lee2019}.
 
 Each enclave in Keystone consists of two main software components: a user-level enclave application (eapp) containing application-specific logic, and a supervisor-level runtime (RT) providing essential operating system abstractions such as exception handling, system call services, and memory management.
-This layered approach reduces the complexity of enclave applications and provides a trusted runtime environment isolated from the untrusted operating system and other processes \cite{dayeol2019keystone}.
+This layered approach reduces the complexity of enclave applications and provides a trusted runtime environment isolated from the untrusted operating system and other processes \cite{Lee2019}.
 
@@
 During execution, the SM mediates context switches between the host and enclave, dynamically adjusting PMP to maintain isolation.
-Destruction securely erases enclave memory to prevent any residual data leakage \cite{dayeol2019keystone}.
+Destruction securely erases enclave memory to prevent any residual data leakage \cite{Lee2019}.
 Keystone also supports optional features such as remote attestation, allowing external verifiers to authenticate enclave state prior to provisioning sensitive data, and can be extended for secure I/O and cryptographic acceleration.
+Figure~\ref{fig:keystone_grouped} illustrates the Keystone enclave design.
+Specifically, Figure~\ref{fig:keystone_overview} shows the architectural components, including the Security Monitor, enclave runtime, and the RISC-V privilege hierarchy.
+Figure~\ref{fig:enclave_lifecycle} depicts the enclave lifecycle phases: creation, execution, and destruction.
 
@@
 Keystone’s modular and open framework leverages the openness of RISC-V to provide strong security guarantees while enabling flexibility for research and customization.
-Its minimal TCB, modular layering (Security Monitor, runtime, eapp), and standards-compliant interfaces make it an attractive platform for developing and evaluating secure applications on modern processors \cite{dayeol2019keystone}.
+Its minimal TCB, modular layering (Security Monitor, runtime, eapp), and standards-compliant interfaces make it an attractive platform for developing and evaluating secure applications on modern processors \cite{Lee2019}.
 To support secure communication and attestation, Keystone enclaves can integrate modern cryptographic primitives, including emerging post-quantum schemes such as Kyber.
@@
 The hardware enforces these permissions such that any access attempt by S-mode or U-mode software that violates a PMP rule results in a trap to M-mode.
-This mechanism ensures that M-mode retains full control over physical memory while being able to delegate controlled access to untrusted software layers~\cite{riscvprivspec,lee2019keystone}.
+This mechanism ensures that M-mode retains full control over physical memory while being able to delegate controlled access to untrusted software layers~\cite{riscvprivspec,Lee2019}.
+The basic hardware architecture of PMP within the RISC-V privilege hierarchy is illustrated in Figure~\ref{fig:pmp-arch}.
 
@@
 Keystone leverages PMP to assign a protected memory region, known as the Enclave Page Memory (EPM), to each enclave.
-This region is configured to be inaccessible to both the untrusted operating system (running in S-mode) and any other software outside the enclave context~\cite{dayeol2019keystone}.
+This region is configured to be inaccessible to both the untrusted operating system (running in S-mode) and any other software outside the enclave context~\cite{Lee2019}.
 During enclave creation, the Keystone Security Monitor (SM)—executing in M-mode—programmatically sets PMP entries to lock the memory region associated with the enclave.
-The SM verifies the integrity of the enclave binary, computes its cryptographic hash, and sets up the memory layout in such a way that only the enclave code has access to its memory contents~\cite{lee2019keystone,keystone2025how}.
+The SM verifies the integrity of the enclave binary, computes its cryptographic hash, and sets up the memory layout in such a way that only the enclave code has access to its memory contents~\cite{Lee2019,keystone2025how}.
+As shown in Figure~\ref{fig:pmp-keystone}, the PMP unit mediates memory access between the OS and the enclave to enforce isolation guarantees.
 
@@
 Conversely, when the enclave resumes execution, the SM reinstates its PMP region.
-This dynamic switching is essential to enforce temporal isolation and prevent time-of-check-to-time-of-use (TOCTTOU) vulnerabilities that may arise during state transitions~\cite{cerberus2022}.
+This dynamic switching is essential to enforce temporal isolation and prevent time-of-check-to-time-of-use (TOCTTOU) vulnerabilities that may arise during state transitions~\cite{lee2022cerberus}.
 
@@
 The fixed number of PMP entries restricts the number of simultaneously protected regions, posing a challenge for complex systems with many isolated domains or subregions.
-Moreover, overlapping PMP regions are resolved based on entry priority (lower-numbered entries take precedence), requiring careful configuration to avoid inadvertent leaks or privilege escalations~\cite{riscvprivspec}.
+Moreover, overlapping PMP regions are resolved based on entry priority (lower-numbered entries take precedence), requiring careful configuration to avoid inadvertent leaks or privilege escalations (see Figure~\ref{fig:pmp-priority})~\cite{riscvprivspec}.
 
@@
 Security researchers have explored various enhancements and formal models to increase the reliability and verifiability of PMP-based isolation.
-For example, Cerberus~\cite{cerberus2022} builds a formally verified memory-sharing abstraction over PMP to enable safe communication between enclaves and untrusted components.
+For example, Cerberus~\cite{lee2022cerberus} builds a formally verified memory-sharing abstraction over PMP to enable safe communication between enclaves and untrusted components.
 Similarly, HexBox~\cite{hexbox2018} leverages PMP for real-time embedded security by associating memory regions with specific control-flow paths in embedded firmware, demonstrating PMP's applicability beyond general-purpose computing.
+Table~\ref{tab:pmp-limitations} summarizes key limitations in PMP and highlights recent research directions aimed at overcoming them.
 
@@
 \centering
+\caption{Summary of PMP limitations and research‑driven enhancements.}
+\label{tab:pmp-limitations}
 \renewcommand{\arraystretch}{1.2}
@@
 \end{tabular}
-\caption{Summary of PMP limitations and research‑driven enhancements.}
-\label{tab:pmp-limitations}
 \end{table}
@@
 This synergy is particularly effective in the RISC-V ecosystem, where the open architecture allows deep customization, formal verification, and academic experimentation~\cite{Survey2023}.
-The combination of PMP, a secure boot mechanism, and a minimal TCB enables Keystone to enforce hardware-rooted enclave isolation without relying on opaque, vendor-specific extensions, setting it apart from alternatives like Intel SGX or ARM TrustZone~\cite{dayeol2019keystone,turn0search0}.
+The combination of PMP, a secure boot mechanism, and a minimal TCB enables Keystone to enforce hardware-rooted enclave isolation without relying on opaque, vendor-specific extensions, setting it apart from alternatives like Intel SGX or ARM TrustZone~\cite{Lee2019,suzaki2021tsperf}.
 
-In conclusion, Physical Memory Protection in RISC-V is a powerful, lightweight, and deterministic mechanism that underpins secure execution in many RISC-V-based platforms.
+Physical Memory Protection in RISC-V is a powerful, lightweight, and deterministic mechanism that underpins secure execution in many RISC-V-based platforms.
 Its integration into frameworks like Keystone enables transparent and verifiable enclave isolation while remaining suitable for constrained systems.
@@
 
-Kyber is a post-quantum key encapsulation mechanism (KEM) designed to resist attacks from both classical and quantum adversaries.
-Its security is based on the hardness of the Module Learning With Errors (Module-LWE) problem, a structured lattice problem considered intractable even for quantum computers.
-Kyber is one of the finalists in the NIST Post-Quantum Cryptography (PQC) competition and has been selected for standardization due to its strong security and performance characteristics \cite{kyber2024}.
-% [SOURCE1]
+Kyber is a post-quantum key encapsulation mechanism (KEM) designed to remain secure against both classical and quantum adversaries.
+It is based on the hardness of the Module Learning With Errors (Module-LWE) problem, a structured lattice problem believed to be difficult to solve even with quantum algorithms \cite{kyber2021}.
+The algorithm has been selected by NIST for standardization in its post-quantum cryptography project, reflecting both its strong theoretical foundation and practical efficiency \cite{kyber2024}.
 
-At its core, Kyber is derived from the Lyubashevsky–Peikert–Regev (LPR) encryption framework, which adapts Learning With Errors (LWE) from integer vectors to more efficient polynomial structures.
-Specifically, Kyber operates in the Module-LWE setting, which balances the structure of Ring-LWE with the flexibility of traditional LWE.
-This results in strong performance and compact key sizes while preserving security \cite{kyber2024,kyber2021}.
-% [SOURCE1, BLOG1]
+Kyber uses polynomial arithmetic in a modular ring to perform its cryptographic operations.
+It is based on the LPR encryption framework \cite{kyber2021}, which adapts the Learning With Errors (LWE) problem from integer vectors to more structured polynomial rings.
+This allows for faster computation and more compact key sizes compared to classical public-key systems.
 
-Kyber supports three security levels:
-\begin{itemize}
-    \item \textbf{Kyber-512}, targeting AES-128-level security,
-    \item \textbf{Kyber-768}, targeting AES-192-level security (recommended),
-    \item \textbf{Kyber-1024}, targeting AES-256-level security.
-\end{itemize}
+Kyber supports three variants, each corresponding to a different level of security roughly equivalent to asymmetric encryption:
+
+\begin{verbatim}
+Kyber-512   → AES-128-level security
+Kyber-768   → AES-192-level security (recommended)
+Kyber-1024  → AES-256-level security
+\end{verbatim}
+
+These parameters determine the number of polynomials, noise levels, and compression factors used internally.
+Figure~\ref{fig:kyber_768} illustrates the Kyber-768 key exchange process, in which two parties derive a shared secret using public key encapsulation and decapsulation.
 
@@
 \includegraphics[width=0.9\linewidth]{figures/kyber_768.png}
-\caption{Kyber768 Key Exchange Algorithm}
+\caption{Kyber-768 Key Exchange Algorithm.}
 \label{fig:kyber_768}
@@
 
+Kyber exposes a compact and simple interface for use in protocols. Its API supports three core operations: key generation, encapsulation, and decapsulation, as shown in Listing~\ref{lst:kem_api}.
+
+\begin{lstlisting}[style=kemstyle, caption={Kyber KEM API}, label={lst:kem_api}]
+(pk, sk)    = KeyGen()           # Generate key pair
+(ct, ss)    = Encapsulate(pk)    # Encrypt shared secret
+ss          = Decapsulate(sk, ct) # Decrypt to recover secret
+\end{lstlisting}
+
+Compared to classical schemes like RSA, Kyber offers more compact key and ciphertext sizes at higher security levels.
+Table~\ref{tab:kyber_sizes} compares the relevant sizes in bytes for Kyber and RSA \cite{kyber2024}.
+
 \begin{table}[h]
 \centering
-\caption{Kyber and RSA: Key and Ciphertext Sizes (in bytes)}
+\caption{Key and Ciphertext Sizes (bytes)}
 \label{tab:kyber_sizes}
+\renewcommand{\arraystretch}{1.2}
 \begin{tabular}{|l|c|c|c|}
@@
 
-While RSA keys are somewhat smaller, Kyber maintains significantly better security in the post-quantum context, and its keys are far more compact than those of many alternative PQC candidates such as Classic McEliece \cite{kyber2021}.
-% [BLOG1]
-
-Kyber is structured as a KEM rather than a traditional public-key encryption (PKE) scheme. Internally, however, it uses a PKE system transformed into a CCA-secure KEM. The KEM exposes a simple API:
-\begin{itemize}
-    \item $(\mathit{pk}, \mathit{sk}) = \texttt{KeyGen}()$
-    \item $(\mathit{ct}, \mathit{ss}) = \texttt{Encapsulate}(\mathit{pk})$
-    \item $\mathit{ss} = \texttt{Decapsulate}(\mathit{sk}, \mathit{ct})$
-\end{itemize}
-
-The encapsulation mechanism makes Kyber especially suitable for protocols like TLS, Signal, and OpenPGP. % [BLOG1]
-
-A simplified pedagogical version called ``Baby Kyber’’ is often used to illustrate the cryptographic structure.
-Baby Kyber uses polynomials over a small modulus and limited degree to show how private and public keys, as well as ciphertexts, are constructed and manipulated.
-Encryption relies on adding small random errors to hide the message, while decryption relies on error tolerance due to coefficient scaling.
-If the message polynomial is scaled appropriately, decryption can recover the original bit string by rounding \cite{kyber2021}.
-% [BLOG1]
-
-In full Kyber, the parameters are chosen to ensure security and correctness:
-\begin{itemize}
-    \item $n = 256$: polynomial degree
-    \item $k \in \{2, 3, 4\}$: number of polynomials (Kyber-512, 768, 1024)
-    \item $q = 3329$: prime modulus
-    \item $\eta_1, \eta_2$: bounds on error distributions
-    \item $d_u, d_v$: compression factors
-\end{itemize}
-
-Kyber’s performance has been optimized for a variety of platforms. Table~\ref{tab:kyber_perf} shows benchmark results on an Intel Haswell CPU with AVX2 vectorization \cite{kyber2024}. % [SOURCE1]
+The parameters that define Kyber’s security and performance are listed in Table~\ref{tab:kyber-params}.
+These include the polynomial degree, modulus, and bounds on the error distributions used to introduce noise in the ciphertexts \cite{kyber2021}.
 
@@
 \centering
-\caption{Kyber Performance on Haswell CPU (AVX2 Optimized)}
-\label{tab:kyber_perf}
-\begin{tabular}{|l|c|c|c|}
-\hline
-\textbf{Variant} & \textbf{KeyGen (cycles)} & \textbf{Encaps (cycles)} & \textbf{Decaps (cycles)} \\
-\hline
-Kyber-512 & 33,856 & 45,200 & 34,572 \\
-Kyber-768 & 52,732 & 67,624 & 53,156 \\
-Kyber-1024 & 73,544 & 97,324 & 79,128 \\
-\hline
+\caption{Kyber Parameters}
+\label{tab:kyber-params}
+\renewcommand{\arraystretch}{1.2}
+\begin{tabular}{ll}
+\toprule
+\textbf{Parameter} & \textbf{Description} \\
+\midrule
+$n = 256$ & Polynomial degree \\
+$k \in \{2,3,4\}$ & Number of polynomials (varies by variant) \\
+$q = 3329$ & Prime modulus for coefficients \\
+$\eta_1, \eta_2$ & Noise bounds in key/ciphertext \\
+$d_u, d_v$ & Compression factors \\
+\bottomrule
 \end{tabular}
@@
 
-Kyber also includes a variant called \texttt{Kyber-90s}, which replaces the use of SHAKE with AES-256 and SHA2 primitives, enabling further acceleration on platforms with AES instruction support \cite{kyber2024}.
-% [SOURCE1]
+Kyber has been optimized for modern platforms.
+Figure~\ref{fig:kyber_perf} shows benchmark cycle counts for each variant on an Intel Haswell CPU using AVX2 vector instructions \cite{kyber2024}.
+These results reflect Kyber’s suitability for high-performance and embedded applications alike.
 
-The security of Kyber ultimately reduces to the hardness of the Shortest Vector Problem (SVP) via the Module-LWE problem.
-Recovering a message or key requires solving an instance of SVP in a structured lattice, a task that remains intractable even with quantum resources\cite{kyber2021}.
-% [BLOG1]
+\begin{figure}[h]
+\centering
+\begin{tikzpicture}
+\begin{axis}[
+    ybar,
+    bar width=12pt,
+    width=12cm,
+    height=7cm,
+    enlarge x limits=0.2,
+    ylabel={Cycles},
+    symbolic x coords={Kyber-512,Kyber-768,Kyber-1024},
+    xtick=data,
+    ymin=0,
+    legend style={at={(0.5,-0.15)}, anchor=north, legend columns=-1},
+    nodes near coords,
+    every node near coord/.append style={
+        font=\small,
+        rotate=90,   % vertical labels
+        anchor=west  % position so they sit neatly above bars
+    },
+    tick label style={font=\small},
+    label style={font=\small},
+    legend cell align={left},
+    grid=major
+]
 
-Due to its efficient structure and compact representation, Kyber has been adopted by numerous organizations:
-\begin{itemize}
-    \item Cloudflare included Kyber in the CIRCL cryptographic library
-    \item Amazon integrated hybrid Kyber modes into AWS KMS
-    \item IBM deployed Kyber in its quantum-secure tape drive
-\end{itemize} % [SOURCE1]
-These deployments demonstrate Kyber’s readiness for practical applications in secure communications, data protection, and cryptographic protocols. 
-The integration of Kyber into TEE frameworks like Keystone is increasingly relevant as systems must remain secure even in the presence of quantum-capable adversaries.
-TEEs benefit from Kyber’s compact keys and efficient operations, making it a strong candidate for enclave-based key exchange and attestation.
+% KeyGen
+\addplot[
+    pattern=north east lines,
+    pattern color=black
+] coordinates {
+    (Kyber-512,33856) (Kyber-768,52732) (Kyber-1024,73544)
+};
+
+% Encaps
+\addplot[
+    pattern=dots,
+    pattern color=black
+] coordinates {
+    (Kyber-512,45200) (Kyber-768,67624) (Kyber-1024,97324)
+};
+
+% Decaps
+\addplot[
+    pattern=crosshatch,
+    pattern color=black
+] coordinates {
+    (Kyber-512,34572) (Kyber-768,53156) (Kyber-1024,79128)
+};
+
+\legend{KeyGen, Encaps, Decaps}
+\end{axis}
+\end{tikzpicture}
+\caption{Kyber Performance on Haswell CPU (AVX2 Optimized)}
+\label{fig:kyber_perf}
+\end{figure}
+
+For platforms that lack efficient SHAKE implementations, Kyber also offers a variant called \texttt{Kyber-90s}, which uses AES-256 and SHA2 instead.
+This allows faster execution on devices with AES hardware acceleration, such as many embedded and mobile processors \cite{kyber2024}.
+
+Kyber has already been deployed in practice.
+Cloudflare integrated Kyber into its CIRCL library to support post-quantum TLS; Amazon added hybrid Kyber modes to its AWS Key Management Service; IBM deployed Kyber in a quantum-resistant tape storage system \cite{kyber2024}.
+These real-world uses demonstrate its maturity and readiness for deployment.
+
+Thanks to its compact keys and efficient operations, Kyber is particularly well-suited for resource-constrained environments such as Trusted Execution Environments (TEEs).
+In this thesis, we evaluate the performance of Kyber within RISC-V-based enclaves and explore best practices for securely configuring post-quantum key exchange in such systems.
 
@@
 First and foremost, Keystone is fully open-source, enabling complete transparency and extensibility.
-This contrasts sharply with the proprietary nature of SGX, TrustZone, and SEV, which limits research innovation and system adaptability \cite{turn0search0}.
-The open stance of Keystone allows researchers to inspect, customize, and formally verify its components, significantly reducing the Trusted Computing Base (TCB) compared to monolithic Secure World kernels or hardware-bound enclaves \cite{turn0search0}.
+This contrasts sharply with the proprietary nature of SGX, TrustZone, and SEV, which limits research innovation and system adaptability \cite{suzaki2021tsperf}.
+The open stance of Keystone allows researchers to inspect, customize, and formally verify its components, significantly reducing the Trusted Computing Base (TCB) compared to monolithic Secure World kernels or hardware-bound enclaves \cite{suzaki2021tsperf}.
+The estimated relative TCB sizes are shown in Figure~\ref{fig:tcb-size}, highlighting Keystone's minimal footprint.
 
@@
 ARM TrustZone operates using a dual-world model: a Secure World and a Normal World, switching through secure monitor calls (SMCs).
-While effective, this design suffers from overhead due to frequent context switching and increased TCB complexity since the Secure World often incorporates a full OS stack (e.g., OP‑TEE) \cite{turn0search0}.
-In contrast, Intel SGX allows enclave creation at user-level with encrypted memory regions but remains constrained by limited enclave size, single-mode operation, and dependency on Intel hardware, hindering portability and flexibility \cite{turn0search0}.
+While effective, this design suffers from overhead due to frequent context switching and increased TCB complexity since the Secure World often incorporates a full OS stack (e.g., OP‑TEE) \cite{suzaki2021tsperf}.
+In contrast, Intel SGX allows enclave creation at user-level with encrypted memory regions but remains constrained by limited enclave size, single-mode operation, and dependency on Intel hardware, hindering portability and flexibility \cite{suzaki2021tsperf}.
+Figure~\ref{fig:ctx-overhead} compares the context switch latency across these systems, showing Keystone with the lowest overhead.
 
@@
 \end{figure}
+
 Keystone effectively synthesizes the strengths of these platforms while avoiding their limitations.
-It supports dynamic enclave creation spanning both user and supervisor modes, with security ensured via low-overhead Physical Memory Protection (PMP) and a lightweight, verifiable Security Monitor operating in M-mode \cite{dayeol2019keystone}.
+It supports dynamic enclave creation spanning both user and supervisor modes, with security ensured via low-overhead Physical Memory Protection (PMP) and a lightweight, verifiable Security Monitor operating in M-mode \cite{Lee2019}.
 This design offers richer enclave functionality with reduced complexity.
@@
 It reports how architectural differences manifest in performance across ecosystems \cite{turn0search5}.
-Broader experimental evaluations, such as those comparing SGX, SEV, and Intel TDX, emphasize their varying overheads in CPU-, memory-, and I/O-bound workloads.
-These comparisons underscore that Keystone—while not the fastest in all cases—offers compelling performance coupled with openness and ease of extension \cite{turn0search2,turn0search10}.
+As shown in Figure~\ref{fig:bench-compare}, Keystone achieves near-native performance on CPU-bound workloads and maintains a balanced profile across memory and I/O operations when compared to SGX and TrustZone.
+
+The evolution of these TEEs over time is visualized in Figure~\ref{fig:tee-timeline}, placing Keystone's introduction in the broader historical context of trusted computing innovation.
+
+Feature-wise, trusted computing surveys place Keystone in a distinct category: unlike SGX and SEV, which emphasize encryption and hardware-rooted attestation, Keystone enables both secure and flexible enclave configurations, attested through PMP and monitor mechanisms, and retains openness—making it uniquely suitable for research-oriented and evolving hardware/software environments \cite{turn0search9}.
+Table~\ref{tab:tee-features} summarizes key differences in TEE capabilities across Keystone, SGX, TrustZone, and SEV.
+
+For these reasons, Keystone represents a promising alternative to closed-source TEEs, especially in academic and experimental contexts.
 
 \begin{figure}[htbp]
+\centering
+
+% --- Benchmark Comparison Chart ---
+\begin{subfigure}[b]{0.95\linewidth}
 \centering
@@
 \addplot[pattern=crosshatch] coordinates {(CPU-bound,95) (Memory-bound,60) (I/O-bound,50)};
-\legend{Keystone,SGX,TrustZone}
+\legend{Keystone, SGX, TrustZone}
 \end{axis}
 \end{tikzpicture}
-\caption{Illustrative TS‑Perf‑style benchmark comparison (REE=100\%).}
+\caption{Illustrative TS‑Perf-style benchmark comparison (REE = 100\%).}
 \label{fig:bench-compare}
-\end{figure}
+\end{subfigure}
 
-\begin{figure}[htbp]
+\vspace{1em}
+
+% --- Timeline ---
+\begin{subfigure}[b]{0.95\linewidth}
 \centering
@@
 \draw[->, thick] (0,0) -- (12,0);
-
 \foreach \x/\year/\name in {0/2004/TrustZone,3/2015/SGX,5/2016/SEV,8/2019/Keystone,10/2020/TDX} {
@@
 \end{tikzpicture}
-\caption{Timeline of selected TEE introductions.}
+\caption{Timeline of selected Trusted Execution Environment (TEE) introductions.}
 \label{fig:tee-timeline}
+\end{subfigure}
+
+\caption{Performance, evolution, and feature comparison of major TEEs.}
+\label{fig:tee-overview}
 \end{figure}
 
-Feature-wise, trusted computing surveys place Keystone in a distinct category: unlike SGX and SEV, which emphasize encryption and hardware-rooted attestation, Keystone enables both secure and flexible enclave configurations, attested through PMP and monitor mechanisms, and retains openness—making it uniquely suitable for research-oriented and evolving hardware/software environments \cite{turn0search9}.
+\vspace{1em}
 
@@
 \toprule
-Feature & Keystone & SGX & TrustZone & SEV \\
+\textbf{Feature} & \textbf{Keystone} & \textbf{SGX} & \textbf{TrustZone} & \textbf{SEV} \\
 \midrule
@@
 
-For these reasons, Keystone represents a promising alternative to closed-source TEEs, especially in academic and experimental contexts.
-
 \section{Performance Challenges in Secure Execution}
@@
 Operations such as arithmetic computations that remain within the processor’s execution pipeline typically show minimal slowdown.
-Empirical studies using TS-perf and Keystone’s own benchmarking tools (e.g., CoreMark and RV8) indicate that when enclave creation and destruction costs are excluded, the performance overhead of TEEs for CPU-intensive tasks is often below 1\%, with Intel SGX, ARM TrustZone, and RISC-V Keystone all exhibiting similar behavior \cite{Suzaki2021,dayeol2019keystone}.
+Empirical studies using TS-perf and Keystone’s own benchmarking tools (e.g., CoreMark and RV8) indicate that when enclave creation and destruction costs are excluded, the performance overhead of TEEs for CPU-intensive tasks is often below 1\%, with Intel SGX, ARM TrustZone, and RISC-V Keystone all exhibiting similar behavior \cite{suzaki2021tsperf,Lee2019}.
 
@@
 Memory isolation mechanisms—particularly those involving encryption, cache partitioning, and page fault handling—can significantly impact runtime performance.
-For instance, in the Keystone framework, applications such as \texttt{miniz} and \texttt{aes} that utilize large memory footprints have shown over 100\% slowdown when cache partitioning is enabled, due to increased cache misses and eviction rates \cite{dayeol2019keystone}.
-Similarly, TS-perf benchmarks report an 11\% reduction in performance for ARM TrustZone under random memory access workloads, largely attributable to memory encryption and hierarchy constraints \cite{Suzaki2021}.
-SGX-based systems experience a sharp decline in performance once the enclave's memory usage exceeds the size of the Enclave Page Cache (EPC), which triggers expensive paging behavior; this trend has been documented comprehensively by SGXGauge \cite{kumar2022sgxgauge}.
-Scientific computing workloads running on SGX and AMD SEV platforms have demonstrated slowdown factors ranging from 1.2× up to 126×, depending on workload characteristics and hardware constraints \cite{akkram2020scientific}.
+For instance, in the Keystone framework, applications such as \texttt{miniz} and \texttt{aes} that utilize large memory footprints have shown over 100\% slowdown when cache partitioning is enabled, due to increased cache misses and eviction rates \cite{Lee2019}.
+Similarly, TS-perf benchmarks report an 11\% reduction in performance for ARM TrustZone under random memory access workloads, largely attributable to memory encryption and hierarchy constraints \cite{suzaki2021tsperf}.
+SGX-based systems experience a sharp decline in performance once the enclave's memory usage exceeds the size of the Enclave Page Cache (EPC), which triggers expensive paging behavior.
+This degradation is illustrated in Figure~\ref{fig:epc-cliff}, where performance drops dramatically beyond the 256~MB EPC boundary.
+This trend has been comprehensively documented by SGXGauge \cite{kumar2022sgxgauge}, with slowdown factors ranging from 1.2× up to 126× depending on the workload and hardware configuration \cite{akkram2020scientific}.
 
@@
 Input/output (I/O) operations also suffer substantial performance penalties in TEE contexts due to the need for secure copying and context switching.
-In Keystone, file I/O performance as measured using IOZone demonstrates a 36.2\% decrease in write throughput and a 40.9\% reduction in read throughput when compared to execution in the Rich Execution Environment (REE).
-Repetitive operations, such as record rewriting, exhibit even greater overhead—up to 55.1\% \cite{dayeol2019keystone}.
-These trends are consistent across other platforms as well; TS-perf reveals that both OP-TEE and SGX-based file operations incur high latency due to their reliance on ECALLs and limited in-enclave storage APIs \cite{Suzaki2021}.
+In Keystone, file I/O performance as measured using IOZone demonstrates a 36.2\% decrease in write throughput and a 40.9\% reduction in read throughput compared to execution in the Rich Execution Environment (REE).
+Repetitive operations, such as record rewriting, exhibit even greater overhead—up to 55.1\% \cite{Lee2019}.
+These results are visualized in Figure~\ref{fig:io-throughput}, which compares enclave and REE performance across several I/O workloads.
+Similar trends have been reported for other platforms; TS-perf reveals that both OP-TEE and SGX-based file operations incur high latency due to reliance on ECALLs and limited in-enclave storage APIs \cite{suzaki2021tsperf}.
 
@@
 For example, \texttt{TEE\_GetREETime()}—which retrieves the time from the REE—requires a context switch and is consequently far slower than \texttt{TEE\_GetSystemTime()}, which accesses an enclave-resident hardware timer.
-The performance difference between these two methods can be as high as 30×, illustrating how even seemingly simple operations may be subject to substantial overhead within a TEE \cite{Suzaki2021}.
+The performance difference between these two methods can be as high as 30×, illustrating how even seemingly simple operations may be subject to substantial overhead within a TEE \cite{suzaki2021tsperf}.
 
@@
 Observations from TS-perf experiments showed unexpected enclave behavior such as thread migration between CPU cores and incorrect CPU utilization reporting, particularly in OP-TEE and Keystone environments.
-These anomalies were traced to a lack of TEE-awareness in OS-level schedulers and, in Keystone's case, bugs in the Eyrie runtime's interrupt handling mechanism \cite{Suzaki2021}.
+These anomalies were traced to a lack of TEE-awareness in OS-level schedulers and, in Keystone's case, bugs in the Eyrie runtime's interrupt handling mechanism \cite{suzaki2021tsperf}.
 
@@
 These steps consume between 2M–7M cycles per page for validation, and an additional 20k–30k cycles for each lifecycle event.
-The use of software-based cryptographic primitives amplifies these costs, though hardware acceleration could substantially mitigate them in future implementations \cite{dayeol2019keystone}.
+The use of software-based cryptographic primitives amplifies these costs, though hardware acceleration could substantially mitigate them in future implementations \cite{Lee2019}.
 

### 2025-08-22 chapters/benchmark.tex
citations & plot amendments are complete for chapters 1 - 2

@@
 \label{tab:hardware-impact}
-\begin{tabular}{@{}lcc@{}}
+\begin{tabular}{@{}lc@{}}
 \toprule
-\textbf{Metric} & \textbf{CPU cores} & \textbf{Memory (MB)} \\
+\textbf{Metric} & \textbf{Hardware scaling} \\
 \midrule
-Workloads improved                      & 2/6 (33\%) & 2/6 (33\%) \\
-Workloads slowed down                   & 2/6 (33\%) & 2/6 (33\%) \\
-Workloads unaffected                    & 2/6 (33\%) & 2/6 (33\%) \\
-Average performance change per step (\%)& -8.71      & -8.71      \\
-Median performance change per core step (\%) & -0.51      & -0.51      \\
+Workloads improved                      & 2/6 (33\%) \\
+Workloads slowed down                   & 2/6 (33\%) \\
+Workloads unaffected                    & 2/6 (33\%) \\
+Average performance change per step (\%) & -8.71     \\
+Median performance change per step (\%)  & -0.51     \\
 \bottomrule

### 2025-08-22 main.tex
new statistical findings & revised sections 4.4 & parts of 5.1
added random forest generation regression plot
citations & plot amendments are complete for chapters 1 - 2

@@
 \usepackage{enumitem}
+\usepackage{multirow}
+\sisetup{table-format=1.3}
+\usepackage{sansmath}
 \sloppy
@@
     tabsize=4
+}
+\lstdefinestyle{latexcode}{
+  basicstyle=\ttfamily\small,
+  keywordstyle=\bfseries\color{blue}, % style for keywords
+  morekeywords={Kyber-512,Kyber-768,Kyber-1024}, % words to highlight
+  frame=single,
+  backgroundcolor=\color{gray!5},
+  numbers=left,
+  numberstyle=\tiny\color{gray!70},
+  numbersep=8pt,
+  showstringspaces=false,
+  breaklines=true,
+  keepspaces=true,
+  tabsize=2,
+  captionpos=b
+}
+
+\lstdefinestyle{kemstyle}{
+  basicstyle=\ttfamily\small,
+  keywordstyle=\bfseries\color{blue!60!black},
+  morekeywords={KeyGen,Encapsulate,Decapsulate},
+  frame=single,
+  backgroundcolor=\color{gray!5},
+  numbers=left,
+  numberstyle=\tiny\color{gray!70},
+  numbersep=8pt,
+  showstringspaces=false,
+  breaklines=true,
+  keepspaces=true, % preserve spaces for alignment
+  tabsize=4,
+  captionpos=b
+}
+
+\lstdefinestyle{params}{
+  basicstyle=\ttfamily\small,
+  escapeinside={(*@}{@*)}, % allow LaTeX between (*@ and @*)
+  frame=single,
+  backgroundcolor=\color{gray!5},
+  numbers=left,
+  numberstyle=\tiny\color{gray!70},
+  numbersep=8pt,
+  showstringspaces=false,
+  breaklines=true,
+  keepspaces=true,
+  tabsize=4,
+  captionpos=b
 }

### 2025-08-22 preamble.tex
citations & plot amendments are complete for chapters 1 - 2

@@
 Die gewonnenen Erkenntnisse tragen dazu bei, den Einsatz von Keystone-Enklaven zu optimieren und ihre praktische Anwendbarkeit in sicherheitskritischen Bereichen auf RISC-V-Architekturen zu verbessern.},
+  Acknowledgements      = {
+    This is the place where you can thank people and institutions, do
+    not try to do this on the title page. The only exception is in
+    case you wrote your thesis while working or staying at a company or abroad. Then you
+},
 Numerische Bibliographie

### 2025-08-23 chapters/system-architecture.tex
chapter 3 complete

@@
 This strict separation ensures that the same benchmark codebase can be executed both natively (without isolation) and securely (within an enclave), without requiring structural changes—thus enabling a fair and consistent comparative analysis.
+
+The experimental procedure followed a systematic pipeline, outlined in Figure~\ref{fig:benchmarking-swimlane}, beginning with benchmark selection and system configuration, followed by dual-mode execution (native and enclave), and concluding with metric collection and comparative analysis.
 
@@
 
+Figure~\ref{fig:integrated_two_column_peripherals_below} provides a unified representation of the development workflow and the virtualized execution environment used in this thesis.
+On the left, the build process begins with the Keystone source code and proceeds through root filesystem generation via Buildroot, orchestration of key components through a \texttt{Make}-based system, and culminates in the launch of the full system within a QEMU-emulated RV64GC RISC-V platform.
+On the right, the runtime stack is depicted in detail: Ubuntu 22.04 LTS acts as the host operating system, supporting QEMU, which emulates the Keystone stack—including the Security Monitor, runtime, and enclave applications.
 \begin{figure}[htbp]
@@
 \begin{tikzpicture}[
-    node distance=1.2cm and 1.5cm,
-    every node/.style={
+    font=\scriptsize,
+    node distance=0.8cm and 1.0cm,
+    box/.style={
         draw,
         rounded corners,
-        align=center,
-        font=\small,
-        minimum width=2.8cm,
-        minimum height=1.2cm,
         fill=gray!15,
-        drop shadow={shadow xshift=0.5ex,shadow yshift=-0.5ex,opacity=0.1}
+        minimum width=3.6cm,
+        minimum height=0.9cm,
+        align=center
+    },
+    stack/.style={
+        draw,
+        rounded corners,
+        fill=gray!15,
+        minimum width=3.0cm,
+        minimum height=0.9cm,
+        align=center
     },
@@
         thick,
-        minimum width=3.5cm,
-        drop shadow={shadow xshift=0.8ex,shadow yshift=-0.8ex,opacity=0.2}
-    },
-    connector/.style={
-        ->,
-        thick,
-        >=stealth,
-        line width=0.6pt
-    }
-]
-
-% Nodes
-\node (host) [fill=gray!30, minimum height=1.4cm] {Ubuntu 22.04 LTS\\ Host OS};
-\node (qemu) [below=of host] {QEMU RV64GC Emulator};
-\node (keystone) [keystone, below=of qemu] {Keystone Stack\\ \footnotesize Security Monitor, Runtime, Enclaves};
-\node (virtio) [fill=gray!20, left=2cm of keystone] {VirtIO Devices\\ (net, storage)};
-\node (plugin) [fill=gray!20, right=2cm of keystone] {Optional Cache Plugin};
-
-% Edges
-\draw[connector] (host) -- (qemu);
-\draw[connector] (qemu) -- (keystone);
-\draw[connector] (virtio) -- (keystone);
-\draw[connector] (plugin) -- (keystone);
-
-% Annotation
-\node[font=\scriptsize, anchor=south west] at ([xshift=0.3cm,yshift=0.3cm]host.north west) {SSH/GDB access from Host};
-
-\end{tikzpicture}
-\caption{Overall virtualized testbed architecture illustrating the host OS, QEMU emulator, Keystone stack, and peripheral components such as VirtIO devices and optional cache plugin.}
-\label{fig:virtualized_testbed_architecture}
-\end{figure}
-
-The build process utilizes a combination of \texttt{Make} and Buildroot.
-Buildroot generates the embedded Linux root filesystem, cross-compiles necessary system libraries, and configures the Linux kernel running inside the QEMU guest.
-The Make-based system orchestrates compilation and deployment of Keystone components such as the runtime, Security Monitor, and kernel modules, ensuring reproducible builds targeting the QEMU RV64 environment.\footnote{Key configuration parameters include the QEMU platform, RV64 CPU architecture, and Buildroot settings optimized for Keystone enclave support.}
-
-\begin{figure}[htbp]
-\centering
-\begin{tikzpicture}[
-    node distance=1.5cm,
-    every node/.style={
-        draw,
-        rounded corners,
-        fill=gray!15,
-        font=\small,
-        minimum width=6cm,
-        minimum height=1.2cm,
-        align=center
+        minimum width=3.0cm
     },
@@
 ]
-% Nodes
-\node (src) {Source Code \\ Keystone, Kernel, Applications};
-\node (buildroot) [below=of src] {Buildroot \\ Rootfs + Kernel};
-\node (make) [below=of buildroot] {Make Orchestration \\ Keystone Runtime + Security Monitor};
-\node (launch) [below=of make] {QEMU Launch};
-\node (bench) [below=of launch] {Run Enclave / Benchmark};
 
-% Arrows
+% --- Left column: Build & Deployment Workflow ---
+\node (src) [box] {Source Code \\ Keystone, Kernel, Apps};
+\node (buildroot) [box, below=of src] {Buildroot \\ Rootfs + Kernel};
+\node (make) [box, below=of buildroot] {Make Orchestration};
+\node (launch) [box, below=of make] {QEMU Launch};
+\node (bench) [box, below=of launch] {Run Enclave / Benchmark};
+
 \foreach \from/\to in {src/buildroot, buildroot/make, make/launch, launch/bench} {
@@
 }
+
+% --- Right column: Testbed Stack ---
+\node (host) [stack, fill=gray!30, minimum height=1.1cm, right=5.0cm of src] {Ubuntu 22.04 LTS\\ Host OS};
+\node (qemu) [stack, below=of host] {QEMU RV64GC Emulator};
+\node (keystone) [stack, keystone, below=of qemu] {Keystone Stack\\ \footnotesize Sec. Monitor, Runtime, Enclaves};
+
+% Peripherals BELOW Keystone, side-by-side
+\node (virtio) [stack, fill=gray!20, below left=0.5cm and 0.5cm of keystone.south] {VirtIO Devices};
+\node (plugin) [stack, fill=gray!20, below right=0.5cm and 0.5cm of keystone.south] {Cache Plugin};
+
+% Testbed arrows
+\draw[arrow] (host) -- (qemu);
+\draw[arrow] (qemu) -- (keystone);
+\draw[arrow] (keystone.south) -- ++(0,-0.25) -| (virtio.north);
+\draw[arrow] (keystone.south) -- ++(0,-0.25) -| (plugin.north);
+
+% Cross-links from workflow to stack (no intersection)
+\draw[arrow, dashed] (launch.east) -- ++(0.8,0) |- (qemu.west);
+\draw[arrow, dashed] (bench.east)  -- ++(1.2,0) |- (keystone.west);
+
+% Annotation
+\node[font=\tiny, anchor=south west] at ([xshift=0.15cm,yshift=0.15cm]host.north west) {SSH/GDB access};
+
 \end{tikzpicture}
-\caption{Build and deployment workflow from source code to running enclave and benchmark execution.}
-\label{fig:build_deployment_workflow}
+\caption{Overview of the build workflow and virtualized Keystone testbed.
+The left column shows the step-by-step process of building and launching enclave applications, starting from source code to benchmark execution.
+The right column depicts the runtime testbed architecture on Ubuntu using QEMU for RISC-V emulation, including the Keystone software stack and supporting peripheral modules (VirtIO and cache plugin).
+Dashed arrows indicate execution-stage linkage between build steps and their corresponding runtime components.}
+\label{fig:integrated_two_column_peripherals_below}
 \end{figure}
 
-Launching the virtualized system is managed through a Makefile-driven workflow defined in the \texttt{run.mk} file, which encapsulates QEMU configuration and execution commands.
-Several runtime parameters are exposed as Makefile variables for flexibility, including:
+The design supports modular and reproducible deployment.
+Peripheral components such as VirtIO devices and an optional cache plugin are attached beneath the enclave runtime, enabling I/O functionality and microarchitectural performance instrumentation.
+Dashed arrows visually link specific build stages to their corresponding runtime targets, providing a clear and traceable path from compilation to execution.
+This integration highlights how the build system and execution environment are tightly coupled to support secure enclave development and testing.\footnote{Key configuration parameters include the QEMU platform target, RV64GC architecture, and Buildroot settings customized for Keystone compatibility.}
 
+Launching the system is handled via a Makefile-based interface, which wraps QEMU commands with tunable runtime parameters.
+The most relevant configuration variables and their default values are summarized in Table~\ref{tab:qemu-params}, including RAM size, core count, and SSH port mapping.
 \begin{table}[h]
 \centering
+\caption{Key QEMU launch parameters and defaults.}
+\label{tab:qemu-params}
 \begin{tabular}{l l p{6cm}}
@@
 \end{tabular}
-\caption{Key QEMU launch parameters and defaults.}
 \end{table}
 
-QEMU is launched with flags specifying the machine type (\texttt{virt}), boot ROM and firmware, kernel and root filesystem images, VirtIO devices for block storage and networking, and network forwarding for SSH access.
-Optionally, a cache plugin \cite{mandour2021cache} can be enabled to provide detailed logging of cache events such as hits, misses, and evictions.
-This plugin offers deeper insight into the memory hierarchy’s behavior during enclave execution, aiding analysis of performance impacts due to cache and memory isolation overhead.
-
-The cache plugin \cite{mandour2021cache} parameters are fully configurable at runtime, including:
-
+Optionally, the cache plugin~\cite{mandour2021cache} can be integrated into the QEMU runtime.
+This plugin enables logging of cache events—such as hits, misses, and evictions—and is particularly useful when analyzing memory performance of the overall system.
+Its configuration options are shown in Listing~\ref{lst:cache_config}.
+%\vspace{5em}
 \begin{lstlisting}[language=bash,caption={Configuration options for cache simulation},label={lst:cache_config}]
@@
 
-For example, to configure the cache plugin with 8 KB L1 instruction and data caches, 64-byte blocks, 4-way associativity, 2 cores, and LRU eviction, the following command is used:
+For example, Listing~\ref{lst:cache_cmd} demonstrates how to configure an L1 cache setup at runtime.
 
-\begin{lstlisting}
+\begin{lstlisting}[language=bash, caption={Configuration command for cache plugin}, label={lst:cache_cmd}]
 make QEMU_PLUGIN_ARGS="dcachesize=8192,dassoc=4,dblksize=64, \
@@
 
-Debugging support is available by enabling the \texttt{KEYSTONE\_DEBUG} flag, which starts QEMU’s built-in GDB server and halts execution at startup for remote debugging.
-After booting the QEMU guest, the Keystone kernel driver is loaded manually using:
+For kernel-level enclave debugging, QEMU’s built-in GDB server can be activated via the \texttt{KEYSTONE\_DEBUG} flag.
+Once the system boots, the Keystone kernel driver is loaded manually using the command in Listing~\ref{lst:modprobe}.
 
-\begin{lstlisting}
+\begin{lstlisting}[language=bash, caption={Loading the Keystone kernel driver}, label={lst:modprobe}]
 modprobe keystone-driver
@@
 
-Additional Makefile targets enable remote command execution over SSH and connection to the RISC-V GDB debugger for low-level inspection of enclave state.
-Debugging requires recompiling Keystone with debugging enabled, typically invoked as:
+In debugging scenarios, the Makefile target \texttt{debug-connect} can be used to attach a GDB session to the running guest.
+A full debugging session typically starts as shown in Listing~\ref{lst:debug}.
 
-\begin{lstlisting}
+\begin{lstlisting}[language=bash, label={lst:debug}, caption={Starting a QEMU session with GDB debugging enabled}]
 KEYSTONE_DEBUG=y make run
@@
 
-The Kyber implementation employed in this work is derived from a well-maintained, lightweight C reference library, chosen for its portability and compatibility with embedded systems.
-This implementation was integrated into the Keystone enclave framework with minimal modification to preserve fidelity to the original cryptographic routines and avoid introducing artifacts that might bias performance measurement.
-Both the IND-CPA (indistinguishability under chosen-plaintext attack) and IND-CCA (indistinguishability under chosen-ciphertext attack) variants of Kyber were considered, with the primary cryptographic operations — key generation, encapsulation, and decapsulation — being individually invoked inside the enclave to facilitate granular performance profiling.
+The Kyber implementation employed in this work is derived from a lightweight, standards-compliant C reference library, selected for its portability and compatibility with embedded systems.
+This implementation was integrated into the Keystone enclave framework with minimal modification to preserve the correctness and performance characteristics of the original routines, while avoiding structural changes that might distort measurement fidelity.
 
-The following critical Kyber functions were separately instrumented and timed to capture their distinct computational demands:
+Both the IND-CPA (indistinguishability under chosen-plaintext attack) and IND-CCA (indistinguishability under chosen-ciphertext attack) security models were considered.
+For performance profiling, the core cryptographic routines—key generation, encapsulation, and decapsulation—were individually invoked within the enclave.
+This allowed for fine-grained instrumentation and measurement of each function’s runtime behavior under enclave-induced constraints.
+
+The specific functions evaluated are shown in Listing~\ref{lst:kyber_api}, grouped by security level:
 
@@
 
+As shown in Listing~\ref{lst:coremark-output}, the CoreMark benchmark outputs the number of iterations per second (IPS), which is a key metric for evaluating system performance.
 \begin{lstlisting}[language=bash, caption={CoreMark sample output used to extract IPS}, label={lst:coremark-output}]
@@
 \centering
+\caption{Summary of representative configurations used in performance evaluation.}
+\label{tab:configurations}
 \footnotesize
@@
 \renewcommand{\arraystretch}{1.2}
-\begin{tabular}{|l|c|c|c|c|c|c|l|}
-\hline
+\begin{tabular}{lccccccl}
+\toprule
 \textbf{Configuration} & \textbf{RAM} & \textbf{Cores} & \textbf{Exec.} &
 \textbf{D‑Cache} & \textbf{I‑Cache} & \textbf{Policy} & \textbf{Assoc./Blk (D/I)} \\
-\hline
+\midrule
 Low‑End Baseline &
@@
 4\,KB & 4\,KB & LRU & 2‑way / 32B,\ 2‑way / 32B \\
-\hline
+%\hline
 Balanced Parallel &
@@
 8\,KB & 8\,KB & FIFO & 4‑way / 64B,\ 4‑way / 64B \\
-\hline
+%\hline
 Mid‑Range Sequential &
@@
 16\,KB & 8\,KB & Random & 8‑way / 32B,\ 4‑way / 64B \\
-\hline
+%\hline
 High‑End Parallel &
@@
 8\,KB & 16\,KB & LRU & 2‑way / 64B,\ 8‑way / 32B \\
-\hline
+%\hline
 Max Capacity Stress &
@@
 32\,KB & 32\,KB & FIFO & 16‑way / 64B,\ 16‑way / 64B \\
-\hline
+\bottomrule
 \end{tabular}
-\caption{Summary of representative configurations used in performance evaluation.}
-\label{tab:configurations}
 \end{table}
@@
 In addition to varying memory capacity, core count, and execution mode (sequential or parallel), each configuration was paired with a distinct cache arrangement to assess the sensitivity of enclave and native workloads to cache architecture.
-The cache parameters are summarised in Table~\ref{tab:configurations} and include:
+The cache parameters are summarised in Table~\ref{tab:configurations} and Listing~\ref{lst:cache_profiles}.
+
+These profiles were designed to represent common cache configurations found in embedded and general-purpose processors, ranging from small, low-associativity caches to larger, highly associative designs.
+The eviction policies (LRU, FIFO, Random) were selected to explore how different replacement strategies affect performance under varying workload characteristics.
 
@@
 
-In addition to these high-level metrics, detailed system-level runtime statistics were gathered using standard Linux facilities such as the \texttt{getrusage()} system call. These included:
+In addition to these high-level metrics, detailed system-level runtime statistics were gathered using standard Linux facilities such as the \texttt{getrusage()} system call.
+
+As shown in Listing~\ref{lst:getrusage}, this call provides insights into CPU time, memory usage, page faults, and context switching events.
+Collecting these indicators was crucial for diagnosing performance behaviors that could not be explained solely by execution time or throughput.
 

### 2025-08-26 chapters/introduction.tex
solo revisions complete

@@
 
-ARM TrustZone~\cite{suzaki2021ts, turn0search8} takes a different approach by partitioning the processor into secure and normal worlds, enabling lightweight isolation for Trusted Applications but with generally weaker isolation guarantees than SGX.
+ARM TrustZone~\cite{suzaki2021tsperf} takes a different approach by partitioning the processor into secure and normal worlds, enabling lightweight isolation for Trusted Applications but with generally weaker isolation guarantees than SGX.
 Its performance impact tends to be lower, yet TrustZone’s capabilities and APIs are less standardized, which complicates broad adoption and benchmarking~\cite{turn0search9}.
@@
 AMD SEV and newer technologies such as SEV-SNP offer memory encryption and integrity protections for virtual machines, providing scalable enclave-like isolation for cloud workloads.
-Coppolino et al.~\cite{turn0search2_coppolino} experimentally evaluated SEV and TDX against SGX, revealing performance and transparency trade-offs that influence their suitability for different application scenarios.
+Coppolino et al.~\cite{coppolino2025experimental} experimentally evaluated SEV and TDX against SGX, revealing performance and transparency trade-offs that influence their suitability for different application scenarios.
 
-In parallel, open-source efforts like Keystone~\cite{Lee2019, dayeol2019keystone} aim to democratize TEE development by providing a modular framework on RISC-V processors, supporting flexible enclave configurations and custom security policies.
+In parallel, open-source efforts like Keystone~\cite{Lee2019} aim to democratize TEE development by providing a modular framework on RISC-V processors, supporting flexible enclave configurations and custom security policies.
 Research into Keystone’s memory models and sharing mechanisms~\cite{yu2022elasticlave, lee2022cerberus} has demonstrated improved performance and formally verified security properties, addressing common TEE bottlenecks such as enclave memory management overhead.
@@
 Benchmarking and performance measurement frameworks are crucial to evaluate these TEEs fairly and holistically.
-TS-perf~\cite{suzaki2021ts} offers a unified performance suite targeting multiple TEE architectures, including SGX, TrustZone, and Keystone, facilitating comparative studies.
+TS-perf~\cite{suzaki2021tsperf} offers a unified performance suite targeting multiple TEE architectures, including SGX, TrustZone, and Keystone, facilitating comparative studies.
 Other benchmarking efforts include SGXGauge~\cite{kumar2022sgxgauge}, which specifically assesses Intel SGX overheads under different workload characteristics, and traditional CPU benchmarks like Dhrystone~\cite{weiss2002dhrystone, york2002benchmarking} and CoreMark~\cite{gal2012exploring}, adapted for enclave contexts.

### 2025-08-30 chapters/background.tex
chapter 3 complete
review complete

@@
 M-mode operates with the highest privilege and possesses unrestricted access to hardware resources, managing critical functions such as interrupt handling, exception processing, and physical memory protection~\cite{Lee2019}.
-Keystone capitalizes on this privilege hierarchy by situating its Security Monitor (SM)—the core Trusted Computing Base (TCB)—within M-mode.
-The SM is responsible for managing enclaves, enforcing strict memory isolation, and mediating access to sensitive operations.
+Keystone capitalizes on this privilege hierarchy by situating its Security Monitor (SM)—the Trusted Computing Base (TCB)—within M-mode.
+The Security Monitor manages enclaves and enforces memory isolation, while the runtime handles routine memory management and access mediation.
 By anchoring the TCB at this highest privilege level, Keystone minimizes the trusted surface exposed to potential attackers.
 
-A key hardware feature leveraged by Keystone is the Physical Memory Protection (PMP) unit, introduced in the RISC-V Privilege Specification v1.10.
+A key hardware feature leveraged by Keystone is the Physical Memory Protection (PMP) unit, introduced in the RISC-V Privilege Specification v1.10 \cite{keystone2025how}.
 PMP empowers M-mode software to define fine-grained access permissions for physical memory regions, specifying which lower privilege levels (U and S) may read, write, or execute code therein~\cite{keystone2025how}.
@@
   \end{tikzpicture}
-  \caption{TrustZone's dual-world architecture with SMC-managed transitions.}
+  \caption{TrustZone's dual-world architecture with secure monitor calls (SMC)-managed transitions.}
   \label{fig:trustzone}
@@
 Use Cases & Cloud security & Mobile security & Research, IoT \\
-Hardware Dependency & Intel CPUs & ARM SoCs & RISC-V \\
+Hardware Dependency & Intel CPUs & ARM SoCs & RISC-V + PMP \\
 \bottomrule
@@
 Keystone leverages the multiple privilege modes defined by the RISC-V specification—user mode (U-mode), supervisor mode (S-mode), and machine mode (M-mode)—to enforce security policies effectively.
-At the core of Keystone’s TCB is the Security Monitor (SM), which executes in M-mode and holds exclusive authority over enclave lifecycle management, memory protection configuration, and access control enforcement.
+At the heart of Keystone’s TCB is the Security Monitor (SM), which executes in M-mode and holds exclusive authority over enclave lifecycle management, memory protection configuration, and access control enforcement.
 The SM uses RISC-V’s Physical Memory Protection (PMP) mechanism to isolate enclave memory regions dynamically, ensuring strict access controls during enclave execution \cite{Lee2019}.
@@
 
-In RISC-V, the PMP mechanism is configured exclusively by machine-mode (M-mode) software—the most privileged execution level in the architecture—and applies access control policies to all lower-privileged software running in supervisor (S-mode) or user mode (U-mode).
+In RISC-V, the Physical Memory Protection (PMP) mechanism is configured exclusively by machine-mode (M-mode) software—the most privileged execution level in the architecture—and applies access control policies to all lower-privileged software running in supervisor (S-mode) or user mode (U-mode).
 Each PMP entry defines a memory region along with a set of permissions (read, write, execute) and an addressing mode.
@@
 This mechanism ensures that M-mode retains full control over physical memory while being able to delegate controlled access to untrusted software layers~\cite{riscvprivspec,Lee2019}.
+Conceptually, M-mode software acts as a trusted monitor or security supervisor, responsible for configuring isolation policies and mediating access between different software privilege levels.
 The basic hardware architecture of PMP within the RISC-V privilege hierarchy is illustrated in Figure~\ref{fig:pmp-arch}.
@@
 \centering
-\begin{tikzpicture}[node distance=2.5cm,>=Stealth]
+\begin{tikzpicture}[node distance=3cm,>=Stealth]
 \tikzstyle{block} = [draw, minimum width=3cm, minimum height=2cm, align=center, font=\small]
@@
 
-A typical RISC-V core supports a fixed number of PMP entries (commonly 8 or 16), each represented by two Control and Status Registers (CSRs): \texttt{pmpcfg} and \texttt{pmpaddr}.
+A typical RISC-V core supports a fixed number of PMP entries (commonly 8 or 16) \cite{keystone2025how}, each represented by two Control and Status Registers (CSRs): \texttt{pmpcfg} and \texttt{pmpaddr}.
 The \texttt{pmpcfg} register specifies the access permissions and address matching mode for each region, while \texttt{pmpaddr} defines the address bounds.
-RISC-V supports several region encoding modes:
+RISC-V supports several region encoding modes as shown in Table~\ref{tab:pmp-modes}, each with distinct trade-offs in terms of flexibility and granularity~\cite{riscvprivspec}.
 
-\begin{itemize}
-    \item \textbf{TOR (Top-of-Range):} Defines a region from the base address of the previous PMP entry up to the current address.
-    \item \textbf{NAPOT (Naturally Aligned Power-Of-Two):} Encodes both the base and size in a single register, enabling efficient representation of power-of-two-sized regions.
-    \item \textbf{NA4 (Naturally Aligned 4-byte):} Encodes a minimal 4-byte memory region.
-\end{itemize}
+% \begin{itemize}
+%     \item \textbf{TOR (Top-of-Range):} Defines a region from the base address of the previous PMP entry up to the current address.
+%     \item \textbf{NAPOT (Naturally Aligned Power-Of-Two):} Encodes both the base and size in a single register, enabling efficient representation of power-of-two-sized regions.
+%     \item \textbf{NA4 (Naturally Aligned 4-byte):} Encodes a minimal 4-byte memory region.
+% \end{itemize}
+
+\begin{table}[htbp]
+\centering
+\caption{PMP address-matching modes.}
+\label{tab:pmp-modes}
+\begin{tabular}{@{}lp{9cm}@{}} % adjust 9cm to fit your layout
+\toprule
+\textbf{Mode} & \textbf{Description} \\ \midrule
+TOR   (Top-of-Range) & Defines a region from the base address of the previous PMP entry up to the current address. \\
+NAPOT (Naturally Aligned Power-of-Two) & Encodes both the base and size in a single register, enabling efficient representation of power-of-two-sized regions. \\
+NA4   (Naturally Aligned 4-byte) & Encodes a minimal 4-byte memory region. \\
+\bottomrule
+\end{tabular}
+\end{table}
 
@@
 Security researchers have explored various enhancements and formal models to increase the reliability and verifiability of PMP-based isolation.
-For example, Cerberus~\cite{lee2022cerberus} builds a formally verified memory-sharing abstraction over PMP to enable safe communication between enclaves and untrusted components.
-Similarly, HexBox~\cite{hexbox2018} leverages PMP for real-time embedded security by associating memory regions with specific control-flow paths in embedded firmware, demonstrating PMP's applicability beyond general-purpose computing.
+For example, Cerberus~\cite{lee2022cerberus} builds a formally verified memory-sharing abstraction over PMP to enable safe communication between enclaves and untrusted components, while HexBox~\cite{hexbox2018} leverages PMP for real-time embedded security by associating memory regions with specific control-flow paths in firmware, demonstrating PMP's applicability beyond general-purpose computing.
+Beyond CPU-managed memory, \textbf{IOPMP} highlights that malicious or compromised I/O devices could also access physical memory, potentially undermining traditional PMP protections, making it essential to extend isolation mechanisms to cover I/O memory as well.
 Table~\ref{tab:pmp-limitations} summarizes key limitations in PMP and highlights recent research directions aimed at overcoming them.
@@
 \includegraphics[width=0.9\linewidth]{figures/kyber_768.png}
-\caption{Kyber-768 Key Exchange Algorithm.}
+\caption{Kyber-768 Key Exchange Algorithm \cite{Pathum_2025}}
 \label{fig:kyber_768}
@@
 \renewcommand{\arraystretch}{1.2}
-\begin{tabular}{|l|c|c|c|}
-\hline
+\begin{tabular}{lccc}
+\toprule
 \textbf{Scheme} & \textbf{Private Key} & \textbf{Public Key} & \textbf{Ciphertext} \\
-\hline
-Kyber-512 & 1632 & 800 & 768 \\
-Kyber-768 & 2400 & 1184 & 1088 \\
-Kyber-1024 & 3168 & 1568 & 1568 \\
-RSA-3072 & 384 & 384 & 384 \\
-RSA-15360 & 1920 & 1920 & 1920 \\
-\hline
+\midrule
+\rowcolor{lightgray} Kyber-512 & 1632 & 800 & 768 \\
+\rowcolor{lightgray} Kyber-768 & 2400 & 1184 & 1088 \\
+\rowcolor{lightgray} Kyber-1024 & 3168 & 1568 & 1568 \\
+\rowcolor{darkgray} RSA-3072 & 384 & 384 & 384 \\
+\rowcolor{darkgray} RSA-15360 & 1920 & 1920 & 1920 \\
+\bottomrule
 \end{tabular}
@@
 Empirical comparisons further highlight Keystone’s competitive performance.
-TS‑Perf, a systematic benchmarking framework, measured execution time across SGX, TrustZone, and RISC‑V Keystone using identical binaries and workloads (e.g., internal APIs, matrix operations, memory and storage access).
+TS‑Perf \cite{suzaki2021tsperf}, a systematic benchmarking framework, measured execution time across SGX, TrustZone, and RISC‑V Keystone using identical binaries and workloads (e.g., internal APIs, matrix operations, memory and storage access).
 It reports how architectural differences manifest in performance across ecosystems \cite{turn0search5}.
@@
 To address these performance issues, researchers have proposed alternative memory models that relax or restructure the traditional enclave-host memory boundaries.
-One such innovation is Elasticlave, a shared-memory model designed for enclaves that reduces isolation costs while preserving strong security properties.
-Elasticlave achieves an order of magnitude better performance compared to traditional spatial isolation models, with overheads as low as 10\% versus 100× to 1000× for spatially isolated configurations \cite{yu2022elasticlave}.
-Similarly, Eleos introduces self-paging within enclaves, allowing the enclave itself to manage virtual memory.
-This eliminates host OS involvement in page fault handling and mitigates context-switching and cache pollution overheads, resulting in significantly improved runtime efficiency \cite{orenbach2023eleos}.
-A related approach, Cerberus, applies formal methods to build a provably secure enclave memory-sharing model on Keystone.
-It enables safe data sharing with external components without compromising security, all while reducing redundancy and memory copying overheads \cite{lee2022cerberus}.
+One such innovation is Elasticlave \cite{yu2022elasticlave}, a shared-memory model designed for enclaves that reduces isolation costs while preserving strong security properties.
+Elasticlave achieves an order of magnitude better performance compared to traditional spatial isolation models, with overheads as low as 10\% versus 100× to 1000× for spatially isolated configurations.
+Similarly, Eleos \cite{orenbach2023eleos} introduces self-paging within enclaves, allowing the enclave itself to manage virtual memory.
+This eliminates host OS involvement in page fault handling and mitigates context-switching and cache pollution overheads, resulting in significantly improved runtime efficiency.
+A related approach, Cerberus \cite{lee2022cerberus}, applies formal methods to build a provably secure enclave memory-sharing model on Keystone.
+It enables safe data sharing with external components without compromising security, all while reducing redundancy and memory copying overheads.
 
 Input/output (I/O) operations also suffer substantial performance penalties in TEE contexts due to the need for secure copying and context switching.
-In Keystone, file I/O performance as measured using IOZone demonstrates a 36.2\% decrease in write throughput and a 40.9\% reduction in read throughput compared to execution in the Rich Execution Environment (REE).
-Repetitive operations, such as record rewriting, exhibit even greater overhead—up to 55.1\% \cite{Lee2019}.
+In Keystone, file I/O performance as measured using IOZone \cite{Lee2019} demonstrates a 36.2\% decrease in write throughput and a 40.9\% reduction in read throughput compared to execution in the Rich Execution Environment (REE).
+Repetitive operations, such as record rewriting, exhibit even greater overhead—up to 55.1\%.
 These results are visualized in Figure~\ref{fig:io-throughput}, which compares enclave and REE performance across several I/O workloads.
-Similar trends have been reported for other platforms; TS-perf reveals that both OP-TEE and SGX-based file operations incur high latency due to reliance on ECALLs and limited in-enclave storage APIs \cite{suzaki2021tsperf}.
+Similar trends have been reported for other platforms; TS-perf reveals that both Open Portable Trusted Execution Environment (OP-TEE) and SGX-based file operations incur high latency due to reliance on ECALLs and limited in-enclave storage APIs \cite{suzaki2021tsperf}.
 
@@
 An additional source of performance variability lies in runtime behavior and process scheduling.
-Observations from TS-perf experiments showed unexpected enclave behavior such as thread migration between CPU cores and incorrect CPU utilization reporting, particularly in OP-TEE and Keystone environments.
+Observations from TS-perf experiments showed unexpected enclave behavior such as thread migration between CPU cores and incorrect CPU utilization reporting, particularly in Open Portable Trusted Execution Environment (OP-TEE) and Keystone environments.
 These anomalies were traced to a lack of TEE-awareness in OS-level schedulers and, in Keystone's case, bugs in the Eyrie runtime's interrupt handling mechanism \cite{suzaki2021tsperf}.
@@
 Monitoring and profiling enclave performance in real-time remains a difficult task.
-TEEMon provides a lightweight performance monitoring framework designed specifically for TEE environments.
-Though it introduces between 5–17\% overhead, it enables continuous visibility into runtime behavior and facilitates the identification of performance bottlenecks that would otherwise be hidden \cite{krahn2020teemon}.
+TEEMon \cite{krahn2020teemon} provides a lightweight performance monitoring framework designed specifically for TEE environments.
+Though it introduces between 5–17\% overhead, it enables continuous visibility into runtime behavior and facilitates the identification of performance bottlenecks that would otherwise be hidden.
 

### 2025-08-30 chapters/benchmark.tex
review complete

@@
 
-Each benchmark was executed both \emph{sequentially} and \emph{in parallel}, ten times each, and the resulting figures compare performance and stability across both modes.
-
+Each benchmark was executed both \emph{sequentially} and \emph{in parallel}, ten times each, allowing us to evaluate both baseline performance and parallel scalability, while repeated runs improve statistical reliability and reduce the influence of outliers.
+
+%The sequential runs (denoted \texttt{Run10\_SEQ}) executed a single instance of the workload in isolation, while the parallel runs (denoted \texttt{Run10}) launched multiple concurrent instances, with results merged to simulate a multi-threaded environment.
+This dual-mode execution provides insights into how well each workload scales with concurrency, as well as the overheads introduced by context switching and resource contention.
 % ------------------------
@@
 % ------------------------
-Figure~\ref{fig:coremark_perf_comparison_log} reports CoreMark's performance profile using the latest data from Run~10 of both sequential and parallel execution modes.
+Figure~\ref{fig:coremark_perf_comparison_log} reports CoreMark's performance profile using the data from Run~10 of both sequential and parallel execution modes.
 Execution time remained consistent, with the sequential case averaging approximately 101.17 seconds of wall-clock time and the parallel case averaging around 103.37 seconds.
@@
         };
-        \legend{Run10, Run10\_SEQ}
+        \legend{Parallel, Sequential}
         \end{axis}
         \end{tikzpicture}
-        \caption{Log‑scale comparison of all key metrics for CoreMark 64M\_1Core: Run10 vs Run10\_SEQ.}
+        \caption{Log‑scale comparison of all key metrics for CoreMark 64M\_1Core: Parallel vs Sequential.}
         \label{fig:coremark_perf_comparison_log}
@@
         };
-        \legend{Run10, Run10\_SEQ}
+        \legend{Parallel, Sequential}
         \end{axis}
         \end{tikzpicture}
-        \caption{Log‑scale comparison of all key metrics for Dhrystone 64M\_1Core: Run10 vs Run10\_SEQ.}
+        \caption{Log‑scale comparison of all key metrics for Dhrystone 64M\_1Core: Parallel vs Sequential.}
         \label{fig:dhrystone_perf_comparison_log}
     \end{subfigure}
-    \caption{Comparison of CoreMark and Dhrystone benchmark performance metrics (log-scale) under Run10 and Run10\_SEQ modes.}
+    \caption{Comparison of CoreMark and Dhrystone benchmark performance metrics (log-scale) under Parallel and Sequential modes.}
 \end{figure}
 
-% ------------------------
-% CPU time figures
-% ------------------------
 Memory behavior was stable and well within constraints, with peak resident set size (RSS) at 1692~KB for the sequential run and 1712~KB for the parallel run—both well below the 64~MB memory cap.
@@
 
-% ------------------------
-% Throughput figures
-% ------------------------
 Figure~\ref{fig:dhrystone_perf_comparison_log} presents detailed results from the Dhrystone benchmark under native conditions.
@@
 
-% ------------------------
-% Memory usage figures
-% ------------------------
 Memory footprint remained small (peak RSS ~1.7~MB) and stable, with 82--86 minor page faults per run.
 
-% ------------------------
-% Context switches figures
-% ------------------------
 Voluntary context switches were minimal (0--1 sequential, 0--3 parallel), while involuntary context switches showed a stark contrast: approximately 350 in the sequential case versus over 25,800 in the parallel case.
 
-% ------------------------
-% Cycle counts figures (Kyber only)
-% ------------------------
 Kyber performance was evaluated using the official \texttt{test\_speed512} binary from the reference implementation's \texttt{ref/} directory.\footnote{See \url{https://github.com/pq-crystals/kyber}.
@@
 
-Across all three workloads, cache behavior was uniformly stable.
-Monitoring via QEMU’s cache plugin confirmed low instruction and data cache miss rates, consistent with compact memory footprints and tight loop-local access patterns.
-Instruction cache performance benefited from small binary sizes and the absence of dynamic control flow, while data cache pressure remained minimal due to statically sized inputs and short-lived stack allocations.
-These characteristics consistently indicate compute-bound, not memory-bound, behavior in the native case.
+Across all three workloads, cache behavior was stable.
+QEMU's cache plugin confirmed low instruction and data cache miss rates, reflecting compact memory footprints and tight loop-local access patterns.
+Instruction streams were \emph{cache-friendly}, thanks to small binaries and predictable control flow, while data cache pressure remained minimal due to static inputs and short-lived stack allocations.
+This indicates that the workloads are primarily compute-bound rather than memory-bound.
 
@@
         };
-        \legend{Native (Run10\_SEQ), Enclave (Run10\_SEQ)}
+        \legend{Native (Sequential), Enclave (Sequential)}
         \end{axis}
         \end{tikzpicture}
-        \caption{Log-scale comparison of Dhrystone 64M\_1Core: Native vs Enclave (Run10\_SEQ).}
+        \caption{Log-scale comparison of Dhrystone 64M\_1Core: Native vs Enclave (Sequential).}
         \label{fig:dhrystone_native_vs_enclave_log}
@@
     };
-    \legend{Native (Run10\_SEQ avg), Enclave (Run10\_SEQ avg)}
+    \legend{Native (Sequential avg), Enclave (Sequential avg)}
     \end{axis}
@@
         \end{tikzpicture}
-        \caption{Log-scale comparison of Dhrystone Enclave (KE): Run10\_SEQ vs Run5*.\\
+        \caption{Log-scale comparison of Dhrystone Enclave: Sequential vs Parallel (5 Runs)*.\\
         \textit{* Run5 includes only partial runs (others failed).}}
@@
         \end{tikzpicture}
-        \caption{Log-scale comparison of CoreMark Enclave (KE) performance: Run10\_SEQ vs Run5*.\\
+        \caption{Log-scale comparison of CoreMark Enclave performance: Sequential vs Parallel (5 Runs)*.\\
         \textit{* Run5 includes only 3 out of 5 completed runs. Runs 4 and 5 failed.}}
@@
     axis x line=none,
-    bar shift=10pt
+    bar shift=10pt,
+    clip=false, % allows legend outside the plot
+    legend style={
+        at={(0.5,1.05)}, % relative to the axis
+        anchor=south,
+        legend columns=1
+    }
 ]
@@
 
-To investigate this, we performed a systematic evaluation of six representative workloads spanning both native and secure execution modes: \texttt{coremark}, \texttt{coremark.ke}, \texttt{dhrystone}, \texttt{enclave.ke}, \texttt{kyber}, and \texttt{kyber.ke}.
+To investigate this, we performed a systematic evaluation of six representative workloads spanning both native and secure execution modes: \texttt{coremark}, \texttt{dhrystone}, and \texttt{kyber} (native execution), as well as \texttt{coremark.ke}, \texttt{dhrystone.ke}, and \texttt{kyber.ke} (executed within a Keystone enclave).
 Each workload was benchmarked across multiple hardware configurations, incrementally increasing the number of CPU cores and the amount of available physical memory.
@@
 
-As shown in Figure~\ref{fig:anova-kruskal}, none of the $p$-values from either ANOVA or Kruskal--Wallis testing fell below the $\alpha = 0.05$ significance threshold.
-This indicates a lack of statistically significant performance variation across hardware scaling steps for all workloads tested.
-The $R^2$ values in subfigure~\ref{fig:r2_grouped} support this, with low model fit observed in both linear and Random Forest regressions, suggesting weak or no explanatory power.
+As shown in Figure~\ref{fig:anova-kruskal}, none of the $p$-values from ANOVA or Kruskal--Wallis testing fell below $\alpha = 0.05$, indicating no statistically significant performance variation across hardware scaling steps.
+The $R^2$ values in subfigure~\ref{fig:r2_grouped} support this, with low fit in both linear and Random Forest regressions, suggesting weak or no explanatory power.
 

### 2025-08-30 chapters/discussion.tex
review complete

@@
 
-This chapter transitions from measurement to meaning: it connects the observed data to the architectural principles underpinning enclave execution, therebyinforming both future research and practical deployment of secure computing systems based on open RISC-V architectures.
+This chapter transitions from measurement to meaning: it connects the observed data to the architectural principles underpinning enclave execution, thereby informing both future research and practical deployment of secure computing systems based on open RISC-V architectures.
 
@@
 However, workloads with more intensive memory access or intricate memory patterns, such as Kyber (a post-quantum cryptographic algorithm), experience much higher overhead.
-The Kyber algorithm, which involves significant pointer dereferencing and array indexing, incurs a significant penalty as every memory access is subject to additional security checks and encryption layers.
-This is compounded by the latency-sensitive nature of cryptographic workloads, where even small delays can result in significant performance degradation.
-The nearly threefold slowdown in enclave execution compared to native execution is primarily due to the added latency from memory encryption and integrity checks, which are inherently more expensive for memory-intensive algorithms.
+Kyber involves frequent pointer dereferencing and array indexing, where each operation depends on the completion of the previous memory access.
+Within an enclave, every memory access triggers additional security checks, including encryption, integrity verification, and, in some cases, TLB flushes on context switches, adding fixed latency per operation.
+
+Cryptographic workloads are particularly sensitive to such latency because the sequential, dependent nature of their computations prevents effective hiding of memory delays.
+Even small delays accumulate across thousands of operations, resulting in a substantial slowdown.
+Consequently, the nearly threefold slowdown observed for Kyber in enclave execution compared to native execution is primarily due to the added latency from memory encryption, integrity checks, and occasional TLB flushes, which disproportionately impact memory-intensive, latency-bound workloads.
 
@@
 \centering
+
+% --- PMP entry usage chart ---
 \begin{tikzpicture}[scale=1, transform shape]
@@
     enlarge x limits=0.15,
-    legend style={at={(0.5,-0.3)},anchor=north,legend columns=2},
+    legend style={at={(0.5,-0.4)},anchor=north,legend columns=2},
     pattern color=black,
@@
 \end{tikzpicture}
-\caption{Fragmentation and misalignment increase PMP entry usage, showing how non-ideal alignment leads to more PMP entries per enclave.}
-\label{fig:pmp-fragmentation-entries}
-\end{figure}
 
-To address overhead and scalability challenges, \emph{enclave multiplexing} offers a promising software-level strategy by consolidating multiple lightweight workloads into a single enclave.
-This reduces the number of PMP entries required by reusing them across tasks, optimizing resource usage and improving scalability.
-An example of a multiplexed enclave task dispatcher is shown in Listing~\ref{lst:enclave-mux}.
+\vspace{1em}
 
-\begin{lstlisting}[language=Python, caption={Multiplexed enclave task dispatcher.}, label={lst:enclave-mux}]
-def enclave_dispatcher(task_id, input_data):
-    if task_id == 1:
-        return process_crypto(input_data)
-    elif task_id == 2:
-        return handle_key_store(input_data)
-    elif task_id == 3:
-        return run_diagnostics(input_data)
-    else:
-        raise ValueError("Unknown task")
-\end{lstlisting}
-
-Furthermore, the trade-off between memory usage and security introduces additional complexity in system design.
-Enclaves need to be carefully sized and aligned to fit within the constraints of PMP's encoding scheme, particularly the NAPOT (Naturally Aligned Power-Of-Two) encoding used to specify memory regions.
-For cryptographic workloads that often require non-contiguous memory layouts, inefficient PMP entry management leads to fragmentation and resource wastage, increasing overhead and limiting concurrent enclaves.
-
-\begin{figure}[htbp]
-\centering
+% --- NAPOT fragmentation diagram ---
 \begin{tikzpicture}[scale=1, transform shape,font=\small]
@@
 \end{tikzpicture}
-\caption{Fragmentation overhead caused by mapping non-contiguous memory regions to NAPOT-aligned blocks. The hatched areas represent wasted padding required for alignment.}
-\label{fig:napot-fragmentation}
+
+\caption{Top: PMP entry usage increases with fragmentation and misalignment.
+Bottom: Fragmentation overhead caused by mapping non-contiguous memory regions to NAPOT-aligned blocks, with hatched areas representing wasted padding.}
+\label{fig:pmp-and-napot-fragmentation}
 \end{figure}
+
+To address overhead and scalability challenges, \emph{enclave multiplexing} offers a promising software-level strategy by consolidating multiple lightweight workloads into a single enclave.
+This reduces the number of PMP entries required by reusing them across tasks, optimizing resource usage and improving scalability.
+An example of a multiplexed enclave task dispatcher is shown in Listing~\ref{lst:enclave-mux}.
+\vspace{8em}
+\begin{lstlisting}[language=Python, caption={Multiplexed enclave task dispatcher.}, label={lst:enclave-mux}]
+def enclave_dispatcher(task_id, input_data):
+    if task_id == 1:
+        return process_crypto(input_data)
+    elif task_id == 2:
+        return handle_key_store(input_data)
+    elif task_id == 3:
+        return run_diagnostics(input_data)
+    else:
+        raise ValueError("Unknown task")
+\end{lstlisting}
+
+To address overhead and scalability challenges, \emph{enclave multiplexing} offers a promising software-level strategy by consolidating multiple lightweight workloads into a single enclave.
+This reduces the number of PMP entries required by reusing them across tasks, optimizing resource usage and improving scalability.
+An example of a multiplexed enclave task dispatcher is shown in Listing~\ref{lst:enclave-mux}.
+
+Furthermore, the trade-off between memory usage and security introduces additional complexity in system design.
+Enclaves need to be carefully sized and aligned to fit within the constraints of PMP's encoding scheme, particularly the NAPOT (Naturally Aligned Power-Of-Two) encoding used to specify memory regions.
+For cryptographic workloads that often require non-contiguous memory layouts, inefficient PMP entry management leads to fragmentation and resource wastage, increasing overhead and limiting concurrent enclaves.
+
+% \begin{figure}[htbp]
+% \centering
+% \begin{tikzpicture}[scale=1, transform shape,font=\small]
+% % Memory bar
+% \draw[thick] (0,0) rectangle (12,1);
+% \node[anchor=south] at (0,1.1) {Base};
+% \node[anchor=south] at (12,1.1) {Top};
+
+% % Needed regions (non-contiguous)
+% \draw[fill=gray!30] (0.5,0.1) rectangle (2.2,0.9) node[pos=.5] {\footnotesize R1};
+% \draw[fill=gray!30] (3.8,0.1) rectangle (4.6,0.9) node[pos=.5] {\footnotesize R2};
+% \draw[fill=gray!30] (7.0,0.1) rectangle (9.1,0.9) node[pos=.5] {\footnotesize R3};
+
+% % NAPOT allocations (outer boxes) and padding (hatched)
+% \draw[thick] (0.25,-0.3) rectangle (2.5,-1.1) node[pos=.5] {\footnotesize NAPOT fit R1};
+% \draw[pattern=north east lines,pattern color=black] (0.25,-0.3) rectangle (0.5,-1.1);
+% \draw[pattern=north east lines,pattern color=black] (2.2,-0.3) rectangle (2.5,-1.1);
+
+% \draw[thick] (3.5,-0.3) rectangle (5.1,-1.1) node[pos=.5] {\footnotesize NAPOT fit R2};
+% \draw[pattern=north east lines,pattern color=black] (3.5,-0.3) rectangle (3.8,-1.1);
+% \draw[pattern=north east lines,pattern color=black] (4.6,-0.3) rectangle (5.1,-1.1);
+
+% \draw[thick] (6.7,-0.3) rectangle (9.3,-1.1) node[pos=.5] {\footnotesize NAPOT fit R3};
+% \draw[pattern=north east lines,pattern color=black] (6.7,-0.3) rectangle (7.0,-1.1);
+% \draw[pattern=north east lines,pattern color=black] (9.1,-0.3) rectangle (9.3,-1.1);
+
+% \node[anchor=west] at (0,-1.5) {\footnotesize Hatched = padding/waste due to power-of-two alignment};
+% \end{tikzpicture}
+% \caption{Fragmentation overhead caused by mapping non-contiguous memory regions to NAPOT-aligned blocks. The hatched areas represent wasted padding required for alignment.}
+% \label{fig:napot-fragmentation}
+% \end{figure}
 
@@
 Consequently, configuring enclave memory regions to maximize contiguity and alignment is paramount; doing so minimizes PMP entry consumption, reduces fragmentation, and alleviates potential bottlenecks in enclave creation and management.
-Developers are advised to meticulously design memory layouts that conform to PMP’s alignment constraints, thereby ensuring that enclave memory mappings are as compact and contiguous as possible (see Figure~\ref{fig:pmp-fragmentation-group}).
+Developers are advised to meticulously design memory layouts that conform to PMP’s alignment constraints, thereby ensuring that enclave memory mappings are as compact and contiguous as possible (see Figure~\ref{fig:pmp-and-napot-fragmentation}).
 
@@
 Given the PMP entry limitations and the observed overheads of enclave context switches, a key deployment recommendation is to limit the number of simultaneously active enclaves to avoid PMP exhaustion, which can lead to explicit enclave creation failures (see Section~\ref{sec:workload-sensitivity}, Chapter~\ref{chap:benchmarking}).
-Practical configurations should, therefore, consolidate workloads into fewer, more capable enclaves when possible—a strategy known as enclave multiplexing.
+Practical configurations should, therefore, consolidate workloads into fewer, more capable enclaves when possible—a strategy known as enclave multiplexing \cite{Occlum2020}.
 This approach not only conserves precious PMP entries but also reduces synchronization overhead and inter-enclave communication latency.

### 2025-08-30 chapters/system-architecture.tex
review complete

@@
 To assess the performance implications of enclave-based isolation, each benchmark was implemented as a standalone enclave application (referred to as an \textit{eapp}), accompanied by a corresponding host application responsible for enclave lifecycle management.
-The host application, running in user space, handles enclave initialization, benchmark loading, invocation of enclave entry points, and inter-domain communication via edge calls.
+The host application, running in user space, handles enclave initialization, benchmark loading, invocation of enclave entry points, and inter-domain communication via edge calls, while the enclave application \texttt{(eapp)} also runs in user mode.
 In contrast, the enclave application is fully isolated by the Keystone runtime and contains only the benchmark logic and associated data structures.
@@
 For kernel-level enclave debugging, QEMU’s built-in GDB server can be activated via the \texttt{KEYSTONE\_DEBUG} flag.
-Once the system boots, the Keystone kernel driver is loaded manually using the command in Listing~\ref{lst:modprobe}.
-
-\begin{lstlisting}[language=bash, caption={Loading the Keystone kernel driver}, label={lst:modprobe}]
-modprobe keystone-driver
-\end{lstlisting}
+Once the system boots, the Keystone kernel driver is loaded manually using the command: \texttt{modprobe keystone-driver}.
 
@@
 The specific functions evaluated are shown in Listing~\ref{lst:kyber_api}, grouped by security level:
-
+\vspace{5em}
 \begin{lstlisting}[language=bash,caption={Kyber cryptographic API functions},label={lst:kyber_api}]
@@
 To ensure measurement accuracy and reduce variability from transient system states, ten iterations were run per configuration with the first iteration discarded as a warm-up phase.
-Execution timing was obtained using the enclave’s internal high-resolution timer API, capturing secure-world runtime exclusively.
-In parallel, external system-level metrics — such as CPU usage, memory footprint, and context switching counts — were gathered from the host environment to provide contextual insights into system overhead and resource utilization.
+Execution timing was obtained using the enclave’s internal high-resolution timer API, capturing only secure-world runtime and thus remaining isolated from background activity in the Linux kernel.
+In parallel, external system-level metrics — including CPU usage, memory footprint, and context switching counts — were gathered from the host environment, reflecting overall system overhead and resource utilization influenced by the kernel and other processes.
 
@@
 Raw benchmark outputs, kernel logs, and cache plugin trace files were archived and documented.
-A complete snapshot of the experimental infrastructure, along with scripts to regenerate results, is provided in Appendix~\ref{appendix:cache-plugin}.
+Complete snapshots of the experimental infrastructure, along with scripts to regenerate results, are publicly available\footnote{\url{https://github.com/coast-moonlight/keystone.git}}.
 This archival effort ensures that all reported results can be independently verified and reproduced by future researchers or evaluators.

### 2025-08-30 preamble.tex
review complete

--- 2025-05-27 preamble.tex
@@
 Masterarbeit,
+Logo-Dateiname        = {uzl-iti-logo-uzl.pdf},
 Verfasst              = {am}{Institut für Technische Informatik},
@@
 Betreuerin            = {Dr.-Ing. Saleh Mulhem},
-Mit Unterstützung von = {Henry Strunck},
+Mit Unterstützung von = {Henrik Strunck, M.Sc},
 Studiengang           = {IT Security},
-Datum                 = {31. August 2025},
-Abstract = {This thesis evaluates the performance of Keystone enclaves, an open-source Trusted Execution Environment (TEE) designed for RISC-V platforms.
-Keystone enables secure computation by isolating sensitive workloads from the rest of the system.
-To assess its performance, the study benchmarks critical system parameters, including the number of CPU cores, cache size, memory allocation, and enclave configuration.
-Using industry-standard benchmarking tools such as lmbench and CoreMark, the research measures execution time, context switch overhead, memory latency, and CPU utilization across a range of hardware and software configurations.
-The experimental analysis identifies key performance bottlenecks that impact the efficiency of enclave execution.
-Based on these findings, the study presents configuration best practices and tuning recommendations tailored to various workload profiles, such as compute-bound, memory-intensive, and mixed applications.
-The insights gained from this evaluation contribute to optimizing the deployment of Keystone enclaves, making them more viable for real-world use cases that demand secure and efficient execution on RISC-V architectures.},
-Zusammenfassung = {Diese Arbeit bewertet die Leistung von Keystone-Enklaven, einer Open-Source Trusted Execution Environment (TEE) für RISC-V-Plattformen.
-Keystone ermöglicht sichere Berechnungen, indem sensible Workloads vom restlichen System isoliert werden.
-Zur Leistungsbewertung werden zentrale Systemparameter wie die Anzahl der CPU-Kerne, die Cache-Größe, die Speicherzuweisung und die Konfiguration der Enklave untersucht.
-Mithilfe standardisierter Benchmarking-Tools wie lmbench und CoreMark werden Ausführungszeit, Kontextwechsel-Overhead, Speicherlatenz und CPU-Auslastung unter verschiedenen Hardware- und Softwarekonfigurationen gemessen.
-Die experimentelle Analyse identifiziert wichtige Leistungsengpässe, die die Effizienz der Enklaven beeinträchtigen.
-Auf Basis dieser Erkenntnisse werden Konfigurationsrichtlinien und Optimierungsempfehlungen vorgestellt, die auf unterschiedliche Workload-Profile zugeschnitten sind, z. B.
-rechenintensive, speicherintensive und gemischte Anwendungen.
-Die gewonnenen Erkenntnisse tragen dazu bei, den Einsatz von Keystone-Enklaven zu optimieren und ihre praktische Anwendbarkeit in sicherheitskritischen Bereichen auf RISC-V-Architekturen zu verbessern.},
+Datum                 = {01. September 2025},
+Abstract = {
+  This thesis evaluates the performance of Keystone enclaves, an open-source Trusted Execution Environment (TEE) for RISC-V platforms.
+Keystone isolates sensitive workloads to enable secure computation, and its performance is assessed across CPU cores, cache sizes, memory allocations, and enclave configurations.
+Using Dhrystone, CoreMark, and cryptographic workloads such as Kyber—a post-quantum key encapsulation mechanism—this work measures execution time, context switch overhead, memory latency, and CPU utilization.
+The results reveal critical bottlenecks and provide best-practice configuration guidelines for compute-bound, memory-intensive, mixed, and post-quantum workloads, supporting efficient and secure deployment of Keystone enclaves on RISC-V architectures.
+},
+Zusammenfassung = {
+  Diese Arbeit bewertet die Leistung von Keystone-Enklaven, einer Open-Source Trusted Execution Environment (TEE) für RISC-V-Plattformen.
+Keystone isoliert sensible Workloads zur sicheren Berechnung, wobei Leistungseinflüsse durch CPU-Kerne, Cache-Größe, Speicherzuweisung und Enklavenkonfiguration untersucht werden.
+Mithilfe von Dhrystone, CoreMark und kryptografischen Workloads wie Kyber—einem postquantischen Schlüssel-Kapselungsverfahren—werden Ausführungszeit, Kontextwechsel-Overhead, Speicherlatenz und CPU-Auslastung gemessen.
+Die Ergebnisse zeigen zentrale Engpässe auf und liefern Konfigurationsrichtlinien für rechenintensive, speicherintensive, gemischte und postquantische Workloads, um Keystone-Enklaven effizient und sicher auf RISC-V-Architekturen einzusetzen.
+},
+Acknowledgements = {
+    I would like to sincerely thank Daniel K. for his insightful guidance and clarifications, which added immense value to this work.
+    I am also grateful to Tom K. for providing the initial log analysis script and for his valuable input on effective visualization approaches.
+    My heartfelt thanks go to my parents for their patience, encouragement, and the opportunities they have provided.  
+    I deeply appreciate my supervisors, Dr.-Ing. Saleh M. and Henrik S. (M.Sc), for their guidance, constructive feedback, and steadfast support throughout this project.  
+    Finally, I extend my gratitude to my fellow Lübeckers and all others who contributed in various ways; their support, though too numerous to mention individually, was invaluable.
+},
 Numerische Bibliographie

### 2025-08-31 draft_presentation.tex
added slides

\documentclass[aspectratio=169]{beamer}
\usetheme{metropolis}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{fontawesome5} % for icons
\usepackage{pifont}       % for checkmarks
\usepackage{xcolor}       % for color definitions
\usepackage{colortbl}     % for \rowcolor, \rowcolors, \cellcolor in tables
\usepackage{tikz}         % for drawing graphics
\usetikzlibrary{patterns,positioning} % enable patterns and positioning for tikz
\usepackage{pgfplots} 
%\pgfplotsset{compat=1.17} % set PGFPlots compatibility

% Colours
\definecolor{keystoneblue}{RGB}{0,77,128}
\definecolor{warmaccent}{RGB}{255,140,0}
\setbeamercolor{structure}{fg=keystoneblue}
\setbeamertemplate{navigation symbols}{}
\setbeamertemplate{footline}[frame number]

% Styling tweaks
\metroset{block=fill,progressbar=frametitle,numbering=fraction}
\renewcommand{\baselinestretch}{1.05}
\setbeamertemplate{frametitle}[default][center]

% Title and Author
\title[Evaluating RISC-V Enclaves]{\textbf{Evaluating RISC-V Enclaves:}\\ Security Without the Slowdown?}
\author{\textbf{Basil Ugbomoiko}}
\institute[UzL]{Universität zu Lübeck \\ Institut für Technische Informatik}
\date{Master’s Thesis Defense — \today}
% Add logo (adjust path as needed)
\titlegraphic{\hfill\includegraphics[height=1.5cm]{uzl-logo.pdf}}
\begin{document}

% Title Page with University Logo (if available)
\begin{frame}
  \titlepage
  \vfill
  \centering\small\textit{In the age of untrusted infrastructure, can security keep up with performance?}
\end{frame}

% Opening Question Frame (Increased vertical space)
\begin{frame}{Opening Question}
\centering
\Huge\bfseries
Can we \textcolor{keystoneblue}{trust} computation on machines we \textcolor{warmaccent}{don't control}? \\[1em]
\large In cloud and edge scenarios, this question is no longer optional — it's urgent.
\end{frame}

\begin{frame}{Problem Statement: Trends}
\begin{itemize}
    \item[\faCloud] Growing reliance on \textcolor{warmaccent}{cloud} and \textcolor{warmaccent}{edge} computing.
    \item[\faUsers] Shared, multi-tenant infrastructure is the norm.
\end{itemize}
\end{frame}

\begin{frame}{Problem Statement: Challenge}
\begin{itemize}
    \item[\faExclamationTriangle] Traditional security assumes a trustworthy OS — \alert{assumption broken}.
    \item[\faLock] Need isolation of critical code/data \textbf{even if the OS is compromised}.
\end{itemize}
\vfill
%\scriptsize\textit{This sets the stage for Trusted Execution Environments (TEEs).}
\end{frame}

% Background on TEEs (Added icons for clarity)
\begin{frame}{Background: TEEs in a Nutshell}
\begin{block}{Definition}
Hardware-based secure “zones” for executing sensitive code and storing sensitive data.
\end{block}
Examples include:
\begin{itemize}
    \item Intel SGX \faMicrochip\ \textit{(Intel's enclave solution for secure execution)} 
    \item ARM TrustZone \faLock\ \textit{(Security for mobile devices)}
    \item AMD SEV \faServer\ \textit{(Virtual machine encryption for server environments)}
    \item Keystone \faKey\ \textit{(Modular, open-source for RISC-V)}
\end{itemize}
\end{frame}

% Research Gap (Emphasized visual structure)
\begin{frame}{Research Gap}
\centering
\Large\bfseries
No comprehensive, benchmarked assessment of \\
\textbf{Keystone} performance overhead with \\
\textit{actionable tuning guidance.}
\end{frame}

% % Research Objectives (Visual appeal with bullet style)
% \begin{frame}{Research Objectives}
% \begin{enumerate}
%     \item Characterize Keystone’s \textbf{architecture} and \textbf{security model}.
%     \item Quantify performance across \textcolor{warmaccent}{varied workloads}.
%     \item Recommend \textbf{best practices} for optimal configuration.
% \end{enumerate}
% \end{frame}

% What is RISC-V? Slide
\begin{frame}{What is RISC-V?}
\begin{itemize}
    \item Open-source, modular processor architecture.
    \item Highly customizable for specialized hardware (ideal for enclaves).
    \item Scalable from embedded systems to high-performance computing.
    \item Growing adoption in academia and industry (e.g., SiFive, Western Digital).
\end{itemize}
\begin{block}{Key Concept}
\textit{RISC-V’s open-source nature enables flexible, secure processor designs for enclave-based systems.}
\end{block}
\end{frame}

% Keystone 
\begin{frame}{Why Keystone?}
\begin{itemize}
    \item Modular, open-source TEE for RISC-V.
    \item Uses Physical Memory Protection (PMP) and minimal Security Monitor for isolation.
    \item Low Trusted Computing Base (TCB) for security and flexibility.
    \item Optimized for secure computation in untrusted environments.
\end{itemize}
\begin{block}{Key Concept}
\textit{Low TCB reduces attack surfaces while enabling high flexibility and performance in enclave-based systems.}
\end{block}
\end{frame}

% Keystone Framework Architecture
\begin{frame}{Keystone Framework Architecture}
\begin{block}{Overview}
    Keystone is designed to be modular, with clear isolation between enclaves and the OS.
\end{block}

%\resizebox{\linewidth}{!}{%
\begin{figure}[htbp]
\centering
\includegraphics[width=\linewidth]{/home/individual-01/Documents/keystone-thesis/figures/keystone_overview.png}
%\caption{Keystone architecture showing components such as the Security Monitor, enclave runtime, and the privilege hierarchy.}
\label{fig:keystone_grouped}
\end{figure}
\end{frame}

% Keystone vs Other TEEs
\begin{frame}{Keystone vs. Other TEEs}
\begin{figure}[htbp]
\centering
\caption{Feature comparison of major TEEs.}
\label{fig:tee-features}
\renewcommand{\arraystretch}{1.3} % more row height
\setlength{\tabcolsep}{5pt}       % more column padding
\rowcolors{2}{gray!5}{white}      % alternating row colors

\begin{tabular}{@{}lcccc@{}}
\toprule
\rowcolor{gray!20}
\textbf{Feature} & \textbf{Keystone} & \textbf{SGX} & \textbf{TrustZone} & \textbf{SEV} \\
\midrule
Open Source            & \cellcolor{green!15}Yes & \cellcolor{red!15}No & \cellcolor{red!15}No & \cellcolor{red!15}No \\
Dynamic Enclaves       & \cellcolor{green!15}Yes & \cellcolor{red!15}No & \cellcolor{yellow!15}Limited & \cellcolor{red!15}No \\
Hardware Dependency    & \cellcolor{green!15}RISC-V & \cellcolor{red!15}Intel & \cellcolor{red!15}ARM & \cellcolor{red!15}AMD \\
TCB Size               & \cellcolor{green!15}Low & \cellcolor{yellow!15}Medium & \cellcolor{red!15}High & \cellcolor{yellow!15}Medium \\
Flexible Configurations& \cellcolor{green!15}High & \cellcolor{red!15}Low & \cellcolor{red!15}Low & \cellcolor{yellow!15}Medium \\
\bottomrule
\end{tabular}
\end{figure}

\end{frame}

\begin{frame}{System Design and Methodology}
    \small
    \begin{itemize}
        \item \textbf{Objective:} Evaluate the performance impact of Keystone's enclave isolation mechanisms on embedded/system-level workloads. \pause
        \item \textbf{Platform:} Emulated RISC-V architecture (RV64GC) using QEMU. \pause
        \item \textbf{Focus:} Quantify computational overhead when using Keystone enclaves vs. traditional non-isolated execution. \pause
    \end{itemize}
    \vspace{0.5em}
    \uncover<4->{\textbf{Platform Configuration: QEMU-based emulation ensures controlled environment for testing.}}
\end{frame}

\begin{frame}{QEMU-Based Virtualized Environment}
    \small
    \begin{itemize}
        \item QEMU emulates RISC-V architecture for repeatability and control. \pause
        \item Ensures consistent test conditions by eliminating hardware-induced variations (thermal throttling, cache behavior). \pause
        \item Configured for RV64GC architecture, as required by Keystone. \pause
    \end{itemize}
    \vspace{0.5em}
    \uncover<4->{\textbf{Why QEMU?} Provides a controlled, consistent environment for repeatability.}
\end{frame}

\begin{frame}{Benchmarking Workflow}
    \small
    \begin{itemize}
        \item \textbf{Define benchmark set} and configure system setup. \pause
        \item \textbf{Dual-mode execution:} Run benchmarks in native mode (without enclave) and enclave mode (with isolation). \pause
        \item \textbf{Metrics collection:} Gather data on execution time, throughput, and overhead. \pause
        \item \textbf{Analysis:} Compare performance across modes and assess overheads. \pause
    \end{itemize}
    \vspace{0.5em}
    %\uncover<5->{\textbf{Visual Workflow:} As shown in Figure \ref{fig:benchmarking-swimlane}, the process starts with benchmarking selection and proceeds through dual-mode execution.}
    \begin{figure}[H]
    \centering
\begin{figure}[htbp]
\centering
\resizebox{\linewidth}{!}{%
\begin{tikzpicture}[
    >=stealth,
    phase/.style={
        font=\bfseries\small,
        text=white,
        minimum height=0.8cm,
        rounded corners,
        inner xsep=6pt,
        text centered
    },
    stepnode/.style={
        draw=black!70,
        thick,
        rounded corners,
        align=center,
        font=\scriptsize,
        minimum height=1.4cm,
        text width=2.8cm,
        inner sep=6pt
    },
    arrow/.style={->, thick, shorten >=2pt, shorten <=2pt}
]

% Top row (Setup & Config) - blue shades
\node[stepnode, fill=blue!15] (s1) at (0,3) {\faTasks\\Define benchmarks};
\node[stepnode, fill=blue!20, right=1.8cm of s1] (s2) {\faMicrochip\\Select hardware};

% Bottom row (Execution & Analysis) - green shades
\node[stepnode, fill=green!15, right=1.8cm of s2] (s3) {\faPlay\\Run native mode};
\node[stepnode, fill=green!20, right=1.8cm of s3] (s4) {\faLock\\Run enclave mode};
\node[stepnode, fill=green!25] (s5) at (0,1) {\faChartBar\\Collect metrics};
\node[stepnode, fill=green!30, right=1.8cm of s5] (s6) {\faSearch\\Analyze overheads};
\node[stepnode, fill=green!40, right=1.8cm of s6] (s7) {\faArchive\\Archive results};

% Arrows top row
\draw[arrow] (s1) -- (s2);
\draw[arrow] (s2) -- (s3);
\draw[arrow] (s3) -- (s4);

% Arrow from top to bottom
\draw[arrow] (s4.south) .. controls +(0,-0.8) and +(0,0.8) .. (s5.north);

% Arrows bottom row
\draw[arrow] (s5) -- (s6);
\draw[arrow] (s6) -- (s7);

% Color legend
\node[draw=black!50, fill=white, rounded corners, align=left, font=\scriptsize, anchor=north west] 
    at ([xshift=1cm,yshift=-0.5cm]s7.south east) (legend) {
    \begin{tikzpicture}[baseline=(current bounding box.center)]
        \node[fill=blue!20, minimum width=0.5cm, minimum height=0.5cm] (b) {};
        \node[right=0.3cm of b, anchor=west] {Setup \& Config};
        \node[fill=green!30, minimum width=0.5cm, minimum height=0.5cm, below=0.2cm of b] (g) {};
        \node[right=0.3cm of g, anchor=west] {Execution \& Analysis};
    \end{tikzpicture}
};

\end{tikzpicture}
}
\caption{Benchmarking and data collection workflow}
\end{figure}
    \caption{Benchmarking and data collection workflow}
    \label{fig:benchmarking-swimlane}
    \end{figure}
\end{frame}

\begin{frame}{Benchmarking Tools}
    \small
    \begin{itemize}
        \item \textbf{Dhrystone:} Measures CPU performance (DMIPS). \pause
        \item \textbf{CoreMark:} Focuses on integer performance, used in embedded systems benchmarking. \pause
        \item \textbf{Two configurations:}
        \begin{itemize}
            \item \textbf{Native Execution:} Benchmark runs as a normal process. \pause
            \item \textbf{Enclave Execution:} Benchmark runs within a Keystone enclave with memory isolation. \pause
        \end{itemize}
    \end{itemize}
    \vspace{0.5em}
    \uncover<5->{\textbf{Key Focus:} These tools provide insights into both CPU-bound and system-level performance.}
\end{frame}

%Results
\section{Synthesis of Key Performance Trends and Hardware Bottlenecks}
\label{sec:synthesis}

\begin{frame}{Enclave Overhead and System Behavior}
\small
\begin{block}{Performance Overhead}
\begin{itemize}
    \item \textbf{Enclave execution} introduces overhead: 10\% to 15\%.
    \item \textbf{Compute-bound workloads} (CoreMark, Dhrystone) see minimal performance loss.
    \item \textbf{Memory-bound workloads} (Kyber) experience severe slowdowns due to added security checks and memory isolation.
\end{itemize}
\end{block}

\begin{block}{Overhead Breakdown}
\begin{itemize}
    \item Memory isolation via PMP adds fixed latency per memory access.
    \item Context switches between host and enclave domains are costly.
\end{itemize}
\end{block}

\centering
\textbf{Key Insight:} Overhead varies based on workload characteristics and memory behavior.
\end{frame}

\begin{frame}{Impact of Scalability Limitations}
\small
\begin{block}{CPU Scaling Limits}
\begin{itemize}
    \item Increasing CPU cores \textbf{benefits multi-threaded tasks}, but gains plateau due to context-switch overhead and enclave thread management.
    \item For example, \texttt{CoreMark} shows marginal improvements beyond 4 cores.
\end{itemize}
\end{block}

\begin{block}{Memory Scaling Constraints}
\begin{itemize}
    \item Increased memory capacity \textbf{improves performance} by reducing page faults.
    \item \textbf{Fragmentation} and \textbf{PMP entry limits} cap the performance improvement.
\end{itemize}
\end{block}

\centering
\textbf{Key Insight:} Hardware resources alone cannot fully predict enclave performance due to system resource allocation limits.
\end{frame}

\begin{frame}{PMP Entry Exhaustion and Multi-Enclave Scalability}
\small
\begin{block}{PMP Resource Limitations}
\begin{itemize}
    \item \textbf{Maximum PMP entries}: 16.
    \item Exhausted PMP entries \textbf{prevent new enclave creation}, causing errors even with abundant system resources.
    \item Failure due to PMP exhaustion: \texttt{SBI error codes}.
\end{itemize}
\end{block}

\begin{block}{Scalability Bottleneck}
\begin{itemize}
    \item In \textbf{multi-tenant environments}, PMP exhaustion limits the number of concurrent enclaves.
    \item This creates scalability issues, especially for high-concurrency workloads.
\end{itemize}
\end{block}

\centering
\textbf{Key Insight:} PMP limitations represent a significant scalability bottleneck for RISC-V enclave-based systems.
\end{frame}

\begin{frame}{Performance Trends and Hardware Bottlenecks}
\centering
\begin{figure}[htbp]
\centering
\begin{tikzpicture}
\begin{axis}[
    ybar,
    bar width=14pt,
    width=10cm,
    height=6cm,
    ymin=0, ymax=120,
    ylabel={Relative performance (\%)},
    symbolic x coords={CoreMark,Dhrystone,Kyber},
    xtick=data,
    ymajorgrids,
    grid style=dashed,
    nodes near coords,
    every node near coord/.append style={font=\footnotesize, /pgf/number format/fixed},
    legend style={at={(0.5,-0.15)}, anchor=north, legend columns=-1}
]

% Enclave: blue, rounded top, transparency
\addplot[
    fill={rgb,255:red,0; green,114; blue,178},
    fill opacity=0.8,
    draw=black!60,
    rounded corners=2pt
] coordinates {(CoreMark,95) (Dhrystone,62) (Kyber,37)};

% Native: orange, rounded top, transparency
\addplot[
    fill={rgb,255:red,230; green,159; blue,0},
    fill opacity=0.8,
    draw=black!60,
    rounded corners=2pt
] coordinates {(CoreMark,100) (Dhrystone,100) (Kyber,100)};

\legend{Enclave,Native}
\end{axis}
\end{tikzpicture}
\caption{Relative performance of enclave vs native execution.}
\end{figure}
\end{frame}

\begin{frame}{Key Takeaways from Performance Trends}
\small
\begin{block}{Main Findings}
\begin{itemize}
    \item Enclave performance suffers primarily in \textbf{memory-bound workloads} due to security checks and context switching.
    \item \textbf{PMP entry limitations} restrict scalability in multi-enclave setups.
    \item Increasing hardware resources does not linearly translate to performance gains due to system-level overheads.
\end{itemize}
\end{block}

\centering
\textbf{Key Insight:} Optimizing both hardware resources and system-level configurations is essential for scalable and efficient enclave execution.
\end{frame}

\begin{frame}{Why Misalignment Bloats PMP Usage}
\small
\vfill

% Problem
\only<1>{
\begin{block}{Problem}
Physical Memory Protection (PMP) entries must cover power‑of‑two sized, naturally aligned regions (NAPOT).
\end{block}
}

% Cause
\only<2>{
\begin{block}{Problem}
Physical Memory Protection (PMP) entries must cover power‑of‑two sized, naturally aligned regions (NAPOT).
\end{block}
\vspace{0.5em}
\begin{block}{Cause}
When a memory region is \textbf{misaligned}:
\begin{itemize}
  \item NAPOT encoding forces padding to the next power‑of‑two boundary.
  \item Padding areas consume PMP coverage even if they are not actual data.
\end{itemize}
\end{block}
% \draw[->, thick] (0,-0.2) -- (0,-0.8); % optional arrow from Problem → Cause
}

% Effect
\only<3>{
\begin{block}{Cause}
When a memory region is \textbf{misaligned}:
\begin{itemize}
  \item NAPOT encoding forces padding to the next power‑of‑two boundary.
  \item Padding areas consume PMP coverage even if they are not actual data.
\end{itemize}
\end{block}
\vspace{0.5em}
\begin{block}{Effect}
\begin{itemize}
  \item A single logical region may require \textbf{multiple PMP entries}.
  \item Fragmentation increases as more regions are misaligned.
\end{itemize}
\end{block}
% \draw[->, thick] (0,-0.2) -- (0,-0.8); % optional arrow from Cause → Effect
}

% Impact
\only<4>{
\begin{block}{Effect}
\begin{itemize}
  \item A single logical region may require \textbf{multiple PMP entries}.
  \item Fragmentation increases as more regions are misaligned.
\end{itemize}
\end{block}
\vspace{0.5em}
\begin{block}{Impact}
\begin{itemize}
  \item Higher PMP entry count $\Rightarrow$ fewer entries left for other regions.
  \item Limits scalability and flexibility of memory protection.
  \item Wastes address space and complicates security policy.
\end{itemize}
\end{block}
}

% Key takeaway
\only<5>{
\centering
\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=1, transform shape,font=\small]

% Memory axis
\draw[thick,->] (0,0) -- (10,0) node[anchor=west]{Memory address};
\foreach \x/\label in {0/0\,KB,1.25/128\,KB,4.25/192\,KB,8/300\,KB}
    \draw (\x,0.1) -- (\x,-0.1) node[anchor=north]{\label};

% === Region 1: contiguous, 1 PMP entry ===
\node[anchor=south] at (1.375,2.1) {\scriptsize 128\,KB aligned};
%\draw[pattern=north east lines,pattern color=black] (0.25,0) rectangle (0.5,1);
\draw[fill=keystoneblue!60] (0.5,0) rectangle (2.2,1);
%\draw[pattern=north east lines,pattern color=black] (2.2,0) rectangle (2.5,1);
\draw[fill=keystoneblue!60,opacity=0.4] (0.5,1) rectangle (2.2,2) node[pos=.5] {\scriptsize 1 entry};

% === Region 2: fragmented, 3 PMP entries ===
\node[anchor=south] at (4.3,4.1) {\scriptsize 192\,KB misaligned};
\draw[pattern=north east lines,pattern color=black] (3.5,0) rectangle (3.8,1);
\draw[fill=warmaccent!60] (3.8,0) rectangle (4.6,1);
\draw[pattern=north east lines,pattern color=black] (4.6,0) rectangle (5.1,1);
\draw[fill=warmaccent!60,opacity=0.4] (3.5,1) rectangle (5.1,4) node[pos=.5] {\scriptsize 3 entries};

% === Region 3: fragmented, 3 PMP entries ===
\node[anchor=south] at (8.0,4.1) {\scriptsize 300\,KB misaligned};
\draw[pattern=north east lines,pattern color=black] (6.7,0) rectangle (7.0,1);
\draw[fill=warmaccent!60] (7.0,0) rectangle (9.1,1);
\draw[pattern=north east lines,pattern color=black] (9.1,0) rectangle (9.3,1);
\draw[fill=warmaccent!60,opacity=0.4] (6.7,1) rectangle (9.3,4) node[pos=.5] {\scriptsize 3 entries};

% Arrows showing cause-effect
%\draw[->,thick] (2.2,0.5) .. controls (2.8,1.5) .. (3.4,2) node[midway,above,sloped]{\scriptsize Padding $\Rightarrow$ More entries};
\draw[->,thick] (5.1,0.5) .. controls (5.4,2) .. (6.6,2.5) node[midway,above,sloped]{\scriptsize Padding $\Rightarrow$ More entries};

% Legend
\begin{scope}[xshift=0cm,yshift=-1.5cm]
\draw[fill=keystoneblue!60] (0,0) rectangle (0.4,0.4) node[right=0.5cm] {Actual region (contiguous)};
\draw[fill=warmaccent!60] (5,0) rectangle (5.4,0.4) node[right=0.5cm] {Actual region (fragmented)};
\draw[pattern=north east lines,pattern color=black] (0,-0.6) rectangle (0.4,-0.2) node[right=0.5cm] {Wasted padding};
\draw[fill=gray!50,opacity=0.4] (5,-0.6) rectangle (5.4,-0.2) node[right=0.5cm] {PMP entry count bar};
\end{scope}

\end{tikzpicture}

\vspace{0.4em}
\scriptsize
\textbf{Takeaway:} Align regions to NAPOT boundaries to minimize padding and preserve PMP entries.
\end{figure}
}

\vfill
\end{frame}

% Optimizations for Keystone Enclaves
\begin{frame}{Optimizations for Keystone Enclaves}
\begin{itemize}
    \item Use contiguous, NAPOT-aligned memory to minimize PMP usage.
    \item Consolidate workloads and limit active enclaves to reduce contention.
    \item Offload non-sensitive tasks to the untrusted domain.
\end{itemize}
\end{frame}

% Real-World Applications of Keystone Enclaves
\begin{frame}{Real-World Applications of Keystone Enclaves}
\begin{itemize}
    \item \textbf{Secure Cloud Computing:} Protects sensitive data in untrusted environments, ensuring confidentiality in multi-tenant clouds.
    \item \textbf{Confidential Data Processing:} Enables secure computations on private data, ideal for financial and healthcare industries.
\end{itemize}
\vspace{2em}
\resizebox{\textwidth}{!}{ % Scale the entire diagram
\begin{tikzpicture}[
    node distance=1.5cm, % Reduce space between nodes
    every node/.style={font=\small, align=center},
    box/.style={draw, rounded corners, thick, fill=blue!15, minimum width=2.8cm, minimum height=1.2cm},
    data/.style={draw, thick, fill=orange!20, rounded corners, minimum width=2.5cm, minimum height=1cm}
]

% Cloud
\node[box, fill=blue!20] (cloud) {\includegraphics[scale=0.05]{example-image} \\ \textbf{Cloud Server} \\ (Untrusted Environment)};

% Lock / Enclave
\node[box, fill=green!20, right=3cm of cloud] (enclave) {\faLock \\ \textbf{Keystone Enclave} \\ Secure Execution};

% Data source
\node[data, left=3cm of cloud] (client) {\faUser \\ \textbf{Client Data}};

% Arrows
\draw[->, thick] (client) -- node[above]{Encrypted Data} (cloud);
\draw[->, thick] (cloud) -- node[above]{Secure Transfer} (enclave);
\draw[->, thick] (enclave) -- ++(0,-2) node[midway,right]{Results (Encrypted)} -| (client.south);

% Optional note
\node[below=1.2cm of cloud, text width=8cm, align=center] 
    {\textit{Sensitive data is encrypted before leaving the client, processed inside the enclave, and never exposed in plaintext to the cloud provider.}};

\end{tikzpicture}
}
\end{frame}

% Future Work and Improvements
\begin{frame}{Future Work and Improvements}
\begin{itemize}
    \item Performance optimizations with hardware extensions.
    \item Multi-enclave systems for more complex use cases.
    %\item Research on side-channel attacks and improved security measures.
    \item Development of better developer tools for enclave management.
\end{itemize}
\end{frame}

% Summary of Findings
\begin{frame}{Key Takeaways}
\begin{itemize}
    \item Keystone offers strong security with minimal overhead for CPU-bound workloads.
    \item Significant overhead for memory bound tasks can be reduced with optimizations.
    \item Customizable settings allow for optimization based on workload type.
\end{itemize}
\end{frame}

% Conclusion
\begin{frame}{Conclusion}
\centering
\Large Keystone can deliver strong security with minimal performance cost — \\
\textit{if workloads are chosen and designed wisely.}
\end{frame}

% Thank You Slide with visual focus
\begin{frame}
\centering
\Huge\textbf{Thank You}\\
\large Questions? \faQuestionCircle
\vfill
\small Universität zu Lübeck — Master’s Thesis Defense
\end{frame}
\end{document}

### 2025-09-01 chapters/discussion.tex
last minute changes

@@
 
-To address overhead and scalability challenges, \emph{enclave multiplexing} offers a promising software-level strategy by consolidating multiple lightweight workloads into a single enclave.
+To address overhead and scalability challenges, \emph{enclave multiplexing} \cite{Occlum2020} offers a promising software-level strategy by consolidating multiple lightweight workloads into a single enclave.
 This reduces the number of PMP entries required by reusing them across tasks, optimizing resource usage and improving scalability.
@@
 
-In terms of software strategies, significant improvements can be made by enhancing the runtime management of enclaves, as outlined in Section~\ref{sec:synthesis} (Chapter~\ref{chap:discussion}).
+In terms of software strategies, significant improvements can be made by enhancing the runtime management of enclaves, as outlined in Section~\ref{sec:synthesis}.
 For example, reducing the frequency of context switches between the enclave and host operating system, and batching enclave invocations, could help mitigate the performance degradation caused by these transitions.
@@
 
-Moreover, careful attention should be given to the design of memory allocation and management strategies in multi-tenant environments, where multiple enclaves may need to coexist on the same system (Section~\ref{sec:implications}, Chapter~\ref{chap:benchmarking}).
+Moreover, careful attention should be given to the design of memory allocation and management strategies in multi-tenant environments, where multiple enclaves may need to coexist on the same system (Section~\ref{sec:implications}).
 In such scenarios, strategies such as enclave multiplexing—where multiple tasks or applications are combined within a single enclave—could significantly reduce the overall resource requirements.

### 2025-09-01 preamble.tex
last minute changes

@@
 Betreuerin            = {Dr.-Ing. Saleh Mulhem},
-Mit Unterstützung von = {Henrik Strunck, M.Sc},
+Mit Unterstützung von = {Henrik Strunck, M.Sc.},
 Studiengang           = {IT Security},
-Datum                 = {01. September 2025},
+Datum                 = {31. August 2025},
 Abstract = {
@@
     My heartfelt thanks go to my parents for their patience, encouragement, and the opportunities they have provided.  
-    I deeply appreciate my supervisors, Dr.-Ing. Saleh M. and Henrik S. (M.Sc), for their guidance, constructive feedback, and steadfast support throughout this project.  
+    I deeply appreciate my supervisors, Dr.-Ing. Saleh M. and Henrik S. (M.Sc.), for their guidance, constructive feedback, and steadfast support throughout this project.  
     Finally, I extend my gratitude to my fellow Lübeckers and all others who contributed in various ways; their support, though too numerous to mention individually, was invaluable.

### 2025-09-02 draft_presentation.tex
new diagram

@@
 \usepackage{pgfplots}
+\usepackage[transitions={dissolve}]{pdfpc}
 %\pgfplotsset{compat=1.17} % set PGFPlots compatibility
@@
 \title[Evaluating RISC-V Enclaves]{\textbf{Evaluating RISC-V Enclaves:}\\ Security Without the Slowdown?}
+\subtitle{Performance Evaluation and Optimization of Keystone Enclaves on RISC-V}
 \author{\textbf{Basil Ugbomoiko}}
@@
 %\scriptsize\textit{This sets the stage for Trusted Execution Environments (TEEs).}
+\begin{figure}[h]
+\centering
+\begin{tikzpicture}[node distance=1.2cm and 1cm, every node/.style={font=\small}]
+    \node[draw, fill=orange!30, minimum width=2.5cm, minimum height=1cm, rounded corners] (app) {\textbf{App}};
+    \node[draw, fill=gray!20, below=of app, minimum width=3cm, minimum height=1cm, rounded corners] (os) {\textbf{Untrusted OS}};
+    \node[draw, fill=blue!15, right=3cm of app, minimum width=2.5cm, minimum height=1cm, rounded corners] (enclave) {\faLock\ \textbf{Enclave}};
+
+    \draw[->, thick, red] (os) -- (app) node[midway, left, align=left, font=\footnotesize] {Possible interference\\ or data snooping};
+    \draw[->, thick, keystoneblue] (app) -- (enclave) node[midway, above, font=\footnotesize] {Trusted Execution};
+    \draw[->, thick, dashed, red] (os) -- (enclave) node[midway, below, font=\footnotesize] {Attempted attack blocked};
+\end{tikzpicture}
+\caption{Enclaves isolate computation from a compromised OS}
+\end{figure}
 \end{frame}
@@
     \begin{itemize}
-        \item \textbf{Dhrystone:} Measures CPU performance (DMIPS). \pause
-        \item \textbf{CoreMark:} Focuses on integer performance, used in embedded systems benchmarking. \pause
-        \item \textbf{Two configurations:}
+        \item \textbf{Dhrystone:} Measures CPU performance (DMIPS).
+        \item \textbf{CoreMark:} Focuses on integer performance, used in embedded systems benchmarking.
+        \begin{block}{Execution Modes}
         \begin{itemize}
-            \item \textbf{Native Execution:} Benchmark runs as a normal process. \pause
-            \item \textbf{Enclave Execution:} Benchmark runs within a Keystone enclave with memory isolation. \pause
+            \item \textbf{Native Execution:} Benchmark runs as a normal process.
+            \item \textbf{Enclave Execution:} Benchmark runs within a Keystone enclave.
         \end{itemize}
+        \end{block}
     \end{itemize}
-    \vspace{0.5em}
-    \uncover<5->{\textbf{Key Focus:} These tools provide insights into both CPU-bound and system-level performance.}
+    %\vspace{0.5em}
+    %\uncover<1->{\textbf{Key Focus:} These tools provide insights into both CPU-bound and system-level performance.}
 \end{frame}
@@
 \begin{scope}[xshift=0cm,yshift=-1.5cm]
-\draw[fill=keystoneblue!60] (0,0) rectangle (0.4,0.4) node[right=0.5cm] {Actual region (contiguous)};
-\draw[fill=warmaccent!60] (5,0) rectangle (5.4,0.4) node[right=0.5cm] {Actual region (fragmented)};
+\draw[fill=keystoneblue!60] (0,0) rectangle (0.4,0.4) node[right=0.5cm] {Contiguous region};
+\draw[fill=warmaccent!60] (5,0) rectangle (5.4,0.4) node[right=0.5cm] {Fragmented region};
 \draw[pattern=north east lines,pattern color=black] (0,-0.6) rectangle (0.4,-0.2) node[right=0.5cm] {Wasted padding};
@@
 \vfill
-\small Universität zu Lübeck — Master’s Thesis Defense
+%\small Universität zu Lübeck — Master’s Thesis Defense
+\includegraphics[height=1cm]{uzl-logo.pdf}
 \end{frame}

### 2025-09-02 draft_presentation.tex
improved slides

@@
 
+% Research Gap (Emphasized visual structure)
+\begin{frame}{Research Gap}
+\centering
+\Large\bfseries
+No comprehensive, benchmarked assessment of \\
+\textbf{Keystone} performance overhead with \\
+\textit{actionable tuning guidance.}
+\end{frame}
+
+\begin{frame}{Contributions}
+\begin{itemize}
+    \item \textbf{Systematic performance evaluation} of Keystone across varied workloads.
+    \item \textbf{Identification of hardware bottlenecks} like PMP entry exhaustion and misaligned memory.
+    \item \textbf{Actionable optimization guidelines} for developers using Keystone TEEs.
+    %\item \textbf{Visual benchmarking workflow} and comparison to alternative TEEs.
+\end{itemize}
+\end{frame}
+
 % Background on TEEs (Added icons for clarity)
@@
 \end{itemize}
-\end{frame}
-
-% Research Gap (Emphasized visual structure)
-\begin{frame}{Research Gap}
-\centering
-\Large\bfseries
-No comprehensive, benchmarked assessment of \\
-\textbf{Keystone} performance overhead with \\
-\textit{actionable tuning guidance.}
 \end{frame}
@@
 % Keystone 
-\begin{frame}{Why Keystone?}
+\begin{frame}{From Problem to Solution: Enter Keystone}
 \begin{itemize}
@@
 \centering
-\caption{Feature comparison of major TEEs.}
-\label{fig:tee-features}
 \renewcommand{\arraystretch}{1.3} % more row height
@@
 \end{tabular}
+\caption{Feature comparison of major TEEs.}
+\label{fig:tee-features}
 \end{figure}
@@
 
-\begin{frame}{Benchmarking Tools}
-    \small
-    \begin{itemize}
-        \item \textbf{Dhrystone:} Measures CPU performance (DMIPS).
-        \item \textbf{CoreMark:} Focuses on integer performance, used in embedded systems benchmarking.
-        \begin{block}{Execution Modes}
-        \begin{itemize}
-            \item \textbf{Native Execution:} Benchmark runs as a normal process.
-            \item \textbf{Enclave Execution:} Benchmark runs within a Keystone enclave.
-        \end{itemize}
-        \end{block}
-    \end{itemize}
-    %\vspace{0.5em}
-    %\uncover<1->{\textbf{Key Focus:} These tools provide insights into both CPU-bound and system-level performance.}
-\end{frame}
+% \begin{frame}{Benchmarking Tools}
+%     \small
+%     \begin{itemize}
+%         \item \textbf{Dhrystone:} Measures CPU performance (DMIPS).
+%         \item \textbf{CoreMark:} Focuses on integer performance, used in embedded systems benchmarking.
+%         \begin{block}{Execution Modes}
+%         \begin{itemize}
+%             \item \textbf{Native Execution:} Benchmark runs as a normal process.
+%             \item \textbf{Enclave Execution:} Benchmark runs within a Keystone enclave.
+%         \end{itemize}
+%         \end{block}
+%     \end{itemize}
+%     %\vspace{0.5em}
+%     %\uncover<1->{\textbf{Key Focus:} These tools provide insights into both CPU-bound and system-level performance.}
+% \end{frame}
 
 %Results
-\section{Synthesis of Key Performance Trends and Hardware Bottlenecks}
+\section{Key Performance Trends and Hardware Bottlenecks}
 \label{sec:synthesis}
@@
 \textbf{Key Insight:} Overhead varies based on workload characteristics and memory behavior.
-\end{frame}
-
-\begin{frame}{Impact of Scalability Limitations}
-\small
-\begin{block}{CPU Scaling Limits}
-\begin{itemize}
-    \item Increasing CPU cores \textbf{benefits multi-threaded tasks}, but gains plateau due to context-switch overhead and enclave thread management.
-    \item For example, \texttt{CoreMark} shows marginal improvements beyond 4 cores.
-\end{itemize}
-\end{block}
-
-\begin{block}{Memory Scaling Constraints}
-\begin{itemize}
-    \item Increased memory capacity \textbf{improves performance} by reducing page faults.
-    \item \textbf{Fragmentation} and \textbf{PMP entry limits} cap the performance improvement.
-\end{itemize}
-\end{block}
-
-\centering
-\textbf{Key Insight:} Hardware resources alone cannot fully predict enclave performance due to system resource allocation limits.
-\end{frame}
-
-\begin{frame}{PMP Entry Exhaustion and Multi-Enclave Scalability}
-\small
-\begin{block}{PMP Resource Limitations}
-\begin{itemize}
-    \item \textbf{Maximum PMP entries}: 16.
-    \item Exhausted PMP entries \textbf{prevent new enclave creation}, causing errors even with abundant system resources.
-    \item Failure due to PMP exhaustion: \texttt{SBI error codes}.
-\end{itemize}
-\end{block}
-
-\begin{block}{Scalability Bottleneck}
-\begin{itemize}
-    \item In \textbf{multi-tenant environments}, PMP exhaustion limits the number of concurrent enclaves.
-    \item This creates scalability issues, especially for high-concurrency workloads.
-\end{itemize}
-\end{block}
-
-\centering
-\textbf{Key Insight:} PMP limitations represent a significant scalability bottleneck for RISC-V enclave-based systems.
 \end{frame}
@@
 
-\begin{frame}{Key Takeaways from Performance Trends}
+\begin{frame}{Impact of Scalability Limitations}
 \small
-\begin{block}{Main Findings}
+\begin{block}{CPU Scaling Limits}
 \begin{itemize}
-    \item Enclave performance suffers primarily in \textbf{memory-bound workloads} due to security checks and context switching.
-    \item \textbf{PMP entry limitations} restrict scalability in multi-enclave setups.
-    \item Increasing hardware resources does not linearly translate to performance gains due to system-level overheads.
+    \item Increasing CPU cores \textbf{benefits multi-threaded tasks}, but gains plateau due to context-switch overhead and enclave thread management.
+    \item For example, \texttt{CoreMark} shows marginal improvements beyond 4 cores.
+\end{itemize}
+\end{block}
+
+\begin{block}{Memory Scaling Constraints}
+\begin{itemize}
+    \item Increased memory capacity \textbf{improves performance} by reducing page faults.
+    \item \textbf{Fragmentation} and \textbf{PMP entry limits} cap the performance improvement.
 \end{itemize}
@@
 \centering
-\textbf{Key Insight:} Optimizing both hardware resources and system-level configurations is essential for scalable and efficient enclave execution.
+\textbf{Key Insight:} Hardware resources alone cannot fully predict enclave performance due to system resource allocation limits.
 \end{frame}
 
-\begin{frame}{Why Misalignment Bloats PMP Usage}
+\begin{frame}{PMP Entry Exhaustion and Multi-Enclave Scalability}
 \small
-\vfill
-
-% Problem
-\only<1>{
-\begin{block}{Problem}
-Physical Memory Protection (PMP) entries must cover power‑of‑two sized, naturally aligned regions (NAPOT).
-\end{block}
-}
-
-% Cause
-\only<2>{
-\begin{block}{Problem}
-Physical Memory Protection (PMP) entries must cover power‑of‑two sized, naturally aligned regions (NAPOT).
-\end{block}
-\vspace{0.5em}
-\begin{block}{Cause}
-When a memory region is \textbf{misaligned}:
+\begin{block}{PMP Resource Limitations}
 \begin{itemize}
-  \item NAPOT encoding forces padding to the next power‑of‑two boundary.
-  \item Padding areas consume PMP coverage even if they are not actual data.
+    \item \textbf{Maximum PMP entries}: 16.
+    \item Exhausted PMP entries \textbf{prevent new enclave creation}, causing errors even with abundant system resources.
+    \item Failure due to PMP exhaustion: \texttt{SBI error codes}.
 \end{itemize}
 \end{block}
-% \draw[->, thick] (0,-0.2) -- (0,-0.8); % optional arrow from Problem → Cause
-}
 
-% Effect
-\only<3>{
-\begin{block}{Cause}
-When a memory region is \textbf{misaligned}:
+\begin{block}{Scalability Bottleneck}
 \begin{itemize}
-  \item NAPOT encoding forces padding to the next power‑of‑two boundary.
-  \item Padding areas consume PMP coverage even if they are not actual data.
+    \item In \textbf{multi-tenant environments}, PMP exhaustion limits the number of concurrent enclaves.
+    \item This creates scalability issues, especially for high-concurrency workloads.
 \end{itemize}
 \end{block}
-\vspace{0.5em}
-\begin{block}{Effect}
+
+\centering
+\textbf{Key Insight:} PMP limitations represent a significant scalability bottleneck for RISC-V enclave-based systems.
+\end{frame}
+
+% \begin{frame}{Key Takeaways from Performance Trends}
+% \small
+% \begin{block}{Main Findings}
+% \begin{itemize}
+%     \item Enclave performance suffers primarily in \textbf{memory-bound workloads} due to security checks and context switching.
+%     \item \textbf{PMP entry limitations} restrict scalability in multi-enclave setups.
+%     \item Increasing hardware resources does not linearly translate to performance gains due to system-level overheads.
+% \end{itemize}
+% \end{block}
+
+% \centering
+% \textbf{Key Insight:} Optimizing both hardware resources and system-level configurations is essential for scalable and efficient enclave execution.
+% \end{frame}
+
+% \begin{frame}{Why Misalignment Bloats PMP Usage}
+% \small
+% \vfill
+
+% % Problem
+% \only<1>{
+% \begin{block}{Problem}
+% Physical Memory Protection (PMP) entries must cover power‑of‑two sized, naturally aligned regions (NAPOT).
+% \end{block}
+% }
+
+% % Cause
+% \only<2>{
+% \begin{block}{Problem}
+% Physical Memory Protection (PMP) entries must cover power‑of‑two sized, naturally aligned regions (NAPOT).
+% \end{block}
+% \vspace{0.5em}
+% \begin{block}{Cause}
+% When a memory region is \textbf{misaligned}:
+% \begin{itemize}
+%   \item NAPOT encoding forces padding to the next power‑of‑two boundary.
+%   \item Padding areas consume PMP coverage even if they are not actual data.
+% \end{itemize}
+% \end{block}
+% % \draw[->, thick] (0,-0.2) -- (0,-0.8); % optional arrow from Problem → Cause
+% }
+
+% % Effect
+% \only<3>{
+% \begin{block}{Cause}
+% When a memory region is \textbf{misaligned}:
+% \begin{itemize}
+%   \item NAPOT encoding forces padding to the next power‑of‑two boundary.
+%   \item Padding areas consume PMP coverage even if they are not actual data.
+% \end{itemize}
+% \end{block}
+% \vspace{0.5em}
+% \begin{block}{Effect}
+% \begin{itemize}
+%   \item A single logical region may require \textbf{multiple PMP entries}.
+%   \item Fragmentation increases as more regions are misaligned.
+% \end{itemize}
+% \end{block}
+% % \draw[->, thick] (0,-0.2) -- (0,-0.8); % optional arrow from Cause → Effect
+% }
+
+% % Impact
+% \only<4>{
+% \begin{block}{Effect}
+% \begin{itemize}
+%   \item A single logical region may require \textbf{multiple PMP entries}.
+%   \item Fragmentation increases as more regions are misaligned.
+% \end{itemize}
+% \end{block}
+% \vspace{0.5em}
+% \begin{block}{Impact}
+% \begin{itemize}
+%   \item Higher PMP entry count $\Rightarrow$ fewer entries left for other regions.
+%   \item Limits scalability and flexibility of memory protection.
+%   \item Wastes address space and complicates security policy.
+% \end{itemize}
+% \end{block}
+% }
+
+% % Key takeaway
+% \only<5>{
+% \centering
+% \begin{figure}[htbp]
+% \centering
+% \begin{tikzpicture}[scale=1, transform shape,font=\small]
+
+% % Memory axis
+% \draw[thick,->] (0,0) -- (10,0) node[anchor=west]{Memory address};
+% \foreach \x/\label in {0/0\,KB,1.25/128\,KB,4.25/192\,KB,8/300\,KB}
+%     \draw (\x,0.1) -- (\x,-0.1) node[anchor=north]{\label};
+
+% % === Region 1: contiguous, 1 PMP entry ===
+% \node[anchor=south] at (1.375,2.1) {\scriptsize 128\,KB aligned};
+% %\draw[pattern=north east lines,pattern color=black] (0.25,0) rectangle (0.5,1);
+% \draw[fill=keystoneblue!60] (0.5,0) rectangle (2.2,1);
+% %\draw[pattern=north east lines,pattern color=black] (2.2,0) rectangle (2.5,1);
+% \draw[fill=keystoneblue!60,opacity=0.4] (0.5,1) rectangle (2.2,2) node[pos=.5] {\scriptsize 1 entry};
+
+% % === Region 2: fragmented, 3 PMP entries ===
+% \node[anchor=south] at (4.3,4.1) {\scriptsize 192\,KB misaligned};
+% \draw[pattern=north east lines,pattern color=black] (3.5,0) rectangle (3.8,1);
+% \draw[fill=warmaccent!60] (3.8,0) rectangle (4.6,1);
+% \draw[pattern=north east lines,pattern color=black] (4.6,0) rectangle (5.1,1);
+% \draw[fill=warmaccent!60,opacity=0.4] (3.5,1) rectangle (5.1,4) node[pos=.5] {\scriptsize 3 entries};
+
+% % === Region 3: fragmented, 3 PMP entries ===
+% \node[anchor=south] at (8.0,4.1) {\scriptsize 300\,KB misaligned};
+% \draw[pattern=north east lines,pattern color=black] (6.7,0) rectangle (7.0,1);
+% \draw[fill=warmaccent!60] (7.0,0) rectangle (9.1,1);
+% \draw[pattern=north east lines,pattern color=black] (9.1,0) rectangle (9.3,1);
+% \draw[fill=warmaccent!60,opacity=0.4] (6.7,1) rectangle (9.3,4) node[pos=.5] {\scriptsize 3 entries};
+
+% % Arrows showing cause-effect
+% %\draw[->,thick] (2.2,0.5) .. controls (2.8,1.5) .. (3.4,2) node[midway,above,sloped]{\scriptsize Padding $\Rightarrow$ More entries};
+% \draw[->,thick] (5.1,0.5) .. controls (5.4,2) .. (6.6,2.5) node[midway,above,sloped]{\scriptsize Padding $\Rightarrow$ More entries};
+
+% % Legend
+% \begin{scope}[xshift=0cm,yshift=-1.5cm]
+% \draw[fill=keystoneblue!60] (0,0) rectangle (0.4,0.4) node[right=0.5cm] {Contiguous region};
+% \draw[fill=warmaccent!60] (5,0) rectangle (5.4,0.4) node[right=0.5cm] {Fragmented region};
+% \draw[pattern=north east lines,pattern color=black] (0,-0.6) rectangle (0.4,-0.2) node[right=0.5cm] {Wasted padding};
+% \draw[fill=gray!50,opacity=0.4] (5,-0.6) rectangle (5.4,-0.2) node[right=0.5cm] {PMP entry count bar};
+% \end{scope}
+
+% \end{tikzpicture}
+
+% \vspace{0.4em}
+% \scriptsize
+% \textbf{Takeaway:} Align regions to NAPOT boundaries to minimize padding and preserve PMP entries.
+% \end{figure}
+% }
+
+% \vfill
+% \end{frame}
+
+\begin{frame}{PMP Entries vs Max Concurrent Enclaves}
+\centering
+\begin{figure}[htbp]
+\centering
+\begin{tikzpicture}
+\begin{axis}[
+    width=10cm,
+    height=6cm,
+    xlabel={PMP Entries},
+    ylabel={Max Concurrent Enclaves},
+    xmin=0, xmax=16,
+    ymin=0, ymax=16,
+    xtick={0,4,8,12,16},
+    ymajorgrids,
+    grid style=dashed,
+    mark options={solid},
+    nodes near coords,
+    every node near coord/.append style={font=\footnotesize, /pgf/number format/fixed},
+    clip=false,
+    cycle list={{teal,mark=*,thick}},
+    tick label style={font=\small},
+    label style={font=\small\bfseries},
+]
+
+% Data plot
+\addplot[
+    color=teal,
+    mark=*,
+    thick
+] coordinates {
+    (0,0) (2,2) (4,4) (6,6) (8,8) (10,8) (12,8) (14,8) (16,8)
+};
+
+% Highlight PMP exhaustion area
+\fill[red!30,opacity=0.4] (axis cs:8,0) rectangle (axis cs:16,16);
+
+% Annotation inside shaded region
+\node[red, align=center, font=\footnotesize\bfseries] at (axis cs:12,12) {PMP Exhaustion \\ Caps at 8 Enclaves};
+
+% Stability boundary line & label
+\draw[dashed, thick, color=blue] (axis cs:4,0) -- (axis cs:4,16);
+\node[anchor=south, font=\footnotesize\bfseries, color=blue] at (axis cs:4,16) {Stability Boundary};
+
+\end{axis}
+\end{tikzpicture}
+\caption{Max enclaves increase with PMP entries until exhaustion caps growth.}
+\end{figure}
+\end{frame}
+
+\begin{frame}{PMP Misalignment: Why It Matters}
+\begin{block}{The Problem}
+PMP entries must describe power-of-two, naturally aligned regions (NAPOT). Misaligned memory allocations cause unnecessary padding.
+\end{block}
+
+\begin{block}{The Effect}
 \begin{itemize}
-  \item A single logical region may require \textbf{multiple PMP entries}.
-  \item Fragmentation increases as more regions are misaligned.
+    \item Misalignment causes a single memory region to consume multiple PMP entries.
+    \item Fragmentation leads to rapid exhaustion of PMP resources (max = 16 entries).
 \end{itemize}
 \end{block}
-% \draw[->, thick] (0,-0.2) -- (0,-0.8); % optional arrow from Cause → Effect
-}
 
-% Impact
-\only<4>{
-\begin{block}{Effect}
-\begin{itemize}
-  \item A single logical region may require \textbf{multiple PMP entries}.
-  \item Fragmentation increases as more regions are misaligned.
-\end{itemize}
-\end{block}
-\vspace{0.5em}
-\begin{block}{Impact}
-\begin{itemize}
-  \item Higher PMP entry count $\Rightarrow$ fewer entries left for other regions.
-  \item Limits scalability and flexibility of memory protection.
-  \item Wastes address space and complicates security policy.
-\end{itemize}
-\end{block}
-}
+\vspace{1em}
+\centering
+\textbf{Next: Visualizing the impact →}
+\end{frame}
 
-% Key takeaway
-\only<5>{
+\begin{frame}{PMP Misalignment: Visualized}
 \centering
@@
 \node[anchor=south] at (1.375,2.1) {\scriptsize 128\,KB aligned};
-%\draw[pattern=north east lines,pattern color=black] (0.25,0) rectangle (0.5,1);
 \draw[fill=keystoneblue!60] (0.5,0) rectangle (2.2,1);
-%\draw[pattern=north east lines,pattern color=black] (2.2,0) rectangle (2.5,1);
 \draw[fill=keystoneblue!60,opacity=0.4] (0.5,1) rectangle (2.2,2) node[pos=.5] {\scriptsize 1 entry};
@@
 % Arrows showing cause-effect
-%\draw[->,thick] (2.2,0.5) .. controls (2.8,1.5) .. (3.4,2) node[midway,above,sloped]{\scriptsize Padding $\Rightarrow$ More entries};
 \draw[->,thick] (5.1,0.5) .. controls (5.4,2) .. (6.6,2.5) node[midway,above,sloped]{\scriptsize Padding $\Rightarrow$ More entries};
@@
 \end{figure}
-}
-
-\vfill
 \end{frame}
@@
 % Cloud
-\node[box, fill=blue!20] (cloud) {\includegraphics[scale=0.05]{example-image} \\ \textbf{Cloud Server} \\ (Untrusted Environment)};
+\node[box, fill=blue!20] (cloud) {\faCloud \\ \textbf{Cloud Server} \\ (Untrusted Environment)};
 
 % Lock / Enclave
-\node[box, fill=green!20, right=3cm of cloud] (enclave) {\faLock \\ \textbf{Keystone Enclave} \\ Secure Execution};
+\node[box, fill=green!20, right=3cm of cloud] (enclave) {\faLock \\ \textbf{Keystone Enclave} \\ Trusted Execution};
 
@@
 
-% Summary of Findings
-\begin{frame}{Key Takeaways}
+% % Summary of Findings
+% \begin{frame}{Key Takeaways}
+% \begin{itemize}
+%     \item Keystone offers strong security with minimal overhead for CPU-bound workloads.
+%     \item Significant overhead for memory bound tasks can be reduced with optimizations.
+%     \item Customizable settings allow for optimization based on workload type.
+% \end{itemize}
+% \end{frame}
+
+\begin{frame}{Summary of Findings}
 \begin{itemize}
-    \item Keystone offers strong security with minimal overhead for CPU-bound workloads.
-    \item Significant overhead for memory bound tasks can be reduced with optimizations.
-    \item Customizable settings allow for optimization based on workload type.
+    \item \textbf{Keystone offers low-overhead enclaves} for CPU-bound workloads.
+    \item \textbf{Memory-bound performance degrades} due to PMP overhead and context switching.
+    \item \textbf{Scalability limited} by fixed PMP entries and misaligned region padding.
+    \item \textbf{Performance can be improved} via alignment and workload partitioning strategies.
 \end{itemize}

### 2025-09-03 draft_presentation.tex
improved best practices section

@@
 % Optimizations for Keystone Enclaves
-\begin{frame}{Optimizations for Keystone Enclaves}
+% \begin{frame}{Optimizations for Keystone Enclaves}
+% \begin{itemize}
+%     \item Use contiguous, NAPOT-aligned memory to minimize PMP usage.
+%     \item Consolidate workloads and limit active enclaves to reduce contention.
+%     \item Offload non-sensitive tasks to the untrusted domain.
+% \end{itemize}
+% \end{frame}
+
+\begin{frame}{Deployment Takeaways}
+\begin{block}{Best Practices}
 \begin{itemize}
-    \item Use contiguous, NAPOT-aligned memory to minimize PMP usage.
-    \item Consolidate workloads and limit active enclaves to reduce contention.
-    \item Offload non-sensitive tasks to the untrusted domain.
+    \item Use NAPOT-aligned, contiguous memory regions.
+    \item Consolidate workloads into fewer enclaves.
+    \item Batch calls to reduce context switches.
 \end{itemize}
+\end{block}
+
+\begin{alertblock}{Key Principle}
+Balance performance, isolation, and PMP resource use.
+\end{alertblock}
 \end{frame}
+
+% \begin{frame}{Memory Configuration Strategies}
+% \begin{block}{Key Practices}
+% \begin{itemize}
+%     \item Use \textbf{NAPOT-aligned, contiguous memory} for enclaves.
+%     \item Avoid fragmentation to conserve PMP entries.
+% \end{itemize}
+% \end{block}
+
+% \begin{block}{Trade-Offs}
+% \begin{itemize}
+%     \item Larger regions improve performance (fewer faults, better locality).
+%     \item But excessive size leads to wasted PMP entries.
+% \end{itemize}
+% \end{block}
+% \end{frame}
+
+% \begin{frame}{CPU Scaling Considerations}
+% \begin{block}{Observations}
+% \begin{itemize}
+%     \item Performance improves up to \textasciitilde{}4 cores.
+%     \item Beyond that, overhead from context switches and sync dominates.
+% \end{itemize}
+% \end{block}
+
+% \begin{alertblock}{Recommendation}
+% Use enclaves for \textbf{compute-bound workloads} (e.g., CoreMark), not memory-bound ones (e.g., Kyber).
+% \end{alertblock}
+% \end{frame}
+
+\begin{frame}{Workload-Aware Enclave Design}
+\begin{block}{Hybrid Cryptographic Workloads}
+\begin{itemize}
+    \item Keep keys inside enclaves.
+    \item Offload bulk crypto outside to reduce overhead.
+\end{itemize}
+\end{block}
+
+\begin{block}{Target Use Cases}
+\begin{itemize}
+    \item Isolate sensitive logic (e.g., ML inference, attestation).
+    \item Avoid enclaves for memory-bound workloads like Kyber.
+\end{itemize}
+\end{block}
+\end{frame}
+
+\begin{frame}{Enclave Concurrency and Multiplexing}
+\begin{block}{Problem}
+Too many active enclaves → PMP exhaustion, context switch cost.
+\end{block}
+\begin{alertblock}{Solution}
+Use enclave multiplexing: combine tasks into fewer enclaves.
+\end{alertblock}
+\begin{block}{Bonus}
+Dynamic PMP reallocation can further improve scalability.
+\end{block}
+\end{frame}
+
+% \begin{frame}{Runtime Optimization}
+% \begin{block}{Key Ideas}
+% \begin{itemize}
+%     \item \textbf{Batch enclave calls} to reduce switch overhead.
+%     \item Use \textbf{dynamic PMP assignment} based on active workloads.
+% \end{itemize}
+% \end{block}
+
+% \begin{block}{Lifecycle Management}
+% \begin{itemize}
+%     \item Reuse enclaves across tasks.
+%     \item Scale resources based on current demand.
+% \end{itemize}
+% \end{block}
+% \end{frame}
 
@@
 % Future Work and Improvements
-\begin{frame}{Future Work and Improvements}
+\begin{frame}{Hardware Outlook \& Future Directions}
+\begin{block}{Limitations}
 \begin{itemize}
-    \item Performance optimizations with hardware extensions.
-    \item Multi-enclave systems for more complex use cases.
-    %\item Research on side-channel attacks and improved security measures.
-    \item Development of better developer tools for enclave management.
+    \item Few PMP entries.
+    \item Rigid NAPOT region encoding.
 \end{itemize}
+\end{block}
+
+\begin{alertblock}{Future Opportunities}
+\begin{itemize}
+    \item More flexible PMP schemes.
+    \item Crypto accelerators for enclave workloads.
+    \item Software-defined memory isolation models.
+\end{itemize}
+\end{alertblock}
 \end{frame}
@@
 
-\begin{frame}{Summary of Findings}
-\begin{itemize}
-    \item \textbf{Keystone offers low-overhead enclaves} for CPU-bound workloads.
-    \item \textbf{Memory-bound performance degrades} due to PMP overhead and context switching.
-    \item \textbf{Scalability limited} by fixed PMP entries and misaligned region padding.
-    \item \textbf{Performance can be improved} via alignment and workload partitioning strategies.
-\end{itemize}
-\end{frame}
+% \begin{frame}{Summary of Best Practices}
+% \begin{block}{Configuration Tips}
+% \begin{itemize}
+%     \item Use \textbf{NAPOT-aligned contiguous memory}.
+%     \item \textbf{Limit concurrent enclaves}; favor multiplexing.
+%     \item Offload non-sensitive tasks.
+%     \item Batch calls, manage PMP dynamically.
+% \end{itemize}
+% \end{block}
+
+% \begin{alertblock}{Design Principles}
+% \textbf{Balance} isolation, performance, and scalability.
+% \end{alertblock}
+% \end{frame}
 

### 2025-09-03 draft_presentation.tex
improved PMP diagram

@@
 
-\begin{frame}{QEMU-Based Virtualized Environment}
-    \small
-    \begin{itemize}
-        \item QEMU emulates RISC-V architecture for repeatability and control. \pause
-        \item Ensures consistent test conditions by eliminating hardware-induced variations (thermal throttling, cache behavior). \pause
-        \item Configured for RV64GC architecture, as required by Keystone. \pause
-    \end{itemize}
-    \vspace{0.5em}
-    \uncover<4->{\textbf{Why QEMU?} Provides a controlled, consistent environment for repeatability.}
-\end{frame}
+% \begin{frame}{QEMU-Based Virtualized Environment}
+%     \small
+%     \begin{itemize}
+%         \item QEMU emulates RISC-V architecture for repeatability and control. \pause
+%         \item Ensures consistent test conditions by eliminating hardware-induced variations (thermal throttling, cache behavior). \pause
+%         \item Configured for RV64GC architecture, as required by Keystone. \pause
+%     \end{itemize}
+%     \vspace{0.5em}
+%     \uncover<4->{\textbf{Why QEMU?} Provides a controlled, consistent environment for repeatability.}
+% \end{frame}
 
@@
     xlabel={PMP Entries},
-    ylabel={Max Concurrent Enclaves},
+    ylabel={Concurrent Enclaves},
     xmin=0, xmax=16,
@@
 
+% Add Yellow Sliding Zone (between 4 and 8 PMP entries)
+\fill[yellow!30, opacity=0.6] (axis cs:4,0) rectangle (axis cs:8,16);
+
 % Annotation inside shaded region
 \node[red, align=center, font=\footnotesize\bfseries] at (axis cs:12,12) {PMP Exhaustion \\ Caps at 8 Enclaves};
+\node[align=center, font=\footnotesize\bfseries, text=black] at (axis cs:6,12) {Sliding Zone};
 

### 2025-09-04 draft_presentation.tex
added overlays

@@
 \begin{frame}{From Problem to Solution: Enter Keystone}
+    \only<1>{
+\begin{block}{What is Keystone?}
 \begin{itemize}
@@
     \item Low Trusted Computing Base (TCB) for security and flexibility.
-    \item Optimized for secure computation in untrusted environments.
+    %\item Optimized for secure computation in untrusted environments.
 \end{itemize}
-\begin{block}{Key Concept}
-\textit{Low TCB reduces attack surfaces while enabling high flexibility and performance in enclave-based systems.}
 \end{block}
-\end{frame}
+    }
+% \begin{block}{Key Concept}
+% \textit{Low TCB reduces attack surfaces while enabling high flexibility and performance in enclave-based systems.}
+% \end{block}
+%\end{frame}
 
 % Keystone Framework Architecture
-\begin{frame}{Keystone Framework Architecture}
-\begin{block}{Overview}
-    Keystone is designed to be modular, with clear isolation between enclaves and the OS.
-\end{block}
-
+%\begin{frame}{Keystone Framework Architecture}
+%     \only<1>{
+% \begin{block}{Overview}
+%     Keystone is designed to be modular, with clear isolation between enclaves and the OS.
+% \end{block}
+%     }
+%     \only<2>{
 %\resizebox{\linewidth}{!}{%
+\only<2>{
 \begin{figure}[htbp]
@@
 \includegraphics[width=\linewidth]{/home/individual-01/Documents/keystone-thesis/figures/keystone_overview.png}
-%\caption{Keystone architecture showing components such as the Security Monitor, enclave runtime, and the privilege hierarchy.}
+\caption{Keystone architecture showing components such as the Security Monitor, enclave runtime, and the privilege hierarchy.}
 \label{fig:keystone_grouped}
 \end{figure}
-\end{frame}
+    }
+%\end{frame}
 
 % Keystone vs Other TEEs
-\begin{frame}{Keystone vs. Other TEEs}
+%\begin{frame}{Keystone vs. Other TEEs}
+\only<3>{
 \begin{figure}[htbp]
@@
 \end{figure}
-
+}
 \end{frame}
@@
 \small
+% First overlay: text only
+\only<1>{
 \begin{block}{Performance Overhead}
 \begin{itemize}
-    \item \textbf{Enclave execution} introduces overhead: 10\% to 15\%.
-    \item \textbf{Compute-bound workloads} (CoreMark, Dhrystone) see minimal performance loss.
-    \item \textbf{Memory-bound workloads} (Kyber) experience severe slowdowns due to added security checks and memory isolation.
+    \item \textbf{Enclave execution} introduces overhead:
+        \begin{itemize}
+            \item Compute-bound workloads like CoreMark and Dhrystone incur moderate overhead of 1.06$\times$ to 1.61$\times$.
+            \item Hybrid workloads (Compute- and memory-bound) like Kyber suffer severe overhead, up to 2.73$\times$.
+        \end{itemize}
+    %\item \textbf{Memory-bound workloads} experience significant penalties due to enclave-induced memory isolation and security checks.
 \end{itemize}
 \end{block}
-
+}
+% Second overlay: add figure
+\only<2>{
 \begin{block}{Overhead Breakdown}
@@
 \end{block}
-
-\centering
-\textbf{Key Insight:} Overhead varies based on workload characteristics and memory behavior.
-\end{frame}
-
-\begin{frame}{Performance Trends and Hardware Bottlenecks}
+}
+% \centering
+% \textbf{Key Insight:} Overhead varies based on workload characteristics and memory behavior.
+% \end{frame}
+% Figure appears only on the second slide of this frame
+\only<3>{
+% \begin{frame}{Performance Trends and Hardware Bottlenecks}
 \centering
@@
 
+% Native: orange, rounded top, transparency
+\addplot[
+    fill={rgb,255:red,230; green,159; blue,0},
+    fill opacity=0.8,
+    draw=black!60,
+    rounded corners=2pt
+] coordinates {(CoreMark,100) (Dhrystone,100) (Kyber,100)};
+
 % Enclave: blue, rounded top, transparency
@@
 
-% Native: orange, rounded top, transparency
-\addplot[
-    fill={rgb,255:red,230; green,159; blue,0},
-    fill opacity=0.8,
-    draw=black!60,
-    rounded corners=2pt
-] coordinates {(CoreMark,100) (Dhrystone,100) (Kyber,100)};
-
-\legend{Enclave,Native}
+\legend{Native,Enclave}
 \end{axis}
@@
 \end{figure}
+}
 \end{frame}
@@
 \small
+\only<1>{
 \begin{block}{PMP Resource Limitations}
@@
 \end{block}
-
+}
+\only<2>{
 \begin{block}{Scalability Bottleneck}
@@
 \end{block}
+}
 
-\centering
-\textbf{Key Insight:} PMP limitations represent a significant scalability bottleneck for RISC-V enclave-based systems.
-\end{frame}
+% \centering
+% \textbf{Key Insight:} PMP limitations represent a significant scalability bottleneck for RISC-V enclave-based systems.
+% \end{frame}
 
@@
 
-\begin{frame}{PMP Entries vs Max Concurrent Enclaves}
-\centering
+% \begin{frame}{PMP Entries vs Max Concurrent Enclaves}
+% \centering
+\only<3>{
 \begin{figure}[htbp]
@@
 \begin{frame}{PMP Misalignment: Why It Matters}
+    \only<1>{
 \begin{block}{The Problem}
@@
 \end{block}
+    }
+%     \only<2>{
+% \begin{block}{The Effect}
+% \begin{itemize}
+%     \item Misalignment causes a single memory region to consume multiple PMP entries.
+%     \item Fragmentation leads to rapid exhaustion of PMP resources (max = 16 entries).
+% \end{itemize}
+% \end{block}
 
-\begin{block}{The Effect}
-\begin{itemize}
-    \item Misalignment causes a single memory region to consume multiple PMP entries.
-    \item Fragmentation leads to rapid exhaustion of PMP resources (max = 16 entries).
-\end{itemize}
-\end{block}
-
-\vspace{1em}
-\centering
-\textbf{Next: Visualizing the impact →}
-\end{frame}
-
-\begin{frame}{PMP Misalignment: Visualized}
-\centering
+% \vspace{1em}
+% \centering
+% \textbf{Next: Visualizing the impact →}
+%\end{frame}
+%     }
+    \only<2>{
+%\begin{frame}{PMP Misalignment: Visualized}
+%\centering
 \begin{figure}[htbp]
@@
 \end{figure}
+    }
 \end{frame}
@@
     \item Batch calls to reduce context switches.
+    \item Adapt enclave usage based on task complexity.
 \end{itemize}
@@
 \begin{alertblock}{Key Principle}
-Balance performance, isolation, and PMP resource use.
+    Balance performance, isolation, and PMP resource usage for optimal efficiency.
 \end{alertblock}
@@
 
-\begin{frame}{Workload-Aware Enclave Design}
-\begin{block}{Hybrid Cryptographic Workloads}
-\begin{itemize}
-    \item Keep keys inside enclaves.
-    \item Offload bulk crypto outside to reduce overhead.
-\end{itemize}
-\end{block}
+% \begin{frame}{Workload-Aware Enclave Design}
+% \begin{block}{Hybrid Cryptographic Workloads}
+% \begin{itemize}
+%     \item Keep keys inside enclaves.
+%     \item Offload bulk crypto operations outside to reduce overhead.
+% \end{itemize}
+% \end{block}
 
-\begin{block}{Target Use Cases}
-\begin{itemize}
-    \item Isolate sensitive logic (e.g., ML inference, attestation).
-    \item Avoid enclaves for memory-bound workloads like Kyber.
-\end{itemize}
-\end{block}
-\end{frame}
+% \begin{block}{Target Use Cases}
+% \begin{itemize}
+%     \item Isolate sensitive logic (e.g., ML inference, attestation).
+%     \item Avoid enclaves for high-memory or parallel tasks. (e.g., bulk cryptographic ops.)
+% \end{itemize}
+% \end{block}
+% \end{frame}
 
-\begin{frame}{Enclave Concurrency and Multiplexing}
-\begin{block}{Problem}
-Too many active enclaves → PMP exhaustion, context switch cost.
-\end{block}
-\begin{alertblock}{Solution}
-Use enclave multiplexing: combine tasks into fewer enclaves.
-\end{alertblock}
-\begin{block}{Bonus}
-Dynamic PMP reallocation can further improve scalability.
-\end{block}
-\end{frame}
+% \begin{frame}{Enclave Concurrency and Multiplexing}
+%     \uncover<1->{
+% \begin{block}{Problem}
+% Too many active enclaves → PMP exhaustion, context switch cost.
+% \end{block}
+%     }
+%     \uncover<2->{
+% \begin{alertblock}{Solution}
+% Use enclave multiplexing: combine tasks into fewer enclaves.
+% \end{alertblock}
+% \begin{block}{Bonus}
+% Dynamic PMP reallocation can further improve scalability.
+% \end{block}
+%     }
+% \end{frame}
 
@@
 \begin{frame}{Hardware Outlook \& Future Directions}
+    \only<1>{
 \begin{block}{Limitations}
@@
 \end{block}
-
+    }
+    \only<2>{
 \begin{alertblock}{Future Opportunities}
@@
 \end{alertblock}
+    }
 \end{frame}

### 2025-09-05 draft_presentation.tex
added new diagrams

@@
 
+\section{Background and System Design}
+
 % Background on TEEs (Added icons for clarity)
@@
 \end{frame}
+
+\section{Experimental Setup}
 
@@
 %Results
-\section{Key Performance Trends and Hardware Bottlenecks}
+\section{Results and Hardware Bottlenecks}
 \label{sec:synthesis}
@@
 
-% Native: orange, rounded top, transparency
+% Native: orange, rounded top, tranRecommendationssparency
 \addplot[
@@
 
+\section{Practical Deployment Strategies and Recommendations}
+
 \begin{frame}{Deployment Takeaways}
@@
 \begin{itemize}
-    \item Use NAPOT-aligned, contiguous memory regions.
-    \item Consolidate workloads into fewer enclaves.
-    \item Batch calls to reduce context switches.
-    \item Adapt enclave usage based on task complexity.
+    \only<1>{\item \faMemory\ \textbf{Align memory precisely:} Use NAPOT-aligned, contiguous regions to simplify PMP configuration.}
+    \only<2>{\item \faProjectDiagram\ \textbf{Minimize enclave count:} Consolidate workloads to reduce PMP entry usage and management overhead.
+    \begin{tikzpicture}[font=\scriptsize,>=stealth]
+
+% --- BEFORE: Multiple enclaves ---
+\node[anchor=west] at (-0.5,2.5) {\textbf{Before: Multiple Enclaves}};
+\foreach \i in {0,2.5,5} {
+  % Enclave box
+  \draw[fill=red!20,thick] (\i,1) rectangle +(2,1);
+  \node at (\i+1,2.15) {Enclave};
+  % Task inside
+  \draw[fill=blue!10,thick] (\i+0.5,1.2) rectangle +(1,0.6);
+  \node at (\i+1,1.5) {Task};
+}
+
+% Arrow between before and after (shorter gap)
+\draw[->,thick] (3.5,0.8) -- (3.5,0.2) node[midway,right]{\textbf{Merge}};
+
+% --- AFTER: Single consolidated enclave (moved up) ---
+\node[anchor=west] at (-0.5,-0.6) {\textbf{After: Single Enclave}};
+\draw[fill=green!20,thick] (0,-1.8) rectangle (7,-0.8);
+\node at (3.5,-2.05) {Single Consolidated Enclave};
+
+% Tasks inside consolidated enclave
+\foreach \i in {0.5,3,5.5} {
+  \draw[fill=blue!10,thick] (\i,-1.6) rectangle +(1,0.6);
+  \node at (\i+0.5,-1.3) {Task};
+}
+
+% Optional legend
+\begin{scope}[shift={(8,0)}]
+  \draw[fill=red!20] (0,1.5) rectangle +(0.5,0.3);
+  \node[right] at (0.6,1.65) {Separate Enclave};
+  \draw[fill=green!20] (0,1) rectangle +(0.5,0.3);
+  \node[right] at (0.6,1.15) {Consolidated Enclave};
+  \draw[fill=blue!10] (0,0.5) rectangle +(0.5,0.3);
+  \node[right] at (0.6,0.65) {Task};
+\end{scope}
+
+\end{tikzpicture}
+
+    }
+    \only<3>{\item \faRetweet\ \textbf{Reduce switching costs:} Batch enclave calls to limit context switches and improve throughput.
+    \begin{figure}[H]
+\centering
+\begin{tikzpicture}[>=stealth,thick,font=\scriptsize,scale=0.9]
+  % === Left: Unbatched ===
+  \fill[green!10] (0,0) rectangle (2.5,2.5);
+  \fill[blue!10] (2.5,0) rectangle (5,2.5);
+
+  \node at (1.25,2.7) {\textbf{Enclave}};
+  \node at (3.75,2.7) {\textbf{Host}};
+
+  \draw[dashed,thick] (2.5,0) -- (2.5,2.5) node[midway,above,sloped];
+
+  \draw[->,red!70!black] (0.8,2.0) -- (3.2,2.0) node[midway,above]{Call 1};
+  \draw[->,red!70!black] (0.8,1.4) -- (3.2,1.4) node[midway,above]{Call 2};
+  \draw[->,red!70!black] (0.8,0.8) -- (3.2,0.8) node[midway,above]{Call 3};
+
+  \node[align=center] at (2.5,-0.6) {\textbf{Unbatched}\\Multiple crossings};
+
+  % === Right: Batched ===
+  \begin{scope}[shift={(6,0)}]
+    \fill[green!10] (0,0) rectangle (2.5,2.5);
+    \fill[blue!10] (2.5,0) rectangle (5,2.5);
+
+    \node at (1.25,2.7) {\textbf{Enclave}};
+    \node at (3.75,2.7) {\textbf{Host}};
+
+    \draw[dashed,thick] (2.5,0) -- (2.5,2.5) node[midway,above,sloped];
+
+\draw[->,red!70!black,thick] (0.8,1.3) -- (2.8,1.3) 
+    node[midway,above]{Batched Call}
+    node[midway,below=4pt,draw,fill=orange!20,rounded corners=2pt,
+         inner xsep=4pt,inner ysep=2pt]{Batch Processor};
+
+    \node[align=center] at (2.5,-0.6) {\textbf{Batched}\\One crossing};
+  \end{scope}
+
+\end{tikzpicture}
+\caption{Unbatched vs Batched enclave-to-host calls}
+\end{figure}
+
+    }
+    \only<4>{\item \faCode\ \textbf{Offload heavy lifting:} Run non-sensitive, high-load tasks outside enclaves for efficiency. \\
+\begin{tikzpicture}[
+    font=\scriptsize,
+    >=Stealth,
+    thick,
+    task/.style={
+        fill=blue!10,
+        rounded corners=1pt,
+        minimum width=2.6cm,
+        minimum height=0.45cm,
+        draw
+    },
+    box/.style={
+        rounded corners=2pt,
+        minimum height=2.4cm,
+        draw,
+        thick
+    }
+]
+
+% --- Enclave ---
+\node[box,fill=green!20,minimum width=3.5cm,anchor=north west,
+      label={[font=\bfseries]above:Enclave}] (enclave) at (0,0) {};
+\node at ([yshift=-0.3cm]enclave.north) {Sensitive Tasks Only};
+
+% Sensitive tasks inside enclave
+\node[task] at ([yshift=-0.9cm]enclave.north) {Crypto Ops};
+\node[task] at ([yshift=-1.5cm]enclave.north) {ML Inference};
+\node[task] at ([yshift=-2.1cm]enclave.north) {Attestation};
+
+% --- Host area ---
+\node[box,fill=gray!15,minimum width=3.5cm,anchor=north west,
+      label={[font=\bfseries]above:Host}] (host) at (5,0) {};
+\node at ([yshift=-0.3cm]host.north) {Non‑Sensitive Tasks};
+
+% Non-sensitive tasks outside
+\node[task] at ([yshift=-0.9cm]host.north) {Bulk Encryption};
+\node[task] at ([yshift=-1.5cm]host.north) {Data Prep};
+
+% --- Arrow & label ---
+\draw[->,ultra thick]
+    ([yshift=-0.9cm]enclave.east) -- ([yshift=-0.9cm]host.west)
+    node[midway,above,font=\scriptsize] {Offload};
+
+% --- Legend ---
+\begin{scope}[shift={(0,-3)}]
+  \draw[fill=green!20,rounded corners=1pt] (0,0) rectangle +(0.35,0.2);
+  \node[right] at (0.45,0.1) {Enclave};
+  \draw[fill=gray!15,rounded corners=1pt] (2,0) rectangle +(0.35,0.2);
+  \node[right] at (2.45,0.1) {Host};
+  \draw[fill=blue!10,rounded corners=1pt] (4,0) rectangle +(0.35,0.2);
+  \node[right] at (4.45,0.1) {Task};
+\end{scope}
+
+\end{tikzpicture}
+    }
 \end{itemize}
@@
 
+\only<5>{
 \begin{alertblock}{Key Principle}
-    Balance performance, isolation, and PMP resource usage for optimal efficiency.
+    \textbf{Optimize the trade-off:} Balance performance, isolation, and PMP resource usage for maximum efficiency.
 \end{alertblock}
+}
 \end{frame}
@@
 
+\section{Future Directions and Conclusion}
 % Future Work and Improvements

### 2025-09-05 draft_presentation.tex
added new diagrams

@@
 \end{tikzpicture}
-\caption{Unbatched vs Batched enclave-to-host calls}
+%\caption{Unbatched vs Batched enclave-to-host calls}
 \end{figure}
@@
 \begin{alertblock}{Key Principle}
-    \textbf{Optimize the trade-off:} Balance performance, isolation, and PMP resource usage for maximum efficiency.
+\textbf{Balance system trade-offs:} Align enclave design with workload needs, considering performance, security, and hardware constraints.
 \end{alertblock}
+\begin{center}
+\begin{tikzpicture}[scale=0.9, every node/.style={font=\scriptsize}, thick]
+    % Define colors
+    \colorlet{perfcol}{blue!60}
+    \colorlet{isocol}{red!60}
+    \colorlet{pmpcol}{green!60}
+
+    % Draw overlapping circles
+    \begin{scope}[blend group=multiply]
+        \fill[perfcol, opacity=0.4] (90:1.4) circle (1.4);
+        \fill[isocol, opacity=0.4] (210:1.4) circle (1.4);
+        \fill[pmpcol, opacity=0.4] (330:1.4) circle (1.4);
+    \end{scope}
+
+    % Circle labels
+    \node[font=\bfseries] at (90:2.3) {Performance};
+    \node[font=\bfseries] at (210:2.3) {Security};
+    \node[font=\bfseries] at (330:2.3) {HW Constraints};
+
+    % Center label
+    \node[fill=white, draw=black!30, rounded corners=2pt, font=\scriptsize, inner sep=2pt] 
+        at (0,0) {\textbf{Balanced Trade-off}};
+
+    % Optional: title above
+    % \node[font=\bfseries\small] at (0,3.2) {Deployment Trade-off Space};
+\end{tikzpicture}
+\end{center}
 }

### 2025-09-05 draft_presentation.tex
added new diagrams

@@
 \small
+\only<1>{
 \begin{block}{CPU Scaling Limits}
 \begin{itemize}
-    \item Increasing CPU cores \textbf{benefits multi-threaded tasks}, but gains plateau due to context-switch overhead and enclave thread management.
+    \item Increasing CPU cores \textbf{benefits multi-threaded tasks}, but gains plateau due to context-switch overhead.
     \item For example, \texttt{CoreMark} shows marginal improvements beyond 4 cores.
@@
 \end{block}
-
+\centering
+\begin{tikzpicture}
+\begin{axis}[
+    ybar,
+    bar width=8pt,
+    width=0.75\linewidth,
+    height=4.2cm,
+    enlarge x limits=0.15,
+    ymin=95, ymax=125,
+    xlabel={CPU cores},
+    ylabel={Rel.\ perf.\ (\%)},
+    symbolic x coords={1,2,3,4,5,6,7,8},
+    xtick=data,
+    ymajorgrids,
+    grid style={dotted, gray!50},
+    font=\scriptsize,
+    tick label style={font=\scriptsize},
+    legend style={
+        font=\scriptsize,
+        at={(0.5,-0.50)},
+        anchor=north,
+        legend columns=2
+    },
+    nodes near coords,
+    every node near coord/.append style={font=\tiny},
+    point meta=explicit symbolic
+]
+
+% Native
+\addplot[fill=blue!40] coordinates {
+    (1,100) [100]
+    (2,108) [108]
+    (3,115) [115]
+    (4,120) [120]
+    (5,122) [122]
+    (6,123) [123]
+    (7,123) [123]
+    (8,123) [123]
+};
+
+% Enclave
+\addplot[fill=green!50!black!40] coordinates {
+    (1,100) [100]
+    (2,102) [102]
+    (3,103) [103]
+    (4,104) [104]
+    (5,106) [106]
+    (6,108) [108]
+    (7,110) [110]
+    (8,110) [110]
+};
+
+\legend{Native, Enclave}
+\end{axis}
+\end{tikzpicture}
+
+}
+\only<2>{
 \begin{block}{Memory Scaling Constraints}
@@
 \end{block}
-
-\centering
-\textbf{Key Insight:} Hardware resources alone cannot fully predict enclave performance due to system resource allocation limits.
+\centering
+\begin{tikzpicture}
+\begin{axis}[
+    ybar,
+    bar width=8pt,
+    width=0.75\linewidth,
+    height=4.2cm,
+    enlarge x limits=0.2,
+    ymin=90, ymax=118,
+    xlabel={Memory (MB)},
+    ylabel={Rel.\ perf.\ (\%)},
+    symbolic x coords={64,128,256,512},
+    xtick=data,
+    ymajorgrids,
+    grid style={dotted, gray!50},
+    font=\scriptsize,
+    tick label style={font=\scriptsize},
+    legend style={
+        font=\scriptsize,
+        at={(0.5,-0.5)},
+        anchor=north,
+        legend columns=1
+    },
+    nodes near coords,
+    every node near coord/.append style={font=\tiny},
+    point meta=explicit symbolic
+]
+
+% Native contiguous
+\addplot[fill=blue!40] coordinates {
+    (64,100) [100]
+    (128,106) [106]
+    (256,112) [112]
+    (512,115) [115]
+};
+
+% Enclave contiguous
+\addplot[fill=green!50!black!40] coordinates {
+    (64,95) [95]
+    (128,98) [98]
+    (256,100) [100]
+    (512,102) [102]
+};
+
+% Enclave fragmented
+\addplot[fill=orange!70!black!40] coordinates {
+    (64,92) [92]
+    (128,94) [94]
+    (256,95) [95]
+    (512,96) [96]
+};
+
+\legend{Native contig., Enclave contig., Enclave frag.}
+\end{axis}
+\end{tikzpicture}
+
+}
+% \centering
+% \textbf{Key Insight:} Hardware resources alone cannot fully predict enclave performance due to system resource allocation limits.
 \end{frame}

### 2025-09-05 draft_presentation.tex
added new diagrams

@@
 \end{block}
-}
-% Second overlay: add figure
-\only<2>{
-\begin{block}{Overhead Breakdown}
-\begin{itemize}
-    \item Memory isolation via PMP adds fixed latency per memory access.
-    \item Context switches between host and enclave domains are costly.
-\end{itemize}
-\end{block}
-}
-% \centering
-% \textbf{Key Insight:} Overhead varies based on workload characteristics and memory behavior.
-% \end{frame}
-% Figure appears only on the second slide of this frame
-\only<3>{
-% \begin{frame}{Performance Trends and Hardware Bottlenecks}
-\centering
-\begin{figure}[htbp]
+
+%\vspace{0.5cm} % Adjust space as needed
+
 \centering
@@
     bar width=14pt,
-    width=10cm,
-    height=6cm,
+    width=0.8\linewidth,
+    height=4cm,
     ymin=0, ymax=120,
-    ylabel={Relative performance (\%)},
+    ylabel={Rel. perf. (\%)},
     symbolic x coords={CoreMark,Dhrystone,Kyber},
@@
     every node near coord/.append style={font=\footnotesize, /pgf/number format/fixed},
-    legend style={at={(0.5,-0.15)}, anchor=north, legend columns=-1}
+    legend style={at={(0.5,-0.35)}, anchor=north, legend columns=-1}
 ]
 
-% Native: orange, rounded top, tranRecommendationssparency
+% Native: orange
 \addplot[
@@
 
-% Enclave: blue, rounded top, transparency
+% Enclave: blue
 \addplot[
@@
 
-\legend{Native,Enclave}
+\legend{Native, Enclave}
 \end{axis}
 \end{tikzpicture}
-\caption{Relative performance of enclave vs native execution.}
-\end{figure}
-}
-\end{frame}
+
+}
+
+% \only<2>{
+% % \begin{frame}{Performance Trends and Hardware Bottlenecks}
+% \centering
+% \begin{figure}[htbp]
+% \centering
+% \begin{tikzpicture}
+% \begin{axis}[
+%     ybar,
+%     bar width=14pt,
+%     width=10cm,
+%     height=6cm,
+%     ymin=0, ymax=120,
+%     ylabel={Relative performance (\%)},
+%     symbolic x coords={CoreMark,Dhrystone,Kyber},
+%     xtick=data,
+%     ymajorgrids,
+%     grid style=dashed,
+%     nodes near coords,
+%     every node near coord/.append style={font=\footnotesize, /pgf/number format/fixed},
+%     legend style={at={(0.5,-0.15)}, anchor=north, legend columns=-1}
+% ]
+
+% % Native: orange, rounded top, tranRecommendationssparency
+% \addplot[
+%     fill={rgb,255:red,230; green,159; blue,0},
+%     fill opacity=0.8,
+%     draw=black!60,
+%     rounded corners=2pt
+% ] coordinates {(CoreMark,100) (Dhrystone,100) (Kyber,100)};
+
+% % Enclave: blue, rounded top, transparency
+% \addplot[
+%     fill={rgb,255:red,0; green,114; blue,178},
+%     fill opacity=0.8,
+%     draw=black!60,
+%     rounded corners=2pt
+% ] coordinates {(CoreMark,95) (Dhrystone,62) (Kyber,37)};
+
+% \legend{Native,Enclave}
+% \end{axis}
+% \end{tikzpicture}
+% \caption{Relative performance of enclave vs native execution.}
+% \end{figure}
+% }
+\end{frame}
+
+% % Second overlay: add figure
+% \only<3>{
+% \begin{block}{Overhead Breakdown}
+% \begin{itemize}
+%     \item Memory isolation via PMP adds fixed latency per memory access.
+%     \item Context switches between host and enclave domains are costly.
+% \end{itemize}
+% \end{block}
+% }
+% \centering
+% \textbf{Key Insight:} Overhead varies based on workload characteristics and memory behavior.
+% \end{frame}
+% Figure appears only on the second slide of this frame
 
@@
     enlarge x limits=0.15,
-    ymin=95, ymax=125,
+    ymin=95, ymax=129,
     xlabel={CPU cores},
@@
 \end{block}
-}
-
-% \centering
-% \textbf{Key Insight:} PMP limitations represent a significant scalability bottleneck for RISC-V enclave-based systems.
-% \end{frame}
-
-% \begin{frame}{Key Takeaways from Performance Trends}
-% \small
-% \begin{block}{Main Findings}
-% \begin{itemize}
-%     \item Enclave performance suffers primarily in \textbf{memory-bound workloads} due to security checks and context switching.
-%     \item \textbf{PMP entry limitations} restrict scalability in multi-enclave setups.
-%     \item Increasing hardware resources does not linearly translate to performance gains due to system-level overheads.
-% \end{itemize}
-% \end{block}
-
-% \centering
-% \textbf{Key Insight:} Optimizing both hardware resources and system-level configurations is essential for scalable and efficient enclave execution.
-% \end{frame}
-
-% \begin{frame}{Why Misalignment Bloats PMP Usage}
-% \small
-% \vfill
-
-% % Problem
-% \only<1>{
-% \begin{block}{Problem}
-% Physical Memory Protection (PMP) entries must cover power‑of‑two sized, naturally aligned regions (NAPOT).
-% \end{block}
-% }
-
-% % Cause
-% \only<2>{
-% \begin{block}{Problem}
-% Physical Memory Protection (PMP) entries must cover power‑of‑two sized, naturally aligned regions (NAPOT).
-% \end{block}
-% \vspace{0.5em}
-% \begin{block}{Cause}
-% When a memory region is \textbf{misaligned}:
-% \begin{itemize}
-%   \item NAPOT encoding forces padding to the next power‑of‑two boundary.
-%   \item Padding areas consume PMP coverage even if they are not actual data.
-% \end{itemize}
-% \end{block}
-% % \draw[->, thick] (0,-0.2) -- (0,-0.8); % optional arrow from Problem → Cause
-% }
-
-% % Effect
-% \only<3>{
-% \begin{block}{Cause}
-% When a memory region is \textbf{misaligned}:
-% \begin{itemize}
-%   \item NAPOT encoding forces padding to the next power‑of‑two boundary.
-%   \item Padding areas consume PMP coverage even if they are not actual data.
-% \end{itemize}
-% \end{block}
-% \vspace{0.5em}
-% \begin{block}{Effect}
-% \begin{itemize}
-%   \item A single logical region may require \textbf{multiple PMP entries}.
-%   \item Fragmentation increases as more regions are misaligned.
-% \end{itemize}
-% \end{block}
-% % \draw[->, thick] (0,-0.2) -- (0,-0.8); % optional arrow from Cause → Effect
-% }
-
-% % Impact
-% \only<4>{
-% \begin{block}{Effect}
-% \begin{itemize}
-%   \item A single logical region may require \textbf{multiple PMP entries}.
-%   \item Fragmentation increases as more regions are misaligned.
-% \end{itemize}
-% \end{block}
-% \vspace{0.5em}
-% \begin{block}{Impact}
-% \begin{itemize}
-%   \item Higher PMP entry count $\Rightarrow$ fewer entries left for other regions.
-%   \item Limits scalability and flexibility of memory protection.
-%   \item Wastes address space and complicates security policy.
-% \end{itemize}
-% \end{block}
-% }
-
-% % Key takeaway
-% \only<5>{
-% \centering
-% \begin{figure}[htbp]
-% \centering
-% \begin{tikzpicture}[scale=1, transform shape,font=\small]
-
-% % Memory axis
-% \draw[thick,->] (0,0) -- (10,0) node[anchor=west]{Memory address};
-% \foreach \x/\label in {0/0\,KB,1.25/128\,KB,4.25/192\,KB,8/300\,KB}
-%     \draw (\x,0.1) -- (\x,-0.1) node[anchor=north]{\label};
-
-% % === Region 1: contiguous, 1 PMP entry ===
-% \node[anchor=south] at (1.375,2.1) {\scriptsize 128\,KB aligned};
-% %\draw[pattern=north east lines,pattern color=black] (0.25,0) rectangle (0.5,1);
-% \draw[fill=keystoneblue!60] (0.5,0) rectangle (2.2,1);
-% %\draw[pattern=north east lines,pattern color=black] (2.2,0) rectangle (2.5,1);
-% \draw[fill=keystoneblue!60,opacity=0.4] (0.5,1) rectangle (2.2,2) node[pos=.5] {\scriptsize 1 entry};
-
-% % === Region 2: fragmented, 3 PMP entries ===
-% \node[anchor=south] at (4.3,4.1) {\scriptsize 192\,KB misaligned};
-% \draw[pattern=north east lines,pattern color=black] (3.5,0) rectangle (3.8,1);
-% \draw[fill=warmaccent!60] (3.8,0) rectangle (4.6,1);
-% \draw[pattern=north east lines,pattern color=black] (4.6,0) rectangle (5.1,1);
-% \draw[fill=warmaccent!60,opacity=0.4] (3.5,1) rectangle (5.1,4) node[pos=.5] {\scriptsize 3 entries};
-
-% % === Region 3: fragmented, 3 PMP entries ===
-% \node[anchor=south] at (8.0,4.1) {\scriptsize 300\,KB misaligned};
-% \draw[pattern=north east lines,pattern color=black] (6.7,0) rectangle (7.0,1);
-% \draw[fill=warmaccent!60] (7.0,0) rectangle (9.1,1);
-% \draw[pattern=north east lines,pattern color=black] (9.1,0) rectangle (9.3,1);
-% \draw[fill=warmaccent!60,opacity=0.4] (6.7,1) rectangle (9.3,4) node[pos=.5] {\scriptsize 3 entries};
-
-% % Arrows showing cause-effect
-% %\draw[->,thick] (2.2,0.5) .. controls (2.8,1.5) .. (3.4,2) node[midway,above,sloped]{\scriptsize Padding $\Rightarrow$ More entries};
-% \draw[->,thick] (5.1,0.5) .. controls (5.4,2) .. (6.6,2.5) node[midway,above,sloped]{\scriptsize Padding $\Rightarrow$ More entries};
-
-% % Legend
-% \begin{scope}[xshift=0cm,yshift=-1.5cm]
-% \draw[fill=keystoneblue!60] (0,0) rectangle (0.4,0.4) node[right=0.5cm] {Contiguous region};
-% \draw[fill=warmaccent!60] (5,0) rectangle (5.4,0.4) node[right=0.5cm] {Fragmented region};
-% \draw[pattern=north east lines,pattern color=black] (0,-0.6) rectangle (0.4,-0.2) node[right=0.5cm] {Wasted padding};
-% \draw[fill=gray!50,opacity=0.4] (5,-0.6) rectangle (5.4,-0.2) node[right=0.5cm] {PMP entry count bar};
-% \end{scope}
-
-% \end{tikzpicture}
-
-% \vspace{0.4em}
-% \scriptsize
-% \textbf{Takeaway:} Align regions to NAPOT boundaries to minimize padding and preserve PMP entries.
-% \end{figure}
-% }
-
-% \vfill
-% \end{frame}
-
-% \begin{frame}{PMP Entries vs Max Concurrent Enclaves}
-% \centering
-\only<3>{
-\begin{figure}[htbp]
+
 \centering
@@
     width=10cm,
-    height=6cm,
+    height=4.5cm,
     xlabel={PMP Entries},
@@
 \end{tikzpicture}
-\caption{Max enclaves increase with PMP entries until exhaustion caps growth.}
-\end{figure}
-}
+}
+
+% \centering
+% \textbf{Key Insight:} PMP limitations represent a significant scalability bottleneck for RISC-V enclave-based systems.
+% \end{frame}
+
+% \begin{frame}{Key Takeaways from Performance Trends}
+% \small
+% \begin{block}{Main Findings}
+% \begin{itemize}
+%     \item Enclave performance suffers primarily in \textbf{memory-bound workloads} due to security checks and context switching.
+%     \item \textbf{PMP entry limitations} restrict scalability in multi-enclave setups.
+%     \item Increasing hardware resources does not linearly translate to performance gains due to system-level overheads.
+% \end{itemize}
+% \end{block}
+
+% \centering
+% \textbf{Key Insight:} Optimizing both hardware resources and system-level configurations is essential for scalable and efficient enclave execution.
+% \end{frame}
+
+% \begin{frame}{Why Misalignment Bloats PMP Usage}
+% \small
+% \vfill
+
+% % Problem
+% \only<1>{
+% \begin{block}{Problem}
+% Physical Memory Protection (PMP) entries must cover power‑of‑two sized, naturally aligned regions (NAPOT).
+% \end{block}
+% }
+
+% % Cause
+% \only<2>{
+% \begin{block}{Problem}
+% Physical Memory Protection (PMP) entries must cover power‑of‑two sized, naturally aligned regions (NAPOT).
+% \end{block}
+% \vspace{0.5em}
+% \begin{block}{Cause}
+% When a memory region is \textbf{misaligned}:
+% \begin{itemize}
+%   \item NAPOT encoding forces padding to the next power‑of‑two boundary.
+%   \item Padding areas consume PMP coverage even if they are not actual data.
+% \end{itemize}
+% \end{block}
+% % \draw[->, thick] (0,-0.2) -- (0,-0.8); % optional arrow from Problem → Cause
+% }
+
+% % Effect
+% \only<3>{
+% \begin{block}{Cause}
+% When a memory region is \textbf{misaligned}:
+% \begin{itemize}
+%   \item NAPOT encoding forces padding to the next power‑of‑two boundary.
+%   \item Padding areas consume PMP coverage even if they are not actual data.
+% \end{itemize}
+% \end{block}
+% \vspace{0.5em}
+% \begin{block}{Effect}
+% \begin{itemize}
+%   \item A single logical region may require \textbf{multiple PMP entries}.
+%   \item Fragmentation increases as more regions are misaligned.
+% \end{itemize}
+% \end{block}
+% % \draw[->, thick] (0,-0.2) -- (0,-0.8); % optional arrow from Cause → Effect
+% }
+
+% % Impact
+% \only<4>{
+% \begin{block}{Effect}
+% \begin{itemize}
+%   \item A single logical region may require \textbf{multiple PMP entries}.
+%   \item Fragmentation increases as more regions are misaligned.
+% \end{itemize}
+% \end{block}
+% \vspace{0.5em}
+% \begin{block}{Impact}
+% \begin{itemize}
+%   \item Higher PMP entry count $\Rightarrow$ fewer entries left for other regions.
+%   \item Limits scalability and flexibility of memory protection.
+%   \item Wastes address space and complicates security policy.
+% \end{itemize}
+% \end{block}
+% }
+
+% % Key takeaway
+% \only<5>{
+% \centering
+% \begin{figure}[htbp]
+% \centering
+% \begin{tikzpicture}[scale=1, transform shape,font=\small]
+
+% % Memory axis
+% \draw[thick,->] (0,0) -- (10,0) node[anchor=west]{Memory address};
+% \foreach \x/\label in {0/0\,KB,1.25/128\,KB,4.25/192\,KB,8/300\,KB}
+%     \draw (\x,0.1) -- (\x,-0.1) node[anchor=north]{\label};
+
+% % === Region 1: contiguous, 1 PMP entry ===
+% \node[anchor=south] at (1.375,2.1) {\scriptsize 128\,KB aligned};
+% %\draw[pattern=north east lines,pattern color=black] (0.25,0) rectangle (0.5,1);
+% \draw[fill=keystoneblue!60] (0.5,0) rectangle (2.2,1);
+% %\draw[pattern=north east lines,pattern color=black] (2.2,0) rectangle (2.5,1);
+% \draw[fill=keystoneblue!60,opacity=0.4] (0.5,1) rectangle (2.2,2) node[pos=.5] {\scriptsize 1 entry};
+
+% % === Region 2: fragmented, 3 PMP entries ===
+% \node[anchor=south] at (4.3,4.1) {\scriptsize 192\,KB misaligned};
+% \draw[pattern=north east lines,pattern color=black] (3.5,0) rectangle (3.8,1);
+% \draw[fill=warmaccent!60] (3.8,0) rectangle (4.6,1);
+% \draw[pattern=north east lines,pattern color=black] (4.6,0) rectangle (5.1,1);
+% \draw[fill=warmaccent!60,opacity=0.4] (3.5,1) rectangle (5.1,4) node[pos=.5] {\scriptsize 3 entries};
+
+% % === Region 3: fragmented, 3 PMP entries ===
+% \node[anchor=south] at (8.0,4.1) {\scriptsize 300\,KB misaligned};
+% \draw[pattern=north east lines,pattern color=black] (6.7,0) rectangle (7.0,1);
+% \draw[fill=warmaccent!60] (7.0,0) rectangle (9.1,1);
+% \draw[pattern=north east lines,pattern color=black] (9.1,0) rectangle (9.3,1);
+% \draw[fill=warmaccent!60,opacity=0.4] (6.7,1) rectangle (9.3,4) node[pos=.5] {\scriptsize 3 entries};
+
+% % Arrows showing cause-effect
+% %\draw[->,thick] (2.2,0.5) .. controls (2.8,1.5) .. (3.4,2) node[midway,above,sloped]{\scriptsize Padding $\Rightarrow$ More entries};
+% \draw[->,thick] (5.1,0.5) .. controls (5.4,2) .. (6.6,2.5) node[midway,above,sloped]{\scriptsize Padding $\Rightarrow$ More entries};
+
+% % Legend
+% \begin{scope}[xshift=0cm,yshift=-1.5cm]
+% \draw[fill=keystoneblue!60] (0,0) rectangle (0.4,0.4) node[right=0.5cm] {Contiguous region};
+% \draw[fill=warmaccent!60] (5,0) rectangle (5.4,0.4) node[right=0.5cm] {Fragmented region};
+% \draw[pattern=north east lines,pattern color=black] (0,-0.6) rectangle (0.4,-0.2) node[right=0.5cm] {Wasted padding};
+% \draw[fill=gray!50,opacity=0.4] (5,-0.6) rectangle (5.4,-0.2) node[right=0.5cm] {PMP entry count bar};
+% \end{scope}
+
+% \end{tikzpicture}
+
+% \vspace{0.4em}
+% \scriptsize
+% \textbf{Takeaway:} Align regions to NAPOT boundaries to minimize padding and preserve PMP entries.
+% \end{figure}
+% }
+
+% \vfill
+% \end{frame}
+
+% \begin{frame}{PMP Entries vs Max Concurrent Enclaves}
+% \centering
+% \only<3>{
+% \begin{figure}[htbp]
+% \centering
+% \begin{tikzpicture}
+% \begin{axis}[
+%     width=10cm,
+%     height=6cm,
+%     xlabel={PMP Entries},
+%     ylabel={Concurrent Enclaves},
+%     xmin=0, xmax=16,
+%     ymin=0, ymax=16,
+%     xtick={0,4,8,12,16},
+%     ymajorgrids,
+%     grid style=dashed,
+%     mark options={solid},
+%     nodes near coords,
+%     every node near coord/.append style={font=\footnotesize, /pgf/number format/fixed},
+%     clip=false,
+%     cycle list={{teal,mark=*,thick}},
+%     tick label style={font=\small},
+%     label style={font=\small\bfseries},
+% ]
+
+% % Data plot
+% \addplot[
+%     color=teal,
+%     mark=*,
+%     thick
+% ] coordinates {
+%     (0,0) (2,2) (4,4) (6,6) (8,8) (10,8) (12,8) (14,8) (16,8)
+% };
+
+% % Highlight PMP exhaustion area
+% \fill[red!30,opacity=0.4] (axis cs:8,0) rectangle (axis cs:16,16);
+
+% % Add Yellow Sliding Zone (between 4 and 8 PMP entries)
+% \fill[yellow!30, opacity=0.6] (axis cs:4,0) rectangle (axis cs:8,16);
+
+% % Annotation inside shaded region
+% \node[red, align=center, font=\footnotesize\bfseries] at (axis cs:12,12) {PMP Exhaustion \\ Caps at 8 Enclaves};
+% \node[align=center, font=\footnotesize\bfseries, text=black] at (axis cs:6,12) {Sliding Zone};
+
+% % Stability boundary line & label
+% \draw[dashed, thick, color=blue] (axis cs:4,0) -- (axis cs:4,16);
+% \node[anchor=south, font=\footnotesize\bfseries, color=blue] at (axis cs:4,16) {Stability Boundary};
+
+% \end{axis}
+% \end{tikzpicture}
+% \caption{Max enclaves increase with PMP entries until exhaustion caps growth.}
+% \end{figure}
+% }
 \end{frame}
@@
 \end{block}
+
+\begin{tikzpicture}[scale=1, transform shape, font=\small]
+
+% Memory bar
+\draw[thick] (0,0) rectangle (12,1);
+\node[anchor=south] at (0,1.1) {\textbf{Base}};
+\node[anchor=south] at (12,1.1) {\textbf{Top}};
+
+% Needed regions (non-contiguous)
+\draw[fill=gray!30] (0.5,0.1) rectangle (2.2,0.9);
+\node[font=\footnotesize] at (1.35,0.5) {\textbf{R1}};
+
+\draw[fill=gray!30] (3.8,0.1) rectangle (4.6,0.9);
+\node[font=\footnotesize] at (4.2,0.5) {\textbf{R2}};
+
+\draw[fill=gray!30] (7.0,0.1) rectangle (9.1,0.9);
+\node[font=\footnotesize] at (8.05,0.5) {\textbf{R3}};
+
+% NAPOT allocations (outer boxes) and padding (hatched)
+\draw[thick, fill=blue!20, draw=blue!80!black] (0.25,-0.3) rectangle (2.5,-1.1);
+\node[font=\footnotesize, align=center] at (1.375,-0.7) {\textbf{NAPOT} \\ \footnotesize fit R1};
+
+\draw[pattern=north east lines, pattern color=black] (0.25,-0.3) rectangle (0.5,-1.1);
+\draw[pattern=north east lines, pattern color=black] (2.2,-0.3) rectangle (2.5,-1.1);
+
+\draw[thick, fill=green!20, draw=green!60!black] (3.5,-0.3) rectangle (5.1,-1.1);
+\node[font=\footnotesize, align=center] at (4.3,-0.7) {\textbf{NAPOT} \\ \footnotesize fit R2};
+
+\draw[pattern=north east lines, pattern color=black] (3.5,-0.3) rectangle (3.8,-1.1);
+\draw[pattern=north east lines, pattern color=black] (4.6,-0.3) rectangle (5.1,-1.1);
+
+\draw[thick, fill=orange!20, draw=orange!70!black] (6.7,-0.3) rectangle (9.3,-1.1);
+\node[font=\footnotesize, align=center] at (8.0,-0.7) {\textbf{NAPOT} \\ \footnotesize fit R3};
+
+\draw[pattern=north east lines, pattern color=black] (6.7,-0.3) rectangle (7.0,-1.1);
+\draw[pattern=north east lines, pattern color=black] (9.1,-0.3) rectangle (9.3,-1.1);
+
+% Legend / note
+\node[anchor=west] at (0,-1.6) {\footnotesize \textbf{Note:} Hatched = padding/waste due to power-of-two alignment};
+
+\end{tikzpicture}
+
     }

### 2025-09-05 draft_presentation.tex
added new diagrams

@@
 \end{itemize}
-\begin{block}{Key Concept}
-\textit{RISC-V’s open-source nature enables flexible, secure processor designs for enclave-based systems.}
-\end{block}
+% \begin{block}{Key Concept}
+% \textit{RISC-V’s open-source nature enables flexible, secure processor designs for enclave-based systems.}
+% \end{block}
+
+\begin{center}
+\scalebox{1}{ % adjust this factor to shrink or enlarge
+\begin{tikzpicture}[font=\scriptsize, >=latex, node distance=0.5cm]
+
+% Base ISA
+\node[draw, fill=blue!20, thick, minimum width=5cm, minimum height=0.8cm, align=center] (isa) {\textbf{RISC-V ISA} \\ Open, stable foundation};
+
+% Optional Extensions layer
+\node[draw, fill=yellow!30, thick, minimum width=4.5cm, minimum height=0.8cm, align=center, above=of isa] (ext) {Optional Extensions \\ (Vector, Crypto, Custom...)};
+
+% Application domains
+\node[draw, fill=orange!30, thick, rounded corners, minimum width=2.8cm, minimum height=0.8cm, align=center, above left=0.9cm and 0.4cm of ext] (embedded) {Embedded};
+\node[draw, fill=green!30, thick, rounded corners, minimum width=2.8cm, minimum height=0.8cm, align=center, above=0.9cm of ext] (hpc) {HPC};
+\node[draw, fill=purple!30, thick, rounded corners, minimum width=2.8cm, minimum height=0.8cm, align=center, above right=0.9cm and 0.4cm of ext] (enclave) {Enclaves};
+
+% Arrows
+\draw[->, thick] (isa) -- (ext);
+\draw[->, thick] (ext.north west) .. controls +(0,0.4) and +(0,-0.4) .. (embedded.south);
+\draw[->, thick] (ext.north) -- (hpc.south);
+\draw[->, thick] (ext.north east) .. controls +(0,0.4) and +(0,-0.4) .. (enclave.south);
+
+\end{tikzpicture}
+} % end scalebox
+\end{center}
 \end{frame}
